

## Mixed distance presets

-	Add a preset for u_mix.
-	Consider adding a complete set of unbiased presets:
  -	u_dep, u_indep, u_mix (for consistency with the paper).
	-	Current implementation of u_mix via "custom" + commensurable is not transparent → simplify.
-	Naming improvements:
	-	Shorten "unbiased_dependent" (too verbose).
	-	Possibly shorten "euclidean_onehot".
	
	
## Categorical distance options (distance_cat)

- Provide a way to list all available categorical distances:
		e.g., a function like getDistMethods() (philentropy-style).
	-	Enable programmatic access to all options for benchmarking:
	  loop over all distance_cat options with mdist.
	-	Consider compatibility/extension with philentropy distances:
  		especially for unsupervised association-based distances.
	-	Replace "tot_var_dist" with "tvd".
	
	
## Supervised distance variants

  -	When a response is provided:
	  -	Default distance_cat should switch to a supervised association-based version.
	  -	Default choice: "tot_var_dist" (or "tvd").
	
	-	Allow any philentropy (or similar) association-based distance in supervised mode.
	-	Handle inconsistent user inputs:
	-	If response + non-association distance (e.g., "matching"):
	  	Either:
	  -	Override with supervised association-based distance + warning, or
	-	Ignore response + warning.
	-	Future extension:
- Allow supervised variants for distance_cont as well.




## Response input handling
-	Allow response to be provided as:
-	Column name (current behavior),  (we may want to keep this)
-	Column vector (e.g., df[,1]) for flexibility.



## Clustering / downstream methods

-	Supporting clustering methods (e.g., spectral, PAM) is useful.

-	Improve spectral_from_dist:
	-	Accept both:
	  -	Distance matrix, and
	  -	Distance object (mdist$distance).
-	Consider expanding built-in methods:
	-	PAM, hierarchical clustering, etc.


## Variable weights

-	Reintroduce or support user-defined variable weights:
	-	Already partially supported via commensurable.
-	Potential uses:
	-	Emphasize certain variables.
	-	Weight continuous vs categorical differently.
	-	Randomized weighting for robustness.
	-	Consider future tuning strategies for weights.



## Variable importance / LOVO framework
-	Current measures:
	-	Alienation coefficient (AC)
	-	Congruence coefficient (CC)
-	Concerns:
	-	AC may be too specific to the paper.
	-	Interpretation may not be obvious for all users.
-	Suggestion:
	-	Make variable importance method-dependent, e.g.:
	-	MDS → AC / CC
	-	PAM → ARI (cluster stability)
	-	Spectral → ARI or similar
	-	General idea:
	-	Evaluate variable importance relative to downstream task.



## Generalization of LOVO analysis
-	Allow specifying a method argument, e.g.:
-	method = (mds, d = 2)
-	method = (pam, k = 4)
-	method = (spectral, k = 4)
-	Compute task-specific importance metrics:
-	MDS → AC / CC
-	Clustering → ARI or similar



## LOVO comparison (lovo_cmp)
-	Rename mad_normalized:
-	Suggested: “relative distance” (clearer interpretation).
-	Extend LOVO comparison to:
-	Support multiple evaluation criteria (depending on downstream method).



## General design direction
Move toward:
-	Consistency with the paper (u_dep, u_indep, u_mix)
-	Transparency of defaults
-	Extensibility (philentropy, supervised variants, downstream methods)
-	User control (weights, methods, metrics)


### Possible issues

### 1. Supervised distance may not be using the response

He computed a supervised distance with:

```{r setup}
#| include: true

set.seed(123)
rm(list = ls())
devtools::load_all("../")
# library(manydist)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
```


```{r}
load("../vignettes/Dartpoints.RData")
D_ab <- mdist(df, response = "Name")
D <- mdist(df[,-1], response = NULL)
sum(abs(D_ab$distance - D$distance))

```

## 1. Association-based + supervised

Association-based variants do not work yet with  null response, except for `tvd`.

Ideally, we should be able to do:

```{r}
D <- mdist(df, distance_cat = "any option from philentropy + tvd + le_and_ho")
```

Then, once that works, we should also be able to combine association-based distances with supervision:

```{r}
D <- mdist(df, response = "VARNAME", distance_cat = "any option from philentropy + tvd + le_and_ho")
```

with "tvd" as the default distance_cat.

## 2. LOVO association-based + supervised

We need to be able to include supervised variants in the LOVO functions.

Simple LOVO

For the simpler lovo_mdist(), this should be relatively straightforward. I would expect something like:

```{r}
lovo_res <- lovo_mdist(df, response = "Name", distance_cat = "tvd")
```

and, more generally, this should also work with different distance_cat variants.

This should then return the LOVO values for the selected distance variant.

LOVO comparison

For lovo_cmp, things may be more difficult, because the supervised variant currently takes a different input structure from the unsupervised one:
	•	supervised currently uses an input that includes a response,
	•	unsupervised uses only the predictor data.

The vignette currently suggests defining a named list with all relevant inputs.

My first idea for comparing supervised and non-supervised variants would be to always include the response field, even for non-supervised variants. Then, as above, the response would simply be ignored whenever a non-association-based distance is used. In this way, we could compare different supervised association-based variants by changing distance_cat, for example:
	•	"tvd"
	•	"le_and_ho"
	•	and so on.

However, this creates some complications:
	•	preset variants should only receive the predictor variables, so the response would have to be ignored there as well;
	•	this approach would still not make it easy to compare association-based non-supervised variants with their supervised counterparts directly.

For example, it would still not be straightforward to compare:
	•	tvd
	•	tvd supervised

or, more generally, any association-based distance in supervised versus non-supervised form.

I think the root of the problem is that the current design uses a single data frame containing the response as one of its columns, instead of separating predictors and labels.

So perhaps we should rethink this and require supervised distances to use separate inputs:
	•	X for the non-label variables
	•	y for the labels

Then the interface would become:

```{r}
mdist(X, response = y, ...)
```


In that setup:
	•	non-supervised methods in methods would simply ignore y;
	•	supervised association-based methods would use it.

Additional LOVO questions

A few smaller questions also came up.

summary() for lovo_cmp
Can we easily obtain tables with the values?

autoplot() for lovo_cmp
When I plot normalized MAD, the y-scale does not seem optimal. The scale appears too large, so the relative differences are hard to see.

For the LOVO comparison function, we would then need to decide whether y should be stored inside each method specification or passed directly when calling the function. I am not yet sure which option is best.


Hierarchical clustering
For hierarchical clustering, which linkage do we use? I assume complete linkage. This should be clearly documented.

Spectral clustering importance
Should there also be a spectral clustering importance measure?

Also, should LOVO default to only one-cluster methods, or to all methods as it currently seems to do?

3. Defaults

I think the default should be commensurable, or equivalently unbiased. Otherwise, the results can be rather unpredictable.

In fact, I would suggest the following defaults for mdist():

```{r}
distance_cont = "manhattan"
scaling_cont  = "std"
distance_cat  = "tvd"
commensurable = TRUE
```

## 4. MCA

When using commensurable scaling, we expect all variants whose Delta matrix is just a constant multiple of matching to become equivalent. So we expect:
	•	matching
	•	eskin
	•	HL
	•	naive

to coincide under commensurability.

However, when I use:

```{r}
distance_cat = "mca"
```


the commensurable distance becomes equal to commensurable "matching".

I do not think this should happen.

MCA uses a Delta matrix depending on the marginals, so unless the category distributions are all equal, the category dissimilarities should still differ according to the observed marginal frequencies.

So this point needs clarification.

## Additional LOVO output question

Can we add summary() or print() methods for lovo and lovo_cmp?

At the moment they return tibbles, but not all information is shown, and the output is not especially targeted.
