---
title: "manydist: Exploring mdist() presets on palmerpenguins"
author: "manydist experiments"
format:
  html:
    toc: true
    code-tools: true
    code-fold: show
    embed-resources: true   
execute:
  echo: true
  warning: FALSE
  message: FALSE
---

> This notebook runs **the update version of** of `mdist`
> on the **palmerpenguins** mixed dataset. It’s designed for quick, iterative testing
> while developing `manydist`. Place this file **outside** your package directory and use
> `devtools::load_all("../manydist")` to load your development code.

## Setup

```{r, echo=FALSE,message=FALSE}
# -- Always use dev sources (not installed pkg) -------------------------------
doc_dir  <- dirname(normalizePath(knitr::current_input()))
pkg_path <- normalizePath(file.path(doc_dir, "../manydist"), mustWork = TRUE)
# 
if ("manydist" %in% loadedNamespaces()) unloadNamespace("manydist")

# load dev sources; export_all helps during development
pkgload::load_all(pkg_path, export_all = TRUE, helpers = TRUE, attach_testthat = FALSE)

# pkgload::load_all("../manydist")

cat("manydist loaded from: ",
    getNamespaceInfo(asNamespace("manydist"), "path"), "\n", sep = "")

# -- Make sure summary.MDist is registered for this session -------------------
# 1) Define a wrapper if it doesn't exist anywhere the generic can see it
if (!exists("summary.MDist", mode = "function")) {
  summary.MDist <- function(object, ...) object$summary(...)
}

# 2) Register it into the manydist namespace so S3 dispatch works even in knit
registerS3method("summary", "MDist", summary.MDist, envir = asNamespace("manydist"))

# 3) Sanity check
cat("Has summary.MDist? ",
    tryCatch(!is.null(getS3method("summary", "MDist")), error = function(e) FALSE),
    "\n", sep = "")
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("palmerpenguins", quietly = TRUE)) install.packages("palmerpenguins")
if (!requireNamespace("cluster", quietly = TRUE)) install.packages("cluster")
if (!requireNamespace("recipes", quietly = TRUE)) install.packages("recipes")
if (!requireNamespace("Rfast", quietly = TRUE)) install.packages("Rfast")

library(devtools)
# Adjust the path to your package root as needed
# devtools::load_all("../manydist")
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
set.seed(123)
```

## Data

```{r}
peng <- palmerpenguins::penguins |>
  dplyr::select(species, island, bill_length_mm, bill_depth_mm,
                flipper_length_mm, body_mass_g, sex) |>
  tidyr::drop_na()

cont_peng = peng |> select(where(is.numeric))
cat_peng = peng |> select(where(is.factor))
```


## Presets to test

I used R6 class object as an output.

```{r}
# preset_structure = tibble(preset = c("gower","unbiased_dependent","euclidean_onehot","gudmm","dkss","mod_gower","custom")) |> 
#   mutate(mix_distance = map(.x=preset,~mdist(x=peng,preset=.x))
         # )

```


##  print and summary

I created an R6 class that contains instruction of what to print (just recalling the `mdist_object_name`) and a summary function that can be recalled via `mdist_object_name$summary()` or `summary(mdist_object_name)` (the latter is S3 style).

```{r}
peng_split=initial_split(peng,prop = .7,strata = species)
tr_peng=training(peng_split)
ts_peng=testing(peng_split)

udep_dist = mdist(x=tr_peng,new_data=ts_peng,preset="unbiased_dependent")
udep_dist$distance




gow_dist = mdist(x=tr_peng,new_data=ts_peng,preset="gower")
gow_dist$distance
gow_dist$distance |> as.matrix()
udep_dist$summary()
udep_dist$distance |> as.matrix()
summary(udep_dist)

```


## leave one variable out assessment (LOVO)

Just like in the paper, I created the `lovo_mdist()` function, that uses the same syntax as `mdist()` and computes the LOVO distances and compares them with the full dist, both in terms of alienation coefficient and MAD

```{r}
lovo_results = lovo_mdist(x=peng,preset="unbiased_dependent")

lovo_results$summary()
print(lovo_results)
summary(lovo_results)

```

## Next up

- homogeneize the overlapping print and summary output (cosmetic)

- think of some plot (would like to have a function that reproduces the plots we have in the paper, for one or more methods...but if it's not painful)
<!-- ## plotting function(s)? -->

<!-- ```{r} -->
<!-- # plot_cmd <- function(d_or_mat, title = "MDS (cmdscale)") { -->
<!-- #   to_matrix <- function(obj) { -->
<!-- #     if (is.null(obj)) return(NULL) -->
<!-- #     if ("R6" %in% class(obj)) { -->
<!-- #       if (!is.null(obj$distance)) return(as.matrix(obj$distance)) -->
<!-- #       if ("to_dist" %in% names(obj)) return(as.matrix(obj$to_dist())) -->
<!-- #       return(NULL) -->
<!-- #     } -->
<!-- #     if (inherits(obj, "dist") || inherits(obj, "dissimilarity")) return(as.matrix(obj)) -->
<!-- #     if (is.matrix(obj)) return(obj) -->
<!-- #     if (is.data.frame(obj)) return(as.matrix(obj)) -->
<!-- #     NULL -->
<!-- #   } -->
<!-- #   m <- to_matrix(d_or_mat) -->
<!-- #   stopifnot(!is.null(m)) -->
<!-- #   coords <- cmdscale(m, k = 2, eig = TRUE) -->
<!-- #   df <- tibble::as_tibble(coords$points) |> rlang::set_names(c("Dim1", "Dim2")) -->
<!-- #   ggplot2::ggplot(df, ggplot2::aes(Dim1, Dim2)) + -->
<!-- #     ggplot2::geom_point() + -->
<!-- #     ggplot2::labs(title = title) -->
<!-- # } -->
<!-- #  -->
<!-- #  -->

<!-- ``` -->



## Session info

```{r}
sessionInfo()
```
