---
title: "Distance-based Learning for Mixed Data: The manydist Package"
author:
  - name: Alfonso Iodice D'Enza
    affiliation: University of Naples Federico II
  - name: Michel van de Velden
    affiliation: Erasmus University Rotterdam
  - name: Angelos Markos
    affiliation: Democritus University of Thrace
  - name: Carlo Cavicchia
    affiliation: Erasmus University Rotterdam
abstract: |
  Computing distances between observations described by mixed-type variables
  -- both numerical and categorical -- is a fundamental task in clustering,
  ordination, and nearest-neighbour methods. Existing distance measures can
  suffer from importance biases: the contribution of a variable may depend on
  its measurement type, distribution, scaling, or number of categories. We
  present the R package manydist, which implements a general framework for
  constructing unbiased mixed-variable distances through multivariate
  additivity and commensurability. We describe the distance construction,
  variable diagnostics, and integration with clustering and the tidymodels
  predictive-modeling ecosystem. The Palmer penguins data illustrate how the
  framework avoids distortions present in standard alternatives.
keywords:
  - distance
  - mixed data
  - commensurability
  - R
header-includes:
  - \usepackage{bm}
---

# Introduction

Measuring the distance or dissimilarity between observations is a fundamental
step in clustering, classification, ordination, and nearest-neighbour methods.
For numerical variables, standard metrics such as Euclidean and Manhattan
distance are widely used. For categorical variables, simple matching and
frequency-based dissimilarities are common. Real data, however, frequently
combine numerical and categorical variables, and defining a distance that
accounts for their different scales and measurement properties is not
straightforward.

A persistent difficulty is *importance bias*: the overall distance can be
unequally affected by variables because of their type, scale, distribution, or
number of categories. Gower's distance [@Gower1971], the most widely used
mixed-variable construction, can favor categorical variables
[@Hennig2013; @foss2016]. Categorical variables with more levels may dominate,
while skewness and scaling choices can unpredictably amplify or attenuate
numerical contributions.

@van2024unbiased formalize unbiased mixed-variable distances through two
properties: *multivariate additivity* and *commensurability*. The R package
manydist [@manydist2026] implements this framework. Its main function,
mdist(), provides a unified entry point that:

- computes variable-wise dissimilarities using a selected numerical
  preprocessing and categorical dissimilarity;
- estimates variable-wise weights that enforce commensurability; and
- aggregates the contributions into an additive distance matrix.

We organize these capabilities as a progression of increasing *awareness* in
distance construction. Scale- and type-aware distances make heterogeneous
variable contributions comparable. Association-aware distances additionally
account for dependence within the numerical and categorical blocks.
Interaction-aware distances incorporate relationships between those blocks,
while response-aware distances use outcome information to construct
task-specific neighbourhoods. The methodological development below focuses
first on scale/type awareness and association awareness; the interaction- and
response-aware extensions enter later through the package interface and
learning workflows.

Beyond distance construction, manydist provides leave-one-variable-out (LOVO)
diagnostics, integration with tidymodels for predictive workflows, and
distance-based interfaces to partitioning around medoids (PAM) and spectral
clustering.

Several R packages provide distances for homogeneous numerical, categorical,
or binary data. @tbl-related-homogeneous summarizes commonly used tools that
serve as building blocks for clustering, ordination, and nearest-neighbour
methods.

```{=latex}
\begingroup
\makeatletter
\let\paperLongtableCaption\LT@caption
\def\LT@caption{\hline\paperLongtableCaption}
\makeatother
```

```{r}
#| label: tbl-related-homogeneous
#| tbl-cap: "R packages providing distances or dissimilarities for homogeneous numerical, categorical, or binary data."
#| tbl-colwidths: [35, 65]
#| echo: false

related_software <- readr::read_csv(
  file.path("data", "related-software.csv"),
  show_col_types = FALSE
)

format_related_software <- function(group) {
  output <- related_software |>
    dplyr::filter(.data$group == .env$group) |>
    dplyr::transmute(
      Software = paste0(
        .data$package,
        "::",
        .data[["function"]],
        "()"
      ),
      Description = .data$description,
      .citation_key = .data$citation_key
    )

  if (knitr::is_latex_output()) {
    middle_cell <- function(value, proportion) {
      width <- paste0(
        "(\\linewidth - 2\\tabcolsep) * ",
        "\\real{", proportion, "}"
      )
      paste0(
        "\\multicolumn{1}{>{\\raggedright\\arraybackslash}m{",
        width,
        "}}{",
        value,
        "}"
      )
    }

    software_labels <- dplyr::if_else(
      stringr::str_detect(output$Software, "_"),
      paste0("\\detokenize{", output$Software, "}"),
      output$Software
    )

    software_entries <- paste0(
      "\\shortstack[l]{",
      software_labels,
      "\\\\[-2pt]{\\footnotesize \\citep{",
      output$.citation_key,
      "}}}"
    )

    output <- output |>
      dplyr::mutate(
        Software = middle_cell(software_entries, "0.3500"),
        Description = middle_cell(.data$Description, "0.6500"),
        Software = dplyr::if_else(
          dplyr::row_number() == 1L,
          .data$Software,
          paste0("\\hline ", .data$Software)
        )
      )
  } else {
    output <- output |>
      dplyr::mutate(
        Software = paste0(
          .data$Software,
          " [@",
          .data$.citation_key,
          "]"
        )
      )
  }

  output |>
    dplyr::select(-.data$.citation_key)
}

knitr::kable(
  format_related_software("homogeneous")
)
```

```{=latex}
\endgroup
```

For numerical matrices, stats::dist() supplies the standard metrics, while
proxy::dist() offers an extensible catalogue and user-defined measures.
parallelDist::parDist() and Rfast::Dist() target larger numerical problems with
parallel or compiled implementations. For categorical and binary data,
klaR::kmodes() uses matching dissimilarities, arules::dissimilarity() supports
sparse transaction data, and nomclust::nomclust() provides frequency-weighted
similarities.

Mixed-data tools are summarized in @tbl-related-mixed. Gower's construction is
implemented in cluster::daisy(), FD::gowdis(), StatMatch::gower.dist(), and
gower::gower_dist(). clustMixType::kproto() balances numerical and categorical
components through the k-prototypes construction [@Huang1998kprototypes].
kmed::distmix() and kdml::dkss() provide additional mixed-data alternatives.

```{=latex}
\begingroup
\makeatletter
\let\paperLongtableCaption\LT@caption
\def\LT@caption{\hline\paperLongtableCaption}
\makeatother
```

```{r}
#| label: tbl-related-mixed
#| tbl-cap: "R packages implementing distances or dissimilarities for mixed numerical and categorical data."
#| tbl-colwidths: [35, 65]
#| echo: false

knitr::kable(
  format_related_software("mixed")
)
```

```{=latex}
\endgroup
```

These tools are useful, but their variable contributions can depend on scale,
prevalence, imbalance, and measurement type. manydist complements them with
additive, commensurable, and association-aware constructions, together with
diagnostic and modeling interfaces.

# A unified mixed-variable distance framework

We follow the notation of @van2024unbiased. Let $n$ be the number of
observations and let $p=p_n+p_c$ be the number of variables, partitioned into
$p_n$ numerical and $p_c$ categorical variables. Observation indices are
$i,l,s$; $j$, $h$, and $k$ index generic, numerical, and categorical
variables, respectively. Observed mixed, numerical, and categorical vectors
are denoted by $\mathbf{x}$, $\mathbf{x}^n$, and $\mathbf{x}^c$; their random
counterparts use $\bm{X}$. The symbol $\mathbf{X}$ is reserved for a data
matrix.

## General mixed-distance setup

For observed vectors $\mathbf{x}_i$ and $\mathbf{x}_l$, manydist constructs an
overall distance by combining contributions from the numerical and categorical
blocks. Using variable-wise dissimilarities $\delta_j$ and weights $w_j$, the
general setup is

$$
\begin{aligned}
d(\mathbf{x}_i,\mathbf{x}_l)
&=\sum_{j=1}^{p}d_j(\mathbf{x}_i,\mathbf{x}_l)\\
&=\sum_{h=1}^{p_n}w_h\,\delta_h^n(\mathbf{x}_i^n,\mathbf{x}_l^n)
 +\sum_{k=1}^{p_c}w_k\,\delta_k^c(\mathbf{x}_i^c,\mathbf{x}_l^c).
\end{aligned}
$$ {#eq-mixed-distance}

The numerical contribution indexed by $h$ can depend on the complete
numerical block, and the categorical contribution indexed by $k$ can depend
on the complete categorical block. The present formulation does not use
cross-type associations inside these component dissimilarities.

## Scale- and type-aware distances

We call a mixed-variable distance *unbiased* when the contribution associated
with each variable does not depend on its measurement type or scale. Two
properties impose this requirement on the general setup.

### Multivariate additivity

For $p$-dimensional observations
$\mathbf{x}_i=(x_{i1},\ldots,x_{ip})$ and
$\mathbf{x}_l=(x_{l1},\ldots,x_{lp})$, a
function $d:\mathcal{X}^p\times\mathcal{X}^p\rightarrow[0,\infty)$ is
multivariate additive if

$$
d(\mathbf{x}_i,\mathbf{x}_l)
=\sum_{j=1}^{p}d_j(\mathbf{x}_i,\mathbf{x}_l),
$$ {#eq-additivity}

where $d_j(\mathbf{x}_i,\mathbf{x}_l)$ is the contribution associated with
variable $j$.
Each component may depend on all elements of the two observations; the index
identifies the associated contribution rather than restricting it to a single
coordinate. This permits additive constructions that incorporate dependencies
between variables. If every $d_j$ is a metric on $\mathcal{X}^p$, their sum is
also a metric.

### Commensurability

Let $\bm{X}_i=(X_{i1},\ldots,X_{ip})$ denote the random vector
corresponding to observation $i$. The variable contributions are commensurable
when, for every $j$ and $i\ne l$,

$$
\mathbb{E}\!\left[d_j(\bm{X}_i,\bm{X}_l)\right]=c,
$$ {#eq-commensurability}

where the particular value of the constant $c$ is unimportant. Additivity
provides an interpretable decomposition; commensurability places those
contributions on equivalent expected scales.

For @eq-mixed-distance, commensurability follows from the population weights

$$
w_h=
\frac{1}{
  \mathbb{E}[\delta_h^n(\bm{X}_i^n,\bm{X}_l^n)]
},
\qquad
w_k=
\frac{1}{
  \mathbb{E}[\delta_k^c(\bm{X}_i^c,\bm{X}_l^c)]
}.
$$ {#eq-population-weights}

In practice, manydist estimates these expectations from the observed pairs:

$$
\begin{aligned}
m_h&=\frac{1}{n(n-1)}\sum_{i\ne l}
\delta_h^n(\mathbf{x}_i^n,\mathbf{x}_l^n),
&\quad w_h&=\frac{1}{m_h},\\
m_k&=\frac{1}{n(n-1)}\sum_{i\ne l}
\delta_k^c(\mathbf{x}_i^c,\mathbf{x}_l^c),
&\quad w_k&=\frac{1}{m_k}.
\end{aligned}
$$ {#eq-empirical-weights}

Thus, each variable-wise distance is divided by its empirical mean before the
contributions are combined. A stabilized form,
$w_j=1/\max(m_j,\varepsilon)$, can be used when dispersion is extremely small.
The formulation also permits constrained weights, such as common weights
within numerical or categorical groups, when only within-type
commensurability is required.

## Association-aware distances: beyond commensurability

Commensurability corrects differences in the expected size of the variable
contributions, but it does not by itself remove redundant information. An
independence-based construction defines the contribution of variable $j$ from
that variable alone,

$$
d_j^{\mathrm{ind}}(\mathbf{x}_i,\mathbf{x}_l)
=w_j\,\delta_j(x_{ij},x_{lj}).
$$ {#eq-independence-distance}

Consequently, when several variables encode the same underlying contrast,
that contrast can contribute repeatedly to the overall distance even after
the variables have been standardized or made commensurable. An
association-based construction instead allows a contribution to depend on the
complete block of variables of the same type,

$$
\begin{aligned}
d_h^{\mathrm{assoc}}(\mathbf{x}_i,\mathbf{x}_l)
  &=w_h\,\delta_h^n(\mathbf{x}_i^n,\mathbf{x}_l^n),\\
d_k^{\mathrm{assoc}}(\mathbf{x}_i,\mathbf{x}_l)
  &=w_k\,\delta_k^c(\mathbf{x}_i^c,\mathbf{x}_l^c).
\end{aligned}
$$ {#eq-association-distance}

Thus, the contribution indexed by $h$ or $k$ need not be a function of that
variable alone, as permitted by multivariate additivity. For numerical
variables, the joint structure can be represented by a covariance or
correlation matrix; for categorical variables, it can be represented by
conditional distributions. The manydist formulation does not use cross-type
associations in these component dissimilarities.

The distinction is illustrated with the Palmer penguins data. The same
six-predictor data set is used throughout the remaining examples.

```{r}
#| label: prepare-penguins

penguins <- palmerpenguins::penguins |>
  dplyr::select(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  tidyr::drop_na()

penguins_x <- penguins |>
  dplyr::select(-species)
penguins_y <- penguins |>
  dplyr::pull(species)
```

```{r}
#| label: fig-independence-association
#| fig-cap: "Scale and association affect distance geometry. In the original units, Euclidean distance is dominated by body mass. Marginal standardization removes the units but retains the correlation and uses both coordinate differences. After whitening, Euclidean distance is equivalent to Mahalanobis distance in the standardized space and accounts for their covariance."
#| fig-subcap:
#|   - "Euclidean distance in the original units ($r = 0.87$)."
#|   - "Standardized Euclidean distance."
#|   - "Mahalanobis distance after whitening."
#| layout: [[1], [1, 1]]
#| fig-width: 4.5
#| fig-height: 2.6
#| fig-pos: "H"
#| echo: false
#| warning: false

dependence_data <- penguins |>
  dplyr::transmute(
    penguin_id = dplyr::row_number(),
    flipper_length_mm,
    body_mass_g,
    flipper_z = as.numeric(scale(.data$flipper_length_mm)),
    body_mass_z = as.numeric(scale(.data$body_mass_g))
  )

standardized_data <- dependence_data |>
  dplyr::select(.data$flipper_z, .data$body_mass_z) |>
  as.matrix()

pca_fit <- stats::prcomp(
  standardized_data,
  center = FALSE,
  scale. = FALSE
)

whitened_scores <- sweep(
  pca_fit$x[, 1:2, drop = FALSE],
  MARGIN = 2,
  STATS = pca_fit$sdev[1:2],
  FUN = "/"
)

dependence_data <- dependence_data |>
  dplyr::mutate(
    pc1 = pca_fit$x[, 1],
    pc2 = pca_fit$x[, 2],
    whitened_pc1 = whitened_scores[, 1],
    whitened_pc2 = whitened_scores[, 2]
  )

pc1_targets <- stats::quantile(
  dependence_data$pc1,
  probs = c(0.2, 0.8)
)

selected_ids <- purrr::map_int(pc1_targets, function(target) {
  dependence_data |>
    dplyr::mutate(target_distance = abs(.data$pc1 - target)) |>
    dplyr::slice_min(
      order_by = .data$target_distance,
      n = 20,
      with_ties = FALSE
    ) |>
    dplyr::slice_min(
      order_by = abs(.data$pc2),
      n = 1,
      with_ties = FALSE
    ) |>
    dplyr::pull(.data$penguin_id)
})

selected_points <- dependence_data |>
  dplyr::filter(.data$penguin_id %in% selected_ids) |>
  dplyr::arrange(.data$flipper_z) |>
  dplyr::mutate(point = c("A", "B"))

raw_segment <- selected_points |>
  dplyr::summarise(
    x = dplyr::first(.data$flipper_length_mm),
    y = dplyr::first(.data$body_mass_g),
    xend = dplyr::last(.data$flipper_length_mm),
    yend = dplyr::last(.data$body_mass_g),
    distance = sqrt((.data$xend - .data$x)^2 + (.data$yend - .data$y)^2),
    xmid = (.data$x + .data$xend) / 2,
    ymid = (.data$y + .data$yend) / 2
  )

standardized_segment <- selected_points |>
  dplyr::summarise(
    x = dplyr::first(.data$flipper_z),
    y = dplyr::first(.data$body_mass_z),
    xend = dplyr::last(.data$flipper_z),
    yend = dplyr::last(.data$body_mass_z),
    distance = sqrt((.data$xend - .data$x)^2 + (.data$yend - .data$y)^2),
    xmid = (.data$x + .data$xend) / 2,
    ymid = (.data$y + .data$yend) / 2
  )

component_segments <- tibble::tibble(
  x = c(standardized_segment$x, standardized_segment$xend),
  y = c(standardized_segment$y, standardized_segment$y),
  xend = c(standardized_segment$xend, standardized_segment$xend),
  yend = c(standardized_segment$y, standardized_segment$yend)
)

mahalanobis_segment <- selected_points |>
  dplyr::summarise(
    x = dplyr::first(.data$whitened_pc1),
    y = dplyr::first(.data$whitened_pc2),
    xend = dplyr::last(.data$whitened_pc1),
    yend = dplyr::last(.data$whitened_pc2),
    distance = sqrt((.data$xend - .data$x)^2 + (.data$yend - .data$y)^2),
    xmid = (.data$x + .data$xend) / 2,
    ymid = (.data$y + .data$yend) / 2
  )

common_score_limits <- c(-3.5, 3.5)
common_score_breaks <- seq(-3, 3, by = 1)

distance_plot_theme <- ggplot2::theme_minimal(base_size = 8) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = 8),
    axis.text = ggplot2::element_text(size = 7),
    plot.margin = ggplot2::margin(3, 4, 3, 3)
  )

raw_plot <- ggplot2::ggplot(
  dependence_data,
  ggplot2::aes(x = .data$flipper_length_mm, y = .data$body_mass_g)
) +
  ggplot2::geom_point(size = 1.1, color = "grey35", alpha = 0.55) +
  ggplot2::stat_ellipse(
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = raw_segment,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.65,
    arrow = grid::arrow(length = grid::unit(0.07, "inches"))
  ) +
  ggplot2::geom_point(
    data = selected_points,
    size = 2.1,
    color = "#A33F3F"
  ) +
  ggplot2::geom_text(
    data = selected_points,
    ggplot2::aes(label = .data$point),
    nudge_y = 170,
    size = 2.7,
    fontface = "bold"
  ) +
  ggplot2::geom_label(
    data = raw_segment,
    ggplot2::aes(
      x = .data$xmid,
      y = .data$ymid,
      label = paste0("d[E]^{raw} == ", round(.data$distance, 0))
    ),
    parse = TRUE,
    inherit.aes = FALSE,
    size = 2.3,
    linewidth = 0,
    fill = "white"
  ) +
  ggplot2::labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)"
  ) +
  distance_plot_theme

independence_plot <- ggplot2::ggplot(
  dependence_data,
  ggplot2::aes(x = .data$flipper_z, y = .data$body_mass_z)
) +
  ggplot2::geom_point(size = 0.65, color = "grey70") +
  ggplot2::stat_ellipse(
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = component_segments,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.55,
    linetype = "dashed",
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = standardized_segment,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.65,
    arrow = grid::arrow(length = grid::unit(0.07, "inches"))
  ) +
  ggplot2::geom_point(
    data = selected_points,
    size = 1.7,
    color = "#A33F3F"
  ) +
  ggplot2::geom_text(
    data = selected_points,
    ggplot2::aes(label = .data$point),
    nudge_y = 0.2,
    size = 2.7,
    fontface = "bold"
  ) +
  ggplot2::geom_label(
    data = standardized_segment,
    ggplot2::aes(
      x = .data$xmid,
      y = .data$ymid,
      label = paste0("d[E]^{z} == ", round(.data$distance, 2))
    ),
    parse = TRUE,
    inherit.aes = FALSE,
    size = 2.3,
    linewidth = 0,
    fill = "white"
  ) +
  ggplot2::scale_x_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::scale_y_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    x = "Flipper length (z-score)",
    y = "Body mass (z-score)"
  ) +
  distance_plot_theme

association_plot <- ggplot2::ggplot(
  dependence_data,
  ggplot2::aes(x = .data$whitened_pc1, y = .data$whitened_pc2)
) +
  ggplot2::geom_point(size = 0.65, color = "grey70") +
  ggplot2::stat_ellipse(
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    color = "#2A7F62"
  ) +
  ggplot2::geom_segment(
    data = mahalanobis_segment,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.65,
    arrow = grid::arrow(length = grid::unit(0.07, "inches"))
  ) +
  ggplot2::geom_point(
    data = selected_points,
    size = 1.7,
    color = "#2A7F62"
  ) +
  ggplot2::geom_text(
    data = selected_points,
    ggplot2::aes(label = .data$point),
    nudge_y = 0.2,
    size = 2.7,
    fontface = "bold"
  ) +
  ggplot2::geom_label(
    data = mahalanobis_segment,
    ggplot2::aes(
      x = .data$xmid,
      y = .data$ymid,
      label = paste0("d[M] == ", round(.data$distance, 2))
    ),
    parse = TRUE,
    inherit.aes = FALSE,
    size = 2.3,
    linewidth = 0,
    fill = "white"
  ) +
  ggplot2::scale_x_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::scale_y_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    x = "Whitened PC 1",
    y = "Whitened PC 2"
  ) +
  distance_plot_theme

raw_plot
independence_plot
association_plot
```

## Numerical variables

In the numerical example, whitening uses the covariance structure to remove
linear redundancy: Euclidean distance in the whitened coordinates is the
Mahalanobis distance in the original standardized coordinates. manydist uses
the same decorrelation principle through principal-component preprocessing,
followed by component-wise commensurability to retain an additive
decomposition.

For the numerical block, the additive construction uses weighted Manhattan
contributions on possibly transformed coordinates:

$$
d_{\mathrm{num}}(\mathbf{x}_i,\mathbf{x}_l)
=\sum_{h=1}^{p_n}w_h
\left|f_h(\mathbf{x}_i)-f_h(\mathbf{x}_l)\right|,
$$ {#eq-numerical-distance}

where $f_h$ may be a univariate or multivariate transformation. Linear
rescaling is absorbed by the weights, but the transformation can still alter
the marginal distribution and therefore the empirical mean distance.

Common transformations include:

- **standard-deviation scaling:** $(x-\bar{x})/s_x$;
- **range scaling:** $(x-x_{\min})/(x_{\max}-x_{\min})$, as in Gower's
  distance;
- **robust scaling:** $(x-x_{\mathrm{Med}})/(x_{Q_3}-x_{Q_1})$.

When numerical variables are linearly related, principal-component scaling
can rotate and rescale the numerical block. If the centered data matrix has
covariance decomposition
$\mathbf{S}=\mathbf{V}\bm{\Lambda}\mathbf{V}^\top$, then
$\mathbf{X}^\ast=\mathbf{X}\mathbf{V}\bm{\Lambda}^{-1/2}$ gives standardized
principal coordinates. This
operation does not require dimensionality reduction; additivity then applies
to the rotated axes.

## Categorical variables

For categorical variables, association enters through conditional
distributions rather than a covariance matrix. The category dissimilarity
matrix provides a common representation for both independence-based and
association-based constructions.

For categorical variable $k$, let $\bm{\Delta}_k$ be the
$q_k\times q_k$ dissimilarity matrix between its categories, and let
$\mathbf{Z}_k$ be
the corresponding $n\times q_k$ indicator matrix. Its observation-wise
distance matrix and expected contribution are

$$
\mathbf{D}_k=\mathbf{Z}_k\bm{\Delta}_k\mathbf{Z}_k^\top,
\qquad
\mathbb{E}[d_k(X_{ik},X_{lk})]
=\bm{p}_k^\top\bm{\Delta}_k\bm{p}_k,
$$ {#eq-categorical-distance}

where $\bm{p}_k$ is the vector of category probabilities. The mean
distance depends on the number and prevalence of categories and on
$\bm{\Delta}_k$; the weights in @eq-empirical-weights restore
commensurability.

Indicator-based dissimilarities treat a binary indicator representation as
numerical. Without scaling, Manhattan distance gives values zero or two.
Hennig--Liao scaling [@Hennig2013], standard-deviation scaling, and direct
category-dissimilarity scaling provide alternative calibrations.

Association-based dissimilarities incorporate relationships among categorical
variables. Let
$\mathbf{R}^{k\mid k'}=(\mathbf{Z}_k^\top \mathbf{Z}_k)^{-1}
\mathbf{Z}_k^\top \mathbf{Z}_{k'}$ be the conditional
distribution of variable $k'$ given the categories of variable $k$.
Dissimilarities can compare rows $\mathbf{r}_a$ and $\mathbf{r}_b$ using total variation
[@ahmad2007k] or a symmetric Kullback--Leibler construction
[@le2005association]. These contributions vanish under independence and grow
with the strength of the association.

# Package structure

The manydist interface follows the three stages of a distance-based analysis:
building a distance, diagnosing the resulting representation, and using it in
a learning workflow. All examples continue with the six-predictor Palmer
penguins data prepared above.

```{r}
#| label: attach-manydist

library(manydist)
```

## Distance building

The central function `mdist()` constructs distances for numerical,
categorical, or mixed-type data. A preset identifies a complete distance
specification. For example, `"u_dep"` combines principal-component scaling of
the numerical block, association-based categorical dissimilarities, and
commensurability weights.

```{r}
#| label: build-u-dep-distance

d_u_dep <- mdist(
  penguins_x,
  preset = "u_dep"
)

d_u_dep
```

The result is an `MDist` object that stores the dissimilarity together with the
fitted preprocessing information and the distance specification. The standard
`dist` representation is available through `to_dist()`.

```{r}
#| label: inspect-u-dep-distance

as.matrix(d_u_dep$to_dist())[1:4, 1:4] |>
  round(digits = 2)
```

Presets provide reproducible shortcuts, while a custom specification exposes
the categorical dissimilarity, numerical preprocessing, and commensurability
choice directly.

```{r}
#| label: build-custom-distance

d_custom <- mdist(
  penguins_x,
  preset = "custom",
  method_cat = "matching",
  method_num = "pc_scores",
  commensurable = TRUE
)
```

The same interface supports extensions to the basic predictor-only
construction. The `response` argument supplies an outcome to response-aware
components without treating it as an ordinary predictor.

```{r}
#| label: build-response-aware-distance

d_response <- mdist(
  penguins,
  response = species,
  preset = "u_dep"
)
```

The `new_data` argument produces rectangular new-to-training dissimilarities.
This distinction between square training distances and rectangular prediction
distances is handled automatically by the workflow interface below.

## Distance diagnostics

A distance is an intermediate data representation, so its sensitivity should
be examined before it is supplied to a learning algorithm. `lovo_mdist()`
implements leave-one-variable-out (LOVO) diagnostics: it computes the full
distance, removes one predictor at a time, recomputes the distance, and
quantifies the resulting change.

```{r}
#| label: diagnose-gower-distance

lovo_gower <- lovo_mdist(
  penguins,
  response = species,
  response_used = FALSE,
  preset = "gower"
)
```

The diagnostics distinguish changes in pairwise dissimilarities from changes
in the geometry they induce. `relative_distance` measures the relative change
in pairwise dissimilarities, whereas `ac_importance` is the alienation between
the full and reduced configurations. Larger values indicate a larger
contribution for both quantities.

```{r}
#| label: inspect-gower-diagnostics

lovo_gower$results |>
  dplyr::select(
    variable,
    relative_distance,
    ac_importance
  ) |>
  dplyr::arrange(dplyr::desc(relative_distance))
```

```{r}
#| label: fig-gower-lovo
#| fig-cap: "Relative leave-one-variable-out contributions under Gower distance. Larger values indicate a larger change in the pairwise dissimilarities when the variable is removed."
#| fig-width: 6
#| fig-height: 3.2

lovo_gower$autoplot(
  metric = "relative_distance",
  reorder = TRUE
) +
  ggplot2::guides(colour = "none")
```

Optional clustering diagnostics compare partitions from the full and reduced
distances using the adjusted Rand index. `compare_lovo_mdist()` applies the
same diagnostics to several distance specifications, making the choice of
distance part of the sensitivity analysis rather than a fixed preprocessing
decision.

## Building distance-based workflows

`step_mdist()` integrates distance construction into a recipe. The fitted step
stores the training data and preprocessing parameters, so the distance is
re-estimated inside each resample and reused when genuinely new observations
are processed. Its `output` argument selects the representation required by
the downstream task.

### Clustering from pairwise distances

For clustering, `output = "pairwise"` produces the square within-training
dissimilarity matrix. It can be consumed by `pam_dist()` or
`spectral_dist()`. The following workflow combines a Gower distance with a
three-cluster partitioning around medoids.

```{r}
#| label: build-pam-workflow

gower_pairwise_recipe <- recipes::recipe(
  ~ .,
  data = penguins_x
) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "pairwise"
  )

pam_fit <- generics::fit(
  pam_dist(num_clusters = 3),
  recipe = gower_pairwise_recipe,
  data = penguins_x
)
```

The species labels were not used to construct the partition. They are added
below only to interpret this illustrative example.

```{r}
#| label: inspect-pam-workflow

pam_assignments <- predict(pam_fit) |>
  dplyr::bind_cols(
    penguins |>
      dplyr::select(species)
  )

pam_assignments |>
  dplyr::count(.pred_cluster, species)
```

The same pairwise recipe can be supplied to `spectral_dist()`, which also
provides access to the fitted spectral embedding.

### Prediction from distances to the training set

For prediction, `output = "distance_to_training"` represents each observation
by its distances to the training observations. `nearest_neighbor_dist()` uses
these columns in a parsnip model specification. Keeping the transformation in
the recipe ensures that assessment and test observations are compared only
with the corresponding training set.

```{r}
#| label: build-knn-workflow

gower_training_recipe <- recipes::recipe(
  species ~ .,
  data = penguins
) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "distance_to_training"
  )

knn_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = 5
) |>
  parsnip::set_engine("manydist")

knn_workflow <- workflows::workflow() |>
  workflows::add_recipe(gower_training_recipe) |>
  workflows::add_model(knn_spec)

knn_workflow
```

The resulting object can be fitted, resampled, and tuned with the standard
tidymodels interface. Model assessment should tune the number of neighbours
and, when it is not fixed in advance, the distance specification under a
common resampling design before the final test set is used.

# Illustration: distance bias in the Palmer penguins dataset

```{r}
#| label: fig-mds-unbiased-dependent
#| fig-cap: "Two-dimensional MDS configuration under the unbiased dependent distance."
#| fig-pos: "H"
#| echo: false

mds_figure <- file.path(
  "generated",
  "figures",
  "fig-mds-unbiased-dependent.pdf"
)

if (file.exists(mds_figure)) {
  knitr::include_graphics(mds_figure)
} else {
  cat("Run Rscript replicate.R to generate this figure.")
}
```

# Conclusion

The conclusion will be migrated after the executable sections establish the
current supported API and feature set.
