---
title: "Distance-based Learning for Mixed Data: The manydist Package"
author:
  - name: Alfonso Iodice D'Enza
    orcid: 0000-0002-6147-0052
    email: iodicede@unina.it
    affiliations:
      - name: University of Naples Federico II
        department: Department of Political Sciences
        address: Via Leopoldo Rodinò 22
        city: Naples
        postal-code: '80138'
        country: Italy
  - name: Michel van de Velden
    orcid: 0000-0002-9807-9057
    email: vandevelden@ese.eur.nl
    affiliations:
      - name: Erasmus University Rotterdam
        department: Econometric Institute, Erasmus School of Economics
        address: Burgemeester Oudlaan 50
        city: Rotterdam
        postal-code: '3062 PA'
        country: The Netherlands
  - name: Angelos Markos
    orcid: 0000-0002-4204-3573
    email: amarkos@eled.duth.gr
    affiliations:
      - name: Democritus University of Thrace
        department: Department of Primary Education, School of Education
        address: Nea Chili
        city: Alexandroupolis
        postal-code: '68132'
        country: Greece
  - name: Carlo Cavicchia
    orcid: 0000-0003-1816-3521
    email: cavicchia@ese.eur.nl
    affiliations:
      - name: Erasmus University Rotterdam
        department: Econometric Institute, Erasmus School of Economics
        address: Burgemeester Oudlaan 50
        city: Rotterdam
        postal-code: '3062 PA'
        country: The Netherlands
abstract: |
  Computing distances between observations described by mixed-type variables
  -- both numerical and categorical -- is a fundamental task in clustering,
  multidimensional scaling, and nearest-neighbour methods. Existing distance
  measures can
  suffer from importance biases: the contribution of a variable may depend on
  its measurement type, distribution, scaling, or number of categories. We
  present the R package manydist, which constructs mixed-variable distances
  that account for such biases. Using the Palmer penguins data, we describe
  distance construction, diagnostics, and benchmarking, showing how
  leave-one-variable-out contributions and comparisons between candidate
  distances make the choice of distance an inspectable part of an analysis
  rather than a fixed preprocessing decision. We then illustrate how to
  integrate distance construction into both unsupervised and supervised
  pipelines. In particular, using a World Development Indicators snapshot, we
  build clustering and classification pipelines in which the preprocessing is
  refit within each resample.
keywords:
  - distance
  - mixed data
  - commensurability
  - R
keywords-formatted:
  - distance
  - mixed data
  - commensurability
  - R
header-includes:
  - \usepackage{bm}
  - \usepackage{needspace}
  - \usepackage{xcolor}
  - \newcommand{\new}[1]{{\color{blue}#1}}
---

## Introduction

Measuring the distance or dissimilarity between observations is a fundamental
step in clustering, classification, multidimensional scaling
[@BorgGroenen2005], and nearest-neighbour methods.
For numerical variables, standard metrics such as Euclidean and Manhattan
distance are widely used. For categorical variables, simple matching and
frequency-based dissimilarities are common. Real data, however, frequently
combine numerical and categorical variables, and defining a distance that
accounts for their different scales and measurement properties is not
straightforward.

A persistent difficulty is *importance bias*: the overall distance can be unequally affected by variables because of their type, scale, distribution, or number of categories. The mechanism is easiest to see for the simple matching dissimilarity, under which two observations differ on a categorical variable whenever they fall in different categories. As the number of categories grows, a random pair is more likely to differ, so the variable's expected contribution increases with the number of levels -- a nominal variable with many categories
can dominate one with few, independently of any relationship to the phenomenon of interest. Gower's distance [@Gower1971], the most widely used mixed-variable construction, rescales each numerical variable to the unit interval by its range; because most observations fall well inside the extremes, the resulting mean pairwise distance is typically smaller than one, which is why Gower tends to favor categorical contributions [@Hennig2013; @foss2016]. Skewed distributions for the numerical variables amplify this effect: a long-tailed numerical variable is compressed further by range scaling, shrinking its contribution even more. These are undesirable factors in the calculation of an overall distance measure: @van2024unbiased describe these problems and formulate a general mixed-variable distance that do not suffer from them.
The proposed unbiased mixed-variable distances have two essential
properties: *multivariate additivity* and *commensurability*. Multivariate additivity requires the overall distance to be the sum of variable-specific contributions, so that each variable enters the total additively. Additivity alone does not make the contributions comparable, since a variable
measured on a wider scale or split into more categories can still dominate the sum. Commensurability adds this missing calibration: the variable-specific distances are defined to be commensurable when the expected contribution of every variable is the same constant, so that no variable is favoured merely by its type, scale, or number of categories. In the general additive formulation, this is achieved by weighting each variable-specific dissimilarity by the reciprocal of its expected value. Together the two properties yield distances that are *unbiased* in the sense that measurement type and scale do not trivially drive a variable's influence.

Common distances such as Euclidean, Manhattan, Hamming, and Gower are independence-based: their variable-specific contributions are constructed without reference to associations among variables. When variables share information, these distances may count the same underlying contrast more than once, introducing redundancy into the distance calculation. Association-aware alternatives incorporate the joint structure of the data: for example, Mahalanobis distance accounts for covariance among numerical variables [@mahalanobis1936], whereas categorical constructions can define category dissimilarities through total variation between conditional distributions [@ahmad2007k; @vandevelden2024categorical].

The \proglang{R} package
\pkg{manydist} [@manydist2026] implements unbiased mixed-variable distances, independence-based and association-aware,
as well as other mixed-variable distances that do not necessarily fall within
the framework established in @van2024unbiased. This article focuses on the
package implementation, variable-level diagnostics and comparison tools, and
the seamless integration of distance computations into learning pipelines. The
package's main function,
\code{mdist()}, provides a unified entry point that:

- computes variable-wise dissimilarities using a selected numerical
  preprocessing and categorical dissimilarity;
- estimates variable-wise weights that enforce commensurability;
- aggregates the contributions into an additive distance matrix;
- retains the fitted preprocessing -- the weights, any rotation, and the
  estimated category dissimilarities -- so that the same distance can be
  reapplied to new observations rather than re-estimated from them.


Beyond distance construction, \pkg{manydist} provides leave-one-variable-out (LOVO)
diagnostics and pairwise comparisons of candidate distances, integration with
\pkg{tidymodels} [@tidymodels] for predictive workflows, and distance-based interfaces to
partitioning around medoids (PAM) and spectral clustering.

The remainder of the article begins with a brief overview of related
\proglang{R} packages (@sec-others). @sec-framework develops the methodological
framework, stating multivariate additivity and commensurability and deriving
the weights that equalize expected variable contributions in the numerical and
categorical blocks. @sec-construction introduces the package interface, centred
on `mdist()` and the `MDist` object that stores the fitted preprocessing.
@sec-diagnostics covers leave-one-variable-out diagnostics and comparisons of
candidate distances. @sec-pipelines shows how distances can be embedded in
\pkg{tidymodels} workflows, illustrated by unsupervised clustering and
supervised classification of World Development Indicators data.
@sec-conclusion concludes the paper.



## Other software {#sec-others}

Several \proglang{R} packages provide distances for homogeneous numerical, categorical,
or binary data. @tbl-related-homogeneous summarizes commonly used tools that
serve as building blocks for clustering, multidimensional scaling, and
nearest-neighbour methods.

<!-- \pkg{manydist} builds on rather than replaces these tools, computing comparable per-block dissimilarities and adding a commensurability layer they do not provide. -->

```{=latex}
\begingroup
\makeatletter
\let\paperLongtableCaption\LT@caption
\def\LT@caption{\hline\paperLongtableCaption}
\makeatother
```

```{r}
#| label: tbl-related-homogeneous
#| tbl-cap: "R packages providing distances or dissimilarities for homogeneous numerical, categorical, or binary data."
#| tbl-colwidths: [35, 65]
#| echo: false

related_software <- readr::read_csv(
  file.path("data", "related-software.csv"),
  show_col_types = FALSE
)

format_related_software <- function(group) {
  output <- related_software |>
    dplyr::filter(.data$group == .env$group) |>
    dplyr::transmute(
      Software = paste0(
        .data$package,
        "::",
        .data[["function"]],
        "()"
      ),
      Description = .data$description,
      .citation_key = .data$citation_key
    )

  if (knitr::is_latex_output()) {
    middle_cell <- function(value, proportion) {
      width <- paste0(
        "(\\linewidth - 2\\tabcolsep) * ",
        "\\real{", proportion, "}"
      )
      paste0(
        "\\multicolumn{1}{>{\\raggedright\\arraybackslash}m{",
        width,
        "}}{",
        value,
        "}"
      )
    }

    software_labels <- dplyr::if_else(
      stringr::str_detect(output$Software, "_"),
      paste0("\\detokenize{", output$Software, "}"),
      output$Software
    )

    software_entries <- paste0(
      "\\shortstack[l]{",
      software_labels,
      "\\\\[-2pt]{\\footnotesize \\citep{",
      output$.citation_key,
      "}}}"
    )

    output <- output |>
      dplyr::mutate(
        Software = middle_cell(software_entries, "0.3500"),
        Description = middle_cell(.data$Description, "0.6500"),
        Software = dplyr::if_else(
          dplyr::row_number() == 1L,
          .data$Software,
          paste0("\\hline ", .data$Software)
        )
      )
  } else {
    output <- output |>
      dplyr::mutate(
        Software = paste0(
          .data$Software,
          " [@",
          .data$.citation_key,
          "]"
        )
      )
  }

  output |>
    dplyr::select(-.data$.citation_key)
}

knitr::kable(
  format_related_software("homogeneous")
)
```

```{=latex}
\endgroup
```

For numerical data, \pkg{stats::dist()} supplies the standard metrics, while
\pkg{proxy::dist()} offers an extensible catalogue and user-defined measures.
\pkg{parallelDist::parDist()} and \pkg{Rfast::Dist()} target larger numerical problems with
parallel or compiled implementations. For categorical and binary data,
\pkg{klaR::kmodes()} uses matching dissimilarities, \pkg{arules::dissimilarity()} supports
sparse transaction data, and \pkg{nomclust::nomclust()} provides frequency-weighted
similarities.

Packages for mixed data are summarized in @tbl-related-mixed and fall into three broad families. The first comprises implementations of Gower's
coefficient: \pkg{cluster::daisy()} (`gower` is one of the implemented metrics), \pkg{FD::gowdis()}, \pkg{StatMatch::gower.dist()}, and \pkg{gower::gower\_dist()}. These are variants of the same underlying construction rather than independent distance measures, sharing its range-standardization scheme and its tendency to give relatively greater influence to categorical variables when many categorical attributes are present or when standardized numerical differences are comparatively small. The second family controls the relative influence of numerical and categorical information through explicit weighting. \pkg{clustMixType::kproto()} combines numerical and categorical dissimilarities within the k-prototypes algorithm [@Huang1998kprototypes], using a weighting parameter $\lambda$ that may be specified or estimated automatically, while \pkg{kmed::distmix()} implements several established mixed-data distances, including the Gower, Podani, Wishart, Huang, Harikumar, and Ahmad formulations. The third family learns an adaptive mixed-data distance from the data. Related to this family, \pkg{kamila::kamila()} balances the numerical and categorical contributions through a weighting estimated from the data rather than specified in advance, and \pkg{clustrd::cluspcamix()} couples dimension reduction to the clustering objective so that the representation and the partition are chosen together. Both address the balance between variable types as part of a clustering procedure, whereas the constructions developed here fix it in the distance itself, before any algorithm is applied. \pkg{kdml::dkss()} constructs a kernel-based distance in which variable-specific bandwidths are estimated from the data and determine each variable's contribution to the resulting metric.

Many existing interfaces for data-adaptive distances compute the distance
directly from all observations supplied to them and do not expose separate
fitting and application operations. By contrast, for supported distance
specifications, \pkg{manydist} explicitly separates these operations. Within
each resample, the numerical preprocessing, category dissimilarities, and
commensurability weights are estimated using only the training data and then
held fixed when computing test-to-training distances. Estimating these
quantities once on the complete data set before resampling would allow the test
observations to influence the learned distance and would therefore introduce
data leakage. @sec-pipelines illustrates this fit-and-apply procedure.

```{=latex}
\begingroup
\makeatletter
\let\paperLongtableCaption\LT@caption
\def\LT@caption{\hline\paperLongtableCaption}
\makeatother
```

```{r}
#| label: tbl-related-mixed
#| tbl-cap: "R packages implementing distances or dissimilarities for mixed numerical and categorical data."
#| tbl-colwidths: [35, 65]
#| echo: false

knitr::kable(
  format_related_software("mixed")
)
```

```{=latex}
\endgroup
```

The packages reviewed above provide a broad collection of
mixed-data distances and specialized distance-based methods: \pkg{manydist}  complements this ecosystem by treating distance construction as a fitted and
inspectable component of an analysis. It combines additive and commensurable
distances with association-aware constructions, diagnostics
for examining variable contributions and comparing candidate distances, and
interfaces for clustering and predictive modeling. For data-adaptive
specifications used within resampling, the numerical preprocessing, category
dissimilarities, and commensurability weights are estimated from the training
data and then held fixed when computing test-to-training distances.

The \pkg{manydist} package is available from the Comprehensive R Archive Network (CRAN)
at \url{https://CRAN.R-project.org/package=manydist}. The package website,
\url{https://alfonsoiodicede.github.io/manydist_package/}, provides reference
documentation, supporting articles, and reproducible versions of the examples
presented below.


## A unified mixed-variable distance framework {#sec-framework}

We follow the notation of @van2024unbiased. Let $I$ be the number of
observations and let $Q=Q_{n}+Q_{c}$ be the number of variables, partitioned
into $Q_{n}$ numerical and $Q_{c}$ categorical variables. Observed mixed, numerical, and categorical vectors
are denoted by $\mathbf{x}$, $\mathbf{x}^n$, and $\mathbf{x}^c$; their random
counterparts use $\bm{X}$. The symbol $\mathbf{X}$ is reserved for the
$I\times Q$ data matrix. 


For observed vectors $\mathbf{x}_i$ and $\mathbf{x}_l$, manydist constructs an
overall distance by combining contributions from the numerical and categorical
blocks. Using variable-wise dissimilarities $\delta_j$ and weights $w_j$, the
general setup is

$$
\begin{aligned}
d(\mathbf{x}_i,\mathbf{x}_l)
&=\sum_{j=1}^{Q}d_j(\mathbf{x}_i,\mathbf{x}_l)\\
&=\sum_{h=1}^{Q_{n}}w_h\,\delta_h^n(\mathbf{x}_i^n,\mathbf{x}_l^n)
 +\sum_{k=1}^{Q_{c}}w_k\,\delta_k^c(\mathbf{x}_i^c,\mathbf{x}_l^c).
\end{aligned}
$$ {#eq-mixed-distance}

The numerical contribution indexed by $h$ can depend on the complete
numerical block, and the categorical contribution indexed by $k$ can depend
on the complete categorical block. The present formulation does not use
cross-type associations inside these component dissimilarities.

```{=latex}
\begingroup
\color{blue}
```

For $Q$-dimensional observations
$\mathbf{x}_i=(x_{i1},\ldots,x_{iQ})$ and
$\mathbf{x}_l=(x_{l1},\ldots,x_{lQ})$, a
function $d:\mathcal{X}^Q\times\mathcal{X}^Q\rightarrow[0,\infty)$ is
*multivariate additive* if

$$
d(\mathbf{x}_i,\mathbf{x}_l)
=\sum_{j=1}^{Q}d_j(\mathbf{x}_i,\mathbf{x}_l),
$$ {#eq-additivity}

where $d_j(\mathbf{x}_i,\mathbf{x}_l)$ is the contribution associated with
variable $j$.
Each component may depend on all elements of the two observations; the index
identifies the associated contribution rather than restricting it to a single
coordinate. This permits additive constructions that incorporate dependencies
between variables. If every $d_j$ is a metric on $\mathcal{X}^Q$, their sum is
also a metric [see @van2024unbiased for the proof].

For the numerical block, the additive construction uses weighted Manhattan
contributions on possibly transformed coordinates:

$$
d_{\mathrm{num}}(\mathbf{x}_i,\mathbf{x}_l)
=\sum_{h=1}^{Q_{n}}w_h
\left|f_h(\mathbf{x}_i)-f_h(\mathbf{x}_l)\right|,
$$ {#eq-numerical-distance}

where $f_h$ may be a univariate or multivariate transformation.
<!-- Standardizing -->
<!-- numerical variables beforehand is therefore unnecessary for the commensurable -->
<!-- constructions, though it remains useful for the non-commensurable baselines. -->

Common transformations of the starting values may be required to deal with different measurement scales:

- **standard-deviation scaling:** $(x-\bar{x})/s_x$, which uses all
  observations rather than only the extremes;
- **range scaling:** $(x-x_{\min})/(x_{\max}-x_{\min})$, as in Gower's
  distance, which preserves the distributional shape but, for the
  non-commensurable presets, is sensitive to outliers through the extremes;
- **robust scaling:** $(x-x_{\mathrm{Med}})/(x_{Q_3}-x_{Q_1})$, which replaces
  the mean and range with their robust counterparts and so resists outliers in those same presets.

For categorical variable $k$ with $q_k$ categories, let
$\mathbf{z}_{ik}$ denote the $q_k$-dimensional indicator vector identifying
the category observed for individual $i$. This vector contains one entry equal
to one, corresponding to the observed category, and all remaining entries equal
to zero. Stacking these vectors gives the $I \times q_k$ indicator matrix
$\mathbf{Z}_k$.

Let $\bm{\Delta}_k$ be a $q_k \times q_k$ matrix whose entry
$\delta_{k,ab}$ specifies the dissimilarity between categories $a$ and
$b$. Typically, $\delta_{k,aa}=0$ and
$\delta_{k,ab}=\delta_{k,ba}$. The observation-level dissimilarity matrix for
variable $k$ is then

\begin{equation}
\mathbf{D}_k
=
\mathbf{Z}_k \bm{\Delta}_k \mathbf{Z}_k^{\top}.
\label{eq:categorical-distance-matrix}
\end{equation}

Its $(i,l)$th entry is

\begin{equation}
[\mathbf{D}_k]_{il}
=
\mathbf{z}_{ik}^{\top}
\bm{\Delta}_k
\mathbf{z}_{lk}
=
\delta_{k,ab},
\qquad
\text{when } x_{ik}=a \text{ and } x_{lk}=b.
\label{eq:categorical-distance-entry}
\end{equation}

Because each indicator vector selects exactly one row and one column of
$\bm{\Delta}_k$, the $(i,l)$th entry of $\mathbf{D}_k$ is simply the
dissimilarity assigned to the categories held by observations $i$ and $l$.
This representation separates category membership, encoded by
$\mathbf{Z}_k$, from the rule used to compare categories, encoded by
$\bm{\Delta}_k$. For simple matching, all distinct categories have
dissimilarity one; in association-aware constructions, the entries of
$\bm{\Delta}_k$ are instead estimated from relationships with other variables
or with a response.

Consequently, the additive contribution of the complete categorical block is

$$
\begin{aligned}
d_{\mathrm{c}}(\mathbf{x}_i,\mathbf{x}_l)
&=\sum_{k=1}^{Q_{c}}w_k\,
\delta_k^c(\mathbf{x}_i^c,\mathbf{x}_l^c)\\
&=\sum_{k=1}^{Q_{c}}w_k\,
\mathbf{z}_{ik}^{\top}\bm{\Delta}_k\mathbf{z}_{lk}.
\end{aligned}
$$ {#eq-categorical-distance}

```{=latex}
\endgroup
```



### Commensurability

We call a mixed-variable distance *unbiased* when the contribution associated
with each variable does not depend on its measurement type or scale, that is, it is multivariate additive and *commensurabile*.





Let $\bm{X}_i=(X_{i1},\ldots,X_{iQ})$ denote the random vector
corresponding to observation $i$. The variable contributions are *commensurable*
when, for every $j$ and $i\ne l$,

$$
\mathbb{E}\!\left[d_j(\bm{X}_i,\bm{X}_l)\right]=c,
$$ {#eq-commensurability}

where the particular value of the constant $c$ is unimportant. Additivity
provides an interpretable decomposition; commensurability places those
contributions on equivalent expected scales.

For @eq-mixed-distance, commensurability follows from the population weights

$$
w_h=
\frac{1}{
  \mathbb{E}[\delta_h^n(\bm{X}_i^n,\bm{X}_l^n)]
},
\qquad
w_k=
\frac{1}{
  \mathbb{E}[\delta_k^c(\bm{X}_i^c,\bm{X}_l^c)]
}.
$$ {#eq-population-weights}

In practice, manydist estimates these expectations from the observed pairs:

$$
\begin{aligned}
m_h&=\frac{1}{I(I-1)}\sum_{i\ne l}
\delta_h^n(\mathbf{x}_i^n,\mathbf{x}_l^n),
&\quad w_h&=\frac{1}{m_h},\\
m_k&=\frac{1}{I(I-1)}\sum_{i\ne l}
\delta_k^c(\mathbf{x}_i^c,\mathbf{x}_l^c),
&\quad w_k&=\frac{1}{m_k}.
\end{aligned}
$$ {#eq-empirical-weights}

Thus, each variable-wise distance is divided by its empirical mean before the
contributions are combined.
Because commensurability divides each variable by its own
empirical mean distance, any constant linear rescaling of a numerical variable
cancels from the result: standard-deviation scaling and no scaling at all yield
the same commensurable distance. Only a transformation that changes the shape of
the marginal distribution -- rather than its location or spread -- alters the
empirical mean distance and hence the commensurable contribution.

\new{TO BE REMOVED The formulation also permits constrained weights, such as common weights within numerical or categorical groups, when only within-type commensurability is required. Under such constrained weights the per-variable cancellation described below no longer applies, and the numerical scaling choice again affects the result.}

Under commensurability the  transformations introduced in @sec-framework  are vacuous. Each is a location-scale map $f_h(x)=(x-a_h)/b_h$, so the offset cancels in the
difference and the scale factor cancels between
$\left|f_h(\mathbf{x}_i)-f_h(\mathbf{x}_l)\right|$ and $m_h$; the three yield
identical distances, and identical to the untransformed variables. The scaling
choice is therefore consequential only for the non-commensurable presets.


For categorical variable $k$, let $\bm{p}_k$ denote the vector of category
probabilities. Its expected unweighted contribution is

$$
\mathbb{E}[\delta_k^c(\bm{X}_i^c,\bm{X}_l^c)]
=\bm{p}_k^\top\bm{\Delta}_k\bm{p}_k,
$$ {#eq-categorical-mean}

which depends on the number and prevalence of categories and on
$\bm{\Delta}_k$. For simple matching, for example,
$\bm{\Delta}_k=\mathbf{1}\mathbf{1}^\top-\mathbf{I}$ and the expected
contribution reduces to $1-\bm{p}_k^\top\bm{p}_k$, which for equally likely
categories equals $(q_k-1)/q_k$. This increases toward one as the number of
categories grows, so a matching variable with many levels contributes more than
one with few -- the categorical counterpart of the scale sensitivity seen for
numerical variables. The weights in @eq-empirical-weights restore
commensurability.

### Association-aware distances

Commensurability corrects differences in the expected magnitude of the
variable contributions, but it does not by itself account for information
shared by several variables. The general formulation above already permits
such associations: a contribution indexed by variable $j$ need not be a
function of $x_{ij}$ and $x_{lj}$ alone, but may depend on the complete
same-type vectors $\mathbf{x}_i^n,\mathbf{x}_l^n$ or
$\mathbf{x}_i^c,\mathbf{x}_l^c$. Association-aware constructions use this
freedom to incorporate joint structure within the numerical and categorical
blocks. The current formulation does not use associations across those two
blocks.

#### Numerical variables

For numerical variables, \pkg{manydist} first standardizes the original
coordinates so that their measurement units do not determine the subsequent
orientation. Let $\mathbf{Z}$ denote the resulting data matrix and write its
sample correlation matrix as

$$
\mathbf{R}=\mathbf{V}\bm{\Lambda}\mathbf{V}^{\top}.
$$

The ordinary principal-component scores are
$\mathbf{T}=\mathbf{Z}\mathbf{V}$. Their columns are uncorrelated, although
they are not generally independent, and have variances given by the diagonal
elements of $\bm{\Lambda}$. With $K$ retained components, the implemented
commensurable numerical contribution is

$$
d_{\mathrm{num}}^{\mathrm{PC}}(\mathbf{x}_i,\mathbf{x}_l)
=\sum_{h=1}^{K}
\frac{|t_{ih}-t_{lh}|}{m_h^{\mathrm{PC}}},
\qquad
m_h^{\mathrm{PC}}
=\frac{1}{I(I-1)}\sum_{r\ne s}|t_{rh}-t_{sh}|.
$$ {#eq-pc-commensurable}

Thus, `method_num = "pc_scores"` performs a rotation rather than an explicit
whitening transformation; `commensurable = TRUE` subsequently places the
absolute-difference contribution of every retained principal coordinate on
the same empirical mean scale. This sequence also makes explicit whitening
unnecessary. If $w_{ih}=t_{ih}/\sqrt{\lambda_h}$ denotes a whitened score,
then

$$
\frac{|w_{ih}-w_{lh}|}
     {\operatorname{mean}_{r\ne s}|w_{rh}-w_{sh}|}
=
\frac{|t_{ih}-t_{lh}|}
     {\operatorname{mean}_{r\ne s}|t_{rh}-t_{sh}|}.
$$ {#eq-whitening-cancels}

Whitening followed by component-wise commensurability therefore gives exactly
the same contribution as PCA rotation followed by commensurability.

The distinction is illustrated in @fig-independence-association using two
strongly correlated measurements from the Palmer penguins data
[@palmerpenguins].

```{r, echo=FALSE}
#| label: prepare-penguins

penguins <- palmerpenguins::penguins |>
  dplyr::select(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  tidyr::drop_na()

penguins_x <- penguins |>
  dplyr::select(-species)
penguins_y <- penguins |>
  dplyr::pull(species)
```

```{r}
#| label: fig-independence-association
#| fig-cap: "Scale correction and association awareness. Raw Manhattan distance is dominated by body mass. Commensurating the original-coordinate contributions removes the scale difference but retains the correlated axes, so shared body-size variation enters through both variables. Rotating standardized data to principal coordinates and commensurating their contributions instead expresses differences along uncorrelated directions, as used by `method_num = \"pc_scores\"` with `commensurable = TRUE`."
#| fig-subcap:
#|   - "Raw coordinates ($r = 0.87$)."
#|   - "Commensurable original coordinates."
#|   - "Commensurable principal coordinates."
#| layout: [[1], [1, 1]]
#| fig-width: 4.5
#| fig-height: 2.6
#| fig-pos: "tbp"
#| echo: false
#| warning: false

dependence_data <- penguins |>
  dplyr::transmute(
    penguin_id = dplyr::row_number(),
    flipper_length_mm,
    body_mass_g,
    flipper_z = as.numeric(scale(.data$flipper_length_mm)),
    body_mass_z = as.numeric(scale(.data$body_mass_g))
  )

standardized_data <- dependence_data |>
  dplyr::select(.data$flipper_z, .data$body_mass_z) |>
  as.matrix()

pca_fit <- stats::prcomp(
  standardized_data,
  center = FALSE,
  scale. = FALSE
)

mean_absolute_pair_distance <- function(x) {
  mean(abs(outer(x, x, "-")))
}

commensurate_columns <- function(x) {
  component_means <- apply(x, 2, mean_absolute_pair_distance)
  sweep(x, MARGIN = 2, STATS = component_means, FUN = "/")
}

commensurable_original <- commensurate_columns(standardized_data)
commensurable_pc <- commensurate_columns(
  pca_fit$x[, 1:2, drop = FALSE]
)

dependence_data <- dependence_data |>
  dplyr::mutate(
    pc1 = pca_fit$x[, 1],
    pc2 = pca_fit$x[, 2],
    flipper_comm = commensurable_original[, 1],
    body_mass_comm = commensurable_original[, 2],
    pc1_comm = commensurable_pc[, 1],
    pc2_comm = commensurable_pc[, 2]
  )

pc1_targets <- stats::quantile(
  dependence_data$pc1,
  probs = c(0.2, 0.8)
)

selected_ids <- purrr::map_int(pc1_targets, function(target) {
  dependence_data |>
    dplyr::mutate(target_distance = abs(.data$pc1 - target)) |>
    dplyr::slice_min(
      order_by = .data$target_distance,
      n = 20,
      with_ties = FALSE
    ) |>
    dplyr::slice_min(
      order_by = abs(.data$pc2),
      n = 1,
      with_ties = FALSE
    ) |>
    dplyr::pull(.data$penguin_id)
})

selected_points <- dependence_data |>
  dplyr::filter(.data$penguin_id %in% selected_ids) |>
  dplyr::arrange(.data$flipper_z) |>
  dplyr::mutate(point = c("A", "B"))

raw_segment <- selected_points |>
  dplyr::summarise(
    x = dplyr::first(.data$flipper_length_mm),
    y = dplyr::first(.data$body_mass_g),
    xend = dplyr::last(.data$flipper_length_mm),
    yend = dplyr::last(.data$body_mass_g),
    distance = abs(.data$xend - .data$x) + abs(.data$yend - .data$y),
    xmid = (.data$x + .data$xend) / 2,
    ymid = (.data$y + .data$yend) / 2
  )

standardized_segment <- selected_points |>
  dplyr::summarise(
    x = dplyr::first(.data$flipper_comm),
    y = dplyr::first(.data$body_mass_comm),
    xend = dplyr::last(.data$flipper_comm),
    yend = dplyr::last(.data$body_mass_comm),
    distance = abs(.data$xend - .data$x) + abs(.data$yend - .data$y),
    xmid = (.data$x + .data$xend) / 2,
    ymid = (.data$y + .data$yend) / 2
  )

association_segment <- selected_points |>
  dplyr::summarise(
    x = dplyr::first(.data$pc1_comm),
    y = dplyr::first(.data$pc2_comm),
    xend = dplyr::last(.data$pc1_comm),
    yend = dplyr::last(.data$pc2_comm),
    distance = abs(.data$xend - .data$x) + abs(.data$yend - .data$y),
    xmid = (.data$x + .data$xend) / 2,
    ymid = (.data$y + .data$yend) / 2
  )

manhattan_segments <- function(segment) {
  tibble::tibble(
    x = c(segment$x, segment$xend),
    y = c(segment$y, segment$y),
    xend = c(segment$xend, segment$xend),
    yend = c(segment$y, segment$yend)
  )
}

raw_components <- manhattan_segments(raw_segment)
standardized_components <- manhattan_segments(standardized_segment)
association_components <- manhattan_segments(association_segment)

common_score_max <- ceiling(max(abs(c(
  commensurable_original,
  commensurable_pc
))))
common_score_limits <- c(-common_score_max, common_score_max)
common_score_breaks <- pretty(common_score_limits, n = 7)

distance_plot_theme <- ggplot2::theme_minimal(base_size = 8) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = 8),
    axis.text = ggplot2::element_text(size = 7),
    plot.margin = ggplot2::margin(3, 4, 3, 3)
  )

raw_plot <- ggplot2::ggplot(
  dependence_data,
  ggplot2::aes(x = .data$flipper_length_mm, y = .data$body_mass_g)
) +
  ggplot2::geom_point(size = 1.1, color = "grey35", alpha = 0.55) +
  ggplot2::stat_ellipse(
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = raw_components,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.55,
    linetype = "dashed",
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = raw_segment,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.65,
    arrow = grid::arrow(length = grid::unit(0.07, "inches"))
  ) +
  ggplot2::geom_point(
    data = selected_points,
    size = 2.1,
    color = "#A33F3F"
  ) +
  ggplot2::geom_text(
    data = selected_points,
    ggplot2::aes(label = .data$point),
    nudge_y = 170,
    size = 2.7,
    fontface = "bold"
  ) +
  ggplot2::geom_label(
    data = raw_segment,
    ggplot2::aes(
      x = .data$xmid,
      y = .data$ymid,
      label = paste0("d[1]^{raw} == ", round(.data$distance, 0))
    ),
    parse = TRUE,
    inherit.aes = FALSE,
    size = 2.3,
    linewidth = 0,
    fill = "white"
  ) +
  ggplot2::labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)"
  ) +
  distance_plot_theme

independence_plot <- ggplot2::ggplot(
  dependence_data,
  ggplot2::aes(x = .data$flipper_comm, y = .data$body_mass_comm)
) +
  ggplot2::geom_point(size = 0.65, color = "grey70") +
  ggplot2::stat_ellipse(
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = standardized_components,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.55,
    linetype = "dashed",
    color = "#A33F3F"
  ) +
  ggplot2::geom_segment(
    data = standardized_segment,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.65,
    arrow = grid::arrow(length = grid::unit(0.07, "inches"))
  ) +
  ggplot2::geom_point(
    data = selected_points,
    size = 1.7,
    color = "#A33F3F"
  ) +
  ggplot2::geom_text(
    data = selected_points,
    ggplot2::aes(label = .data$point),
    nudge_y = 0.2,
    size = 2.7,
    fontface = "bold"
  ) +
  ggplot2::geom_label(
    data = standardized_segment,
    ggplot2::aes(
      x = .data$xmid,
      y = .data$ymid,
      label = paste0("d[1]^{ind} == ", round(.data$distance, 2))
    ),
    parse = TRUE,
    inherit.aes = FALSE,
    size = 2.3,
    linewidth = 0,
    fill = "white"
  ) +
  ggplot2::scale_x_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::scale_y_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    x = "Flipper contribution",
    y = "Mass contribution"
  ) +
  distance_plot_theme

association_plot <- ggplot2::ggplot(
  dependence_data,
  ggplot2::aes(x = .data$pc1_comm, y = .data$pc2_comm)
) +
  ggplot2::geom_point(size = 0.65, color = "grey70") +
  ggplot2::stat_ellipse(
    type = "norm",
    level = 0.68,
    linewidth = 0.5,
    color = "#2A7F62"
  ) +
  ggplot2::geom_segment(
    data = association_components,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.55,
    linetype = "dashed",
    color = "#2A7F62"
  ) +
  ggplot2::geom_segment(
    data = association_segment,
    ggplot2::aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend),
    inherit.aes = FALSE,
    linewidth = 0.65,
    arrow = grid::arrow(length = grid::unit(0.07, "inches"))
  ) +
  ggplot2::geom_point(
    data = selected_points,
    size = 1.7,
    color = "#2A7F62"
  ) +
  ggplot2::geom_text(
    data = selected_points,
    ggplot2::aes(label = .data$point),
    nudge_y = 0.2,
    size = 2.7,
    fontface = "bold"
  ) +
  ggplot2::geom_label(
    data = association_segment,
    ggplot2::aes(
      x = .data$xmid,
      y = .data$ymid,
      label = paste0("d[1]^{PC} == ", round(.data$distance, 2))
    ),
    parse = TRUE,
    inherit.aes = FALSE,
    size = 2.3,
    linewidth = 0,
    fill = "white"
  ) +
  ggplot2::scale_x_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::scale_y_continuous(
    limits = common_score_limits,
    breaks = common_score_breaks,
    expand = ggplot2::expansion(mult = 0)
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    x = "PC1 contribution",
    y = "PC2 contribution"
  ) +
  distance_plot_theme

raw_plot
independence_plot
association_plot
```


The first two panels of @fig-independence-association separate scale correction
from association awareness. Commensurability equalizes the two marginal
contributions in the middle panel, but their strong correlation remains. The
last panel applies @eq-pc-commensurable: PCA fixes an orientation in which the
coordinates are uncorrelated, and commensurability then equalizes their mean
absolute-difference contributions. Unlike Euclidean distance after whitening,
which equals Mahalanobis distance, this is an additive Manhattan distance in
the principal-component orientation. It is related to the Mahalanobis idea of
using covariance structure, but it is not Mahalanobis distance.

Equalizing the retained principal-coordinate contributions necessarily gives
more relative weight to directions with little variation. This is useful when
small departures from a strong relationship are informative, but it can also
amplify low-variance noise. Explained variance is therefore not itself a
measure of substantive importance. By default all available components are
retained; `ncomp` or `threshold` can restrict the construction to a leading
subspace when dimensionality reduction is justified. Such restriction is a
modeling decision: dropping components also reduces the expected total
contribution of the numerical block relative to the categorical block.

#### Categorical variables

For categorical variables, association enters through conditional
distributions rather than a covariance matrix. The category-dissimilarity
matrix provides a common representation for independence- and
association-based constructions, following @vandevelden2024categorical.
Indicator-based dissimilarities instead treat a binary indicator representation
as numerical; without scaling, Manhattan distance gives values zero or two and
ignores both category prevalence and the number of categories.

For an association-based construction, let
$\mathbf{R}^{k\mid k'}=(\mathbf{Z}_k^\top\mathbf{Z}_k)^{-1}
\mathbf{Z}_k^\top\mathbf{Z}_{k'}$ be the conditional distribution of variable
$k'$ given the categories of variable $k$. Dissimilarities between two
categories $a$ and $b$ of variable $k$ compare the corresponding rows
$\mathbf{r}_a$ and $\mathbf{r}_b$, using total variation [@ahmad2007k] or a
symmetric Kullback--Leibler construction [@le2005association]. Under
independence, every category of $k$ induces the same conditional distribution
of $k'$, so the rows coincide and the association-based category
dissimilarities vanish. They increase as the categories become more distinct
in what they imply about the other categorical variables.

This construction is conceptually different from PCA rotation: it changes the
dissimilarities between categories rather than changing the coordinate system.
Commensurability subsequently divides each categorical-variable contribution
by its empirical mean, placing the association-based categorical contributions
on the same expected scale as the commensurable principal-coordinate
contributions. Without a response, the construction requires at least two
categorical variables; with only one, the fitted specification falls back to
simple matching.

#### Response-aware categorical distances

The same conditional-distribution construction accommodates an outcome, as
proposed and evaluated by @vandevelden2024categorical. When a response $y$ is
supplied, $\mathbf{R}^{k\mid y}$ records its distribution within each category
of predictor $k$. Two categories are then dissimilar to the extent that they
imply different response distributions. This is the response-awareness used by
the `response` argument of `mdist()`, and it is the only route by which an
outcome enters a distance in \pkg{manydist}. It also makes a response-aware
categorical dissimilarity available with a single categorical predictor. The
numerical block currently has no response-aware counterpart: its PCA rotation
uses predictor associations only.

```{=latex}
\Needspace{8\baselineskip}
```

## Distance construction {#sec-construction}

The \pkg{manydist} interface separates the specification of a distance from its subsequent diagnostic and modeling uses. This section introduces the construction of distances with `mdist()`; the following sections build diagnostics and learning pipelines on the resulting objects. The examples continue with the six-predictor Palmer penguins data prepared above.

```{r}
#| label: attach-manydist

library("manydist")
```

The central function `mdist()` constructs distances for numerical, categorical, or mixed-type data. At its most basic, a distance is specified through the data and three components:

* `method_num`, which determines the preprocessing of the numerical variables;
* `method_cat`, which determines the dissimilarity used for categorical variables; and
* `commensurable`, which determines whether the component contributions are rescaled to have equal expected contributions before they are combined.


The principal arguments are summarized below. The default specification uses standard-deviation scaling for the numerical variables, total-variation dissimilarities for the categorical variables, and commensurability.

| Argument | Role | Default | Main options |
|---|---|---|---|
| `x` | Numerical, categorical, or mixed-type data | required | --- |
| `method_num` | Numerical preprocessing | `"std"` | `"none"`, `"std"`, `"range"`, `"robust"`, `"pc_scores"` |
| `method_cat` | Categorical dissimilarity | `"tvd"` | `"matching"`, `"tvd"`, and further association-, independence-, and indicator-based methods |
| `commensurable` | Equalize expected component contributions | `TRUE` | `TRUE`, `FALSE` |

: Principal arguments of the clustering function {#tbl-args tbl-colwidths="[18,22,10,50]"}

The complete set of categorical methods can be inspected with `dist_methods_tbl_cat()`. Additional arguments control particular constructions, such as the number of retained principal components when `method_num = "pc_scores"`; these are documented in the package reference.

Because `preset = "custom"` is the default, these components can be specified directly without explicitly requesting a custom preset. For example,

```{r}
#| label: build-custom-distance

d_custom <- mdist(
  penguins_x,
  method_cat = "matching",
  method_num = "pc_scores",
  commensurable = TRUE
)
```

constructs a commensurable additive distance using matching dissimilarities for the categorical variables and principal-component preprocessing for the numerical block.

For the commensurable constructions, any per-variable affine rescaling of a numerical variable cancels when its contribution is divided by its empirical mean distance. Consequently, `"std"`, `"range"`, and `"robust"` produce the same distance when `commensurable = TRUE`; their choice matters for non-commensurable constructions. Multivariate transformations such as `"pc_scores"` are different because they alter the coordinate system rather than simply rescaling variables individually.

```{r}
#| label: verify-scaling-vacuity

purrr::map(
  c("std", "range", "robust"),
  ~ mdist(
    penguins_x,
    method_num = .x,
    method_cat = "matching",
    commensurable = TRUE
  )$to_dist()
) |>
  (\(d) c(
    all.equal(d[[1]], d[[2]]),
    all.equal(d[[1]], d[[3]])
  ))()
```

### Distance presets

For commonly used specifications, `mdist()` provides named presets. A preset fixes the relevant components and, where necessary, any additional computation required by that distance. The presets serve two purposes. The `u_` family provides convenient, reproducible specifications within the unbiased-distance framework developed above. The remaining presets provide familiar or recently proposed comparison distances. Some of these use constructions that cannot be expressed solely as a combination of `method_num`, `method_cat`, and `commensurable`, and therefore require dedicated preset implementations.

@tbl-presets summarizes the presets used in this article.

| Preset | Numerical preprocessing | Categorical dissimilarity | Additive | Unbiased | Response-aware |
|:---|:---|:---|:--:|:--:|:--:|
| `gower` | range scaling to unit interval | simple matching | yes | no | no |
| `euclidean` | standard-deviation scaling | indicator (one-hot) coding | no | no | no |
| `hl` | standard-deviation scaling | Hennig--Liao indicator scaling | no | no | no |
| `mod_gower` | modified Gower [@liu2024] | -- | yes | no | no |
| `gudmm` | generalized multi-aspect metric [@mousavi2023] | -- | no | no | no |
| `dkss` | kernel product similarity [@ghashti2024] | -- | no | no | no |
| `u_indep` | standard-deviation scaling | simple matching | yes | yes | no |
| `u_dep` | principal-component scaling | total variation distance | yes | yes | yes |
| `u_mix` | standard-deviation scaling | total variation distance | yes | yes | yes |

: Distance presets used in this article. Additive distances satisfy multivariate additivity and unbiased distances additionally enforce commensurability. Response-awareness affects the categorical dissimilarity only and applies when an outcome is supplied; otherwise the corresponding presets reduce to their predictor-only form. Because commensurability cancels per-variable affine scaling, the numerical preprocessing listed for `u_indep` and `u_mix` does not affect the resulting distance. Principal-component preprocessing in `u_dep` is multivariate and therefore does affect the distance. {#tbl-presets tbl-colwidths="[12,28,26,10,10,14]"}

For example, `u_dep` combines principal-component preprocessing of the numerical block, association-based categorical dissimilarities, and commensurability weights:

```{r}
#| label: build-u-dep-distance

d_u_dep <- mdist(
  penguins_x,
  preset = "u_dep"
)

d_u_dep
```

The result is an `MDist` object that contains the computed dissimilarity together with the fitted distance specification and preprocessing information. In particular, it retains the commensurability weights, any principal-component rotation, and the estimated category dissimilarities. Retaining these quantities allows the same fitted distance construction to be applied to new observations rather than re-estimated from them. The standard R `dist` representation is returned by `to_dist()`.

```{r}
#| label: inspect-u-dep-distance

as.matrix(d_u_dep$to_dist())[1:4, 1:4] |>
  round(digits = 2)
```

The basic construction can also be extended with outcome information. The `response` argument supplies an outcome without treating it as an ordinary predictor. For response-aware specifications, the outcome shapes the categorical dissimilarities and therefore produces task-specific neighbourhoods. The numerical preprocessing remains predictor-only. For specifications that are not response-aware, the argument is ignored.

```{r}
#| label: build-response-aware-distance

d_response <- mdist(
  penguins,
  response = species,
  preset = "u_dep"
)
```

Finally, the `new_data` argument computes rectangular new-to-training dissimilarities. New observations are measured using the preprocessing fitted to the training data -- including the stored weights, rotation, and category dissimilarities -- rather than having these quantities re-estimated from the new observations. This fitted construction is essential in predictive settings and forms the basis of the resampling workflow developed in @sec-pipelines.


## Distance diagnostics and benchmarking {#sec-diagnostics}

With several plausible distance specifications available for mixed data, it can be informative to examine how the resulting dissimilarities depend on the variables included and how differences between distances propagate to subsequent analyses. \pkg{manydist} provides diagnostics at two levels. The first compares dissimilarity matrices directly. The second considers their consequences for particular downstream procedures, namely multidimensional scaling (MDS) and clustering. These levels answer different questions: the former concerns the distance itself, whereas the latter is conditional on the subsequent analysis being considered.

### Leave-one-variable-out diagnostics

`lovo_mdist()` implements leave-one-variable-out (LOVO) diagnostics for a single distance specification. It first computes the distance using all variables and then removes each variable in turn, recomputes the distance, and quantifies the resulting change.

```{r}
#| label: diagnose-gower-distance

lovo_gower <- lovo_mdist(
  penguins_x,
  preset = "gower"
)
```

The LOVO diagnostics can be read in two stages. The direct distance diagnostics quantify changes in the dissimilarity matrix itself:

- `mad_importance` is the mean absolute difference between the full dissimilarity matrix and the matrix obtained after leaving out a variable.
- `relative_distance` is the normalized version of `mad_importance`, allowing the relative LOVO contributions of the variables to be compared on a common scale.

The downstream diagnostics instead quantify what the change in distance does to a particular subsequent analysis:

- `mds_congruence` (equivalently, `cc_importance`) and `ac_importance` compare classical MDS representations of the full and reduced distances.
- When clustering is requested, `pam_importance`, `hclust_importance`, or `spectral_importance` measure the change in the corresponding clustering partition as one minus the adjusted Rand index (ARI).

We first consider the direct distance diagnostics.

```{r}
#| label: inspect-gower-distance-diagnostics

lovo_gower$results |>
  dplyr::select(
    variable,
    mad_importance,
    relative_distance
  ) |>
  dplyr::arrange(dplyr::desc(relative_distance))
```

Figure @fig-gower-lovo displays the relative LOVO contributions for the Gower distance. It provides a direct view of which variables most affect the pairwise dissimilarities when omitted.

```{r}
#| label: fig-gower-lovo
#| fig-cap: "Relative leave-one-variable-out contributions under Gower distance. Larger values indicate a larger relative change in the pairwise dissimilarities when the variable is omitted."
#| fig-width: 6
#| fig-height: 3.2

ggplot2::autoplot(
  lovo_gower,
  metric = "relative_distance",
  reorder = TRUE
)
```

LOVO can also describe the consequences of removing a variable for a subsequent distance-based analysis. For MDS, `lovo_mdist()` applies classical multidimensional scaling to the full and leave-one-variable-out dissimilarity matrices. By default, two-dimensional configurations are used, with the number of dimensions controlled by `dims`. The congruence coefficient compares the interpoint distances of the resulting configurations. Its corresponding alienation coefficient is

$$
a_j = \sqrt{1-c_j^2},
$$

where $c_j$ is the congruence coefficient between the full and leave-one-variable-out MDS configurations. Alienation close to zero indicates highly congruent configurations, whereas larger values indicate a greater change in that particular MDS representation.

These MDS diagnostics should be distinguished from `mad_importance` and `relative_distance`. A variable can change the original dissimilarity matrix substantially without producing a comparable change in a low-dimensional MDS representation, or conversely. The quantities do not provide competing estimates of the same effect: they describe different objects. MDS-based diagnostics are therefore most relevant when such a representation is itself of interest.

Clustering provides another optional downstream diagnostic. When `cluster_k` is supplied, `lovo_mdist()` applies the specified clustering method to the full and reduced distances and compares the resulting partitions using ARI. The corresponding LOVO clustering importance is defined as $1-\mathrm{ARI}$, so zero indicates an unchanged partition and larger values indicate a greater change after the variable is omitted. For example, PAM-based diagnostics for a three-cluster solution are obtained with

```{r}
#| label: diagnose-gower-pam

lovo_gower_pam <- lovo_mdist(
  penguins_x,
  preset = "gower",
  cluster_k = 3,
  cluster_methods = "pam"
)
```

As with the MDS diagnostics, clustering-based LOVO measures are conditional on the downstream procedure: their interpretation depends on the clustering method and the specified number of clusters.

### Comparing LOVO diagnostics across distance specifications

The same LOVO analysis can be repeated for several distance specifications using `compare_lovo_mdist()`. Each element of `methods` defines one distance specification, and the function applies `lovo_mdist()` separately to each one before combining the results. This makes it possible to examine whether the effect of leaving out a variable depends on how the distance itself is constructed.

Here we compare four specifications and request three-cluster PAM diagnostics:

```{r}
#| label: compare-lovo-pam

lovo_comparison <- compare_lovo_mdist(
  x = penguins_x,
  methods = list(
    Gower = list(preset = "gower"),
    Euclidean = list(preset = "euclidean"),
    `Unbiased independent` = list(preset = "u_indep"),
    `Unbiased dependent` = list(preset = "u_dep")
  ),
  cluster_k = 3,
  cluster_methods = "pam"
)
```

Figure @fig-compare-lovo-pam shows the corresponding PAM-based LOVO importance. Each value is $1-\mathrm{ARI}$ between the partition obtained from all variables and the partition obtained after leaving out the indicated variable. The comparison is therefore made *within* each distance specification: larger values indicate that omitting the variable changes its three-cluster PAM solution more strongly. No species labels are used.

```{r}
#| label: fig-compare-lovo-pam
#| fig-cap: "Effect of leaving out each variable on three-cluster PAM solutions under four distance specifications. Values are one minus the adjusted Rand index between the full and leave-one-variable-out partitions; larger values indicate a greater change in the partition."
#| fig-width: 6.3
#| fig-height: 3.6

ggplot2::autoplot(
  lovo_comparison,
  metric = "pam_importance",
  reorder = TRUE
)
```

The same comparison can instead be displayed for `mad_importance`, `relative_distance`, or the MDS-based diagnostics, depending on which aspect of the LOVO analysis is relevant.

For association-based specifications, leaving out a variable can change not only the available information but also the construction of the remaining contributions. Under `u_dep`, for example, removing either categorical variable leaves only one categorical variable. Because the association-based categorical construction requires at least two categorical variables, the reduced specification falls back to matching. LOVO results in such cases should therefore be interpreted with this change in construction in mind.

### Comparing candidate distance specifications

LOVO asks how a particular distance changes when one variable is omitted. A different question is how complete distance specifications differ when they are applied to the same data. `benchmark_mdist()` addresses this second question. It computes a set of candidate distances and then compares every pair of distances that could be computed.

Candidate specifications may be supplied directly or generated with `all_dist_method_specs()`. For example, the following code compares the Gower, unbiased independent, and unbiased dependent presets:

```{r}
#| label: benchmark-candidate-distances

candidate_specs <- all_dist_method_specs(
  mode = "presets_only",
  preset = c("gower", "u_indep", "u_dep")
)

set.seed(2026)

distance_benchmark <- benchmark_mdist(
  x = penguins_x,
  specs = candidate_specs,
  cluster_k = 3,
  cluster_methods = "pam"
)

benchmark_comparisons(distance_benchmark)
```

The comparisons again fall into direct distance measures and measures associated with particular downstream analyses.

For two candidate distances $d_a$ and $d_b$, `mad` is the mean absolute difference between their pairwise dissimilarities. Because this quantity retains the overall scale of the two distances, `relative_distance` provides the symmetric scale-adjusted comparison

$$
r(d_a,d_b)
=
\frac{
2\operatorname{mean}(|d_a-d_b|)
}{
\operatorname{mean}(|d_a|)
+
\operatorname{mean}(|d_b|)
}.
$$ {#eq-benchmark-relative-distance}

Thus, `mad` describes the absolute magnitude of the differences between the two dissimilarity matrices, whereas `relative_distance` expresses that difference relative to their overall magnitudes.

The MDS-based comparisons address a different question. Classical MDS is applied separately to the two distances, using `dims = 2` by default, and the resulting configurations are compared using the congruence coefficient and its corresponding alienation coefficient. These statistics concern similarity of the chosen MDS representations rather than direct similarity of the original dissimilarity matrices.

Clustering comparisons are optional. When `cluster_k` is specified, each requested clustering method is applied to every candidate distance, and partitions produced by the same clustering method are compared across pairs of distance specifications using ARI. An ARI of one indicates identical partitions, while lower values indicate increasingly different assignments. These values measure agreement between the resulting partitions; they do not indicate that either distance or partition is preferable.

The pairwise results can be extracted with `benchmark_comparisons()` as above or displayed using the `autoplot()` method. For example, Figure @fig-benchmark-relative-distance shows the symmetric relative-distance comparison among the three candidate specifications.

```{r}
#| label: fig-benchmark-relative-distance
#| fig-cap: "Pairwise symmetric relative distances among three candidate distance specifications. Smaller values indicate more similar dissimilarity matrices after accounting for their overall magnitude."
#| fig-width: 5.5
#| fig-height: 4.2

ggplot2::autoplot(
  distance_benchmark,
  metric = "relative_distance"
)
```

The same display can be requested for raw MAD, MDS congruence or alienation, and, when clustering has been requested, ARI.

The available diagnostics should not be interpreted as alternative estimates of a single quantity. MAD and relative distance compare dissimilarity matrices directly; congruence and alienation compare a particular classical MDS representation; and ARI compares partitions obtained from a specified clustering procedure. Which comparisons are informative therefore depends on how the distance is intended to be used. The diagnostics describe different aspects of candidate distance specifications rather than providing, by themselves, a rule for selecting a preferred distance.


## Distance-based learning pipelines {#sec-pipelines}

Distance-based methods are often described as if the distance matrix were a fixed input to the analysis. This is appropriate when the distance is fully specified in advance, but not when its construction involves quantities estimated from the data. In \pkg{manydist}, for example, commensurability weights, principal-component rotations, and some categorical dissimilarities are estimated from the observations used to fit the distance. In resampling and prediction, these quantities must therefore be estimated from the training data only and then held fixed when new observations are processed. Otherwise, information from assessment or test observations can enter the distance construction before model evaluation.

\pkg{manydist} integrates this fit-and-apply principle with \pkg{tidymodels}, allowing distance construction to become part of a learning workflow rather than a preprocessing operation performed once in advance. The interfaces considered here cover two settings: unsupervised clustering and supervised nearest-neighbour classification. MDS is also a downstream use of a distance, as considered in @sec-diagnostics, but it is not part of the learning workflows illustrated below.

### Integrating fitted distances into learning workflows

In the \pkg{tidymodels} framework, a *recipe* is an object that specifies preprocessing operations to be estimated from training data and subsequently applied to new data. `step_mdist()` adds fitted distance construction as one such preprocessing step.

`step_mdist()` estimates the distance construction from the training data and stores the quantities needed to apply it subsequently, including commensurability weights, any principal-component rotation, and estimated category dissimilarities. During resampling, these quantities are re-estimated separately within each analysis set and then held fixed when its assessment observations are processed. For response-aware specifications, the outcome from the analysis set may also enter the fitted categorical dissimilarities, while assessment and test outcomes are not used.

The representation required from the fitted distance depends on the learning task. In clustering, all observations in the current data set are compared with one another, so the downstream method uses a square pairwise distance matrix. In nearest-neighbour prediction, each new observation is compared only with the training observations, producing a rectangular new-to-training distance matrix. The `output` argument of `step_mdist()` selects between these representations: `"pairwise"` for clustering and `"distance_to_training"` for nearest-neighbour prediction. The resulting recipes can be paired with `pam_dist()` or `spectral_dist()` for clustering and with `nearest_neighbor_dist()` for nearest-neighbour learning.

The packaged `wdi_2022` data provide a fixed snapshot of five numerical World Development Indicators and three World Bank classifications for 181 countries [@worldbank2022wdi]. We use these data to illustrate both learning settings: first to compare unsupervised country partitions and then to predict World Bank income group.

```{r}
#| label: prepare-wdi-data

data("wdi_2022", package = "manydist")

wdi_clustering <- wdi_2022 |>
  dplyr::select(-country)

distance_presets <- c(
  gower = "gower",
  euclidean = "euclidean",
  u_indep = "u_indep",
  u_dep = "u_dep",
  u_mix = "u_mix"
)
```

```{r}
#| label: prepare-wdi-labels
#| echo: false

distance_labels <- c(
  gower = "Gower",
  euclidean = "Euclidean",
  u_indep = "Unbiased independent",
  u_dep = "Unbiased dependent",
  u_mix = "Unbiased mixed"
)
```

For the supervised task, countries labelled `"Not classified"` are removed. Country is an identifier, and lending category is excluded because it is administratively related to income group. The two-country North America level is pooled with Latin America and the Caribbean to reduce the risk that a categorical level is absent from an analysis fold.

```{r}
#| label: prepare-wdi-classification

wdi_classification <- wdi_2022 |>
  dplyr::filter(income != "Not classified") |>
  dplyr::mutate(
    income = forcats::fct_drop(income),
    region = forcats::fct_collapse(
      region,
      Americas = c(
        "North America",
        "Latin America & Caribbean"
      )
    )
  ) |>
  dplyr::select(-country, -lending)

wdi_classification |>
  dplyr::count(income)
```

A stratified split reserves 25% of the countries for final evaluation. The remaining training partition supplies a common set of five stratified folds for evaluating every distance and neighbourhood size.

```{r}
#| label: split-wdi-classification

set.seed(2026)

wdi_split <- rsample::initial_split(
  wdi_classification,
  prop = 0.75,
  strata = income
)

wdi_train <- rsample::training(wdi_split)

set.seed(2026)

wdi_folds <- rsample::vfold_cv(
  wdi_train,
  v = 5,
  strata = income
)
```

```{r}
#| label: prepare-neighbor-grid

neighbor_grid <- tibble::tibble(
  neighbors = c(1L, 3L, 5L, 7L, 9L, 15L)
)
```

### Why the distance must be refitted during resampling

The practical consequence of treating distance construction as a fitted preprocessing operation can be demonstrated in the supervised task. @tbl-refitting-leakage compares the same five-neighbour classifier under two cross-validation procedures for three commensurable specifications. In the first, a single distance is fitted on the complete training partition and then subsetted by fold. In the second, the distance is fitted separately within each analysis fold and then applied to that fold's assessment observations. The voting rule and resamples are otherwise identical, so the comparison isolates the effect of fitting the distance before rather than within resampling.

The intended workflow can be written directly with `step_mdist()`:

```{r}
#| label: show-honest-arm
#| eval: false

honest_cv <- function(preset, k) {
  workflows::workflow() |>
    workflows::add_recipe(
      recipes::recipe(income ~ ., data = wdi_train) |>
        step_mdist(
          recipes::all_predictors(),
          preset = preset,
          output = "distance_to_training"
        )
    ) |>
    workflows::add_model(
      nearest_neighbor_dist(mode = "classification", neighbors = k) |>
        parsnip::set_engine("manydist")
    ) |>
    tune::fit_resamples(
      resamples = wdi_folds,
      metrics = yardstick::metric_set(yardstick::bal_accuracy)
    ) |>
    tune::collect_metrics(summarize = FALSE) |>
    dplyr::filter(.metric == "bal_accuracy") |>
    dplyr::pull(.estimate)
}
```

The following code reproduces both procedures explicitly so that their difference can be quantified.

```{r}
#| label: compare-refitting-leakage
#| cache: true
#| echo: false

response_aware_presets <- c("u_dep", "u_mix")
leakage_presets <- c("u_indep", "u_mix", "u_dep")

knn_vote <- function(d_block, train_y, k) {
  apply(d_block, 1, function(row) {
    nn <- order(row)[seq_len(k)]
    names(sort(table(train_y[nn]), decreasing = TRUE))[1]
  })
}

fit_mdist <- function(data, preset, new_data = NULL) {
  args <- if (preset %in% response_aware_presets) {
    list(data, response = quote(income))
  } else {
    list(dplyr::select(data, -income))
  }
  args$preset <- preset
  if (!is.null(new_data)) {
    args$new_data <- new_data
  }
  do.call(mdist, args)
}

fold_accuracy <- function(d_block, train_y, truth, k) {
  pred <- knn_vote(d_block, train_y, k)
  yardstick::bal_accuracy_vec(
    truth = truth,
    estimate = factor(pred, levels = levels(wdi_train$income))
  )
}

leaky_cv <- function(preset, k) {
  m_all <- as.matrix(fit_mdist(wdi_train, preset)$to_dist())

  purrr::map_dbl(wdi_folds$splits, function(split) {
    fit_id <- split$in_id
    out_id <- rsample::complement(split)
    fold_accuracy(
      m_all[out_id, fit_id, drop = FALSE],
      wdi_train$income[fit_id],
      wdi_train$income[out_id],
      k
    )
  })
}

honest_cv <- function(preset, k) {
  purrr::map_dbl(wdi_folds$splits, function(split) {
    fit_data <- rsample::analysis(split)
    out_data <- rsample::assessment(split)

    d_block <- unclass(
      fit_mdist(
        fit_data,
        preset,
        new_data = dplyr::select(out_data, -income)
      )$distance
    )

    stopifnot(
      nrow(d_block) == nrow(out_data),
      ncol(d_block) == nrow(fit_data)
    )

    fold_accuracy(
      d_block,
      fit_data$income,
      out_data$income,
      k
    )
  })
}

set.seed(2026)

leakage_comparison <- tibble::tibble(preset = leakage_presets) |>
  dplyr::mutate(
    leaky = purrr::map(preset, leaky_cv, k = 5L),
    honest = purrr::map(preset, honest_cv, k = 5L),
    single_matrix = purrr::map_dbl(leaky, mean),
    refit_per_fold = purrr::map_dbl(honest, mean),
    optimism = single_matrix - refit_per_fold,
    paired_se = purrr::map2_dbl(
      leaky,
      honest,
      ~ stats::sd(.x - .y) / sqrt(length(.x))
    )
  )
```

```{=latex}
\Needspace{8\baselineskip}
```

```{r}
#| label: tbl-refitting-leakage
#| tbl-cap: "Cross-validated balanced accuracy for WDI income classification with five neighbours, computing the distance once on the training partition against refitting it on each analysis set. The standard error is that of the paired fold-wise differences."
#| echo: false

knitr::kable(
  leakage_comparison |>
    dplyr::transmute(
      Distance = unname(distance_labels[preset]),
      `Single matrix` = single_matrix,
      `Refit per fold` = refit_per_fold,
      Optimism = optimism,
      `Paired SE` = paired_se
    ),
  digits = 3
)
```

In this example, the single-matrix procedure gives higher cross-validated balanced accuracy for all three specifications, although most differences are small relative to the fold-to-fold variability. The gap is largest for `u_dep`, whose fitted preprocessing includes a principal-component rotation. The point of the example is not the magnitude for this particular data set, but that the distance contains estimated quantities and therefore belongs inside the resampling procedure.

The same issue can affect comparisons among candidate distances. @tbl-refitting-leakage-grid repeats the two procedures across the neighbourhood grid and records the rank of each specification under each approach.

```{r}
#| label: compare-refitting-leakage-grid
#| cache: true
#| warning: false
#| echo: false

set.seed(2026)

leakage_grid <- tidyr::expand_grid(
  preset = leakage_presets,
  k = neighbor_grid$neighbors
) |>
  dplyr::mutate(
    single_matrix = purrr::map2_dbl(preset, k, ~ mean(leaky_cv(.x, .y))),
    refit_per_fold = purrr::map2_dbl(preset, k, ~ mean(honest_cv(.x, .y)))
  ) |>
  dplyr::group_by(k) |>
  dplyr::mutate(
    leaky_rank = rank(-single_matrix),
    honest_rank = rank(-refit_per_fold)
  ) |>
  dplyr::ungroup()
```

```{=latex}
\Needspace{10\baselineskip}
```

```{r}
#| label: tbl-refitting-leakage-grid
#| tbl-cap: "Rank of each specification by cross-validated balanced accuracy at every neighbourhood size, under a single training-partition distance and under fold-wise refitting. The two approaches agree except at one and five neighbours."
#| echo: false

leakage_grid |>
  dplyr::mutate(Distance = unname(distance_labels[preset])) |>
  dplyr::select(Distance, k, leaky_rank, honest_rank) |>
  tidyr::pivot_longer(
    c(leaky_rank, honest_rank),
    names_to = "Approach",
    values_to = "rank"
  ) |>
  dplyr::mutate(
    Approach = dplyr::if_else(
      Approach == "leaky_rank",
      "Single matrix",
      "Refit per fold"
    )
  ) |>
  tidyr::pivot_wider(
    names_from = Distance,
    values_from = rank
  ) |>
  dplyr::arrange(k, dplyr::desc(Approach)) |>
  dplyr::rename(Neighbours = k) |>
  knitr::kable()
```

The two procedures rank the three specifications identically at four of the six neighbourhood sizes. They differ at one neighbour, where the leading two exchange places, and at five neighbours, where `u_dep` moves from first to last. Neither difference occurs at the neighbourhood size selected by any of the three specifications in this example. With five folds and 133 training countries, the comparison is illustrative; its purpose is to show that fixing an estimated distance before resampling can change both estimated performance and comparisons among candidate specifications.

### Unsupervised country clustering

The clustering workflows contain no outcome, so every candidate distance is constructed from the observed variables alone. To focus on the effects of the distance and clustering method, the number of clusters is fixed at four throughout. This value is illustrative rather than an estimated optimum.

One recipe is defined for each distance specification. Here `output = "pairwise"` instructs `step_mdist()` to provide the square distance matrix required by the clustering method.

```{r}
#| label: build-wdi-clustering-recipes

clustering_recipes <- purrr::map(
  distance_presets,
  ~ recipes::recipe(~ ., data = wdi_clustering) |>
    step_mdist(
      recipes::all_predictors(),
      preset = .x,
      output = "pairwise"
    )
)
```

Each distance is combined with partitioning around medoids (PAM) and spectral clustering. Both methods use the same pairwise representation produced by the corresponding recipe. For example, the following calls fit both clustering methods using `u_mix`.

```{r}
#| label: fit-example-wdi-clustering
set.seed(2026)

wdi_pam_fit <- generics::fit(
  pam_dist(num_clusters = 4),
  recipe = clustering_recipes$u_mix,
  data = wdi_clustering
)

wdi_spectral_fit <- generics::fit(
  spectral_dist(num_clusters = 4, nstart = 50),
  recipe = clustering_recipes$u_mix,
  data = wdi_clustering
)
```

```{r}
#| label: fit-wdi-clustering-grid
#| echo: false

fit_clustering_pipeline <- function(distance, method) {
  model <- if (method == "PAM") {
    pam_dist(num_clusters = 4)
  } else {
    spectral_dist(num_clusters = 4, nstart = 50)
  }

  fitted <- generics::fit(
    model,
    recipe = clustering_recipes[[distance]],
    data = wdi_clustering
  )

  assignment <- predict(fitted)$.pred_cluster |>
    as.integer()

  silhouette <- cluster::silhouette(
    assignment,
    stats::as.dist(fitted$train_matrix)
  )

  tibble::tibble(
    distance = distance,
    method = method,
    fit = list(fitted),
    cluster = list(assignment),
    average_silhouette = mean(silhouette[, "sil_width"])
  )
}

set.seed(2026)

clustering_results <- tidyr::expand_grid(
  distance = names(distance_presets),
  method = c("PAM", "Spectral")
) |>
  purrr::pmap_dfr(fit_clustering_pipeline) |>
  dplyr::mutate(
    distance_label = unname(distance_labels[distance])
  )
```

Average silhouette width describes separation within each fitted distance--clustering combination. Because each silhouette is calculated using the distance that produced its partition, silhouette values are interpreted within a distance rather than as a criterion for ranking different distance metrics. The ARI in @tbl-wdi-clustering-summary instead compares the PAM and spectral assignments under the same distance. Thus, silhouette concerns separation within a fitted solution, whereas ARI concerns agreement between two partitions.

```{=latex}
\Needspace{11\baselineskip}
```

```{r}
#| label: tbl-wdi-clustering-summary
#| tbl-cap: "Four-cluster WDI solutions under five distances and two clustering methods."
#| echo: false

clustering_summary <- clustering_results |>
  dplyr::select(
    distance,
    distance_label,
    method,
    average_silhouette,
    cluster
  ) |>
  tidyr::pivot_wider(
    names_from = method,
    values_from = c(average_silhouette, cluster),
    names_sep = "_"
  ) |>
  dplyr::mutate(
    method_ari = purrr::map2_dbl(
      cluster_PAM,
      cluster_Spectral,
      aricode::ARI
    )
  ) |>
  dplyr::arrange(match(distance, names(distance_presets)))

knitr::kable(
  clustering_summary |>
    dplyr::transmute(
      Distance = distance_label,
      `PAM silhouette` = average_silhouette_PAM,
      `Spectral silhouette` = average_silhouette_Spectral,
      `PAM--spectral ARI` = method_ari
    ),
  digits = 3
)
```

The cross-method ARI in @tbl-wdi-clustering-summary holds the distance fixed and compares the two clustering methods. The complementary comparison in @fig-wdi-clustering-ari instead holds the clustering method fixed and compares partitions obtained under different distances. Higher ARI values indicate more similar country assignments.

```{r}
#| label: benchmark-wdi-clustering-ari
#| echo: false

wdi_benchmark_specs <- all_dist_method_specs(
  mode = "presets_only",
  preset = unname(distance_presets)
) |>
  dplyr::mutate(
    label = unname(distance_labels[preset])
  ) |>
  dplyr::arrange(match(preset, unname(distance_presets)))

set.seed(2026)

wdi_distance_benchmark <- benchmark_mdist(
  x = wdi_clustering,
  specs = wdi_benchmark_specs,
  cluster_k = 4,
  cluster_methods = c("pam", "spectral"),
  spectral_nstart = 50
)
```

```{r}
#| label: fig-wdi-clustering-ari
#| fig-cap: "Pairwise adjusted Rand indices for four-cluster WDI partitions. Within each panel, the clustering method is held fixed while the distance varies; only one triangle is shown."
#| fig-width: 6.5
#| fig-height: 4.8
#| fig-pos: "H"
#| echo: false

ggplot2::autoplot(
  wdi_distance_benchmark,
  metric = "ari",
  digits = 2
) +
  ggplot2::labs(fill = "ARI") +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold")
  )
```

Together, these summaries show separately how the clustering assignments vary with the clustering method and with the distance specification. They are descriptive comparisons rather than a rule for choosing a universally preferred clustering solution.

### Supervised income classification

For supervised learning, \pkg{manydist} provides `nearest_neighbor_dist()`, a nearest-neighbour model that predicts each new observation from its distances to the training observations. Accordingly, the recipes below use `output = "distance_to_training"`.

Because these recipes include `income` as an outcome, response-aware specifications (`u_dep` and `u_mix`) use the income labels from the current analysis set when their categorical dissimilarities are fitted. The labels from assessment folds and from the final held-out data are not used in this preprocessing. The other three specifications remain response-independent.

One recipe is defined for each distance and paired with the same nearest-neighbour model. The number of neighbours is tuned over the grid defined above.

```{r}
#| label: build-wdi-classification-workflows

classification_recipes <- purrr::map(
  distance_presets,
  ~ recipes::recipe(income ~ ., data = wdi_train) |>
    step_mdist(
      recipes::all_predictors(),
      preset = .x,
      output = "distance_to_training"
    )
)

knn_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = tune::tune()
) |>
  parsnip::set_engine("manydist")

classification_workflows <- workflowsets::workflow_set(
  preproc = classification_recipes,
  models = list(knn = knn_spec)
)
```

The workflow set allows the distance specification and neighbourhood size to be evaluated under the same resampling design. The number of neighbours is tuned within each workflow, while the distance is compared across workflows. Balanced accuracy is used as the primary metric because the four income groups have different sizes.

```{r}
#| label: tune-wdi-classification-workflows

set.seed(2026)

classification_results <- workflowsets::workflow_map(
  classification_workflows,
  fn = "tune_grid",
  seed = 2026,
  resamples = wdi_folds,
  grid = neighbor_grid,
  metrics = yardstick::metric_set(
    yardstick::bal_accuracy,
    yardstick::accuracy
  ),
  control = tune::control_grid(save_pred = TRUE),
  verbose = FALSE
)
```

`rank_results()` identifies the best neighbourhood size within each distance workflow and then ranks those workflows using balanced accuracy.

```{r}
#| label: rank-wdi-classification-workflows

classification_ranking <- workflowsets::rank_results(
  classification_results,
  rank_metric = "bal_accuracy",
  select_best = TRUE
)
```

```{r}
#| label: summarize-wdi-classification-workflows
#| echo: false

classification_metrics <- purrr::map_dfr(
  classification_results$wflow_id,
  function(wflow_id) {
    workflowsets::extract_workflow_set_result(
      classification_results,
      wflow_id
    ) |>
      tune::collect_metrics() |>
      dplyr::mutate(wflow_id = wflow_id)
  }
)

classification_summary <- classification_ranking |>
  dplyr::filter(.metric == "bal_accuracy") |>
  dplyr::select(
    wflow_id,
    .config,
    mean,
    std_err,
    rank
  ) |>
  dplyr::left_join(
    classification_metrics |>
      dplyr::filter(.metric == "bal_accuracy") |>
      dplyr::select(
        wflow_id,
        .config,
        neighbors
      ),
    by = c("wflow_id", ".config")
  ) |>
  dplyr::mutate(
    distance = sub("_knn$", "", wflow_id),
    distance_label = unname(distance_labels[distance])
  ) |>
  dplyr::arrange(rank)
```

```{r}
#| label: tbl-wdi-classification-ranking
#| tbl-cap: "Best nearest-neighbour configuration for each WDI distance, ranked by five-fold cross-validated balanced accuracy."
#| echo: false

knitr::kable(
  classification_summary |>
    dplyr::transmute(
      Distance = distance_label,
      Neighbours = neighbors,
      `Balanced accuracy` = mean,
      `Standard error` = std_err,
      Rank = rank
    ),
  digits = 3
)
```

The leading configurations are separated by less than one standard error. The results distinguish the commensurable specifications from the two baselines more clearly than they distinguish the commensurable specifications from one another; in particular, the two response-aware presets do not outperform `u_indep` in this example.

Figure @fig-wdi-knn-tuning shows the complete tuning profiles rather than only the best configuration under each distance.

```{r}
#| label: fig-wdi-knn-tuning
#| fig-cap: "Five-fold cross-validated balanced accuracy for WDI income classification. Each curve tunes the number of neighbours under a different distance specification, with ribbons giving plus or minus one standard error. All five bands overlap at every neighbourhood size."
#| fig-width: 6.4
#| fig-height: 3.7
#| fig-pos: "H"
#| echo: false

classification_metrics |>
  dplyr::filter(.metric == "bal_accuracy") |>
  dplyr::mutate(
    distance = sub("_knn$", "", wflow_id),
    distance_label = unname(distance_labels[distance])
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = neighbors,
      y = mean,
      color = distance_label
    )
  ) +
  ggplot2::geom_ribbon(
    ggplot2::aes(
      ymin = mean - std_err,
      ymax = mean + std_err,
      fill = distance_label
    ),
    alpha = 0.09,
    colour = NA,
    show.legend = FALSE
  ) +
  ggplot2::geom_line(
    ggplot2::aes(linetype = distance_label),
    linewidth = 0.6
  ) +
  ggplot2::geom_point(size = 1.8) +
  ggplot2::scale_x_continuous(
    breaks = neighbor_grid$neighbors
  ) +
  ggplot2::scale_color_brewer(
    palette = "Dark2",
    name = "Distance"
  ) +
  ggplot2::scale_fill_brewer(
    palette = "Dark2"
  ) +
  ggplot2::scale_linetype_manual(
    values = c(
      "Euclidean" = "22",
      "Gower" = "42",
      "Unbiased dependent" = "solid",
      "Unbiased independent" = "solid",
      "Unbiased mixed" = "solid"
    ),
    name = "Distance"
  ) +
  ggplot2::labs(
    x = "Number of neighbours",
    y = "Cross-validated balanced accuracy"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::guides(
    color = ggplot2::guide_legend(nrow = 2, byrow = TRUE),
    linetype = ggplot2::guide_legend(nrow = 2, byrow = TRUE)
  ) +
  ggplot2::theme(legend.position = "top")
```

Only after the distance specification and number of neighbours have been selected is the held-out partition used. `last_fit()` refits the selected recipe and nearest-neighbour model on the complete training partition and evaluates them on the held-out countries.

```{=latex}
\Needspace{6\baselineskip}
```

```{r}
#| label: finalize-wdi-classification-workflow

selected_workflow_id <- classification_ranking |>
  dplyr::filter(.metric == "bal_accuracy") |>
  dplyr::slice_min(
    rank,
    n = 1,
    with_ties = FALSE
  ) |>
  dplyr::pull(wflow_id)

selected_results <- workflowsets::extract_workflow_set_result(
  classification_results,
  selected_workflow_id
)

best_neighbors <- tune::select_best(
  selected_results,
  metric = "bal_accuracy"
)

final_wdi_workflow <- workflowsets::extract_workflow(
  classification_results,
  selected_workflow_id
) |>
  tune::finalize_workflow(best_neighbors)

final_wdi_results <- tune::last_fit(
  final_wdi_workflow,
  split = wdi_split,
  metrics = yardstick::metric_set(
    yardstick::bal_accuracy,
    yardstick::accuracy
  )
)

final_wdi_metrics <- tune::collect_metrics(final_wdi_results)

final_wdi_metrics
```

```{r}
#| label: summarize-final-wdi-classification
#| echo: false

selected_distance_label <- classification_summary |>
  dplyr::filter(wflow_id == selected_workflow_id) |>
  dplyr::pull(distance_label)

heldout_balanced_accuracy <- final_wdi_metrics |>
  dplyr::filter(.metric == "bal_accuracy") |>
  dplyr::pull(.estimate)

heldout_accuracy <- final_wdi_metrics |>
  dplyr::filter(.metric == "accuracy") |>
  dplyr::pull(.estimate)
```

Cross-validation selected the `r selected_distance_label` distance with
`r best_neighbors$neighbors` neighbours. On the held-out countries, the
selected workflow attained balanced accuracy
`r sprintf("%.3f", heldout_balanced_accuracy)` and ordinary accuracy
`r sprintf("%.3f", heldout_accuracy)`. Balanced accuracy averages class-specific performance across the four groups, whereas ordinary accuracy is the overall proportion of correctly classified countries. Given the modest held-out class counts, these values complete the workflow illustration rather than establish a general performance ranking.

```{r}
#| label: inspect-wdi-classification-predictions

wdi_test_predictions <- tune::collect_predictions(
  final_wdi_results
)

wdi_confusion <- yardstick::conf_mat(
  wdi_test_predictions,
  truth = income,
  estimate = .pred_class
)
```

```{r}
#| label: tbl-wdi-confusion
#| tbl-cap: "Held-out WDI income-group confusion matrix. Rows are predicted classes and columns are observed classes."
#| echo: false

income_abbreviations <- c(
  "High income" = "High",
  "Low income" = "Low",
  "Lower middle income" = "Lower middle",
  "Upper middle income" = "Upper middle"
)

confusion_table <- wdi_confusion$table |>
  as.data.frame() |>
  tibble::as_tibble() |>
  dplyr::mutate(
    Prediction = dplyr::recode(
      as.character(Prediction),
      !!!income_abbreviations
    ),
    Truth = dplyr::recode(
      as.character(Truth),
      !!!income_abbreviations
    )
  ) |>
  tidyr::pivot_wider(
    names_from = Truth,
    values_from = Freq
  )

knitr::kable(confusion_table)
```

Together, the two applications illustrate how the same fitted-distance interface supports different learning tasks. In clustering, `step_mdist()` supplies a pairwise representation that can be combined with alternative clustering methods. In classification, it supplies new-to-training distances so that distance specification and neighbourhood size can be evaluated under a common resampling design, with all fitted preprocessing---including any response-aware categorical dissimilarities---restricted to the appropriate analysis data.


## Conclusion {#sec-conclusion}

\pkg{manydist} implements the unbiased mixed-variable distance framework of
@van2024unbiased and extends it into a workflow for constructing, diagnosing,
and applying distances. The central function `mdist()` combines a numerical
preprocessing, a categorical dissimilarity, and commensurability weights into an
additive distance, with presets that range from familiar baselines to
association- and response-aware constructions. Because the resulting `MDist`
object retains its fitted preprocessing, the same distance can be reapplied to
new observations, which is what allows the package to move from a single
distance matrix to resampled and predictive workflows without leakage.
@tbl-refitting-leakage quantifies what that discipline is worth. Computing the distance once and resampling it afterwards yields an optimistic cross-validated estimate, clearly so for the one specification whose preprocessing includes a principal-component rotation, and by smaller amounts elsewhere.

Two applications illustrated the interface. The Palmer penguins data showed how
scale and association shape the distance geometry, and how leave-one-variable-out
diagnostics and pairwise comparisons of candidate distances make the choice of
distance an explicit,
inspectable part of an analysis rather than a fixed preprocessing decision. The
World Development Indicators snapshot then placed the distance inside
\pkg{tidymodels} pipelines: as a component compared through clustering agreement
and separation, and as a tuned choice evaluated jointly with neighbourhood size
under a common resampling design. In both settings the diagnostics indicate
whether a substantive conclusion is robust to the choice of distance or depends
on it. **The \pkg{tidymodels} integration eliminates the need to write and maintain manual loops by automating repetitive model-fitting and evaluation tasks within a consistent framework. This makes the code cleaner and easier to reproduce, while simplifying preprocessing, parameter tuning, resampling, and performance comparison across multiple models.**

The framework has deliberate limits. An unbiased distance treats all variables
equivalently by construction; it is a neutral starting point, not a claim of
optimality, and a variable can and should be reweighted when subject-matter
knowledge justifies it. The component dissimilarities use associations within
the numerical and categorical blocks but not across them, so cross-type
interactions are not represented. **Response-aware
constructions are likewise confined to the categorical block, so an outcome
cannot yet shape the numerical contributions.** The diagnostics are sensitivity analyses
rather than model-selection rules: they show how much an analysis rests on the
choice of distance, but they do not, on their own, select one.

These boundaries also mark the natural directions for further work: cross-type
interaction-aware distances, response-aware constructions for the numerical block, and
scaling the commensurable distances to larger problems through the compiled
and parallel back-ends already available for homogeneous data. By making distance
construction, diagnosis, and use a single interface, \pkg{manydist} aims to make
the choice of a mixed-variable distance a transparent and reproducible step in
distance-based learning.

## Computational details {.unnumbered}

The results in this article were obtained using \proglang{R} 4.5.2 with
\pkg{manydist} 0.5.1, \pkg{tidymodels} 1.4.1, \pkg{dplyr} 1.2.0 and \pkg{palmerpenguins} 0.1.1.
\proglang{R} itself and all packages used are available from CRAN at
\url{https://CRAN.R-project.org/}. The development version of \pkg{manydist}
and its documentation are hosted at
\url{https://alfonsoiodicede.github.io/manydist_package/}.
