---
title: "Benchmarking distance choices"
author: "Alfonso Iodice D'Enza"
format:
  html:
    number-sections: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Mixed-type data do not determine a unique dissimilarity. Different choices
about scaling, categorical comparisons, commensurability, and associations can
produce different geometries and, consequently, different learning results.

`manydist` provides three helpers for studying these choices systematically:

- `all_dist_method_specs()` constructs tables of valid specifications;
- `benchmark_mdist()` evaluates `mdist()` once for every row and compares all
  successful distances;
- `benchmark_comparisons()` extracts the pairwise diagnostics.

This guide shows how to define a manageable candidate set, run it safely,
inspect failures, compare the resulting dissimilarities, and optionally assess
their clustering solutions. The benchmark is a sensitivity-analysis tool; the
final choice should still be based on the intended downstream task and an
appropriate validation design.

## Setup

```{r}
#| label: setup

library(manydist)
library(dplyr)
library(tidyr)
library(ggplot2)
```

We use complete cases from `palmerpenguins`. The species label is retained for
later use, but the first benchmark is deliberately unsupervised and uses only
the predictors.

```{r}
#| label: prepare-data

penguins_small <- palmerpenguins::penguins |>
  dplyr::select(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  tidyr::drop_na()

penguin_x <- penguins_small |>
  dplyr::select(-species)

dplyr::glimpse(penguin_x)
```

## Constructing a candidate set

`all_dist_method_specs()` returns one specification per row. Preset rows use
the `preset` column, whereas component rows describe a custom combination of
`method_cat`, `method_num`, and `commensurable`.

```{r}
#| label: inspect-preset-specs

all_dist_method_specs(mode = "presets_only") |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable
  )
```

It is usually better to begin with a small set of substantively different
choices than to run every possible combination. Here we compare four presets
and two custom specifications.

```{r}
#| label: define-candidates

candidate_specs <- dplyr::bind_rows(
  all_dist_method_specs(
    mode = "presets_only",
    preset = c("gower", "u_indep", "euclidean", "hl")
  ),
  tibble::tibble(
    spec_type = "component",
    preset = "custom",
    method_cat = "matching",
    method_num = "std",
    commensurable = c(FALSE, TRUE)
  )
) |>
  dplyr::mutate(
    label = dplyr::case_when(
      preset == "gower" ~ "Gower",
      preset == "u_indep" ~ "Unbiased independent",
      preset == "euclidean" ~ "Euclidean",
      preset == "hl" ~ "Hennig--Liao",
      commensurable ~ "Matching + z scores (weighted)",
      TRUE ~ "Matching + z scores"
    )
  )

candidate_specs
```

This explicit table is useful for reproducibility: it records exactly which
distance definitions entered the comparison.

## Running the benchmark

`benchmark_mdist()` applies `mdist()` to each specification and compares every
pair of successful results. An error in one row does not stop the remaining
computations. By default, `cluster_k = NULL`, so no clustering is performed.

```{r}
#| label: run-benchmark

distance_benchmark <- benchmark_mdist(
  x = penguin_x,
  specs = candidate_specs
)

distance_benchmark |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable,
    ok,
    error
  )
```

The returned `MDistBenchmark` object remains a tibble and adds three columns:

- `result` contains the `MDist` object, or the captured error;
- `ok` indicates whether computation succeeded;
- `error` contains the error message for an unsuccessful row.

```{r}
#| label: benchmark-status

distance_benchmark |>
  dplyr::count(ok)
```

Failed specifications remain easy to inspect.

```{r}
#| label: inspect-failures

distance_benchmark |>
  dplyr::filter(!ok) |>
  dplyr::select(
    preset,
    method_cat,
    method_num,
    commensurable,
    error
  )
```

This behavior is especially helpful for larger sensitivity analyses, where
some methods may be incompatible with the variables in a particular data set.

## Comparing successful distances

`benchmark_comparisons()` returns one row for each unique pair of successful
specifications. Four diagnostics are always available:

- `mad` is the mean absolute difference between the corresponding pairwise
  dissimilarities;
- `relative_distance` divides MAD by the average magnitude of the two
  distances;
- `mds_congruence` compares their classical MDS configurations;
- `alienation` is $\sqrt{1-c^2}$, where $c$ is MDS congruence.

The symmetric relative distance is

$$
\frac{2\operatorname{mean}(|d_a-d_b|)}
{\operatorname{mean}(|d_a|)+\operatorname{mean}(|d_b|)}.
$$

It is zero for identical distances and does not depend on how many
specifications are included in the benchmark.

```{r}
#| label: inspect-pairwise-diagnostics

pairwise_diagnostics <- benchmark_comparisons(distance_benchmark)

pairwise_diagnostics |>
  dplyr::select(
    method_1,
    method_2,
    mad,
    relative_distance,
    mds_congruence,
    alienation
  ) |>
  dplyr::arrange(dplyr::desc(relative_distance)) |>
  dplyr::slice_head(n = 10)
```

Raw MAD retains differences in absolute scale. Relative distance makes those
differences comparable, while congruence and alienation describe changes in
the induced geometry. The measures are complementary rather than competing
rankings.

`autoplot()` displays any diagnostic as an annotated triangular heatmap. The
optional `label` column in `candidate_specs` supplies the axis labels.

::: {.panel-tabset}

### Relative distance

```{r}
#| label: plot-relative-distance
#| fig-alt: "Triangular heatmap of pairwise relative distances among six distance specifications."
#| fig-width: 9
#| fig-height: 6.5

ggplot2::autoplot(
  distance_benchmark,
  metric = "relative_distance",
  digits = 2
)
```

### Alienation

```{r}
#| label: plot-alienation
#| fig-alt: "Triangular heatmap of pairwise alienation coefficients among six distance specifications."
#| fig-width: 9
#| fig-height: 6.5

ggplot2::autoplot(
  distance_benchmark,
  metric = "alienation",
  digits = 2
)
```

:::

## Optional clustering comparisons

Supplying `cluster_k` asks the benchmark to cluster every successful distance.
The same value of $K$ and the same clustering method are used within each
pairwise comparison. Available methods are `"pam"`, `"hclust"`, and
`"spectral"`. If `cluster_k` is omitted or `NULL`, this potentially expensive
step is skipped.

```{r}
#| label: benchmark-clustering-solutions

set.seed(2026)

clustering_benchmark <- benchmark_mdist(
  x = penguin_x,
  specs = candidate_specs,
  cluster_k = 3,
  cluster_methods = c("pam", "spectral"),
  spectral_nstart = 50
)

benchmark_comparisons(clustering_benchmark) |>
  dplyr::select(
    method_1,
    method_2,
    ari_pam,
    ari_spectral
  ) |>
  dplyr::arrange(ari_pam) |>
  dplyr::slice_head(n = 10)
```

The adjusted Rand indices compare two unsupervised partitions; the penguin
species labels are not used. An ARI of one indicates identical assignments,
whereas smaller values indicate greater sensitivity to the distance
specification.

```{r}
#| label: plot-clustering-agreement
#| fig-alt: "Faceted triangular heatmaps of pairwise adjusted Rand indices for PAM and spectral clustering."
#| fig-width: 10
#| fig-height: 6

ggplot2::autoplot(
  clustering_benchmark,
  metric = "ari",
  digits = 2
)
```

## Response-aware candidates

Some specifications can use a response when constructing the dissimilarity.
They can be listed with `mode = "response_aware_only"`.

```{r}
#| label: response-aware-specs

response_specs <- all_dist_method_specs(
  mode = "response_aware_only"
) |>
  dplyr::filter(
    spec_type == "preset",
    preset %in% c("u_dep", "u_mix")
  )

response_specs
```

When the response is supplied to `benchmark_mdist()`, it is used only by
response-aware components and is not treated as an ordinary predictor.

```{r}
#| label: response-aware-benchmark

response_benchmark <- benchmark_mdist(
  x = penguins_small,
  response = species,
  specs = response_specs
)

response_benchmark |>
  dplyr::select(preset, ok, error)

benchmark_comparisons(response_benchmark)
```

If response-aware distances are compared by predictive performance, their
construction must occur inside each resample. Constructing them once from the
full outcome vector would leak assessment information into training.

The recipe interface handles this separation directly. When a response-aware
specification is used in `step_mdist()`, the step discovers the single recipe
outcome and fits the response-aware profiles using only the current analysis
set. The fitted profiles are then applied to assessment observations without
using their outcomes. Set `response_used = FALSE` when the comparison should
use the predictor-only form of a response-aware specification.

## Scaling up a comparison

The same workflow can be expanded by filtering the full specification table,
as in the unevaluated example below. Keep in mind that $M$ successful
specifications produce $M(M-1)/2$ pairwise rows, and clustering adds one fitted
partition for every distance and clustering-method combination.

```{r}
#| label: larger-grid
#| eval: false

larger_specs <- all_dist_method_specs(
  mode = "full",
  method_cat = c("matching", "tvd"),
  method_num = c("std", "robust"),
  commensurable = c(FALSE, TRUE)
)

larger_benchmark <- benchmark_mdist(
  x = penguins_small,
  response = species,
  specs = larger_specs
)
```

Before expanding the grid, decide what the benchmark is intended to measure:

- for clustering, compare stability, separation, or external labels when
  legitimately available;
- for prediction, compare candidates through resampling and preserve a final
  test set;
- for exploratory analysis, inspect whether substantive conclusions are
  stable across plausible distance definitions;
- always keep unsuccessful rows and their messages as part of the audit trail.

The clustering and supervised-learning guides show how `MDist` specifications
enter those downstream workflows.
