---
title: "Benchmarking distance choices"
author: "Alfonso Iodice D'Enza"
format:
  html:
    number-sections: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Mixed-type data do not determine a unique dissimilarity. Different choices
about scaling, categorical comparisons, commensurability, and associations can
produce different geometries and, consequently, different learning results.

`manydist` provides two helpers for studying these choices systematically:

- `all_dist_method_specs()` constructs tables of valid specifications;
- `benchmark_mdist()` evaluates `mdist()` once for every row of such a table.

This guide shows how to define a manageable candidate set, run it safely,
inspect failures, and compare the resulting dissimilarities. The benchmark
constructs distances; the final choice should still be based on the intended
downstream task and an appropriate validation design.

## Setup

```{r}
#| label: setup

library(manydist)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

We use complete cases from `palmerpenguins`. The species label is retained for
later use, but the first benchmark is deliberately unsupervised and uses only
the predictors.

```{r}
#| label: prepare-data

penguins_small <- palmerpenguins::penguins |>
  dplyr::select(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  tidyr::drop_na()

penguin_x <- penguins_small |>
  dplyr::select(-species)

dplyr::glimpse(penguin_x)
```

## Constructing a candidate set

`all_dist_method_specs()` returns one specification per row. Preset rows use
the `preset` column, whereas component rows describe a custom combination of
`method_cat`, `method_num`, and `commensurable`.

```{r}
#| label: inspect-preset-specs

all_dist_method_specs(mode = "presets_only") |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable
  )
```

It is usually better to begin with a small set of substantively different
choices than to run every possible combination. Here we compare four presets
and two custom specifications.

```{r}
#| label: define-candidates

candidate_specs <- dplyr::bind_rows(
  all_dist_method_specs(
    mode = "presets_only",
    preset = c("gower", "u_indep", "euclidean", "hl")
  ),
  tibble::tibble(
    spec_type = "component",
    preset = "custom",
    method_cat = "matching",
    method_num = "std",
    commensurable = c(FALSE, TRUE)
  )
)

candidate_specs
```

This explicit table is useful for reproducibility: it records exactly which
distance definitions entered the comparison.

## Running the benchmark

`benchmark_mdist()` applies `mdist()` to each specification. An error in one
row does not stop the remaining computations.

```{r}
#| label: run-benchmark

distance_benchmark <- benchmark_mdist(
  x = penguin_x,
  specs = candidate_specs
)

distance_benchmark |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable,
    ok,
    error
  )
```

The returned tibble adds three columns:

- `result` contains the `MDist` object, or the captured error;
- `ok` indicates whether computation succeeded;
- `error` contains the error message for an unsuccessful row.

```{r}
#| label: benchmark-status

distance_benchmark |>
  dplyr::count(ok)
```

Failed specifications remain easy to inspect.

```{r}
#| label: inspect-failures

distance_benchmark |>
  dplyr::filter(!ok) |>
  dplyr::select(
    preset,
    method_cat,
    method_num,
    commensurable,
    error
  )
```

This behavior is especially helpful for larger sensitivity analyses, where
some methods may be incompatible with the variables in a particular data set.

## Summarising the resulting geometries

The successful `MDist` objects can be extracted from the `result` list-column
and analysed like any other computed object.

```{r}
#| label: prepare-distance-summaries

distance_values <- function(x) {
  distance_matrix <- as.matrix(x$distance)
  distance_matrix[lower.tri(distance_matrix)]
}

successful_distances <- distance_benchmark |>
  dplyr::filter(ok) |>
  dplyr::mutate(
    method = dplyr::if_else(
      spec_type == "preset",
      preset,
      paste(
        method_cat,
        method_num,
        dplyr::if_else(
          commensurable,
          "commensurable",
          "not commensurable"
        ),
        sep = " / "
      )
    ),
    values = purrr::map(result, distance_values),
    mean_distance = purrr::map_dbl(values, mean),
    sd_distance = purrr::map_dbl(values, stats::sd)
  )

successful_distances |>
  dplyr::select(method, mean_distance, sd_distance)
```

The absolute scale can differ across specifications. Geometry can therefore
also be compared through the association between corresponding pairwise
dissimilarities. The following example uses Gower dissimilarity as a reference.

```{r}
#| label: compare-with-gower

gower_values <- successful_distances |>
  dplyr::filter(preset == "gower") |>
  dplyr::pull(values) |>
  purrr::pluck(1)

geometry_agreement <- successful_distances |>
  dplyr::transmute(
    method,
    correlation_with_gower = purrr::map_dbl(
      values,
      ~ stats::cor(.x, gower_values)
    )
  )

geometry_agreement
```

```{r}
#| label: plot-geometry-agreement
#| fig-alt: "Bar chart of the correlation between each candidate dissimilarity and Gower dissimilarity."

geometry_agreement |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = stats::reorder(method, correlation_with_gower),
      y = correlation_with_gower
    )
  ) +
  ggplot2::geom_col(fill = "#2C7FB8") +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = NULL,
    y = "Correlation with Gower pairwise dissimilarities"
  ) +
  ggplot2::theme_minimal()
```

A high correlation indicates similar pairwise ordering, not necessarily
equivalent downstream performance. Scale summaries and geometry agreement are
diagnostics, not selection criteria on their own.

## Response-aware candidates

Some specifications can use a response when constructing the dissimilarity.
They can be listed with `mode = "response_aware_only"`.

```{r}
#| label: response-aware-specs

response_specs <- all_dist_method_specs(
  mode = "response_aware_only"
) |>
  dplyr::filter(
    spec_type == "preset",
    preset %in% c("u_dep", "u_mix")
  )

response_specs
```

When the response is supplied to `benchmark_mdist()`, it is used only by
response-aware components and is not treated as an ordinary predictor.

```{r}
#| label: response-aware-benchmark

response_benchmark <- benchmark_mdist(
  x = penguins_small,
  response = species,
  specs = response_specs
)

response_benchmark |>
  dplyr::select(preset, ok, error)
```

If response-aware distances are compared by predictive performance, their
construction must occur inside each resample. Constructing them once from the
full outcome vector would leak assessment information into training.

The recipe interface handles this separation directly. When a response-aware
specification is used in `step_mdist()`, the step discovers the single recipe
outcome and fits the response-aware profiles using only the current analysis
set. The fitted profiles are then applied to assessment observations without
using their outcomes. Set `response_used = FALSE` when the comparison should
use the predictor-only form of a response-aware specification.

## Scaling up a comparison

The same workflow can be expanded by filtering the full specification table.
For example:

```{r}
#| label: larger-grid
#| eval: false

larger_specs <- all_dist_method_specs(
  mode = "full",
  method_cat = c("matching", "tvd"),
  method_num = c("std", "robust"),
  commensurable = c(FALSE, TRUE)
)

larger_benchmark <- benchmark_mdist(
  x = penguins_small,
  response = species,
  specs = larger_specs
)
```

Before expanding the grid, decide what the benchmark is intended to measure:

- for clustering, compare stability, separation, or external labels when
  legitimately available;
- for prediction, compare candidates through resampling and preserve a final
  test set;
- for exploratory analysis, inspect whether substantive conclusions are
  stable across plausible distance definitions;
- always keep unsuccessful rows and their messages as part of the audit trail.

The clustering and supervised-learning guides show how `MDist` specifications
enter those downstream workflows.
