---
title: "Distance specifications"
author: "Alfonso Iodice D'Enza"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---




## Overview

`manydist` provides tools for constructing dissimilarities/distances for numerical, categorical, and mixed-type data.

The main function is `mdist()`: the user can either specify the distance via a `preset` or customize it. 
The custom specification requires:

- `method_cat` controls the categorical dissimilarity;
- `method_num` controls the preprocessing of numerical variables;
- `commensurable` controls whether variable-wise contributions should be, on average, the same;


## Setup

```{r}
library(manydist)
library(dplyr)
library(tidyr)
library(recipes)
```

We use the `palmerpenguins` data for illustration.

```{r}
if (requireNamespace("palmerpenguins", quietly = TRUE)) {
  data("penguins", package = "palmerpenguins")

  penguins_small <- palmerpenguins::penguins |>
    dplyr::select(-year)|>
    tidyr::drop_na()
}
```

The data contain numerical variables, such as bill length and body mass, and categorical variables, such as island and sex.

```{r}
skim_cols <- c(
  "skim_type",
  "skim_variable",
  "complete_rate",
  "numeric.mean",
  "numeric.sd",
  "factor.n_unique"
)

skimr::skim(penguins_small) |>
  dplyr::as_tibble() |>
  dplyr::select(dplyr::any_of(skim_cols)) |>
  kableExtra::kbl(format = "html", digits = 2) |>
  kableExtra::kable_styling(full_width = FALSE)

```

## Computing a mixed-type dissimilarity

The simplest way to compute a mixed-type dissimilarity is to use a `preset`.

For example, the `"gower"` preset computes a Gower-type dissimilarity for mixed numerical and categorical data.

```{r}
d_gow <- penguins_small |> dplyr::select(-species) |> mdist(preset = "gower")

d_gow
```



The result is an `MDist` object. Internally, `MDist` is implemented as an
R6 object, which stores the computed dissimilarity together with metadata and
methods for extracting or reusing the result.
The dissimilarity can be extracted as a standard `dist` object using
`to_dist()`. For display, we convert it to a matrix and show only the first
few rows and columns.

```{r}
d_gow_dist <- d_gow$to_dist()

d_gow_mat <- as.matrix(d_gow_dist)

d_gow_mat[1:5, 1:5] |>
  round(digits = 2) 
```

The `"gower"` preset gives the same dissimilarities as `cluster::daisy()` with
`metric = "gower"`.

```{r}
d_daisy <- cluster::daisy(
  penguins_small |> dplyr::select(-species),
  metric = "gower"
)

all.equal(
  as.matrix(d_gow$to_dist()),
  as.matrix(d_daisy),
  tolerance = 1e-10
)
```


The `preset` argument provides shortcuts for common distance specifications.
Some presets are classical mixed-type distances, while others are designed to
construct unbiased or association-aware dissimilarities.

```{r}
#| echo: false

available_presets <- tibble::tibble(
  preset = c(
    "gower",
    "euclidean",
    "u_indep",
    "u_dep",
    "u_mix",
    "hl",
    "gudmm",
    "dkss",
    "mod_gower"
  ),
  description = c(
    "Gower-type mixed-data dissimilarity.",
    "Euclidean distance after one-hot encoding categorical variables.",
    "Unbiased distance with independent variable-wise contributions.",
    "Unbiased association-aware distance.",
    "Mixed unbiased association-aware distance.",
    "Heterogeneous Euclidean-type distance.",
    "Generalized distance for mixed data.",
    "Distance for mixed data based on distributional comparisons.",
    "Modified Gower-type dissimilarity."
  )
)

available_presets |>
  knitr::kable()
```

For the complete list of supported specifications and their argument
combinations, see:

```{r, eval=FALSE}
?mdist
```


## Custom specification

Presets are convenient, but users can also define custom specifications.

In a custom specification, `method_cat` controls the categorical dissimilarity and `method_num` controls the preprocessing applied to numerical variables.

```{r}
d_custom <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "custom",
  method_cat = "matching",
  method_num = "std",
  commensurable = TRUE
)

d_custom
```

Here, categorical variables are compared using matching dissimilarities, numerical variables are standardized, and variable-wise contributions are made commensurable. Note that pairwise distances are always computed via Manhattan distance, with the exection of 
`euclidean` and `hl` presets.


## Response-aware specifications

Some distance specifications can use a response variable when constructing the
dissimilarity. In these cases, the response is supplied through the `response`
argument of `mdist()`.

The response is not treated as an ordinary predictor. Instead, it is removed
from the predictor set and used only by the components that support
response-aware construction.

Response-aware construction is available in two ways:

1. through response-aware presets;
2. through response-aware categorical methods in custom specifications.



The response-aware presets can be inspected with
`all_dist_method_specs(mode = "response_aware_only")`.

```{r}
response_aware_presets <- all_dist_method_specs(mode = "response_aware_only") |>
  dplyr::filter(spec_type == "preset") |>
  dplyr::distinct(preset) |>
  dplyr::arrange(preset)

response_aware_presets |>
  kableExtra::kbl(
    format = "html",
    col.names = "Response-aware preset"
  ) |>
  kableExtra::kable_styling(full_width = FALSE)
```

For example, the `"u_dep"` preset can use the response when constructing the
dissimilarity.

```{r}
d_response <- mdist(
  penguins_small,
  response = species,
  preset = "u_dep"
)

d_response
```

The response can also be supplied as a character string.

```{r}
d_response_chr <- mdist(
  penguins_small,
  response = "species",
  preset = "u_dep"
)

d_response_chr
```

In both calls, `species` is used as the response variable and is not included
among the predictors.

The recipe interface provides the same behavior without repeating the outcome
name in `step_mdist()`. When a recipe contains exactly one outcome,
response-aware specifications automatically use the analysis-set outcomes
during `prep()`. The fitted response-aware profiles are then reused when
assessment or test observations are baked, without using their outcomes. Set
`response_used = FALSE` in `step_mdist()` to request a predictor-only distance.



Response-aware construction can also be used in custom specifications. In this
case, `preset = "custom"` is combined with a response-aware categorical method.

The following table lists the response-aware categorical methods implemented
directly in `manydist`.

```{r}
dist_methods_tbl() |>
  dplyr::filter(
    argument == "method_cat",
    response_aware,
    engine == "manydist") |> 
  dplyr::select(method) |> 
  kableExtra::kbl(full_width = FALSE)
```

In addition to the methods implemented directly in `manydist`, categorical
dissimilarities available through `philentropy` can also be used in
response-aware custom specifications.


<!-- ## Euclidean preset -->

<!-- The `"euclidean"` preset computes Euclidean distances after converting categorical variables to one-hot dummy variables. -->

<!-- ```{r} -->
<!-- d_euclidean <- mdist( -->
<!--   penguins_small |> dplyr::select(-species), -->
<!--   preset = "euclidean" -->
<!-- ) -->

<!-- d_euclidean -->
<!-- ``` -->

<!-- This preset replaces the former `"euclidean_onehot"` name. -->

<!-- ## Comparing presets -->

<!-- Different presets encode different assumptions about how numerical and categorical variables should contribute to the final dissimilarity. -->

<!-- ```{r} -->
<!-- d_u_indep <- mdist( -->
<!--   penguins_small |> dplyr::select(-species), -->
<!--   preset = "u_indep" -->
<!-- ) -->

<!-- d_u_mix <- mdist( -->
<!--   penguins_small |> dplyr::select(-species), -->
<!--   preset = "u_mix" -->
<!-- ) -->

<!-- d_hl <- mdist( -->
<!--   penguins_small |> dplyr::select(-species), -->
<!--   preset = "hl" -->
<!-- ) -->

<!-- d_u_indep -->
<!-- d_u_mix -->
<!-- d_hl -->
<!-- ``` -->

<!-- The available presets include: -->

<!-- ```{r} -->
<!-- c( -->
<!--   "custom", -->
<!--   "gower", -->
<!--   "euclidean", -->
<!--   "unbiased_dependent", -->
<!--   "u_dep", -->
<!--   "u_indep", -->
<!--   "u_mix", -->
<!--   "hl", -->
<!--   "gudmm", -->
<!--   "dkss", -->
<!--   "mod_gower" -->
<!-- ) -->
<!-- ``` -->

<!-- Some presets are intended mainly for train-train dissimilarities and may not support distances from new observations to training observations. -->

<!-- ## Distances from new observations to training observations -->

<!-- `mdist()` can also compute distances from new observations to the training observations through the `new_data` argument. -->

<!-- ```{r} -->
<!-- set.seed(123) -->

<!-- split_id <- sample(seq_len(nrow(penguins_small)), size = floor(0.75 * nrow(penguins_small))) -->

<!-- penguins_train <- penguins_small[split_id, ] -->
<!-- penguins_test  <- penguins_small[-split_id, ] -->

<!-- d_new <- mdist( -->
<!--   x = penguins_train |> dplyr::select(-species), -->
<!--   new_data = penguins_test |> dplyr::select(-species), -->
<!--   preset = "gower" -->
<!-- ) -->

<!-- d_new -->
<!-- ``` -->

<!-- The resulting dissimilarity is rectangular: rows correspond to test observations and columns correspond to training observations. -->

<!-- ```{r} -->
<!-- dim(as.matrix(d_new$distance)) -->
<!-- ``` -->

<!-- ## Using `manydist` in recipe workflows -->

<!-- The function `step_mdist()` provides a recipe interface to `mdist()`. It replaces selected predictors with a distance-based representation. -->

<!-- In supervised workflows, the response is defined on the left-hand side of the recipe formula. Therefore, when we select `recipes::all_predictors()`, the response variable is not used as a predictor in the distance computation. -->

<!-- ```{r} -->
<!-- rec <- recipes::recipe(species ~ ., data = penguins_small) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "gower", -->
<!--     output = "distance_to_training" -->
<!--   ) -->

<!-- rec_prep <- recipes::prep(rec, training = penguins_small) -->

<!-- baked <- recipes::bake(rec_prep, new_data = penguins_small) -->

<!-- baked |> -->
<!--   dplyr::slice_head(n = 5) -->
<!-- ``` -->

<!-- The resulting data contain the outcome variable, `species`, together with one distance column for each training observation. These columns are named `dist_1`, `dist_2`, and so on. -->

<!-- ## Custom specifications in recipes -->

<!-- The same interface can be used with a custom distance specification. -->

<!-- ```{r} -->
<!-- rec_custom <- recipes::recipe(species ~ ., data = penguins_small) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "custom", -->
<!--     method_cat = "matching", -->
<!--     method_num = "std", -->
<!--     commensurable = TRUE, -->
<!--     output = "distance_to_training" -->
<!--   ) -->

<!-- rec_custom_prep <- recipes::prep(rec_custom, training = penguins_small) -->

<!-- baked_custom <- recipes::bake(rec_custom_prep, new_data = penguins_small) -->

<!-- baked_custom |> -->
<!--   dplyr::slice_head(n = 5) -->
<!-- ``` -->

<!-- In this example, `method_cat = "matching"` specifies the categorical dissimilarity, while `method_num = "std"` specifies standardization of numerical variables. -->

<!-- ## Pairwise dissimilarities in recipes -->

<!-- For distance-based clustering workflows, `step_mdist()` can return the within-training pairwise dissimilarity matrix by setting `output = "pairwise"`. -->

<!-- In this case, the recipe is intended for training-only workflows. -->

<!-- ```{r} -->
<!-- penguins_predictors <- penguins_small |> -->
<!--   dplyr::select(-species) -->

<!-- rec_pairwise <- recipes::recipe(~ ., data = penguins_predictors) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "gower", -->
<!--     output = "pairwise" -->
<!--   ) -->

<!-- rec_pairwise_prep <- recipes::prep(rec_pairwise, training = penguins_predictors) -->

<!-- pairwise_dist <- recipes::bake( -->
<!--   rec_pairwise_prep, -->
<!--   new_data = penguins_predictors -->
<!-- ) -->

<!-- pairwise_dist |> -->
<!--   dplyr::slice_head(n = 5) -->
<!-- ``` -->

<!-- ## Euclidean distances in recipes -->

<!-- The `"euclidean"` preset can also be used inside a recipe. Categorical predictors are one-hot encoded internally before the Euclidean distance is computed. -->

<!-- ```{r} -->
<!-- rec_euclidean <- recipes::recipe(species ~ ., data = penguins_small) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "euclidean", -->
<!--     output = "distance_to_training" -->
<!--   ) -->

<!-- rec_euclidean_prep <- recipes::prep(rec_euclidean, training = penguins_small) -->

<!-- baked_euclidean <- recipes::bake(rec_euclidean_prep, new_data = penguins_small) -->

<!-- baked_euclidean |> -->
<!--   dplyr::slice_head(n = 5) -->
<!-- ``` -->

<!-- ## Distance-based nearest-neighbour prediction -->

<!-- The distance representation produced by `step_mdist()` can be used in downstream supervised learning workflows. -->

<!-- Here we illustrate a nearest-neighbour classifier using the distance-to-training representation. The recipe first converts the original mixed-type predictors into distances from each observation to the training observations. The model is then tuned over the number of neighbours. -->

<!-- ```{r} -->
<!-- library(tidymodels) -->

<!-- set.seed(123) -->

<!-- penguin_split <- rsample::initial_split( -->
<!--   penguins_small, -->
<!--   prop = 0.75, -->
<!--   strata = species -->
<!-- ) -->

<!-- penguin_train <- rsample::training(penguin_split) -->
<!-- penguin_test  <- rsample::testing(penguin_split) -->

<!-- set.seed(123) -->

<!-- penguin_folds <- rsample::vfold_cv( -->
<!--   penguin_train, -->
<!--   v = 5, -->
<!--   strata = species -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- rec_knn <- recipes::recipe(species ~ ., data = penguin_train) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "gower", -->
<!--     output = "distance_to_training" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- knn_spec <- nearest_neighbor_dist( -->
<!--   neighbors = tune::tune() -->
<!-- ) |> -->
<!--   parsnip::set_mode("classification") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- knn_wf <- workflows::workflow() |> -->
<!--   workflows::add_recipe(rec_knn) |> -->
<!--   workflows::add_model(knn_spec) -->

<!-- knn_grid <- tibble::tibble( -->
<!--   neighbors = c(1, 3, 5, 7, 9) -->
<!-- ) -->

<!-- set.seed(123) -->

<!-- knn_res <- tune::tune_grid( -->
<!--   knn_wf, -->
<!--   resamples = penguin_folds, -->
<!--   grid = knn_grid, -->
<!--   metrics = yardstick::metric_set(accuracy) -->
<!-- ) -->

<!-- knn_res |> -->
<!--   tune::collect_metrics() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- best_knn <- knn_res |> -->
<!--   tune::select_best(metric = "accuracy") -->

<!-- best_knn -->
<!-- ``` -->

<!-- ```{r} -->
<!-- final_knn_wf <- knn_wf |> -->
<!--   tune::finalize_workflow(best_knn) -->

<!-- final_knn_fit <- final_knn_wf |> -->
<!--   parsnip::fit(data = penguin_train) -->

<!-- knn_pred <- predict(final_knn_fit, new_data = penguin_test) |> -->
<!--   dplyr::bind_cols(penguin_test |> dplyr::select(species)) -->

<!-- yardstick::accuracy(knn_pred, truth = species, estimate = .pred_class) -->
<!-- ``` -->

<!-- This example uses `preset = "gower"`, but the same workflow can be used with any `mdist()` preset or with a custom specification based on `method_cat`, `method_num`, and `commensurable`. -->

<!-- ## Distance-based PAM clustering -->

<!-- The same dissimilarities can be used for clustering. For clustering workflows, `step_mdist()` can return the within-training pairwise dissimilarity matrix by setting `output = "pairwise"`. -->

<!-- ```{r} -->
<!-- penguins_predictors <- penguins_small |> -->
<!--   dplyr::select(-species) -->

<!-- rec_pam <- recipes::recipe(~ ., data = penguins_predictors) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "gower", -->
<!--     output = "pairwise" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- pam_spec <- pam_dist( -->
<!--   num_clusters = 3 -->
<!-- ) -->

<!-- pam_fit <- generics::fit( -->
<!--   pam_spec, -->
<!--   recipe = rec_pam, -->
<!--   data = penguins_predictors -->
<!-- ) -->

<!-- pam_pred <- predict(pam_fit) -->

<!-- pam_pred |> -->
<!--   dplyr::slice_head(n = 10) -->
<!-- ``` -->

<!-- Since the true species labels are available in this example, we can inspect the relationship between the PAM clusters and the species labels. -->

<!-- ```{r} -->
<!-- pam_pred |> -->
<!--   dplyr::bind_cols(penguins_small |> dplyr::select(species)) |> -->
<!--   dplyr::count(.pred_cluster, species) -->
<!-- ``` -->

<!-- ## Distance-based spectral clustering -->

<!-- Spectral clustering can be used in the same way, by combining a pairwise distance recipe with the corresponding model specification. -->

<!-- ```{r} -->
<!-- rec_spectral <- recipes::recipe(~ ., data = penguins_predictors) |> -->
<!--   step_mdist( -->
<!--     recipes::all_predictors(), -->
<!--     preset = "gower", -->
<!--     output = "pairwise" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- spectral_spec <- spectral_dist( -->
<!--   num_clusters = 3 -->
<!-- ) -->

<!-- spectral_fit <- generics::fit( -->
<!--   spectral_spec, -->
<!--   recipe = rec_spectral, -->
<!--   data = penguins_predictors -->
<!-- ) -->

<!-- spectral_pred <- predict(spectral_fit) -->

<!-- spectral_pred |> -->
<!--   dplyr::slice_head(n = 10) -->
<!-- ``` -->

<!-- Again, because this is an illustrative labelled data set, we can compare the resulting clusters with the observed species labels. -->

<!-- ```{r} -->
<!-- spectral_pred |> -->
<!--   dplyr::bind_cols(penguins_small |> dplyr::select(species)) |> -->
<!--   dplyr::count(.pred_cluster, species) -->
<!-- ``` -->


<!-- ## A small benchmark example -->

<!-- The function `benchmark_mdist()` applies `mdist()` repeatedly over a grid of distance specifications. This is useful for comparing alternative presets and custom specifications in a systematic way. -->

<!-- For a small example, we restrict the grid to a few specifications. -->

<!-- ```{r} -->
<!-- small_specs <- dplyr::bind_rows( -->
<!--   all_dist_method_specs(mode = "presets_only") |> -->
<!--     dplyr::filter(preset %in% c("gower", "u_dep", "u_indep", "euclidean")), -->

<!--   all_dist_method_specs( -->
<!--     mode = "full", -->
<!--     method_cat = c("matching", "tvd"), -->
<!--     method_num = c("std", "robust"), -->
<!--     commensurable = c(TRUE, FALSE) -->
<!--   ) |> -->
<!--     dplyr::filter(spec_type == "component") -->
<!-- ) -->

<!-- small_specs -->
<!-- ``` -->

<!-- ```{r} -->
<!-- bench_small <- benchmark_mdist( -->
<!--   penguins_small, -->
<!--   response = species, -->
<!--   specs = small_specs -->
<!-- ) -->

<!-- bench_small |> -->
<!--   dplyr::select( -->
<!--     spec_type, -->
<!--     preset, -->
<!--     method_cat, -->
<!--     method_num, -->
<!--     commensurable, -->
<!--     ok, -->
<!--     error -->
<!--   ) -->
<!-- ``` -->

<!-- The `ok` column reports whether the distance computation completed successfully. Failed specifications are kept in the output, with the corresponding error message stored in `error`. -->

<!-- ```{r} -->
<!-- bench_small |> -->
<!--   dplyr::count(ok) -->
<!-- ``` -->

<!-- When all specifications are successful, the resulting `MDist` objects are stored in the `result` list-column and can be inspected or reused. -->

<!-- ```{r} -->
<!-- bench_small |> -->
<!--   dplyr::filter(ok) |> -->
<!--   dplyr::select(spec_type, preset, method_cat, method_num, commensurable, result) |> -->
<!--   dplyr::slice_head(n = 3) -->
<!-- ``` -->

<!-- ## Leave-one-variable-out diagnostics -->

<!-- `lovo_mdist()` computes leave-one-variable-out diagnostics for distance-based variable importance. -->

<!-- It first computes the full dissimilarity matrix. Then it removes one predictor at a time, recomputes the dissimilarity matrix, and compares the reduced matrix with the full one. -->

<!-- ```{r} -->
<!-- lovo_gower <- lovo_mdist( -->
<!--   penguins_small, -->
<!--   response = species, -->
<!--   response_used = FALSE, -->
<!--   preset = "gower" -->
<!-- ) -->

<!-- lovo_gower -->
<!-- ``` -->

<!-- The result can be summarized. -->

<!-- ```{r} -->
<!-- summary(lovo_gower) -->
<!-- ``` -->

<!-- The relative contribution of each variable can be visualized with `autoplot()`. -->

<!-- ```{r} -->
<!-- lovo_gower$autoplot( -->
<!--   metric = "relative_distance", -->
<!--   reorder = TRUE -->
<!-- ) -->
<!-- ``` -->

<!-- When `response_used = FALSE`, the response column is removed before computing distances. When `response_used = TRUE`, the response can be used by response-aware distance specifications, but it is still not treated as a predictor in the leave-one-variable-out loop. -->

<!-- ```{r} -->
<!-- lovo_response <- lovo_mdist( -->
<!--   penguins_small, -->
<!--   response = species, -->
<!--   response_used = TRUE, -->
<!--   preset = "u_dep" -->
<!-- ) -->

<!-- lovo_response -->
<!-- ``` -->

<!-- ## Inspecting method specifications -->

<!-- `all_dist_method_specs()` returns a table of distance specifications that can be used for benchmarking or sensitivity analysis. -->

<!-- ```{r} -->
<!-- specs <- all_dist_method_specs() -->

<!-- specs |> -->
<!--   dplyr::select( -->
<!--     spec_type, -->
<!--     preset, -->
<!--     method_cat, -->
<!--     method_num, -->
<!--     commensurable -->
<!--   ) |> -->
<!--   dplyr::slice_head(n = 10) -->
<!-- ``` -->

<!-- The same table can be passed to `benchmark_mdist()`. -->

<!-- ```{r} -->
<!-- bench <- benchmark_mdist( -->
<!--   penguins_small, -->
<!--   response = species, -->
<!--   specs = specs -->
<!-- ) -->

<!-- bench |> -->
<!--   dplyr::select( -->
<!--     spec_type, -->
<!--     preset, -->
<!--     method_cat, -->
<!--     method_num, -->
<!--     commensurable, -->
<!--     ok, -->
<!--     error -->
<!--   ) |> -->
<!--   dplyr::slice_head(n = 10) -->
<!-- ``` -->
