---
title: "Distance-based clustering workflows"
author: "Alfonso Iodice D'Enza"
format:
  html:
    number-sections: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Distance-based clustering separates two decisions:

1. how observations are compared;
2. how the resulting dissimilarities are partitioned.

This separation is particularly useful for mixed-type data. `manydist`
constructs the dissimilarity, while clustering methods such as partitioning
around medoids (PAM) and spectral clustering operate on that common
representation.

This guide covers both direct use of an `MDist` object and the integrated
recipe interface:

- `step_mdist(output = "pairwise")` constructs within-training
  dissimilarities;
- `pam_dist()` fits a PAM clustering;
- `spectral_dist()` fits a spectral clustering and exposes its embedding.

## Setup

```{r}
#| label: setup

library(manydist)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(recipes)
```

We use mixed numerical and categorical predictors from `palmerpenguins`.
Species is retained only for post-hoc interpretation; it is not used to form
the clusters.

```{r}
#| label: prepare-data

penguins_small <- palmerpenguins::penguins |>
  dplyr::select(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  tidyr::drop_na()

penguin_x <- penguins_small |>
  dplyr::select(-species)

dplyr::glimpse(penguin_x)
```

## Starting from an `MDist` object

The direct workflow is useful when a clustering function already accepts a
`dist` object. Here, `mdist()` constructs a Gower dissimilarity and
`cluster::pam()` performs the partitioning.

```{r}
#| label: direct-gower-distance

d_gower <- mdist(
  penguin_x,
  preset = "gower"
)

d_gower
```

```{r}
#| label: direct-pam-fit

pam_direct <- cluster::pam(
  x = d_gower$to_dist(),
  k = 3,
  diss = TRUE
)

pam_direct
```

Cluster numbers have no intrinsic ordering. A contingency table is therefore
more informative than comparing the numeric cluster labels directly.

```{r}
#| label: direct-pam-table

tibble::tibble(
  cluster = factor(pam_direct$clustering),
  species = penguins_small$species
) |>
  dplyr::count(cluster, species)
```

The species labels are used here only to interpret an example whose classes
are known. In an ordinary unsupervised analysis, cluster validation should
instead rely on stability, separation, and subject-matter interpretation.

## A recipe for pairwise dissimilarities

The integrated workflow begins with a recipe. For clustering,
`output = "pairwise"` tells `step_mdist()` to replace the original predictors
with the square within-training dissimilarity matrix.

```{r}
#| label: pairwise-recipe

gower_pairwise_recipe <- recipes::recipe(
  ~ .,
  data = penguin_x
) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "pairwise"
  )

gower_pairwise_recipe
```

`output = "pairwise"` is intended for training a clustering model. Baking
genuinely new observations in this mode is not supported directly, because
new-to-training dissimilarities are rectangular. The fitted clustering models
handle that prediction step internally.

## PAM with `pam_dist()`

`pam_dist()` defines the clustering specification. The model is fitted with
the recipe and the original predictor data.

```{r}
#| label: pam-workflow-fit

pam_spec <- pam_dist(num_clusters = 3)

pam_fit <- generics::fit(
  pam_spec,
  recipe = gower_pairwise_recipe,
  data = penguin_x
)

pam_fit
```

Calling `predict()` without `new_data` returns the training assignments.

```{r}
#| label: pam-training-predictions

pam_assignments <- predict(pam_fit) |>
  dplyr::bind_cols(
    penguins_small |>
      dplyr::select(species)
  )

pam_assignments |>
  dplyr::count(.pred_cluster, species)
```

Distances to the fitted medoids are also available.

```{r}
#| label: pam-medoid-distances

predict(pam_fit, type = "dist") |>
  dplyr::slice_head(n = 6)
```

### Assigning new observations

For a genuine train/test illustration, we fit PAM on one subset and assign
each held-out observation to its nearest fitted medoid.

```{r}
#| label: clustering-split

set.seed(2026)

cluster_split <- rsample::initial_split(
  penguins_small,
  prop = 0.8,
  strata = species
)

cluster_train <- rsample::training(cluster_split)
cluster_test <- rsample::testing(cluster_split)

cluster_train_x <- cluster_train |>
  dplyr::select(-species)

cluster_test_x <- cluster_test |>
  dplyr::select(-species)
```

```{r}
#| label: pam-new-data

train_pairwise_recipe <- recipes::recipe(
  ~ .,
  data = cluster_train_x
) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "pairwise"
  )

pam_train_fit <- generics::fit(
  pam_dist(num_clusters = 3),
  recipe = train_pairwise_recipe,
  data = cluster_train_x
)

pam_test_assignments <- predict(
  pam_train_fit,
  new_data = cluster_test_x
)

pam_test_assignments |>
  dplyr::bind_cols(cluster_test |> dplyr::select(species)) |>
  dplyr::count(.pred_cluster, species)
```

The fitted `step_mdist()` preprocessor is reused, so held-out observations are
compared with the training observations on the training scale.

## Spectral clustering with `spectral_dist()`

Spectral clustering converts dissimilarities into a Gaussian affinity,
constructs a spectral embedding, and partitions that embedding with k-means.
The same pairwise recipe can be reused.

```{r}
#| label: spectral-fit

set.seed(2026)

spectral_spec <- spectral_dist(
  num_clusters = 3,
  sigma = NULL,
  nstart = 50
)

spectral_fit <- generics::fit(
  spectral_spec,
  recipe = gower_pairwise_recipe,
  data = penguin_x
)

spectral_fit
```

With `sigma = NULL`, the bandwidth is chosen from the median pairwise
dissimilarity. Training assignments are returned by the default prediction
type.

```{r}
#| label: spectral-training-predictions

spectral_assignments <- predict(spectral_fit) |>
  dplyr::bind_cols(
    penguins_small |>
      dplyr::select(species)
  )

spectral_assignments |>
  dplyr::count(.pred_cluster, species)
```

The spectral coordinates can be extracted with `type = "embed"`.

```{r}
#| label: spectral-embedding

spectral_embedding <- predict(
  spectral_fit,
  type = "embed"
) |>
  dplyr::bind_cols(
    spectral_assignments |>
      dplyr::select(.pred_cluster, species)
  )

spectral_embedding |>
  dplyr::slice_head(n = 6)
```

```{r}
#| label: plot-spectral-embedding
#| fig-alt: "Scatter plot of the first two spectral coordinates, coloured by cluster and shaped by penguin species."

spectral_embedding |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = dim_1,
      y = dim_2,
      colour = .pred_cluster,
      shape = species
    )
  ) +
  ggplot2::geom_point(alpha = 0.75, size = 2) +
  ggplot2::labs(
    x = "Spectral dimension 1",
    y = "Spectral dimension 2",
    colour = "Cluster"
  ) +
  ggplot2::theme_minimal()
```

## Comparing distance choices

Changing the dissimilarity can change the clustering even when the partitioning
method and number of clusters remain fixed. A compact comparison can reuse the
same workflow with several presets.

```{r}
#| label: compare-pam-presets

fit_pam_preset <- function(preset) {
  rec <- recipes::recipe(~ ., data = penguin_x) |>
    step_mdist(
      recipes::all_predictors(),
      preset = preset,
      output = "pairwise"
    )

  generics::fit(
    pam_dist(num_clusters = 3),
    recipe = rec,
    data = penguin_x
  )
}

pam_comparison <- tibble::tibble(
  preset = c("gower", "hl", "euclidean")
) |>
  dplyr::mutate(
    fit = purrr::map(preset, fit_pam_preset),
    assignments = purrr::map(fit, predict),
    adjusted_rand = purrr::map_dbl(
      assignments,
      ~ aricode::ARI(
        as.integer(penguins_small$species),
        as.integer(.x$.pred_cluster)
      )
    )
  )

pam_comparison |>
  dplyr::select(preset, adjusted_rand)
```

The adjusted Rand index is possible here because a known external label is
available. It should not be interpreted as an internal clustering criterion,
and it should not be optimised against labels that are meant to remain hidden.
For a broader, auditable set of distance specifications, see the
[benchmarking guide](benchmarking_distance_choices.html).

## Practical guidance

- Use `output = "pairwise"` for training-only clustering representations.
- Keep outcomes or reference labels outside the clustering recipe unless a
  deliberately response-aware analysis is intended. If a recipe formula does
  contain an outcome, response-aware specifications use it by default; set
  `response_used = FALSE` to retain an outcome-independent clustering
  distance.
- Treat cluster labels as nominal identifiers: cluster 1 is not inherently
  smaller or better than cluster 2.
- Set the random seed when fitting spectral clustering.
- Compare plausible distance definitions as part of the clustering
  sensitivity analysis, not as an afterthought.
- For new observations, call `predict()` on the fitted `pam_dist` or
  `spectral_dist` object so that the training preprocessor is reused.
