---
title: "Nearest-neighbor learning workflows"
author: "Alfonso Iodice D'Enza"
format:
  html:
    number-sections: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Nearest-neighbour prediction is defined by a dissimilarity: a new observation
is compared with the training observations, and its prediction is formed from
the closest cases. For mixed-type predictors, the distance definition is
therefore part of the model.

`manydist` integrates this idea with tidymodels through:

- `step_mdist(output = "distance_to_training")`, which replaces the original
  predictors with distances to the training observations;
- `nearest_neighbor_dist()`, a parsnip model specification that consumes those
  distance columns.

This guide builds, tunes, and evaluates a mixed-data nearest-neighbour
classifier while keeping all preprocessing inside the resampling workflow.

## Setup

```{r}
#| label: setup

library(manydist)
library(dplyr)
library(tidyr)
library(ggplot2)
library(recipes)
library(rsample)
library(parsnip)
library(workflows)
library(tune)
library(yardstick)
```

We predict penguin species from four numerical and two categorical variables.

```{r}
#| label: prepare-data

penguins_small <- palmerpenguins::penguins |>
  dplyr::select(
    species,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    island,
    sex
  ) |>
  tidyr::drop_na()

dplyr::glimpse(penguins_small)
```

## Creating a final test set

The test set is held aside before any tuning. Stratification preserves the
species proportions in both partitions.

```{r}
#| label: initial-split

set.seed(2026)

penguin_split <- rsample::initial_split(
  penguins_small,
  prop = 0.75,
  strata = species
)

penguin_train <- rsample::training(penguin_split)
penguin_test <- rsample::testing(penguin_split)

dplyr::bind_rows(
  penguin_train |>
    dplyr::count(species) |>
    dplyr::mutate(partition = "training"),
  penguin_test |>
    dplyr::count(species) |>
    dplyr::mutate(partition = "test")
) |>
  dplyr::select(partition, species, n)
```

The test observations are used only after the workflow and the number of
neighbours have been selected.

## A distance-to-training recipe

The recipe outcome is `species`. Selecting `all_predictors()` keeps it out of
the predictor set. Because Gower distance is not response-aware, `species` is
not used in the distance calculation below.

For response-aware specifications such as `u_dep` and `u_mix`, however,
`step_mdist()` automatically discovers the single recipe outcome and uses the
outcomes from the current analysis set when the step is prepared. The fitted
profiles are reused for assessment and test observations; their outcomes are
never required or used. Set `response_used = FALSE` to construct a
predictor-only version of a response-aware distance.

```{r}
#| label: distance-recipe

gower_recipe <- recipes::recipe(
  species ~ .,
  data = penguin_train
) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "distance_to_training"
  )

gower_recipe
```

During `prep()`, the step fits the distance preprocessor and stores the
training predictors. During `bake()`, it creates columns named `dist_1`,
`dist_2`, and so on.

```{r}
#| label: inspect-distance-representation

prepared_gower_recipe <- recipes::prep(
  gower_recipe,
  training = penguin_train
)

training_distances <- recipes::bake(
  prepared_gower_recipe,
  new_data = penguin_train
)

training_distances |>
  dplyr::select(species, dplyr::starts_with("dist_")) |>
  dplyr::slice_head(n = 5) |>
  dplyr::select(1:7)
```

For training data, the distance block is square. For new data, it is
rectangular: each row is a new observation and each column is one training
observation.

```{r}
#| label: inspect-test-distance-dimensions

test_distances <- recipes::bake(
  prepared_gower_recipe,
  new_data = penguin_test
)

c(
  test_rows = nrow(test_distances),
  training_distance_columns = sum(
    startsWith(names(test_distances), "dist_")
  )
)
```

This check is useful for understanding the representation, but a fitted
workflow normally performs `prep()` and `bake()` automatically.

## Defining the nearest-neighbour model

`nearest_neighbor_dist()` uses precomputed distance columns. We mark
`neighbors` for tuning and explicitly select the `manydist` engine.

```{r}
#| label: knn-specification

knn_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = tune::tune()
) |>
  parsnip::set_engine("manydist")

knn_spec
```

The recipe and model specification are combined in a workflow.

```{r}
#| label: knn-workflow

knn_workflow <- workflows::workflow() |>
  workflows::add_recipe(gower_recipe) |>
  workflows::add_model(knn_spec)

knn_workflow
```

## Tuning the number of neighbours

Cross-validation estimates performance using only the training partition.
Because the distance transformation lives inside the recipe, every assessment
fold is compared only with the corresponding analysis fold. This prevents
preprocessing leakage.

```{r}
#| label: cross-validation-folds

set.seed(2026)

penguin_folds <- rsample::vfold_cv(
  penguin_train,
  v = 5,
  strata = species
)

neighbor_grid <- tibble::tibble(
  neighbors = c(1L, 3L, 5L, 7L, 9L, 15L)
)
```

```{r}
#| label: tune-neighbors

set.seed(2026)

knn_results <- tune::tune_grid(
  knn_workflow,
  resamples = penguin_folds,
  grid = neighbor_grid,
  metrics = yardstick::metric_set(yardstick::accuracy),
  control = tune::control_grid(save_pred = TRUE)
)

knn_metrics <- tune::collect_metrics(knn_results)

knn_metrics
```

```{r}
#| label: plot-tuning-results
#| fig-alt: "Line plot of cross-validated classification accuracy against the number of neighbours."

knn_metrics |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = neighbors,
      y = mean
    )
  ) +
  ggplot2::geom_line(colour = "#2C7FB8") +
  ggplot2::geom_point(size = 2, colour = "#2C7FB8") +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      ymin = mean - std_err,
      ymax = mean + std_err
    ),
    width = 0.4
  ) +
  ggplot2::scale_x_continuous(breaks = neighbor_grid$neighbors) +
  ggplot2::labs(
    x = "Number of neighbours",
    y = "Cross-validated accuracy"
  ) +
  ggplot2::theme_minimal()
```

The best value is selected using the same metric.

```{r}
#| label: select-best-neighbors

best_neighbors <- tune::select_best(
  knn_results,
  metric = "accuracy"
)

best_neighbors
```

## Final fit and test-set evaluation

The selected value is inserted into the workflow, which is then fitted once to
all training observations.

```{r}
#| label: final-fit

final_knn_workflow <- tune::finalize_workflow(
  knn_workflow,
  best_neighbors
)

final_knn_fit <- parsnip::fit(
  final_knn_workflow,
  data = penguin_train
)

final_knn_fit
```

Class predictions and class probabilities can both be requested for the held
out observations.

```{r}
#| label: test-predictions

test_predictions <- predict(
  final_knn_fit,
  new_data = penguin_test
) |>
  dplyr::bind_cols(
    predict(
      final_knn_fit,
      new_data = penguin_test,
      type = "prob"
    )
  ) |>
  dplyr::bind_cols(
    penguin_test |>
      dplyr::select(species)
  )

test_predictions |>
  dplyr::slice_head(n = 8)
```

```{r}
#| label: test-accuracy

yardstick::accuracy(
  test_predictions,
  truth = species,
  estimate = .pred_class
)
```

```{r}
#| label: test-confusion-matrix

yardstick::conf_mat(
  test_predictions,
  truth = species,
  estimate = .pred_class
)
```

The test result estimates the performance of the entire selected workflow:
Gower preprocessing, the distance-to-training representation, and the tuned
nearest-neighbour rule.

## Using a custom distance specification

Presets are not required. The same workflow can use a custom categorical
dissimilarity, numerical preprocessing method, and commensurability choice.

```{r}
#| label: custom-distance-recipe

custom_recipe <- recipes::recipe(
  species ~ .,
  data = penguin_train
) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "custom",
    method_cat = "matching",
    method_num = "std",
    commensurable = TRUE,
    output = "distance_to_training"
  )

custom_workflow <- workflows::workflow() |>
  workflows::add_recipe(custom_recipe) |>
  workflows::add_model(knn_spec)

custom_workflow
```

To compare this workflow fairly with the Gower workflow, tune both on the same
resampling folds and with the same metric. The distance specification is a
modeling choice, so selecting it after looking at the final test result would
give an optimistic assessment. The
[benchmarking guide](benchmarking_distance_choices.html) shows how to
construct auditable tables of candidate distance definitions.

## Practical guidance

- Split off the final test set before selecting the distance or the number of
  neighbours.
- Keep `step_mdist()` inside the recipe so its preprocessor is re-estimated
  within each resample.
- Response-aware specifications automatically use the analysis-fold outcome;
  use `response_used = FALSE` when an outcome-independent distance is wanted.
- Use `output = "distance_to_training"` for prediction; `output = "pairwise"`
  is intended for clustering.
- Remember that each fitted analysis set produces a different number of
  distance columns. The workflow handles this automatically.
- Tune the distance definition and the number of neighbours under the same
  resampling design when both are uncertain.
- Request class probabilities when the application needs calibrated decisions
  or probability-based performance metrics, not only hard class labels.
