---
author: Alfonso Iodice D'Enza
title: Distance-based learning with manydist and tidymodels
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r } 
#| label: setup 
#| include: true

set.seed(123) 
rm(list = ls()) 
devtools::load_all("../") 
#library(manydist) 
library(tidyverse) 
library(tidymodels)
library(palmerpenguins) 
```

# Introduction

The **manydist** package provides a flexible framework for distance-based
learning with continuous, categorical, and mixed-type predictors. Its
core function `mdist()` computes mixed-variable dissimilarities under a
range of presets and custom specifications, while the `MDist` class
stores both the resulting distance matrix and the associated metadata.
The package also integrates with the **tidymodels** ecosystem through a
recipe step, `step_mdist()`, and a custom distance-based nearest-neighbour
model, `nearest_neighbor_dist()`.

In this vignette we use the `palmerpenguins` data to show:

1. How to compute mixed-type distances with `mdist()` in base R.
2. How to use `step_mdist()` to obtain a distance-based representation in a recipe.
3. How to fit and tune a k-nearest neighbours model based on manydist
   via `nearest_neighbor_dist()` and `tune_grid()`.
4. How the same distance machinery can be used consistently across
   direct and tidymodels-based workflows.

The focus is on *classification* of penguin species, but the same ideas
apply to other supervised or unsupervised tasks that rely on pairwise
distances.

# Data: Palmer Penguins

We start from the `penguins` data and remove rows with missing values.

```{r penguins}
penguins_clean <- penguins |>
  drop_na()

penguins_clean |>
  count(species)
```



# Variable influence diagnostics with `lovo_mdist()`

The package also provides a leave-one-variable-out diagnostic through
`lovo_mdist()`. For each predictor, the function recomputes the distance
after removing that variable and compares the resulting structure with
the full-distance solution.

```{r lovo}
#| cache: true
lovo_res <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "unbiased_dependent"
)

summary(lovo_res)
```


# Comparing LOVO diagnostics across methods

The function `compare_lovo_mdist()` extends the leave-one-variable-out
diagnostic to multiple distance specifications. This is useful when we
want to assess how variable contributions change across alternative
mixed-type distances.

Here we compare two presets:

- `"gower"`, a standard mixed-type distance,
- `"unbiased_dependent"`, an association-aware distance that enforces
  commensurability.

The output is an `MDistLOVOCompare` object that supports `print()`,
`summary()`, and `autoplot()`.

```{r lovo-compare}
#| cache: true
lovo_cmp <- compare_lovo_mdist(
  x = penguins_clean |> select(-species),
  methods = list(
    gower = list(
      preset = "gower"
    ),
    u_dep = list(
      preset = "unbiased_dependent"
    ),
    custom = list(
      distance_cont = "manhattan",
      distance_cat  = "HL",
      commensurable = FALSE
    )
  )
)



lovo_cmp

```

### Summary chunk

```{r}
summary(lovo_cmp)
```

### plot chunk


```{r}
#| cache: true
autoplot(lovo_cmp, metric = "mad_normalized")
```

## K nearest neighbours

We create a stratified train/test split:

```{r split} 
set.seed(123) 
peng_split <- initial_split(penguins_clean, strata = species) 
peng_train <- training(peng_split) 
peng_test  <- testing(peng_split)
dim(peng_train) 
dim(peng_test)
```

# Base R: using `mdist()` directly

```{r base-mdist}
x_train <- peng_train |>
  select(-species)
x_test  <- peng_test |>
  select(-species)
y_train <- peng_train$species
y_test  <- peng_test$species
```

```{r base-mdist-compute} 
d_obj <- mdist( x = x_train, new_data = x_test, preset = "unbiased_dependent" )
d_obj
```



```{r base-mdist-inspect}
class(d_obj)
distance_matrix <- as.matrix(d_obj$distance)
dim(distance_matrix)
```


```{r base-knn-helper} 
distance_matrix <- as.matrix(d_obj$distance)
n_neighbours = 5
nn_idx <- apply(distance_matrix, 1, function(d) order(d)[1:n_neighbours])
pred_base <- apply(nn_idx, 2, function(idx) {
  names(which.max(table(y_train[idx])))
})

mean(pred_base == y_test)

```


# Recipes: distance-based representation with `step_mdist()`

```{r rec-mdist} 
peng_rec <- recipe(species ~ ., data = peng_train)|> 
  step_mdist(all_predictors(), preset = "unbiased_dependent")

peng_rec
```


```{r rec-mdist-prep}
prepped_rec <- peng_rec |>
  prep()
```

```{r rec-mdist-bake} 
train_dist <- bake(prepped_rec, new_data = peng_train) 
test_dist  <- bake(prepped_rec, new_data = peng_test)
```

# Distance-based KNN with parsnip

```{r model-spec} 
knn_spec <- nearest_neighbor_dist( mode ="classification", neighbors = 5 ) |> 
  set_engine("precomputed")

knn_spec
```

```{r workflow}
peng_wf <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_spec)

peng_wf
```

```{r fit-workflow} 
peng_fit <- peng_wf |>   fit(data = peng_train)
```

```{r predict-prob-class} 
peng_pred <- predict(peng_fit, new_data = peng_test, type = "prob") |> 
  bind_cols(predict(peng_fit, new_data =peng_test, type = "class"), peng_test |> select(species))

peng_pred |>  head()

peng_pred |>
  accuracy(truth = species, estimate = .pred_class)
```



# Tuning the number of neighbors

```{r tune-spec} 
knn_tune_spec <- nearest_neighbor_dist( mode ="classification", neighbors = tune() ) |>  set_engine("precomputed")

knn_tune_spec
```

```{r tune-workflow}
peng_wf_tune <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_tune_spec)

```


```{r tune-grid}
#| cache: true
set.seed(123) 
peng_folds <- vfold_cv(peng_train, v = 5, strata = species)

k_grid <- tibble(neighbors = seq(3, 15, by = 2)) 
k_grid


peng_metrics <- metric_set(accuracy)

tuned_knn <- tune_grid(
  peng_wf_tune,
  resamples = peng_folds,
  grid = k_grid,
  metrics = peng_metrics
)

```


```{r tune-results} 
collect_metrics(tuned_knn) 
show_best(tuned_knn,metric = "accuracy")

autoplot(tuned_knn)
```

```{r finalize}
best_k <- select_best(tuned_knn, metric = "accuracy")

final_wf <- peng_wf_tune |>
  finalize_workflow(best_k)

final_fit <- final_wf |>
  fit(data = peng_train)

final_pred <- predict(final_fit, new_data = peng_test, type = "class") |>
  bind_cols(peng_test |>
  select(species))

final_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Summary

In this vignette we showed how **manydist** can be used for
distance-based learning on mixed data using `mdist()`, `step_mdist()`,
and `nearest_neighbor_dist()` within the tidymodels ecosystem.
