---
author: Alfonso Iodice D'Enza
title: Distance-based learning with manydist and tidymodels
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: true

set.seed(123)
rm(list = ls())
devtools::load_all("../")
# library(manydist)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
```

# Introduction

The manydist package provides a flexible framework for distance-based
learning with continuous, categorical, and mixed-type predictors. Its
core function `mdist()` computes mixed-variable dissimilarities under a
range of presets and custom specifications, while the `MDist` class
stores both the resulting distance matrix and the associated metadata.

The package also integrates with the tidymodels ecosystem through a
recipe step, `step_mdist()`, a custom distance-based nearest-neighbour
model, `nearest_neighbor_dist()`, a distance-based partitioning around
medoids implementation, `pam_dist()`, and a distance-based spectral
clustering implementation, `spectral_dist()`.

This vignette illustrates the main functionalities of the package:

1. computing mixed-type distances with `mdist()`;
2. assessing variable influence with `lovo_mdist()`;
3. comparing diagnostics across methods with `compare_lovo_mdist()`;
4. creating distance-based representations with `step_mdist()`;
5. fitting and tuning a KNN model with `nearest_neighbor_dist()`;
6. fitting a distance-based PAM clustering model with `pam_dist()`;
7. fitting a distance-based spectral clustering model with `spectral_dist()`.

The focus is on classification of penguin species, but the same ideas
apply to other supervised or unsupervised tasks that rely on pairwise
distances.

# Data

We start from the `penguins` data and remove rows with missing values.

```{r penguins}
penguins_clean <- palmerpenguins::penguins |>
  drop_na()

penguins_clean |>
  dplyr::count(species)
```

# Computing mixed-type distances with `mdist()`

The main entry point of the package is `mdist()`, which computes a
mixed-variable dissimilarity according to a chosen preset or custom
specification.

```{r mdist-core}
d_obj_all <- mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep"
)

d_obj_all
```

The resulting object stores the distance matrix together with the
distance specification.

```{r mdist-core-inspect}
class(d_obj_all)
dim(base::as.matrix(d_obj_all$distance))
```

# Variable diagnostics with `lovo_mdist()`

The function `lovo_mdist()` provides a leave-one-variable-out diagnostic.
For each predictor, the function recomputes the distance after removing
that variable and compares the resulting structure with the full-distance
solution.

```{r lovo}
#| cache: false
lovo_res <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep"
)

summary(lovo_res)
```

A graphical summary is also available through `autoplot()`.

```{r lovo-plot}
autoplot(lovo_res, "mad_normalized")
```

## LOVO with clustering-based diagnostics

In addition to distance- and MDS-based diagnostics, `lovo_mdist()`
can optionally assess how much the clustering solution changes when each
variable is removed. This is activated by specifying the number of
clusters through `cluster_k`.

In the next example, we compute leave-one-variable-out diagnostics for
the same distance specification as above, and compare the full-data
partition with the leave-one-out partitions using both PAM and
hierarchical clustering. Agreement is measured through the adjusted Rand
index (ARI), while `pam_importance` and `hclust_importance` are defined
as `1 - ARI`, so that larger values indicate stronger influence of the
left-out variable.

```{r lovo-cluster}
#| cache: false
lovo_res_clust <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep",
  cluster_k = 3
)

summary(lovo_res_clust)
```

The clustering-based importance measures can be visualized with
`autoplot()` in the same way as the original LOVO diagnostics.

```{r lovo-cluster-plot-pam}
autoplot(lovo_res_clust, metric = "pam_importance")
```

```{r lovo-cluster-plot-hclust}
autoplot(lovo_res_clust, metric = "hclust_importance")
```

If needed, only one clustering method can be requested.

```{r lovo-cluster-pam-only}
lovo_res_pam <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep",
  cluster_k = 3,
  cluster_methods = "pam"
)

autoplot(lovo_res_pam, metric = "pam_importance")
```

# Comparing distance specifications with `compare_lovo_mdist()`

The function `compare_lovo_mdist()` extends the LOVO diagnostic to
multiple distance specifications. This makes it possible to compare how
variable contributions change across alternative mixed-type distances.

In the example below, we compare:

- `"gower"`, a standard mixed-type distance;
- `"u_dep"`, an association-aware distance that enforces
  commensurability;
- a custom specification based on Manhattan distance for numeric
  variables and Hennig--Liao distance for categorical variables.

```{r lovo-compare}
#| cache: true
lovo_cmp <- compare_lovo_mdist(
  x = penguins_clean |> select(-species),
  methods = list(
    gower = list(preset = "gower"),
    u_dep = list(preset = "u_dep"),
    custom = list(
      distance_cont = "manhattan",
      distance_cat  = "HL",
      commensurable = FALSE
    )
  )
)

lovo_cmp
```

```{r lovo-compare-summary}
summary(lovo_cmp)
```

```{r lovo-compare-plot}
#| cache: true
autoplot(lovo_cmp, metric = "mad_normalized")
```

# Distance-based KNN in base R

We now move to a simple predictive setting and create a stratified
train/test split.

```{r split}
set.seed(123)
peng_split <- initial_split(penguins_clean, strata = species)
peng_train <- training(peng_split)
peng_test  <- testing(peng_split)

dim(peng_train)
dim(peng_test)
```

We separate predictors and outcome:

```{r base-mdist}
x_train <- peng_train |>
  select(-species)
x_test  <- peng_test |>
  select(-species)
y_train <- peng_train$species
y_test  <- peng_test$species
```

We compute train-to-test distances with `mdist()`.

```{r base-mdist-compute}
test_d_obj <- mdist(
  x = x_train,
  new_data = x_test,
  preset = "u_dep"
)

test_d_obj
```

```{r base-mdist-inspect}
distance_matrix <- base::as.matrix(test_d_obj$distance)
dim(distance_matrix)
```

These distances can be used directly in a minimal KNN implementation.

```{r base-knn}
n_neighbours <- 5

nn_idx <- apply(distance_matrix, 1, function(d) order(d)[1:n_neighbours])

pred_base <- apply(nn_idx, 2, function(idx) {
  names(which.max(table(y_train[idx])))
})

mean(pred_base == y_test)
```

# Distance-based representations with `step_mdist()`

To integrate with `tidymodels`, manydist provides the recipe step
`step_mdist()`, which computes dissimilarities relative to the training
set.

- `"distance_to_training"` (default): a rectangular test-to-training
  distance matrix, suited to supervised learners such as KNN;
- `"pairwise"`: a square within-sample distance matrix, suited to
  clustering and other unsupervised methods.

```{r rec-mdist}
peng_rec <- recipe(species ~ ., data = peng_train) |>
  step_mdist(all_predictors(), preset = "u_dep")

peng_rec
```

```{r rec-mdist-prep}
prepped_rec <- peng_rec |>
  prep()
```

As a quick sanity check, the distance matrix obtained through the recipe
matches the one computed directly with `mdist()`.

```{r rec-mdist-bake}
train_dist <- bake(prepped_rec, new_data = peng_train)
prepped_test_d_obj <- bake(prepped_rec, new_data = peng_test)

sum(prepped_test_d_obj |> select(-species) - test_d_obj$distance)
```

# Distance-based KNN with `nearest_neighbor_dist()`

The package also provides a custom parsnip model,
`nearest_neighbor_dist()`, with a `"precomputed"` engine.

```{r model-spec}
knn_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = 5
) |>
  set_engine("precomputed")

knn_spec
```

We can combine the recipe and the model in a workflow.

```{r workflow}
peng_wf <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_spec)

peng_wf
```

```{r fit-workflow}
peng_fit <- peng_wf |>
  fit(data = peng_train)
```

Predictions and performance evaluation are then straightforward.

```{r predict-prob-class}
peng_pred <- predict(peng_fit, new_data = peng_test, type = "prob") |>
  bind_cols(
    predict(peng_fit, new_data = peng_test, type = "class"),
    peng_test |> select(species)
  )

peng_pred |>
  head()
```

```{r eval-acc}
peng_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Tuning the number of neighbors

Because `nearest_neighbor_dist()` is registered as a parsnip model, its
hyperparameters can be tuned using `tune_grid()`.

```{r tune-spec}
knn_tune_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = tune()
) |>
  set_engine("precomputed")

knn_tune_spec
```

```{r tune-workflow}
peng_wf_tune <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_tune_spec)
```

```{r tune-grid}
#| cache: true
set.seed(123)
peng_folds <- vfold_cv(peng_train, v = 5, strata = species)

k_grid <- tibble(neighbors = seq(3, 15, by = 2))
k_grid

peng_metrics <- metric_set(accuracy)

tuned_knn <- tune_grid(
  peng_wf_tune,
  resamples = peng_folds,
  grid = k_grid,
  metrics = peng_metrics
)
```

```{r tune-results}
collect_metrics(tuned_knn)
show_best(tuned_knn, metric = "accuracy")

autoplot(tuned_knn)
```

After selecting the best value of `k`, the workflow can be finalized and
re-fitted on the training set.

```{r finalize}
best_k <- select_best(tuned_knn, metric = "accuracy")

final_wf <- peng_wf_tune |>
  finalize_workflow(best_k)

final_fit <- final_wf |>
  fit(data = peng_train)

final_pred <- predict(final_fit, new_data = peng_test, type = "class") |>
  bind_cols(peng_test |> select(species))

final_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Distance-based clustering with `pam_dist()` and `spectral_dist()`

For clustering, `step_mdist()` is used with `output = "pairwise"`, so
that the within-training pairwise dissimilarity matrix is computed and
passed to the clustering method.

We first define a recipe that computes the pairwise dissimilarity matrix
on the training set.

```{r rec-clust}
peng_rec_clust <- recipe(~ ., data = peng_train |> select(-species)) |>
  step_mdist(
    all_predictors(),
    preset = "u_dep",
    output = "pairwise"
  )

peng_rec_clust
```

## Distance-based PAM clustering with `pam_dist()`

We define a PAM clustering specification with three clusters.

```{r pam-spec}
pam_spec <- pam_dist(num_clusters = 3)

pam_spec
```

The model is then fitted by combining the specification with the recipe
and the training predictors.

```{r pam-fit}
pam_fit <- generics::fit(
  pam_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

pam_fit
```

The fitted object returns cluster memberships for the training set.

```{r pam-predict-train}
pam_train_pred <- predict(pam_fit)

pam_train_pred |>
  head()
```

Since the true penguin species are known, we can compare the obtained
partition with the observed classes using the adjusted Rand index (ARI).

```{r pam-train-ari}
mclust::adjustedRandIndex(
  as.integer(pam_train_pred$.pred_cluster),
  as.integer(peng_train$species)
)
```

For new observations, `pam_dist()` computes distances from the test set
to the training set, extracts the distances to the fitted medoids, and
assigns each observation to its nearest medoid.

```{r pam-predict-test}
pam_test_pred <- predict(
  pam_fit,
  new_data = peng_test |> select(-species)
)

pam_test_pred |>
  head()
```

The corresponding ARI on the test set is:

```{r pam-test-ari}
mclust::adjustedRandIndex(
  as.integer(pam_test_pred$.pred_cluster),
  as.integer(peng_test$species)
)
```

If desired, the distances from each test observation to the medoids can
also be extracted directly.

```{r pam-test-dists}
predict(
  pam_fit,
  new_data = peng_test |> select(-species),
  type = "dist"
) |>
  head()
```

## Distance-based spectral clustering with `spectral_dist()`

The usage of `spectral_dist()` parallels that of `pam_dist()`: the model
is fitted on the pairwise training dissimilarity matrix computed by
`step_mdist()`, and predictions for new observations are obtained by
extending the spectral embedding to the test set.

We define a spectral clustering specification with three clusters.

```{r spectral-spec}
spectral_spec <- spectral_dist(
  num_clusters = 3
)

spectral_spec
```

The model is fitted in the same way as `pam_dist()`.

```{r spectral-fit}
spectral_fit <- generics::fit(
  spectral_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

spectral_fit
```

The fitted object returns cluster memberships for the training set.

```{r spectral-predict-train}
spectral_train_pred <- predict(spectral_fit)

spectral_train_pred |>
  head()
```

As for PAM, the resulting partition can be compared with the observed
penguin species using the adjusted Rand index.

```{r spectral-train-ari}
mclust::adjustedRandIndex(
  as.integer(spectral_train_pred$.pred_cluster),
  as.integer(peng_train$species)
)
```

Predictions for new observations are obtained through the stored
distance preprocessor and an out-of-sample extension of the spectral
embedding.

```{r spectral-predict-test}
spectral_test_pred <- predict(
  spectral_fit,
  new_data = peng_test |> select(-species)
)

spectral_test_pred |>
  head()
```

The corresponding ARI on the test set is:

```{r spectral-test-ari}
mclust::adjustedRandIndex(
  as.integer(spectral_test_pred$.pred_cluster),
  as.integer(peng_test$species)
)
```

If desired, the spectral embedding coordinates can also be extracted
directly.

```{r spectral-embed-train}
predict(spectral_fit, type = "embed") |>
  head()
```

```{r spectral-embed-test}
predict(
  spectral_fit,
  new_data = peng_test |> select(-species),
  type = "embed"
) |>
  head()
```

# Summary

In this vignette we showed how **manydist** can be used for
distance-based learning on mixed data through:

- mixed-type distance computation with `mdist()`;
- variable diagnostics with `lovo_mdist()`;
- method comparison with `compare_lovo_mdist()`;
- recipe-based representations with `step_mdist()`;
- distance-based KNN modeling and tuning with `nearest_neighbor_dist()`;
- distance-based PAM clustering with `pam_dist()`;
- distance-based spectral clustering with `spectral_dist()`.
