---
author: Alfonso Iodice D'Enza
title: Distance-based learning with manydist and tidymodels
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: true

set.seed(123)
rm(list = ls())
devtools::load_all("../")
# library(manydist)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
```

# Introduction

The **manydist** package provides a flexible framework for distance-based
learning with continuous, categorical, and mixed-type predictors. Its
core function `mdist()` computes mixed-variable dissimilarities under a
range of presets and custom specifications, while the `MDist` class
stores both the resulting distance matrix and the associated metadata.
The package also integrates with the **tidymodels** ecosystem through a
recipe step, `step_mdist()`, and a custom distance-based nearest-neighbour
model, `nearest_neighbor_dist()`.

This vignette illustrates the main functionalities of the package:

1. computing mixed-type distances with `mdist()`;
2. assessing variable influence with `lovo_mdist()`;
3. comparing diagnostics across methods with `compare_lovo_mdist()`;
4. creating distance-based representations with `step_mdist()`;
5. fitting and tuning a KNN model with `nearest_neighbor_dist()` and `tune_grid()`.

The focus is on *classification* of penguin species, but the same ideas
apply to other supervised or unsupervised tasks that rely on pairwise
distances.

# Data

We start from the `penguins` data and remove rows with missing values.

```{r penguins}
penguins_clean <- penguins |>
  drop_na()

penguins_clean |>
  count(species)
```

# Computing mixed-type distances with `mdist()`

The main entry point of the package is `mdist()`, which computes a
mixed-variable dissimilarity according to a chosen preset or custom
specification.

```{r mdist-core}
d_obj_all <- mdist(
  x = penguins_clean |> select(-species),
  preset = "unbiased_dependent"
)

d_obj_all
```

The resulting object stores the distance matrix together with the
distance specification.

```{r mdist-core-inspect}
class(d_obj_all)
dim(as.matrix(d_obj_all$distance))
```

# Variable diagnostics with `lovo_mdist()`

The function `lovo_mdist()` provides a leave-one-variable-out diagnostic.
For each predictor, the function recomputes the distance after removing
that variable and compares the resulting structure with the full-distance
solution.

```{r lovo}
#| cache: false
lovo_res <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "unbiased_dependent"
)

summary(lovo_res)
```

A graphical summary is also available through `autoplot()`.

```{r lovo-plot}
autoplot(lovo_res, "mad_normalized")
```

# Comparing distance specifications with `compare_lovo_mdist()`

The function `compare_lovo_mdist()` extends the LOVO diagnostic to
multiple distance specifications. This makes it possible to compare how
variable contributions change across alternative mixed-type distances.

In the example below, we compare:

- `"gower"`, a standard mixed-type distance;
- `"unbiased_dependent"`, an association-aware distance that enforces
  commensurability;
- a custom specification based on Manhattan distance for numeric
  variables and Hennig–Liao distance for categorical variables.

```{r lovo-compare}
#| cache: true
lovo_cmp <- compare_lovo_mdist(
  x = penguins_clean |> select(-species),
  methods = list(
    gower = list(
      preset = "gower"
    ),
    u_dep = list(
      preset = "unbiased_dependent"
    ),
    custom = list(
      distance_cont = "manhattan",
      distance_cat  = "HL",
      commensurable = FALSE
    )
  )
)

lovo_cmp
```

```{r lovo-compare-summary}
summary(lovo_cmp)
```

```{r lovo-compare-plot}
#| cache: true
autoplot(lovo_cmp, metric = "mad_normalized")
```

# Distance-based KNN in base R

We now move to a simple predictive setting and create a stratified
train/test split.

```{r split}
set.seed(123)
peng_split <- initial_split(penguins_clean, strata = species)
peng_train <- training(peng_split)
peng_test  <- testing(peng_split)

dim(peng_train)
dim(peng_test)
```

We separate predictors and outcome:

```{r base-mdist}
x_train <- peng_train |>
  select(-species)
x_test  <- peng_test |>
  select(-species)
y_train <- peng_train$species
y_test  <- peng_test$species
```

We compute train-to-test distances with `mdist()`.

```{r base-mdist-compute}
d_obj <- mdist(
  x = x_train,
  new_data = x_test,
  preset = "unbiased_dependent"
)

d_obj
```

```{r base-mdist-inspect}
distance_matrix <- as.matrix(d_obj$distance)
dim(distance_matrix)
```

These distances can be used directly in a minimal KNN implementation.

```{r base-knn}
n_neighbours <- 5

nn_idx <- apply(distance_matrix, 1, function(d) order(d)[1:n_neighbours])

pred_base <- apply(nn_idx, 2, function(idx) {
  names(which.max(table(y_train[idx])))
})

mean(pred_base == y_test)
```

# Distance-based representations with `step_mdist()`

To integrate with `tidymodels`, manydist provides the recipe step
`step_mdist()`, which transforms the predictor set into a
distance-based representation.

```{r rec-mdist}
peng_rec <- recipe(species ~ ., data = peng_train) |>
  step_mdist(all_predictors(), preset = "unbiased_dependent")

peng_rec
```

```{r rec-mdist-prep}
prepped_rec <- peng_rec |>
  prep()
```

```{r rec-mdist-bake}
train_dist <- bake(prepped_rec, new_data = peng_train)
test_dist  <- bake(prepped_rec, new_data = peng_test)
```

# Distance-based KNN with `nearest_neighbor_dist()`

The package also provides a custom parsnip model,
`nearest_neighbor_dist()`, with a `"precomputed"` engine.

```{r model-spec}
knn_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = 5
) |>
  set_engine("precomputed")

knn_spec
```

We can combine the recipe and the model in a workflow.

```{r workflow}
peng_wf <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_spec)

peng_wf
```

```{r fit-workflow}
peng_fit <- peng_wf |>
  fit(data = peng_train)
```

Predictions and performance evaluation are then straightforward.

```{r predict-prob-class}
peng_pred <- predict(peng_fit, new_data = peng_test, type = "prob") |>
  bind_cols(
    predict(peng_fit, new_data = peng_test, type = "class"),
    peng_test |> select(species)
  )

peng_pred |>
  head()
```

```{r eval-acc}
peng_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Tuning the number of neighbors

Because `nearest_neighbor_dist()` is registered as a parsnip model, its
hyperparameters can be tuned using `tune_grid()`.

```{r tune-spec}
knn_tune_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = tune()
) |>
  set_engine("precomputed")

knn_tune_spec
```

```{r tune-workflow}
peng_wf_tune <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_tune_spec)
```

```{r tune-grid}
#| cache: true
set.seed(123)
peng_folds <- vfold_cv(peng_train, v = 5, strata = species)

k_grid <- tibble(neighbors = seq(3, 15, by = 2))
k_grid

peng_metrics <- metric_set(accuracy)

tuned_knn <- tune_grid(
  peng_wf_tune,
  resamples = peng_folds,
  grid = k_grid,
  metrics = peng_metrics
)
```

```{r tune-results}
collect_metrics(tuned_knn)
show_best(tuned_knn, metric = "accuracy")

autoplot(tuned_knn)
```

After selecting the best value of `k`, the workflow can be finalized and
re-fitted on the training set.

```{r finalize}
best_k <- select_best(tuned_knn, metric = "accuracy")

final_wf <- peng_wf_tune |>
  finalize_workflow(best_k)

final_fit <- final_wf |>
  fit(data = peng_train)

final_pred <- predict(final_fit, new_data = peng_test, type = "class") |>
  bind_cols(peng_test |> select(species))

final_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Summary

In this vignette we showed how **manydist** can be used for
distance-based learning on mixed data through:

- mixed-type distance computation with `mdist()`;
- variable diagnostics with `lovo_mdist()`;
- method comparison with `compare_lovo_mdist()`;
- recipe-based representations with `step_mdist()`;
- distance-based KNN modeling and tuning with `nearest_neighbor_dist()`.
