---
author: Alfonso Iodice D'Enza
title: Distance-based learning with manydist and tidymodels
subtitle: development log
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: true

set.seed(123)
rm(list = ls())
devtools::load_all("../")
# library(manydist)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(kableExtra)
```




This document consolidates the current comments and open issues into a single working to-do list, removing redundancies and grouping related points together.

##  Presets and defaults

###  Presets


-  Consider adding a complete set of unbiased presets for consistency with the paper: 
    - ✅  `u_dep` 
    - ✅  `u_indep`
    - ✅  `u_mix`
  

###  Naming

-   ✅ Shorten `"unbiased_dependent"` if possible.
  - it is now `u_dep`
  
- 🟡 Consider shortening `"euclidean_onehot"`.
  
  - not sure about this: the name is self-explanatory, one could just use `euclidean`, but it would not be clear the encoding for the categorical...

###  Defaults

✅  Reconsider the default settings of `mdist()`.

 default values:

```{r, eval=FALSE}
preset="custom"
distance_cont = "manhattan"
scaling_cont  = "std"
distance_cat  = "tvd"
commensurable = TRUE
```


##  Categorical distances

This section tracks the ongoing refactoring of the categorical-distance machinery, mainly through `cdist()`, `cat_delta()`, and `cat_custom_delta()`.

The previous CRAN implementation has been frozen under:
- `cdist_old_cran`
- `cat_delta_old_cran`
- `cat_custom_delta_old_cran`

This makes it possible to check backward compatibility while extending the new implementation.

###  Backward-compatibility checks

The first step was to verify that the refactored code reproduces the previous behavior where it should.

```{r}
load(file = "Dartpoints.RData")

x_cat <- df |>
  dplyr::select(-Name) |>
  dplyr::select(where(is.factor))

all.equal(
  base::as.matrix(cdist_old_cran(x_cat, method = "tvd", response = NULL)$distance_mat),
  base::as.matrix(cdist(x_cat,          method = "tvd", response = NULL)$distance_mat)
)

all.equal(
  base::as.matrix(cdist_old_cran(x_cat, method = "matching")$distance_mat),
  base::as.matrix(cdist(x_cat,          method = "matching")$distance_mat)
)

all.equal(
  base::as.matrix(cdist_old_cran(x_cat, method = "tvd", response = df$Name)$distance_mat),
  base::as.matrix(cdist(x_cat,          method = "tvd", response = df$Name)$distance_mat)
)
```

At this stage, the new implementation is being checked incrementally against the frozen CRAN version before introducing broader extensions.


### Available distances and benchmarking

A key step in the redesign was to make the distance options in `manydist` easier to inspect programmatically. This is useful for at least three reasons:

- it makes the package interface more transparent;
- it supports systematic benchmarking across alternative specifications;
- it clarifies which methods are internally implemented and which rely on external engines such as `philentropy`.

::: {.callout-note appearance="minimal" icon=false}
Provide a way to list all available categorical distances. Consider a helper like `getDistMethods()` in the style of `philentropy`, and allow programmatic access to all options for benchmarking loops.
:::

A first implementation strategy is to rely on:

```{r}
philentropy::getDistMethods()
```

which already provides a full programmatic list of philentropy-supported categorical measures.

However, this alone is not sufficient for `manydist`, because the package exposes a broader interface than the philentropy set alone. In particular, the available methods must also distinguish:

- whether they refer to `distance_cont`, `distance_cat`, `scaling_cont`, or `preset`;
- whether they apply to numeric, categorical, or mixed data;
- whether they are association-based, independence-based, indicator-scaling based, or mixed-block based;
- whether they are response-aware;
- which engine computes them.

To address this, I introduced a helper returning a tibble of method metadata:

```{r}
dist_methods_tbl() |> kbl()
```

The goal of this helper is descriptive rather than operational. It acts as a compact dictionary of the currently available distance-related options in `manydist`, including continuous distances, categorical distances, scaling options, and mixed-data presets.

The current columns record, for each method:

- `method`: the method name;
- `argument`: the argument of `mdist()` to which the method belongs;
- `data_type`: whether it refers to numeric, categorical, or mixed data;
- `distance_basis`: the main methodological family;
- `response_aware`: whether the method can make use of a supplied response;
- `engine`: whether the method is computed by `manydist`, `philentropy`, `kdml`, or through an external port.

This makes the package interface much easier to inspect programmatically. For example, one can now retrieve only the categorical distances, only response-aware methods, or only methods computed through a given engine.

The next step was to move from a descriptive inventory to an actionable benchmarking grid. For that purpose, I added a second helper:

```{r}
selected_specs = all_dist_method_specs(mode = "response_aware",scaling_cont="std",distance_cat=c("jensen-shannon","tvd"))
```



- `dist_methods_tbl()` is descriptive: it tells the user what is available;
- `all_dist_method_specs()` is actionable: it builds a tibble of valid `mdist()` specifications that can actually be run.

Each row of `all_dist_method_specs()` represents one candidate specification for `mdist()`. Depending on the row, this may correspond either to:

- a preset-based specification;
- or a component-based specification combining `distance_cont`, `distance_cat`, `scaling_cont`, and `commensurable`.

This provides the basis for benchmarking workflows, because the specification tibble can be passed directly to a benchmarking wrapper.

To make this concrete, I added:

```{r}
bench_res = benchmark_mdist(df,response = "Name",specs = selected_specs)
bench_res
```



which applies `mdist()` over the rows of a specification tibble, typically created with `all_dist_method_specs()`. In this way, benchmarking no longer depends on ad hoc loops written by the user; instead, the package now has a more explicit infrastructure for method enumeration and comparison.

Overall, this redesign separates three related but distinct tasks:

- method description (`dist_methods_tbl()`);
- method-specification generation (`all_dist_method_specs()`);
- repeated execution for benchmarking (`benchmark_mdist()`).

This makes the package interface more transparent and should also help future extensions, documentation, and systematic comparison studies.

###  Validation of philentropy-based distances

The refactored `cdist()` now supports the philentropy family more systematically.

Initial spot checks:

```{r}
D_hel <- cdist(x_cat, method = "hellinger", commensurable = FALSE)
D_js  <- cdist(x_cat, method = "jensen-shannon", commensurable = FALSE)
D_euc <- cdist(x_cat, method = "euclidean", commensurable = FALSE)
```

Full screening over all philentropy methods:

```{r}
phil_methods <- philentropy::getDistMethods()

phil_checks <- purrr::map_dfr(phil_methods, function(m) {
  out <- tryCatch({
    D <- cdist(x_cat, method = m, commensurable = FALSE)
    mat <- base::as.matrix(D$distance_mat)
    Delta <- base::as.matrix(D$delta)

    tibble::tibble(
      method = m,
      ok = TRUE,
      dist_any_na = any(is.na(mat)),
      dist_any_inf = any(is.infinite(mat)),
      dist_sym = isTRUE(all.equal(mat, t(mat), tolerance = 1e-10)),
      delta_any_na = any(is.na(Delta)),
      delta_any_inf = any(is.infinite(Delta)),
      delta_sym = isTRUE(all.equal(Delta, t(Delta), tolerance = 1e-10)),
      dist_min = min(mat, na.rm = TRUE),
      dist_max = max(mat, na.rm = TRUE)
    )
  }, error = function(e) {
    tibble::tibble(
      method = m,
      ok = FALSE,
      dist_any_na = NA,
      dist_any_inf = NA,
      dist_sym = NA,
      delta_any_na = NA,
      delta_any_inf = NA,
      delta_sym = NA,
      dist_min = NA_real_,
      dist_max = NA_real_,
      error = conditionMessage(e)
    )
  })

  if (!"error" %in% names(out)) out$error <- NA_character_
  out
})

phil_checks
```

The same check under commensurability:

```{r}
phil_checks_comm <- purrr::map_dfr(phil_methods, function(m) {
  out <- tryCatch({
    D <- cdist(x_cat, method = m, commensurable = TRUE)
    mat <- base::as.matrix(D$distance_mat)
    Delta <- base::as.matrix(D$delta)

    tibble::tibble(
      method = m,
      ok = TRUE,
      dist_any_na = any(is.na(mat)),
      dist_any_inf = any(is.infinite(mat)),
      dist_sym = isTRUE(all.equal(mat, t(mat), tolerance = 1e-10)),
      delta_any_na = any(is.na(Delta)),
      delta_any_inf = any(is.infinite(Delta)),
      delta_sym = isTRUE(all.equal(Delta, t(Delta), tolerance = 1e-10)),
      dist_min = min(mat, na.rm = TRUE),
      dist_max = max(mat, na.rm = TRUE)
    )
  }, error = function(e) {
    tibble::tibble(
      method = m,
      ok = FALSE,
      dist_any_na = NA,
      dist_any_inf = NA,
      dist_sym = NA,
      delta_any_na = NA,
      delta_any_inf = NA,
      delta_sym = NA,
      dist_min = NA_real_,
      dist_max = NA_real_,
      error = conditionMessage(e)
    )
  })

  if (!"error" %in% names(out)) out$error <- NA_character_
  out
})

phil_checks_comm
```

These checks establish which philentropy distances behave well inside the manydist framework, both with and without commensurability.

::: {.callout-note appearance="minimal" icon=false}
Original comment:
Extend `mdist()` so that unsupervised association-based distances work beyond `tvd`, ideally including `le_and_ho`, relevant philentropy-based options, and other compatible association-based distances.
:::

The current outcome is that philentropy-based distances can now be validated systematically within `cdist()`, and this provides the basis for extending the supported association-based categorical distances beyond `tvd`.


###  Validation of `le_and_ho`

`le_and_ho` was checked separately because it does not pass through the generic philentropy wrapper.

```{r}
D_lh <- cdist(x_cat, method = "le_and_ho", commensurable = FALSE)

mat_lh <- base::as.matrix(D_lh$distance_mat)
Delta_lh <- base::as.matrix(D_lh$delta)

any(is.na(mat_lh))
any(is.infinite(mat_lh))
isTRUE(all.equal(mat_lh, t(mat_lh), tolerance = 1e-10))

any(is.na(Delta_lh))
any(is.infinite(Delta_lh))
isTRUE(all.equal(Delta_lh, t(Delta_lh), tolerance = 1e-10))

range(mat_lh, na.rm = TRUE)
range(Delta_lh, na.rm = TRUE)
```

At the unsupervised level, `le_and_ho` behaves correctly:
- the distance matrix is finite;
- the distance matrix is symmetric;
- the Delta matrix is finite;
- the Delta matrix is symmetric.

This confirms that `le_and_ho` can be treated as a valid association-based categorical distance within the current `cdist()` framework.

::: {.callout-note appearance="minimal" icon=false}
Original comment:
Extend `mdist()` so that unsupervised association-based distances work beyond `tvd`, ideally including `le_and_ho`.
:::

This specific check supports that extension, at least in the unsupervised setting.

###  Supervised vs unsupervised profile-based distances

The current redesign distinguishes between:

- **response-irrelevant methods**, which ignore `response` and issue a warning;
- **profile-based methods**, which use:
  - unsupervised profiles when `response = NULL`,
  - supervised category-by-response profiles when `response` is supplied.

This logic is now implemented through the interaction between:

- `cdist()`,

- `cat_delta()`,

- `cat_custom_delta()`.

::: {.callout-note appearance="minimal" icon=false}
Clarify and implement the intended supervised behavior. When `response = NULL`, `distance_cat = "tvd"` should behave as the ordinary unsupervised version; when `response` is provided, it should switch to the supervised association-based version. Also consider extending this logic to other association-based categorical distances.
:::

#### `tvd`

```{r}
tvd_unsup <- cdist(x_cat, method = "tvd", response = NULL)
tvd_sup   <- cdist(x_cat, method = "tvd", response = df$Name)

mat_unsup  <- base::as.matrix(tvd_unsup$distance_mat)
mat_sup    <- base::as.matrix(tvd_sup$distance_mat)

delta_unsup <- base::as.matrix(tvd_unsup$delta)
delta_sup   <- base::as.matrix(tvd_sup$delta)

c(
  unsup_na   = any(is.na(mat_unsup)),
  unsup_inf  = any(is.infinite(mat_unsup)),
  unsup_sym  = isTRUE(all.equal(mat_unsup, t(mat_unsup), tolerance = 1e-10)),
  sup_na     = any(is.na(mat_sup)),
  sup_inf    = any(is.infinite(mat_sup)),
  sup_sym    = isTRUE(all.equal(mat_sup, t(mat_sup), tolerance = 1e-10))
)

sum(abs(mat_unsup - mat_sup))
sum(abs(delta_unsup - delta_sup))

isTRUE(all.equal(unname(diag(delta_unsup)), rep(0, nrow(delta_unsup))))
isTRUE(all.equal(unname(diag(delta_sup)), rep(0, nrow(delta_sup))))
```

These checks confirm that:

- both unsupervised and supervised `tvd` are valid;
- both remain symmetric;
- supervision changes both the distance matrix and the Delta matrix;
- both Delta matrices retain zero diagonal.

#### `gifi_chi2`

```{r}
gifi_unsup <- cdist(x_cat, method = "gifi_chi2", response = NULL)
gifi_sup   <- cdist(x_cat, method = "gifi_chi2", response = df$Name)

mat_unsup   <- base::as.matrix(gifi_unsup$distance_mat)
mat_sup     <- base::as.matrix(gifi_sup$distance_mat)
delta_unsup <- base::as.matrix(gifi_unsup$delta)
delta_sup   <- base::as.matrix(gifi_sup$delta)

c(
  unsup_na  = any(is.na(mat_unsup)),
  unsup_inf = any(is.infinite(mat_unsup)),
  unsup_sym = isTRUE(all.equal(mat_unsup, t(mat_unsup), tolerance = 1e-10)),
  sup_na    = any(is.na(mat_sup)),
  sup_inf   = any(is.infinite(mat_sup)),
  sup_sym   = isTRUE(all.equal(mat_sup, t(mat_sup), tolerance = 1e-10))
)

sum(abs(mat_unsup - mat_sup))
sum(abs(delta_unsup - delta_sup))

all(diag(delta_unsup) == 0)
all(diag(delta_sup) == 0)

range(mat_unsup, na.rm = TRUE)
range(mat_sup, na.rm = TRUE)

range(delta_unsup, na.rm = TRUE)
range(delta_sup, na.rm = TRUE)
```

These checks show that supervised and unsupervised logic also works for `gifi_chi2`, with valid and symmetric outputs in both cases.

#### `le_and_ho` and philentropy-based profile distances

A first representative check uses `le_and_ho` and `jensen-shannon`:

```{r}
lh_unsup <- cdist(x_cat, method = "le_and_ho", response = NULL)
lh_sup   <- cdist(x_cat, method = "le_and_ho", response = df$Name)

js_unsup <- cdist(x_cat, method = "jensen-shannon", response = NULL)
js_sup   <- cdist(x_cat, method = "jensen-shannon", response = df$Name)
```

Difference checks:

```{r}
sum(abs(base::as.matrix(lh_unsup$distance_mat) - base::as.matrix(lh_sup$distance_mat)))
sum(abs(base::as.matrix(lh_unsup$delta) - base::as.matrix(lh_sup$delta)))

sum(abs(base::as.matrix(js_unsup$distance_mat) - base::as.matrix(js_sup$distance_mat)))
sum(abs(base::as.matrix(js_unsup$delta) - base::as.matrix(js_sup$delta)))
```

These checks are intended to verify that:

- the supervised profile-based branch is active;
- the supervised version does not collapse to the unsupervised one.



###  Response handling policy

Current design choice:

- profile-based categorical distances use `response` when provided;
- non-profile-based methods ignore `response` and emit a warning.

This is currently the cleanest policy, and it avoids silently overriding the user’s chosen distance.

::: {.callout-note appearance="minimal" icon=false}
Original comment:
When the user provides `response` together with a non-association-based distance such as `"matching"`, decide whether to ignore `response` with a warning or override to a supervised association-based distance.
:::

Current choice:
- **ignore `response` and warn**.

Illustrative checks:

```{r, eval=FALSE}
chk1 <- cdist(x_cat, method = "matching", response = df$Name)
chk2 <- cdist(x_cat, method = "mca", response = df$Name)
chk3 <- cdist(x_cat, method = "tvd", response = df$Name)
chk4 <- cdist(x_cat, method = "le_and_ho", response = df$Name)
```

This policy is consistent with the current separation between:
- response-irrelevant methods;
- profile-based methods.



###  Naming cleanup

The naming transition from `"tot_var_dist"` to `"tvd"` is now part of the cleanup and should be made consistent throughout the package and documentation.

::: {.callout-note appearance="minimal" icon=false}
Original comment:
Replace `"tot_var_dist"` with `"tvd"` consistently.
:::

Current direction:
- normalize `"tot_var_dist"` to `"tvd"` internally;
- use `"tvd"` consistently in examples, checks, and documentation;
- progressively remove legacy references where no longer needed.



##  LOVO framework and outputs

### LOVO with supervised distances

Enable supervised association-based distances in `lovo_mdist()`.

```{r}
mdist(df, response = Name)
mdist(df, response = "Name")
lovo_mdist(df, response = Name, distance_cat = "tvd")
lovo_mdist(df, response = "Name", distance_cat = "tvd")



lovo_res <- lovo_mdist(df, response = Name, distance_cat = "tvd")
lovo_res$results


lovo_comm <- lovo_mdist(df, response = Name, distance_cat = "tvd", commensurable = TRUE)
lovo_nocomm <- lovo_mdist(df, response = Name, distance_cat = "tvd", commensurable = FALSE)
lovo_comm$results

lovo_nocomm$results
```



```{r}
lovo_res_unsup <- lovo_mdist(df |> select(-Name), distance_cat = "tvd")
summary(lovo_res_unsup)
```
Added  the argument `response_used` that can be TRUE or FALSE, if the user want's to compare what the response brings to the table

```{r}
l1 <- lovo_mdist(df, response = Name, distance_cat = "tvd", response_used = FALSE)
l2 <- lovo_mdist(df |> dplyr::select(-Name), distance_cat = "tvd", response_used = FALSE)

all.equal(l1$results, l2$results)
```


###  LOVO comparison design

Rework `lovo_cmp` / `compare_lovo_mdist()` so that supervised and non-supervised methods can be compared in a unified framework.

A possible interface is:

```{r}
#compare_lovo_mdist(X, y, methods)
```

with: - non-supervised methods ignoring `y`; 

- supervised association-based methods using `y`.

Need to decide whether: 

- `y` should be passed directly to the comparison function, or - `y` should be embedded inside each method specification.

###  Variable-importance logic

Generalize LOVO so that importance depends on the downstream task.

Possible structure: - MDS → AC / CC - PAM → ARI or cluster stability metric - Spectral clustering → ARI or similar - More generally: importance should reflect the target use of the distance

###  LOVO metrics and naming

-   Reconsider whether Alienation Coefficient (AC) should remain prominent, since it is closely tied to the paper and may be hard to interpret for general users.
-   Rename `mad_normalized` to something clearer, for example:
    -   `"relative distance"`

###  LOVO plots and output

-   Improve the y-scale in `autoplot()` for normalized MAD / relative distance.
-   Add targeted `summary()` and/or `print()` methods for:
    -   `lovo`
    -   `lovo_cmp`
-   Ensure summary output can produce usable tables, not just raw tibbles.


### `lovo_mdist()` output-method sanity checks

```{r}
# load(file = "Dartpoints.RData")

# core LOVO object without clustering diagnostics
lovo_basic <- lovo_mdist(
  df,
  response = Name,
  distance_cat = "tvd",
  response_used = TRUE,
  commensurable = FALSE
)

# LOVO object with clustering diagnostics
lovo_cluster <- lovo_mdist(
  df,
  response = Name,
  distance_cat = "tvd",
  response_used = TRUE,
  commensurable = FALSE,
  cluster_k = 3
)

check_tbl <- tibble::tibble(
  check = character(),
  passed = logical(),
  detail = character()
)

add_check <- function(tbl, check, expr) {
  out <- tryCatch(
    {
      val <- force(expr)
      tibble::tibble(
        check = check,
        passed = isTRUE(val),
        detail = if (isTRUE(val)) "TRUE" else paste(capture.output(print(val)), collapse = " ")
      )
    },
    error = function(e) {
      tibble::tibble(
        check = check,
        passed = FALSE,
        detail = conditionMessage(e)
      )
    }
  )
  dplyr::bind_rows(tbl, out)
}

# 1. renamed metrics exist in results
check_tbl <- add_check(
  check_tbl,
  "results contain relative_distance",
  "relative_distance" %in% names(lovo_basic$results)
)

check_tbl <- add_check(
  check_tbl,
  "results contain mds_congruence",
  "mds_congruence" %in% names(lovo_basic$results)
)

# 2. backward-compatible metric names still exist
check_tbl <- add_check(
  check_tbl,
  "results retain mad_normalized for compatibility",
  "mad_normalized" %in% names(lovo_basic$results)
)

check_tbl <- add_check(
  check_tbl,
  "results retain cc_importance for compatibility",
  "cc_importance" %in% names(lovo_basic$results)
)

# 3. new and old aliases are equal
check_tbl <- add_check(
  check_tbl,
  "relative_distance equals mad_normalized",
  all.equal(lovo_basic$results$relative_distance, lovo_basic$results$mad_normalized)
)

check_tbl <- add_check(
  check_tbl,
  "mds_congruence equals cc_importance",
  all.equal(lovo_basic$results$mds_congruence, lovo_basic$results$cc_importance)
)

# 4. print() runs without error
check_tbl <- add_check(
  check_tbl,
  "print() works on basic LOVO object",
  {
    capture.output(print(lovo_basic))
    TRUE
  }
)

check_tbl <- add_check(
  check_tbl,
  "print() works on clustering LOVO object",
  {
    capture.output(print(lovo_cluster))
    TRUE
  }
)

# 5. summary() runs without error
check_tbl <- add_check(
  check_tbl,
  "summary() works on basic LOVO object",
  {
    capture.output(summary(lovo_basic))
    TRUE
  }
)

check_tbl <- add_check(
  check_tbl,
  "summary() works on clustering LOVO object",
  {
    capture.output(summary(lovo_cluster))
    TRUE
  }
)

# 6. autoplot() returns ggplot objects for key metrics
p_rel <- lovo_basic$autoplot(metric = "relative_distance")
p_mad <- lovo_basic$autoplot(metric = "mad_importance")
p_mds <- lovo_basic$autoplot(metric = "mds_congruence")
p_ac  <- lovo_basic$autoplot(metric = "ac_importance")

check_tbl <- add_check(
  check_tbl,
  "autoplot(relative_distance) returns ggplot",
  inherits(p_rel, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(mad_importance) returns ggplot",
  inherits(p_mad, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(mds_congruence) returns ggplot",
  inherits(p_mds, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(ac_importance) returns ggplot",
  inherits(p_ac, "ggplot")
)

# 7. clustering metrics only plot when available
p_pam <- lovo_cluster$autoplot(metric = "pam_importance")
p_hc  <- lovo_cluster$autoplot(metric = "hclust_importance")

check_tbl <- add_check(
  check_tbl,
  "autoplot(pam_importance) works when clustering is enabled",
  inherits(p_pam, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(hclust_importance) works when clustering is enabled",
  inherits(p_hc, "ggplot")
)

# 8. clustering columns absent/present as expected
check_tbl <- add_check(
  check_tbl,
  "basic LOVO object has no clustering importance columns",
  !any(c("pam_importance", "hclust_importance") %in% names(lovo_basic$results))
)

check_tbl <- add_check(
  check_tbl,
  "clustering LOVO object has clustering importance columns",
  all(c("pam_importance", "hclust_importance") %in% names(lovo_cluster$results))
)

check_tbl
```

```{r}
# optional visual check of the updated defaults
p_rel
```


### `compare_lovo_mdist()` sanity checks

```{r}

cmp_methods <- list(
  tvd_sup = lovo_method_spec(
    response = Name,
    distance_cat = "tvd",
    response_used = TRUE,
    commensurable = FALSE
  ),
  tvd_unsup = lovo_method_spec(
    response = Name,
    distance_cat = "tvd",
    response_used = FALSE,
    commensurable = FALSE
  )
)


cmp_basic <- compare_lovo_mdist(
  x = df,
  methods = cmp_methods
)


cmp_cluster <- compare_lovo_mdist(
  x = df,
  methods = cmp_methods ,
  cluster_k = 3
)

check_tbl <- tibble::tibble(
  check = character(),
  passed = logical(),
  detail = character()
)

add_check <- function(tbl, check, expr) {
  out <- tryCatch(
    {
      val <- force(expr)
      tibble::tibble(
        check = check,
        passed = isTRUE(val),
        detail = if (isTRUE(val)) "TRUE" else paste(capture.output(print(val)), collapse = " ")
      )
    },
    error = function(e) {
      tibble::tibble(
        check = check,
        passed = FALSE,
        detail = conditionMessage(e)
      )
    }
  )
  dplyr::bind_rows(tbl, out)
}

# 1. object creation
check_tbl <- add_check(
  check_tbl,
  "compare_lovo_mdist() returns MDistLOVOCompare",
  inherits(cmp_basic, "MDistLOVOCompare")
)

# 2. methods column present and correct
check_tbl <- add_check(
  check_tbl,
  "results contain expected method labels",
  setequal(as.character(unique(cmp_basic$results$method)), c("tvd_sup", "tvd_unsup"))
)

# 3. updated metric names available
check_tbl <- add_check(
  check_tbl,
  "results contain relative_distance",
  "relative_distance" %in% names(cmp_basic$results)
)

check_tbl <- add_check(
  check_tbl,
  "results contain mds_congruence",
  "mds_congruence" %in% names(cmp_basic$results)
)

# 4. backward-compatible aliases retained
check_tbl <- add_check(
  check_tbl,
  "results retain mad_normalized",
  "mad_normalized" %in% names(cmp_basic$results)
)

check_tbl <- add_check(
  check_tbl,
  "results retain cc_importance",
  "cc_importance" %in% names(cmp_basic$results)
)

# 5. aliases are equal
check_tbl <- add_check(
  check_tbl,
  "relative_distance equals mad_normalized",
  all.equal(cmp_basic$results$relative_distance, cmp_basic$results$mad_normalized)
)

check_tbl <- add_check(
  check_tbl,
  "mds_congruence equals cc_importance",
  all.equal(cmp_basic$results$mds_congruence, cmp_basic$results$cc_importance)
)

# 6. no clustering columns without cluster_k
check_tbl <- add_check(
  check_tbl,
  "no clustering columns when cluster_k is NULL",
  !any(c("ari_pam", "pam_importance", "ari_hclust", "hclust_importance") %in%
         names(cmp_basic$results))
)

# 7. clustering columns present with cluster_k
check_tbl <- add_check(
  check_tbl,
  "clustering columns added when cluster_k is supplied",
  all(c("ari_pam", "pam_importance", "ari_hclust", "hclust_importance") %in%
        names(cmp_cluster$results))
)

# 8. print/summary work
check_tbl <- add_check(
  check_tbl,
  "print() works on compare object",
  {
    capture.output(print(cmp_basic))
    TRUE
  }
)

check_tbl <- add_check(
  check_tbl,
  "summary() works on compare object",
  {
    capture.output(summary(cmp_basic))
    TRUE
  }
)

# 9. autoplot works for updated metrics
p_rel <- autoplot(cmp_basic, metric = "relative_distance")
p_mad <- autoplot(cmp_basic, metric = "mad_importance")
p_mds <- autoplot(cmp_basic, metric = "mds_congruence")

check_tbl <- add_check(
  check_tbl,
  "autoplot(relative_distance) returns ggplot",
  inherits(p_rel, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(mad_importance) returns ggplot",
  inherits(p_mad, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(mds_congruence) returns ggplot",
  inherits(p_mds, "ggplot")
)

# 10. clustering autoplot works when enabled
p_pam <- autoplot(cmp_cluster, metric = "pam_importance")
p_hclust <- autoplot(cmp_cluster, metric = "hclust_importance")

check_tbl <- add_check(
  check_tbl,
  "autoplot(pam_importance) works when clustering is enabled",
  inherits(p_pam, "ggplot")
)

check_tbl <- add_check(
  check_tbl,
  "autoplot(hclust_importance) works when clustering is enabled",
  inherits(p_hclust, "ggplot")
)

check_tbl
```

```{r}
# optional visual check
p_rel
```
::: {.callout-note appearance="minimal" icon=false}
This section verifies that the renamed metrics are present, backward-compatible aliases are retained, and the updated `print()`, `summary()`, and `autoplot()` methods behave correctly both with and without clustering diagnostics.
:::

### Additional LOVO questions

-   Should there be a spectral clustering importance measure?
-   Should LOVO default to one-cluster methods only, or to all available methods? -> cluster only if number of clusters is provided.

## Downstream clustering support

### Spectral clustering

Improve `spectral_from_dist()` so that it accepts both: - a distance matrix; - a distance object such as `mdist$distance`.

###  Built-in downstream methods

Consider expanding built-in support for: - PAM - hierarchical clustering - spectral clustering - other downstream methods using distances

###  Hierarchical clustering documentation

Document clearly which linkage is used in hierarchical clustering workflows. Current assumption: complete linkage. This should be confirmed and made explicit.

##  Variable weights

-   Reintroduce or expand support for user-defined variable weights.
-   Potential uses include:
    -   emphasizing specific variables;
    -   weighting continuous and categorical blocks differently;
    -   randomized weighting for robustness analyses;
    -   future tuning strategies for weights.

Need to clarify how this should interact with commensurability.

##  MCA issue

Investigate the behavior of `distance_cat = "mca"` under commensurability.

Expected: - methods whose Delta matrix is only a constant multiple of matching should collapse under commensurability, e.g.: - matching - eskin - HL - naive

Unexpected: - `mca` currently appears to collapse to commensurable matching as well.

Why this is problematic: - MCA uses marginal frequencies in its Delta matrix; - therefore, unless category distributions are exactly balanced, commensurable `mca` should not become identical to commensurable matching.

This issue needs clarification and likely a code check.

## Documentation and transparency


### Specific documentation needs

-   Document clearly which distances are:
    -   association-based;
    -   supervised-capable;
    -   independent of the response.
-   Document what happens when `response` is supplied with a distance that does not use it.
-   Document hierarchical clustering defaults.
-   Document LOVO output and interpretation more clearly.

