---
author: Alfonso Iodice D'Enza
title: Distance-based learning with manydist and tidymodels
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: true

set.seed(123)
rm(list = ls())
devtools::load_all("../")
# library(manydist)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
```

# Introduction

The manydist package provides a flexible framework for distance-based
learning with continuous, categorical, and mixed-type predictors. Its
core function `mdist()` computes mixed-variable dissimilarities under a
range of presets and custom specifications, while the `MDist` class
stores both the resulting distance matrix and the associated metadata.

The package also integrates with the tidymodels ecosystem through a
recipe step, `step_mdist()`, a custom distance-based nearest-neighbour
model, `nearest_neighbor_dist()`, a distance-based partitioning around
medoids implementation, `pam_dist()`, and a distance-based spectral
clustering implementation, `spectral_dist()`.

This vignette illustrates the main functionalities of the package:

1. computing mixed-type distances with `mdist()`;
2. assessing variable influence with `lovo_mdist()`;
3. comparing diagnostics across methods with `compare_lovo_mdist()`;
4. creating distance-based representations with `step_mdist()`;
5. fitting and tuning a KNN model with `nearest_neighbor_dist()`;
6. fitting a distance-based PAM clustering model with `pam_dist()`;
7. fitting a distance-based spectral clustering model with `spectral_dist()`.

The focus is on classification of penguin species, but the same ideas
apply to other supervised or unsupervised tasks that rely on pairwise
distances.

# Data

We start from the `penguins` data and remove rows with missing values.

```{r penguins}
penguins_clean <- palmerpenguins::penguins |>
  drop_na()

penguins_clean |>
  dplyr::count(species)
```

# Computing mixed-type distances with `mdist()`

The main entry point of the package is `mdist()`, which computes a
mixed-variable dissimilarity according to a chosen preset or custom
specification.

```{r mdist-core}
d_obj_all <- mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep"
)

summary(d_obj_all)
```

The resulting object stores the distance matrix together with the
distance specification.

```{r mdist-core-inspect}
class(d_obj_all)
dim(base::as.matrix(d_obj_all$distance))
```

# Variable diagnostics with `lovo_mdist()`

The function `lovo_mdist()` provides a leave-one-variable-out diagnostic.
For each predictor, the function recomputes the distance after removing
that variable and compares the resulting structure with the full-distance
solution.

```{r lovo}
#| cache: false
lovo_res <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep"
)

summary(lovo_res)
```

A graphical summary is also available through `autoplot()`.

```{r lovo-plot}
autoplot(lovo_res, "mad_normalized")
```

## LOVO with clustering-based diagnostics

In addition to distance- and MDS-based diagnostics, `lovo_mdist()`
can optionally assess how much the clustering solution changes when each
variable is removed. This is activated by specifying the number of
clusters through `cluster_k`.

In the next example, we compute leave-one-variable-out diagnostics for
the same distance specification as above, and compare the full-data
partition with the leave-one-out partitions using both PAM and
hierarchical clustering. Agreement is measured through the adjusted Rand
index (ARI), while `pam_importance` and `hclust_importance` are defined
as `1 - ARI`, so that larger values indicate stronger influence of the
left-out variable.

```{r lovo-cluster}
#| cache: false
lovo_res_clust <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep",
  cluster_k = 3
)

summary(lovo_res_clust)
```

The clustering-based importance measures can be visualized with
`autoplot()` in the same way as the original LOVO diagnostics.

```{r lovo-cluster-plot-pam}
autoplot(lovo_res_clust, metric = "pam_importance")
```

```{r lovo-cluster-plot-hclust}
autoplot(lovo_res_clust, metric = "hclust_importance")
```

If needed, only one clustering method can be requested.

```{r lovo-cluster-pam-only}
lovo_res_pam <- lovo_mdist(
  x = penguins_clean |> select(-species),
  preset = "u_dep",
  cluster_k = 3,
  cluster_methods = "pam"
)

autoplot(lovo_res_pam, metric = "pam_importance")
```

# Comparing distance specifications with `compare_lovo_mdist()`

The function `compare_lovo_mdist()` extends the LOVO diagnostic to
multiple distance specifications. This makes it possible to compare how
variable contributions change across alternative mixed-type distances.

In the example below, we compare:

- `"gower"`, a standard mixed-type distance;
- `"u_dep"`, an association-aware distance that enforces
  commensurability;
- a custom specification based on Manhattan distance for numeric
  variables and Hennig--Liao distance for categorical variables.

```{r lovo-compare}
#| cache: true
lovo_cmp <- compare_lovo_mdist(
  x = penguins_clean |> select(-species),
  methods = list(
    gower = list(preset = "gower"),
    u_dep = list(preset = "u_dep"),
    custom = list(
      distance_cont = "manhattan",
      distance_cat  = "HL",
      commensurable = FALSE
    )
  )
)

lovo_cmp
```

```{r lovo-compare-summary}
summary(lovo_cmp)
```

```{r lovo-compare-plot}
#| cache: true
autoplot(lovo_cmp, metric = "mad_normalized")
```

# Distance-based KNN in base R

We now move to a simple predictive setting and create a stratified
train/test split.

```{r split}
set.seed(123)
peng_split <- initial_split(penguins_clean, strata = species)
peng_train <- training(peng_split)
peng_test  <- testing(peng_split)

dim(peng_train)
dim(peng_test)
```

We separate predictors and outcome:

```{r base-mdist}
x_train <- peng_train |>
  select(-species)
x_test  <- peng_test |>
  select(-species)
y_train <- peng_train$species
y_test  <- peng_test$species
```

We compute train-to-test distances with `mdist()`.

```{r base-mdist-compute}
test_d_obj <- mdist(
  x = x_train,
  new_data = x_test,
  preset = "u_dep"
)

summary(test_d_obj)
```

```{r base-mdist-inspect}
distance_matrix <- base::as.matrix(test_d_obj$distance)
dim(distance_matrix)
```

These distances can be used directly in a minimal KNN implementation.

```{r base-knn}
n_neighbours <- 5

nn_idx <- apply(distance_matrix, 1, function(d) order(d)[1:n_neighbours])

pred_base <- apply(nn_idx, 2, function(idx) {
  names(which.max(table(y_train[idx])))
})

mean(pred_base == y_test)
```

# Distance-based representations with `step_mdist()`

To integrate with `tidymodels`, manydist provides the recipe step
`step_mdist()`, which computes dissimilarities relative to the training
set.

- `"distance_to_training"` (default): a rectangular test-to-training
  distance matrix, suited to supervised learners such as KNN;
- `"pairwise"`: a square within-sample distance matrix, suited to
  clustering and other unsupervised methods.

```{r rec-mdist}
peng_rec <- recipe(species ~ ., data = peng_train) |>
  step_mdist(all_predictors(), preset = "u_dep")

peng_rec
```

```{r rec-mdist-prep}
prepped_rec <- peng_rec |>
  prep()
```

As a quick sanity check, the distance matrix obtained through the recipe
matches the one computed directly with `mdist()`.

```{r rec-mdist-bake}
train_dist <- bake(prepped_rec, new_data = peng_train)
prepped_test_d_obj <- bake(prepped_rec, new_data = peng_test)

sum(prepped_test_d_obj |> select(-species) - test_d_obj$distance)
```

# Distance-based KNN with `nearest_neighbor_dist()`

The package also provides a custom parsnip model,
`nearest_neighbor_dist()`, with a `"precomputed"` engine.

```{r model-spec}
knn_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = 5
) |>
  set_engine("precomputed")

knn_spec
```

We can combine the recipe and the model in a workflow.

```{r workflow}
peng_wf <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_spec)

peng_wf
```

```{r fit-workflow}
peng_fit <- peng_wf |>
  fit(data = peng_train)
```

Predictions and performance evaluation are then straightforward.

```{r predict-prob-class}
peng_pred <- predict(peng_fit, new_data = peng_test, type = "prob") |>
  bind_cols(
    predict(peng_fit, new_data = peng_test, type = "class"),
    peng_test |> select(species)
  )

peng_pred |>
  head()
```

```{r eval-acc}
peng_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Tuning the number of neighbors

Because `nearest_neighbor_dist()` is registered as a parsnip model, its
hyperparameters can be tuned using `tune_grid()`.

```{r tune-spec}
knn_tune_spec <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = tune()
) |>
  set_engine("precomputed")

knn_tune_spec
```

```{r tune-workflow}
peng_wf_tune <- workflow() |>
  add_recipe(peng_rec) |>
  add_model(knn_tune_spec)
```

```{r tune-grid}
#| cache: true
set.seed(123)
peng_folds <- vfold_cv(peng_train, v = 5, strata = species)

k_grid <- tibble(neighbors = seq(3, 15, by = 2))
k_grid

peng_metrics <- metric_set(accuracy)

tuned_knn <- tune_grid(
  peng_wf_tune,
  resamples = peng_folds,
  grid = k_grid,
  metrics = peng_metrics
)
```

```{r tune-results}
collect_metrics(tuned_knn)
show_best(tuned_knn, metric = "accuracy")

autoplot(tuned_knn)
```

After selecting the best value of `k`, the workflow can be finalized and
re-fitted on the training set.

```{r finalize}
best_k <- select_best(tuned_knn, metric = "accuracy")

final_wf <- peng_wf_tune |>
  finalize_workflow(best_k)

final_fit <- final_wf |>
  fit(data = peng_train)

final_pred <- predict(final_fit, new_data = peng_test, type = "class") |>
  bind_cols(peng_test |> select(species))

final_pred |>
  accuracy(truth = species, estimate = .pred_class)
```

# Distance-based clustering with `pam_dist()` and `spectral_dist()`

For clustering, `step_mdist()` is used with `output = "pairwise"`, so
that the within-training pairwise dissimilarity matrix is computed and
passed to the clustering method.

We first define a recipe that computes the pairwise dissimilarity matrix
on the training set.

```{r rec-clust}
peng_rec_clust <- recipe(~ ., data = peng_train |> select(-species)) |>
  step_mdist(
    all_predictors(),
    preset = "u_dep",
    output = "pairwise"
  )

peng_rec_clust
```

## Distance-based PAM clustering with `pam_dist()`

We define a PAM clustering specification with three clusters.

```{r pam-spec}
pam_spec <- pam_dist(num_clusters = 3)

pam_spec
```

The model is then fitted by combining the specification with the recipe
and the training predictors.

```{r pam-fit}
pam_fit <- generics::fit(
  pam_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

pam_fit
```

The fitted object returns cluster memberships for the training set.

```{r pam-predict-train}
pam_train_pred <- predict(pam_fit)

pam_train_pred |>
  head()
```

Since the true penguin species are known, we can compare the obtained
partition with the observed classes using the adjusted Rand index (ARI).

```{r pam-train-ari}
mclust::adjustedRandIndex(
  as.integer(pam_train_pred$.pred_cluster),
  as.integer(peng_train$species)
)
```

For new observations, `pam_dist()` computes distances from the test set
to the training set, extracts the distances to the fitted medoids, and
assigns each observation to its nearest medoid.

```{r pam-predict-test}
pam_test_pred <- predict(
  pam_fit,
  new_data = peng_test |> select(-species)
)

pam_test_pred |>
  head()
```

The corresponding ARI on the test set is:

```{r pam-test-ari}
mclust::adjustedRandIndex(
  as.integer(pam_test_pred$.pred_cluster),
  as.integer(peng_test$species)
)
```

If desired, the distances from each test observation to the medoids can
also be extracted directly.

```{r pam-test-dists}
predict(
  pam_fit,
  new_data = peng_test |> select(-species),
  type = "dist"
) |>
  head()
```

## Distance-based spectral clustering with `spectral_dist()`

The usage of `spectral_dist()` parallels that of `pam_dist()`: the model
is fitted on the pairwise training dissimilarity matrix computed by
`step_mdist()`, and predictions for new observations are obtained by
extending the spectral embedding to the test set.

We define a spectral clustering specification with three clusters.

```{r spectral-spec}
spectral_spec <- spectral_dist(
  num_clusters = 3
)

spectral_spec
```

The model is fitted in the same way as `pam_dist()`.

```{r spectral-fit}
spectral_fit <- generics::fit(
  spectral_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

spectral_fit
```

The fitted object returns cluster memberships for the training set.

```{r spectral-predict-train}
spectral_train_pred <- predict(spectral_fit)

spectral_train_pred |>
  head()
```

As for PAM, the resulting partition can be compared with the observed
penguin species using the adjusted Rand index.

```{r spectral-train-ari}
mclust::adjustedRandIndex(
  as.integer(spectral_train_pred$.pred_cluster),
  as.integer(peng_train$species)
)
```

Predictions for new observations are obtained through the stored
distance preprocessor and an out-of-sample extension of the spectral
embedding.

```{r spectral-predict-test}
spectral_test_pred <- predict(
  spectral_fit,
  new_data = peng_test |> select(-species)
)

spectral_test_pred |>
  head()
```

The corresponding ARI on the test set is:

```{r spectral-test-ari}
mclust::adjustedRandIndex(
  as.integer(spectral_test_pred$.pred_cluster),
  as.integer(peng_test$species)
)
```

If desired, the spectral embedding coordinates can also be extracted
directly.

```{r spectral-embed-train}
predict(spectral_fit, type = "embed") |>
  head()
```

```{r spectral-embed-test}
predict(
  spectral_fit,
  new_data = peng_test |> select(-species),
  type = "embed"
) |>
  head()
```


## Sanity check: direct distance matrix versus recipe-based clustering

Before illustrating the two clustering interfaces, we verify that the
recipe-based workflow is consistent with the direct use of `mdist()`.

We first compute the training dissimilarity matrix directly with
`mdist()`.

```{r clust-direct-distance}
d_train_obj <- mdist(
  x = peng_train |> select(-species),
  preset = "u_dep"
)

D_train_direct <- as.matrix(d_train_obj$distance)
dim(D_train_direct)
```

We then fit PAM directly on this matrix.

```{r clust-direct-pam}
pam_direct <- cluster::pam(
  x = stats::as.dist(D_train_direct),
  k = 3,
  diss = TRUE
)

pam_direct_clust <- pam_direct$clustering
head(pam_direct_clust)
```

Now we fit the recipe-based `pam_dist()` model.

```{r clust-recipe-pam}
pam_spec <- pam_dist(num_clusters = 3)

pam_fit <- generics::fit(
  pam_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

pam_recipe_clust <- predict(pam_fit)$.pred_cluster |>
  as.integer()

head(pam_recipe_clust)
```

The two PAM partitions coincide up to label permutation. We check this
through the adjusted Rand index.

```{r clust-check-pam}
mclust::adjustedRandIndex(
  pam_direct_clust,
  pam_recipe_clust
)
```

We can do the same for spectral clustering. First, fit the model
directly from the distance matrix.

```{r clust-direct-spectral}
spectral_direct_spec <- spectral_dist(num_clusters = 3)

spectral_direct_fit <- .fit_spectral_dist(
  spec = spectral_direct_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

spectral_direct_clust <- spectral_direct_fit$clusters
head(spectral_direct_clust)
```

For a stricter direct check based only on the distance matrix, we can
also reconstruct the spectral solution from `D_train_direct`.

```{r clust-direct-spectral-matrix}
aff <- .affinity_from_dist(D_train_direct, sigma = NULL)
A <- aff$A

d <- rowSums(A)
Dhalf_inv <- diag(ifelse(d > 0, 1 / sqrt(d), 0), nrow(A), ncol(A))
S <- Dhalf_inv %*% A %*% Dhalf_inv

ev <- eigen(S, symmetric = TRUE)
U <- ev$vectors[, seq_len(3), drop = FALSE]
U_norm <- .row_normalize(U)

set.seed(123)
spectral_direct_matrix <- stats::kmeans(U_norm, centers = 3, nstart = 50)$cluster

head(spectral_direct_matrix)
```

Now fit the recipe-based `spectral_dist()` model.

```{r clust-recipe-spectral}
spectral_spec <- spectral_dist(num_clusters = 3)

spectral_fit <- generics::fit(
  spectral_spec,
  recipe = peng_rec_clust,
  data = peng_train |> select(-species)
)

spectral_recipe_clust <- predict(spectral_fit)$.pred_cluster |>
  as.integer()

head(spectral_recipe_clust)
```

Again, we compare the resulting partitions via the adjusted Rand index.

```{r clust-check-spectral}
mclust::adjustedRandIndex(
  spectral_direct_matrix,
  spectral_recipe_clust
)
```

An ARI equal to 1 confirms that the clustering obtained through the
recipe-based pipeline matches the one obtained by first computing the
distance matrix with `mdist()` and then applying the clustering method
directly.

## Consistency checks on continuous data

In the special case of purely continuous predictors, the downstream
interfaces provided by **manydist** should be consistent with the
corresponding standard methods applied directly to the original data
matrix under Euclidean geometry. For simplicity, we illustrate these
checks on continuous data only, since Euclidean distance provides a
natural and unambiguous reference in that setting. We then verify this
for distance computation, nearest neighbours, PAM clustering, and
spectral clustering.
We start by restricting the penguin data to continuous predictors only.

```{r sanity-continuous-setup}
#| warning: false

library(class)
library(cluster)
library(kernlab)
library(mclust)

peng_train_cont <- peng_train |>
  select(species, where(is.numeric))

peng_test_cont <- peng_test |>
  select(species, where(is.numeric))

x_train_cont <- peng_train_cont |> select(-species)
x_test_cont  <- peng_test_cont  |> select(-species)
y_train_cont <- peng_train_cont$species
y_test_cont  <- peng_test_cont$species

X_train <- as.matrix(x_train_cont)
X_test  <- as.matrix(x_test_cont)

k_knn   <- 5
k_clust <- 3
```

### Euclidean distances: direct computation versus `mdist()`

We first verify that `mdist()` reproduces ordinary Euclidean distances on
continuous data when used with Euclidean distance, no scaling, and no
commensurability correction.

```{r sanity-distance-direct}
D_train_eucl <- as.matrix(stats::dist(X_train))

.euclid_block <- function(x_new, x_ref) {
  x_new <- as.matrix(x_new)
  x_ref <- as.matrix(x_ref)

  n_new <- nrow(x_new)
  all_x <- rbind(x_new, x_ref)
  D_all <- as.matrix(stats::dist(all_x))

  D_all[seq_len(n_new), n_new + seq_len(nrow(x_ref)), drop = FALSE]
}

D_test_train_eucl <- .euclid_block(X_test, X_train)

d_train_mdist <- mdist(
  x = x_train_cont,
  distance_cont = "euclidean",
  scaling_cont = "none",
  commensurable = FALSE
)

D_train_mdist <- as.matrix(d_train_mdist$distance)

d_test_mdist <- mdist(
  x = x_train_cont,
  new_data = x_test_cont,
  distance_cont = "euclidean",
  scaling_cont = "none",
  commensurable = FALSE
)

D_test_train_mdist <- as.matrix(d_test_mdist$distance)
```

The maximum absolute difference is negligible on the training set and
zero on the test-to-training distance block.

```{r sanity-distance-check}
distance_check <- tibble::tibble(
  dataset = c("train", "test"),
  check = c(
    "base Euclidean vs mdist() max abs diff",
    "test-to-train Euclidean vs mdist() max abs diff"
  ),
  value = c(
    max(abs(D_train_eucl - D_train_mdist)),
    max(abs(D_test_train_eucl - D_test_train_mdist))
  )
)

distance_check
```

### KNN: direct Euclidean KNN versus `nearest_neighbor_dist()`

We now compare standard KNN on the original continuous predictors with
the `manydist` interface based on Euclidean dissimilarities.

```{r sanity-knn-direct}
knn_direct_pred <- class::knn(
  train = X_train,
  test  = X_test,
  cl    = y_train_cont,
  k     = k_knn
)
```

The corresponding `manydist` pipeline uses `step_mdist()` to compute the
Euclidean test-to-training distance representation.

```{r sanity-knn-manydist}
rec_knn_cont <- recipe(species ~ ., data = peng_train_cont) |>
  step_mdist(
    all_predictors(),
    distance_cont = "euclidean",
    scaling_cont = "none",
    commensurable = FALSE
  )

knn_spec_cont <- nearest_neighbor_dist(
  mode = "classification",
  neighbors = k_knn
) |>
  set_engine("precomputed")

wf_knn_cont <- workflow() |>
  add_recipe(rec_knn_cont) |>
  add_model(knn_spec_cont)

knn_fit_cont <- fit(wf_knn_cont, data = peng_train_cont)

knn_manydist_pred <- predict(
  knn_fit_cont,
  new_data = peng_test_cont,
  type = "class"
)$.pred_class
```

We compare the direct and manydist-based predictions through exact
agreement and report the manydist classification accuracy on the test
set.

```{r sanity-knn-check}
knn_check <- tibble::tibble(
  dataset = c("test", "test"),
  check = c(
    "direct vs manydist exact agreement",
    "manydist accuracy"
  ),
  value = c(
    mean(knn_direct_pred == knn_manydist_pred),
    yardstick::accuracy_vec(y_test_cont, knn_manydist_pred)
  )
)

knn_check
```

### PAM: direct Euclidean PAM versus `pam_dist()`

For PAM clustering, the direct reference is `cluster::pam()` applied to
the original continuous feature matrix with Euclidean metric.

```{r sanity-pam-direct}
pam_direct_fit <- cluster::pam(
  x = X_train,
  k = k_clust,
  metric = "euclidean"
)

pam_direct_train <- pam_direct_fit$clustering
```

The corresponding `manydist` interface is based on a pairwise Euclidean
dissimilarity matrix produced through `step_mdist(..., output = "pairwise")`.

```{r sanity-pam-manydist}
rec_clust_cont <- recipe(~ ., data = x_train_cont) |>
  step_mdist(
    all_predictors(),
    distance_cont = "euclidean",
    scaling_cont = "none",
    commensurable = FALSE,
    output = "pairwise"
  )

pam_spec_cont <- pam_dist(num_clusters = k_clust)

pam_manydist_fit <- generics::fit(
  pam_spec_cont,
  recipe = rec_clust_cont,
  data = x_train_cont
)

pam_manydist_train <- predict(pam_manydist_fit)$.pred_cluster |>
  as.integer()
```

We compare the two training partitions via the adjusted Rand index.

```{r sanity-pam-check}
pam_check <- tibble::tibble(
  dataset = "train",
  check = "direct vs manydist partition ARI",
  value = mclust::adjustedRandIndex(pam_direct_train, pam_manydist_train)
)

pam_check
```

### Spectral clustering: direct Gaussian-kernel clustering versus `spectral_dist()`

For spectral clustering, we compare `spectral_dist()` with a direct
spectral clustering benchmark on the original continuous data using
`kernlab::specc()` and a Gaussian kernel. The Gaussian bandwidth is
matched to the median Euclidean distance used by the manydist
implementation.

```{r sanity-spectral-direct}
sigma_sc <- stats::median(D_train_eucl[upper.tri(D_train_eucl)])
gamma_kernlab <- 1 / (2 * sigma_sc^2)

set.seed(123)
spectral_direct_train <- kernlab::specc(
  x = X_train,
  centers = k_clust,
  kernel = "rbfdot",
  kpar = list(sigma = gamma_kernlab)
)

spectral_direct_train <- as.integer(spectral_direct_train)
```

The corresponding `manydist` interface uses the same number of clusters
and bandwidth, and is fitted on the pairwise Euclidean dissimilarity
matrix produced by `step_mdist()`.

```{r sanity-spectral-manydist}
spectral_spec_cont <- spectral_dist(
  num_clusters = k_clust,
  sigma = sigma_sc
)

set.seed(123)
spectral_manydist_fit <- generics::fit(
  spectral_spec_cont,
  recipe = rec_clust_cont,
  data = x_train_cont
)

spectral_manydist_train <- predict(spectral_manydist_fit)$.pred_cluster |>
  as.integer()
```

We again compare the two training partitions via the adjusted Rand
index.

```{r sanity-spectral-check}
spectral_check <- tibble::tibble(
  dataset = "train",
  check = "direct vs manydist partition ARI",
  value = mclust::adjustedRandIndex(spectral_direct_train, spectral_manydist_train)
)

spectral_check
```

### Summary of consistency checks

We finally collect the results into a single summary table.

```{r sanity-summary}
sanity_summary <- tibble::tibble(
  method = c(
    "Distance matrix",
    "Distance matrix",
    "KNN",
    "KNN",
    "PAM",
    "Spectral"
  ),
  dataset = c(
    "train",
    "test",
    "test",
    "test",
    "train",
    "train"
  ),
  check = c(
    "base Euclidean vs mdist() max abs diff",
    "test-to-train Euclidean vs mdist() max abs diff",
    "direct vs manydist exact agreement",
    "manydist accuracy",
    "direct vs manydist partition ARI",
    "direct vs manydist partition ARI"
  ),
  value = c(
    distance_check$value[1],
    distance_check$value[2],
    knn_check$value[1],
    knn_check$value[2],
    pam_check$value,
    spectral_check$value
  )
)

sanity_summary
```

These results confirm that, on purely continuous data, `mdist()`
reproduces Euclidean geometry essentially exactly, and that the
downstream `manydist` interfaces are consistent with the corresponding
standard methods applied directly to the original data matrix.


# Summary

In this vignette we showed how **manydist** can be used for
distance-based learning on mixed data through:

- mixed-type distance computation with `mdist()`;
- variable diagnostics with `lovo_mdist()`;
- method comparison with `compare_lovo_mdist()`;
- recipe-based representations with `step_mdist()`;
- distance-based KNN modeling and tuning with `nearest_neighbor_dist()`;
- distance-based PAM clustering with `pam_dist()`;
- distance-based spectral clustering with `spectral_dist()`.
