---
title: "On the use of presets"
author: "Alfonso Iodice D'Enza"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
library(manydist)
library(dplyr)
library(tidyr)
library(kableExtra)
```

We use the `palmerpenguins` data for illustration.

```{r}
penguins_small <- palmerpenguins::penguins |>
  dplyr::select(-year) |>
  tidyr::drop_na()
```

## Two types of presets

Some presets are mainly convenient shortcuts for combinations of `mdist()`
arguments. For example, the `"u_dep"` preset corresponds to an unbiased
association-aware distance construction using categorical association
information, numerical preprocessing, and commensurability.

The available presets can be listed from `dist_methods_tbl()`.

```{r}
dist_methods_tbl() |>
  dplyr::filter(argument == "preset", method != "custom") |>
  dplyr::select(
    preset = method,
    response_aware,
    engine
  ) |>
  kableExtra::kbl()
```

The `engine` column helps distinguish presets implemented directly in
`manydist` from presets based on external constructions or ports. For example,
`"u_indep"`, `"u_dep"`, `"u_mix"`, and `"hl"` are implemented directly in
`manydist`, whereas `"dkss"` uses the `kdml` implementation and `"gudmm"` and
`"mod_gower"` are external ports.

## Presets as argument shortcuts

Consider the `"u_dep"` preset, which computes an unbiased association-aware
distance measure.

```{r}
x_penguins <- penguins_small |>
  dplyr::select(-species)

d_u_dep_preset <- mdist(
  x_penguins,
  preset = "u_dep"
)

d_u_dep_explicit <- mdist(
  x_penguins,
  method_cat = "tvd",
  method_num = "pc_scores",
  commensurable = TRUE
)

all.equal(
  as.matrix(d_u_dep_preset$distance),
  as.matrix(d_u_dep_explicit$distance)
)
```

The result shows that, in this case, the preset is equivalent to the
corresponding explicit specification of `method_cat`, `method_num`, and
`commensurable`.

## Presets based on specific distance constructions

Other presets are not merely combinations of the basic arguments above. Instead,
they correspond to specific distance constructions, implementations, or external
ports. For example, the `"mod_gower"` preset computes a modified version of
Gower's distance.

```{r}
d_mod_gower <- mdist(
  x_penguins,
  preset = "mod_gower"
)

d_mod_gower
```

Similarly, presets such as `"gower"`, `"gudmm"`, and `"dkss"` refer to specific
distance constructions. In these cases, the preset identifies the distance
method itself, rather than simply selecting a combination of `method_cat`,
`method_num`, and `commensurable`.

The `"euclidean"` preset is slightly different: it uses a dedicated
preprocessing strategy, converting categorical variables into one-hot dummy
variables before computing Euclidean distances on the column-wise standardized
dataset.

We can compare it with the standard Euclidean distance obtained from `dist()` on
the numerical variables only, provided that the numerical variables are
standardized first.

```{r}
x_penguins_num <- penguins_small |>
  dplyr::select(where(is.numeric))

d_euc_mdist <- mdist(
  x_penguins_num,
  preset = "euclidean"
)

d_euc_dist <- dist(
  x_penguins_num |>
    dplyr::mutate(
      dplyr::across(
        dplyr::everything(),
        ~ as.numeric(scale(.x))
      )
    ),
  method = "euclidean"
)

all.equal(
  as.matrix(d_euc_mdist$distance),
  as.matrix(d_euc_dist)
)
```

The default remains `method_num = "std"`, so the numerical-only Euclidean
preset agrees with `dist()` applied after standardization. When all predictors
are numeric, however, `method_num` can be overridden. For example,
`method_num = "none"` skips numerical preprocessing and reproduces ordinary
Euclidean distance on the original variables.

```{r}
d_euc_unscaled <- mdist(
  x_penguins_num,
  preset = "euclidean",
  method_num = "none"
)

d_euc_dist_unscaled <- dist(
  x_penguins_num,
  method = "euclidean"
)

all.equal(
  as.matrix(d_euc_unscaled$distance),
  as.matrix(d_euc_dist_unscaled)
)

```

For numerical-only inputs, the supported choices are `"none"`, `"std"`,
`"range"`, `"robust"`, and `"pc_scores"`. This override is restricted to
numerical-only data. When categorical variables are present, the `"euclidean"`
preset retains its standard preprocessing strategy: categorical variables are
one-hot encoded and `method_num = "std"` is used.
