---
title: "distance-based clustering workflows"
author: "Alfonso Iodice D'Enza"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

`manydist` provides tools for constructing dissimilarities for numerical, categorical, and mixed-type data.

The main function is `mdist()`. It computes a dissimilarity object that can be used in distance-based learning workflows, including clustering, nearest-neighbour prediction, and diagnostic analyses.

The package supports both preset specifications and custom specifications. In the current interface:

- `method_cat` controls the categorical dissimilarity;
- `method_num` controls the preprocessing of numerical variables;
- `commensurable` controls whether variable-wise contributions are placed on a comparable scale;
- `preset` selects a predefined distance specification.

The numerical metric is no longer a user-facing argument. For most presets, numerical variables are compared through Manhattan-type contributions. The exceptions are presets such as `"hl"` and `"euclidean"`, where the Euclidean structure is part of the definition.

The `"euclidean"` preset replaces the former `"euclidean_onehot"` name. It computes Euclidean distances after one-hot encoding categorical variables.

## Setup

```{r}
library(manydist)
library(dplyr)
library(tidyr)
library(recipes)
```

We use the `palmerpenguins` data for illustration.

```{r}
if (requireNamespace("palmerpenguins", quietly = TRUE)) {
  data("penguins", package = "palmerpenguins")

  penguins_small <- palmerpenguins::penguins |>
    dplyr::select(
      species,
      bill_length_mm,
      bill_depth_mm,
      flipper_length_mm,
      body_mass_g,
      island,
      sex
    ) |>
    tidyr::drop_na()
}
```

The data contain numerical variables, such as bill length and body mass, and categorical variables, such as island and sex.

```{r}
dplyr::glimpse(penguins_small)
```

## Computing a mixed-type dissimilarity

The simplest way to compute a mixed-type dissimilarity is to use a preset.

For example, the `"gower"` preset computes a Gower-type dissimilarity for mixed numerical and categorical data.

```{r}
d_gower <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "gower"
)

d_gower
```

The resulting object stores the dissimilarity matrix and metadata about the distance specification.

```{r}
class(d_gower)
```

The distance matrix can be extracted with `to_dist()`.

```{r}
as.matrix(d_gower$to_dist())[1:5, 1:5]
```

## Response-aware dissimilarities

Some categorical dissimilarities can use a response variable. When a response is supplied to `mdist()`, it is used for response-aware distance construction and is removed from the predictors.

Therefore, in the following example, `species` is used as the response and is not also treated as a predictor.

```{r}
d_response <- mdist(
  penguins_small,
  response = species,
  preset = "u_dep"
)

d_response
```

The same can be written by passing the response as a character string.

```{r}
d_response_chr <- mdist(
  penguins_small,
  response = "species",
  preset = "u_dep"
)

d_response_chr
```

## Custom distance specifications

Presets are convenient, but users can also define custom specifications.

In a custom specification, `method_cat` controls the categorical dissimilarity and `method_num` controls the preprocessing applied to numerical variables.

```{r}
d_custom <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "custom",
  method_cat = "matching",
  method_num = "std",
  commensurable = TRUE
)

d_custom
```

Here, categorical variables are compared using matching dissimilarities, numerical variables are standardized, and variable-wise contributions are made commensurable.

A response-aware custom specification can also be used.

```{r}
d_custom_response <- mdist(
  penguins_small,
  response = species,
  preset = "custom",
  method_cat = "tvd",
  method_num = "std",
  commensurable = TRUE
)

d_custom_response
```

If only one categorical predictor is available, methods that require more than one categorical variable may switch to a simpler alternative internally.

## Euclidean preset

The `"euclidean"` preset computes Euclidean distances after converting categorical variables to one-hot dummy variables.

```{r}
d_euclidean <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "euclidean"
)

d_euclidean
```

This preset replaces the former `"euclidean_onehot"` name.

## Comparing presets

Different presets encode different assumptions about how numerical and categorical variables should contribute to the final dissimilarity.

```{r}
d_u_indep <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "u_indep"
)

d_u_mix <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "u_mix"
)

d_hl <- mdist(
  penguins_small |> dplyr::select(-species),
  preset = "hl"
)

d_u_indep
d_u_mix
d_hl
```

The available presets include:

```{r}
c(
  "custom",
  "gower",
  "euclidean",
  "unbiased_dependent",
  "u_dep",
  "u_indep",
  "u_mix",
  "hl",
  "gudmm",
  "dkss",
  "mod_gower"
)
```

Some presets are intended mainly for train-train dissimilarities and may not support distances from new observations to training observations.

## Distances from new observations to training observations

`mdist()` can also compute distances from new observations to the training observations through the `new_data` argument.

```{r}
set.seed(123)

split_id <- sample(seq_len(nrow(penguins_small)), size = floor(0.75 * nrow(penguins_small)))

penguins_train <- penguins_small[split_id, ]
penguins_test  <- penguins_small[-split_id, ]

d_new <- mdist(
  x = penguins_train |> dplyr::select(-species),
  new_data = penguins_test |> dplyr::select(-species),
  preset = "gower"
)

d_new
```

The resulting dissimilarity is rectangular: rows correspond to test observations and columns correspond to training observations.

```{r}
dim(as.matrix(d_new$distance))
```

## Using `manydist` in recipe workflows

The function `step_mdist()` provides a recipe interface to `mdist()`. It replaces selected predictors with a distance-based representation.

In supervised workflows, the response is defined on the left-hand side of the recipe formula. Therefore, when we select `recipes::all_predictors()`, the response variable is not used as a predictor in the distance computation.

```{r}
rec <- recipes::recipe(species ~ ., data = penguins_small) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "distance_to_training"
  )

rec_prep <- recipes::prep(rec, training = penguins_small)

baked <- recipes::bake(rec_prep, new_data = penguins_small)

baked |>
  dplyr::slice_head(n = 5)
```

The resulting data contain the outcome variable, `species`, together with one distance column for each training observation. These columns are named `dist_1`, `dist_2`, and so on.

## Custom specifications in recipes

The same interface can be used with a custom distance specification.

```{r}
rec_custom <- recipes::recipe(species ~ ., data = penguins_small) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "custom",
    method_cat = "matching",
    method_num = "std",
    commensurable = TRUE,
    output = "distance_to_training"
  )

rec_custom_prep <- recipes::prep(rec_custom, training = penguins_small)

baked_custom <- recipes::bake(rec_custom_prep, new_data = penguins_small)

baked_custom |>
  dplyr::slice_head(n = 5)
```

In this example, `method_cat = "matching"` specifies the categorical dissimilarity, while `method_num = "std"` specifies standardization of numerical variables.

## Pairwise dissimilarities in recipes

For distance-based clustering workflows, `step_mdist()` can return the within-training pairwise dissimilarity matrix by setting `output = "pairwise"`.

In this case, the recipe is intended for training-only workflows.

```{r}
penguins_predictors <- penguins_small |>
  dplyr::select(-species)

rec_pairwise <- recipes::recipe(~ ., data = penguins_predictors) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "pairwise"
  )

rec_pairwise_prep <- recipes::prep(rec_pairwise, training = penguins_predictors)

pairwise_dist <- recipes::bake(
  rec_pairwise_prep,
  new_data = penguins_predictors
)

pairwise_dist |>
  dplyr::slice_head(n = 5)
```

## Euclidean distances in recipes

The `"euclidean"` preset can also be used inside a recipe. Categorical predictors are one-hot encoded internally before the Euclidean distance is computed.

```{r}
rec_euclidean <- recipes::recipe(species ~ ., data = penguins_small) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "euclidean",
    output = "distance_to_training"
  )

rec_euclidean_prep <- recipes::prep(rec_euclidean, training = penguins_small)

baked_euclidean <- recipes::bake(rec_euclidean_prep, new_data = penguins_small)

baked_euclidean |>
  dplyr::slice_head(n = 5)
```

## Distance-based nearest-neighbour prediction

The distance representation produced by `step_mdist()` can be used in downstream supervised learning workflows.

Here we illustrate a nearest-neighbour classifier using the distance-to-training representation. The recipe first converts the original mixed-type predictors into distances from each observation to the training observations. The model is then tuned over the number of neighbours.

```{r}
library(tidymodels)

set.seed(123)

penguin_split <- rsample::initial_split(
  penguins_small,
  prop = 0.75,
  strata = species
)

penguin_train <- rsample::training(penguin_split)
penguin_test  <- rsample::testing(penguin_split)

set.seed(123)

penguin_folds <- rsample::vfold_cv(
  penguin_train,
  v = 5,
  strata = species
)
```

```{r}
rec_knn <- recipes::recipe(species ~ ., data = penguin_train) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "distance_to_training"
  )
```

```{r}
knn_spec <- nearest_neighbor_dist(
  neighbors = tune::tune()
) |>
  parsnip::set_mode("classification")
```

```{r}
knn_wf <- workflows::workflow() |>
  workflows::add_recipe(rec_knn) |>
  workflows::add_model(knn_spec)

knn_grid <- tibble::tibble(
  neighbors = c(1, 3, 5, 7, 9)
)

set.seed(123)

knn_res <- tune::tune_grid(
  knn_wf,
  resamples = penguin_folds,
  grid = knn_grid,
  metrics = yardstick::metric_set(accuracy)
)

knn_res |>
  tune::collect_metrics()
```

```{r}
best_knn <- knn_res |>
  tune::select_best(metric = "accuracy")

best_knn
```

```{r}
final_knn_wf <- knn_wf |>
  tune::finalize_workflow(best_knn)

final_knn_fit <- final_knn_wf |>
  parsnip::fit(data = penguin_train)

knn_pred <- predict(final_knn_fit, new_data = penguin_test) |>
  dplyr::bind_cols(penguin_test |> dplyr::select(species))

yardstick::accuracy(knn_pred, truth = species, estimate = .pred_class)
```

This example uses `preset = "gower"`, but the same workflow can be used with any `mdist()` preset or with a custom specification based on `method_cat`, `method_num`, and `commensurable`.

## Distance-based PAM clustering

The same dissimilarities can be used for clustering. For clustering workflows, `step_mdist()` can return the within-training pairwise dissimilarity matrix by setting `output = "pairwise"`.

```{r}
penguins_predictors <- penguins_small |>
  dplyr::select(-species)

rec_pam <- recipes::recipe(~ ., data = penguins_predictors) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "pairwise"
  )
```

```{r}
pam_spec <- pam_dist(
  num_clusters = 3
)

pam_fit <- generics::fit(
  pam_spec,
  recipe = rec_pam,
  data = penguins_predictors
)

pam_pred <- predict(pam_fit)

pam_pred |>
  dplyr::slice_head(n = 10)
```

Since the true species labels are available in this example, we can inspect the relationship between the PAM clusters and the species labels.

```{r}
pam_pred |>
  dplyr::bind_cols(penguins_small |> dplyr::select(species)) |>
  dplyr::count(.pred_cluster, species)
```

## Distance-based spectral clustering

Spectral clustering can be used in the same way, by combining a pairwise distance recipe with the corresponding model specification.

```{r}
rec_spectral <- recipes::recipe(~ ., data = penguins_predictors) |>
  step_mdist(
    recipes::all_predictors(),
    preset = "gower",
    output = "pairwise"
  )
```

```{r}
spectral_spec <- spectral_dist(
  num_clusters = 3
)

spectral_fit <- generics::fit(
  spectral_spec,
  recipe = rec_spectral,
  data = penguins_predictors
)

spectral_pred <- predict(spectral_fit)

spectral_pred |>
  dplyr::slice_head(n = 10)
```

Again, because this is an illustrative labelled data set, we can compare the resulting clusters with the observed species labels.

```{r}
spectral_pred |>
  dplyr::bind_cols(penguins_small |> dplyr::select(species)) |>
  dplyr::count(.pred_cluster, species)
```


## A small benchmark example

The function `benchmark_mdist()` applies `mdist()` repeatedly over a grid of distance specifications. This is useful for comparing alternative presets and custom specifications in a systematic way.

For a small example, we restrict the grid to a few specifications.

```{r}
small_specs <- dplyr::bind_rows(
  all_dist_method_specs(mode = "presets_only") |>
    dplyr::filter(preset %in% c("gower", "u_dep", "u_indep", "euclidean")),

  all_dist_method_specs(
    mode = "full",
    method_cat = c("matching", "tvd"),
    method_num = c("std", "robust"),
    commensurable = c(TRUE, FALSE)
  ) |>
    dplyr::filter(spec_type == "component")
)

small_specs
```

```{r}
bench_small <- benchmark_mdist(
  penguins_small,
  response = species,
  specs = small_specs
)

bench_small |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable,
    ok,
    error
  )
```

The `ok` column reports whether the distance computation completed successfully. Failed specifications are kept in the output, with the corresponding error message stored in `error`.

```{r}
bench_small |>
  dplyr::count(ok)
```

When all specifications are successful, the resulting `MDist` objects are stored in the `result` list-column and can be inspected or reused.

```{r}
bench_small |>
  dplyr::filter(ok) |>
  dplyr::select(spec_type, preset, method_cat, method_num, commensurable, result) |>
  dplyr::slice_head(n = 3)
```

## Leave-one-variable-out diagnostics

`lovo_mdist()` computes leave-one-variable-out diagnostics for distance-based variable importance.

It first computes the full dissimilarity matrix. Then it removes one predictor at a time, recomputes the dissimilarity matrix, and compares the reduced matrix with the full one.

```{r}
lovo_gower <- lovo_mdist(
  penguins_small,
  response = species,
  response_used = FALSE,
  preset = "gower"
)

lovo_gower
```

The result can be summarized.

```{r}
summary(lovo_gower)
```

The relative contribution of each variable can be visualized with `autoplot()`.

```{r}
lovo_gower$autoplot(
  metric = "relative_distance",
  reorder = TRUE
)
```

When `response_used = FALSE`, the response column is removed before computing distances. When `response_used = TRUE`, the response can be used by response-aware distance specifications, but it is still not treated as a predictor in the leave-one-variable-out loop.

```{r}
lovo_response <- lovo_mdist(
  penguins_small,
  response = species,
  response_used = TRUE,
  preset = "u_dep"
)

lovo_response
```

## Inspecting method specifications

`all_dist_method_specs()` returns a table of distance specifications that can be used for benchmarking or sensitivity analysis.

```{r}
specs <- all_dist_method_specs()

specs |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable
  ) |>
  dplyr::slice_head(n = 10)
```

The same table can be passed to `benchmark_mdist()`.

```{r}
bench <- benchmark_mdist(
  penguins_small,
  response = species,
  specs = specs
)

bench |>
  dplyr::select(
    spec_type,
    preset,
    method_cat,
    method_num,
    commensurable,
    ok,
    error
  ) |>
  dplyr::slice_head(n = 10)
```

