---
title: "Assessing variable contributions to distances"
author: "Alfonso Iodice D'Enza"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    number-sections: true
    embed-resources: true
    css: article-style.css
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
library(manydist)
library(dplyr)
library(tidyr)
library(kableExtra)
```

We use the `palmerpenguins` data for illustration.

```{r}
penguins_small <- palmerpenguins::penguins |>
  dplyr::select(-year) |>
  tidyr::drop_na()
```

## Leave-one-variable-out diagnostics

`lovo_mdist()` computes leave-one-variable-out diagnostics for assessing variable contributions to a distance construction.

The function first computes the full dissimilarity matrix using all predictors. Then, one predictor at a time, it removes a variable, recomputes the dissimilarity matrix, and compares the reduced dissimilarity with the full one.

Variables whose removal strongly changes the dissimilarity matrix, the induced configuration, or the resulting clustering structure can be interpreted as having a larger contribution to the distance construction.

```{r}
lovo_gower <- lovo_mdist(
  penguins_small,
  response = species,
  response_used = FALSE,
  preset = "gower"
)

lovo_gower
```

The result can be summarized as follows.

```{r}
summary(lovo_gower)
```

The main results are stored in the `$results` field.

```{r}
lovo_gower$results |>
  kableExtra::kbl(format = "html", digits = 3) |>
  kableExtra::kable_styling(full_width = FALSE)
```

## Interpreting the output

The output of `lovo_mdist()` contains several diagnostics, each capturing a different aspect of the change induced by removing a variable.

The main diagnostics can be interpreted as follows.

```{r}
#| echo: false
metric_description <- tibble::tibble(
  metric = c(
    "mad_importance",
    "relative_distance",
    "mds_congruence",
    "cc_importance",
    "ac_importance"
  ),
  interpretation = c(
    "Mean absolute difference between the full dissimilarity matrix and the leave-one-variable-out dissimilarity matrix.",
    "Normalized version of `mad_importance`; it reports the relative contribution of each variable.",
    "Congruence between the MDS configuration obtained from the full distance and the one obtained after removing the variable.",
    "Congruence coefficient between the full and reduced configurations.",
    "Alienation coefficient between the full and reduced configurations."
  ),
  contribution_direction = c(
    "Larger values indicate a larger contribution.",
    "Larger values indicate a larger contribution.",
    "Smaller values indicate a larger contribution.",
    "Smaller values indicate a larger contribution.",
    "Larger values indicate a larger contribution."
  )
)

metric_description |>
  kableExtra::kbl(format = "html") |>
  kableExtra::kable_styling(full_width = FALSE)
```

The metrics are complementary. A variable may have a large effect on the pairwise dissimilarities without substantially changing the geometric configuration. Conversely, a variable may have a modest effect on the average distances but still affect the configuration induced by the distance matrix.

## Visualising variable contributions

The relative contribution of each variable can be visualized with `autoplot()`.

```{r}
lovo_gower$autoplot(
  metric = "relative_distance",
  reorder = TRUE
)
```

Other metrics can be visualized by changing the `metric` argument. For example, `ac_importance` displays the alienation coefficient.

```{r}
lovo_gower$autoplot(
  metric = "ac_importance",
  reorder = TRUE
)
```

The congruence coefficient can also be displayed. Since this is an agreement measure, smaller values indicate that removing the variable produces a larger change in the induced configuration.

```{r}
lovo_gower$autoplot(
  metric = "cc_importance",
  reorder = TRUE
)
```

## Clustering-based LOVO diagnostics

`lovo_mdist()` can also assess how removing each variable affects cluster partitions derived from the distance matrix.

To compute clustering-based diagnostics, the number of clusters must be supplied through `cluster_k`. The clustering methods are controlled by `cluster_methods`. Available options are `"pam"`, `"hclust"`, and `"spectral"`.

In the following example, we compute clustering-based diagnostics with three clusters.

```{r}
lovo_gower_cluster <- lovo_mdist(
  penguins_small,
  response = species,
  response_used = FALSE,
  preset = "gower",
  cluster_k = 3,
  cluster_methods = c("pam", "hclust", "spectral")
)

lovo_gower_cluster
```

The corresponding results include ARI-based diagnostics for each clustering method.

```{r}
lovo_gower_cluster$results |>
  dplyr::select(
    variable,
    variable_type,
    ari_pam,
    pam_importance,
    ari_hclust,
    hclust_importance,
    ari_spectral,
    spectral_importance
  ) |>
  kableExtra::kbl(format = "html", digits = 3) |>
  kableExtra::kable_styling(full_width = FALSE)
```

The ARI columns compare the partition obtained from the full distance matrix with the partition obtained after removing each variable. Since ARI is an agreement measure, smaller values indicate a stronger change in the clustering structure.

The corresponding importance columns are defined as `1 - ARI`. Therefore, larger values of `pam_importance`, `hclust_importance`, and `spectral_importance` indicate a larger contribution to the clustering structure induced by the distance.

```{r}
#| echo: false
cluster_metric_description <- tibble::tibble(
  metric = c(
    "ari_pam",
    "ari_hclust",
    "ari_spectral",
    "pam_importance",
    "hclust_importance",
    "spectral_importance"
  ),
  interpretation = c(
    "ARI between PAM clusters from the full and reduced distances.",
    "ARI between hierarchical clusters from the full and reduced distances.",
    "ARI between spectral clusters from the full and reduced distances.",
    "PAM-based importance, computed as `1 - ari_pam`.",
    "Hierarchical-clustering importance, computed as `1 - ari_hclust`.",
    "Spectral-clustering importance, computed as `1 - ari_spectral`."
  ),
  contribution_direction = c(
    "Smaller values indicate a larger contribution.",
    "Smaller values indicate a larger contribution.",
    "Smaller values indicate a larger contribution.",
    "Larger values indicate a larger contribution.",
    "Larger values indicate a larger contribution.",
    "Larger values indicate a larger contribution."
  )
)

cluster_metric_description |>
  kableExtra::kbl(format = "html") |>
  kableExtra::kable_styling(full_width = FALSE)
```

For example, the PAM-based importance can be visualized as follows.

```{r}
lovo_gower_cluster$autoplot(
  metric = "pam_importance",
  reorder = TRUE
)
```

The same diagnostic can be inspected for hierarchical clustering.

```{r}
lovo_gower_cluster$autoplot(
  metric = "hclust_importance",
  reorder = TRUE
)
```

Or for spectral clustering.

```{r}
lovo_gower_cluster$autoplot(
  metric = "spectral_importance",
  reorder = TRUE
)
```

The three cluster-based diagnostics need not rank variables in exactly the same way. A variable may be important for the partition induced by one clustering method and less important for another. This makes clustering-based LOVO diagnostics useful for assessing whether variable contributions are stable across clustering algorithms.

## Response-aware LOVO diagnostics

When `response_used = FALSE`, the response column is removed before computing distances. When `response_used = TRUE`, the response can be used by response-aware distance specifications, but it is still not treated as a predictor in the leave-one-variable-out loop.

```{r}
lovo_response <- lovo_mdist(
  penguins_small,
  response = species,
  response_used = TRUE,
  preset = "u_dep"
)

lovo_response
```

The response-aware diagnostics can be inspected in the same way.

```{r}
lovo_response$results |>
  kableExtra::kbl(format = "html", digits = 3) |>
  kableExtra::kable_styling(full_width = FALSE)
```

For example, the relative-distance contribution can be visualized as follows.

```{r}
lovo_response$autoplot(
  metric = "relative_distance",
  reorder = TRUE
)
```

## Comparing LOVO diagnostics across distance specifications

The function `compare_lovo_mdist()` compares leave-one-variable-out diagnostics across several distance specifications.

Instead of running `lovo_mdist()` separately for each specification, the user supplies a named list of methods. Each element of the list contains arguments passed to `lovo_mdist()`, such as `preset`, `method_cat`, `method_num`, and `commensurable`.

For each distance specification, `compare_lovo_mdist()` computes the full distance, recomputes the leave-one-variable-out distances, extracts the diagnostics, and combines the results into a single `MDistLOVOCompare` object.

```{r}
lovo_compare <- compare_lovo_mdist(
  penguins_small,
  response = "species",
  response_used = FALSE,
  methods = list(
    gower = list(preset = "gower"),
    u_indep = list(preset = "u_indep"),
    custom_matching = list(
      preset = "custom",
      method_cat = "matching",
      method_num = "std",
      commensurable = TRUE
    )
  )
)

lovo_compare
```

The combined results are stored in the `$results` field. The table contains one row for each combination of distance specification and left-out variable.

```{r}
lovo_compare$results |>
  dplyr::select(
    method,
    variable,
    variable_type,
    relative_distance,
    mad_importance,
    mds_congruence,
    ac_importance
  ) |>
  kableExtra::kbl(format = "html", digits = 3) |>
  kableExtra::kable_styling(full_width = FALSE)
```

The summary reports the range and average of the main LOVO diagnostics within each distance specification.

```{r}
summary(lovo_compare)
```

The comparison object also has an `autoplot()` method. This makes it possible to compare the contribution profiles induced by different distance specifications.

```{r}
lovo_compare$autoplot(
  metric = "relative_distance",
  reorder = TRUE
)
```

A comparison can also include clustering-based diagnostics by passing `cluster_k` and, optionally, `cluster_methods`.

```{r}
lovo_compare_cluster <- compare_lovo_mdist(
  penguins_small,
  response = "species",
  response_used = FALSE,
  cluster_k = 3,
  cluster_methods = c("pam", "hclust", "spectral"),
  methods = list(
    gower = list(preset = "gower"),
    u_indep = list(preset = "u_indep"),
    custom_matching = list(
      preset = "custom",
      method_cat = "matching",
      method_num = "std",
      commensurable = TRUE
    )
  )
)

lovo_compare_cluster
```

The resulting object includes the ARI-based diagnostics for the selected clustering methods.

```{r}
lovo_compare_cluster$results |>
  dplyr::select(
    method,
    variable,
    variable_type,
    pam_importance,
    hclust_importance,
    spectral_importance
  ) |>
  kableExtra::kbl(format = "html", digits = 3) |>
  kableExtra::kable_styling(full_width = FALSE)
```

For example, the following plot compares PAM-based cluster importance across the selected distance specifications.

```{r}
lovo_compare_cluster$autoplot(
  metric = "pam_importance",
  reorder = TRUE
)
```

The comparison is useful because variable contributions are not absolute properties of the data alone. They also depend on the distance specification. A variable that is highly influential under one distance may have a smaller contribution under another specification, especially when the two distances encode different assumptions about scaling, variable type, commensurability, or association.


<!-- ## Summary -->

<!-- `lovo_mdist()` provides complementary diagnostics for assessing variable contributions to distance construction. -->

<!-- Distance-based diagnostics measure how much the dissimilarity matrix changes when a variable is removed. Configuration-based diagnostics assess how much the induced MDS representation changes. Clustering-based diagnostics compare the partitions obtained from the full and reduced distances, using PAM, hierarchical clustering, or spectral clustering. -->

<!-- Together, these diagnostics help identify variables that play an important role in the distance representation and in the downstream structures derived from it. -->
