---
title: "Revised real-data evaluation for spectral clustering of mixed data"
subtitle: "Reproducible supplementary analysis for the first revision"
author: "Supplementary real-data code"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: false
---

This document defines the revised real-data evaluation.  It directly addresses
the reviewers' requests for a broader set of applications, a consolidated
dataset-by-method comparison, fully specified optimization settings, and a
separate assessment of whether the interaction term changes the empirical
partition.  Class labels are removed before clustering and are used only to set
the known number of clusters and to compute the adjusted Rand index (ARI).
For the South African Heart case study, the labels are additionally
used in a separate supervised diagnostic that compares a main-effects
classifier with a classifier containing numerical--categorical interactions.
Neither classifier contributes to the dissimilarities, the interaction gate,
or the clustering solution.

# Analysis parameters

```{r}
#| label: analysis-parameters

analysis_parameters <- list(
  analysis_version = "R1-real-v1",

  execution = list(
    # A normal render is read-only. Change this to TRUE to produce missing
    # dataset artifacts.
    run_experiments = FALSE,
    datasets = c(
      "Star", "Automobile", "Australian_credit", "Palmer_penguins",
      "Statlog_heart", "SAheart"
    ),
    # UCI source files are cached locally only when this switch is TRUE.
    allow_download = FALSE,
    output_directory = "results/R1_real_data",
    raw_data_directory = "data/real_data",
    seed_base = 27100000L,
    show_progress = TRUE
  ),

  preprocessing = list(
    numeric_missing = "median",
    categorical_missing = "explicit_Missing_level",
    standardize_numeric = TRUE,
    missing_level = "Missing"
  ),

  distance = list(
    preset = "custom",
    method_num = "pc_scores",
    method_cat = "tvd",
    commensurable = FALSE,
    interaction = FALSE,
    gamma = 1,
    prop_nn = 0.10,
    score = "ba",
    decision = "prior_corrected",
    gate_permutations = 499L,
    gate_alpha = 0.01,
    gate_seed_offset = 31L,
    ahmad_dey_bins = 5L
  ),

  neighbour_sensitivity = list(
    enabled = TRUE,
    proportions = c(0.02, 0.05, 0.10, 0.20)
  ),

  spectral = list(
    embedding_dimension = "known number of classes K",
    affinity = "global Gaussian",
    bandwidth = "median positive off-diagonal dissimilarity",
    kmeans_nstart = 100L,
    kmeans_iter_max = 100L,
    kmeans_algorithm = "Hartigan-Wong"
  ),

  algorithm_benchmarks = list(
    kprototypes = list(
      package = "clustMixType",
      nstart = 100L,
      iter_max = 100L,
      type = "huang",
      lambda = NULL
    ),
    kamila = list(
      package = "kamila",
      num_init = 100L,
      max_iter = 100L,
      con_init_method = "runif",
      cat_bw = 0.025
    )
  ),

  methods = c(
    D_mint = TRUE,
    D_m = TRUE,
    Gower = TRUE,
    Modified_Gower = TRUE,
    Euclidean_onehot = TRUE,
    Ahmad_Dey = TRUE,
    Kprototypes = TRUE,
    KAMILA = TRUE
  ),

  sources = list(
    Automobile = "https://archive.ics.uci.edu/ml/machine-learning-databases/autos/imports-85.data",
    Australian_credit = "https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/australian/australian.dat",
    Statlog_heart = "https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/heart/heart.dat"
  )
)
```

The two stochastic competitors are each fitted **once per dataset**, with 100
internal initializations.  The reported ARI is the ARI of the solution selected
by the algorithm, not an average of ARIs from separate clustering runs.  This
implements the reviewers' requested distinction between repeated independent
runs and internal random starts.

# Reproducible setup

```{r}
#| label: setup

`%||%` <- function(x, y) if (is.null(x)) y else x

input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) "")
if (!length(input_file) || is.na(input_file[[1L]]) || !nzchar(input_file[[1L]])) {
  local_candidate <- file.path(getwd(), "R1_real_data.qmd")
  input_file <- if (file.exists(local_candidate)) local_candidate else file.path(
    getwd(), "spectral_cluster_experiments", "spectral_stco_revision_r1",
    "R1_real_data.qmd"
  )
}
input_file <- normalizePath(input_file[[1L]], mustWork = TRUE)
revision_directory <- dirname(input_file)
repository_root <- dirname(dirname(revision_directory))
full_directory <- file.path(revision_directory, "full")
moon_directory <- file.path(revision_directory, "moon_scaling")

required_packages <- c("devtools", "digest", "knitr", "mclust", "nnet")
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages)) {
  stop("Missing required packages: ", paste(missing_packages, collapse = ", "))
}

devtools::load_all(file.path(repository_root, "manydist"), quiet = TRUE)

source_files <- c(
  bridge = file.path(revision_directory, "R", "manydist_bridge.R"),
  gate_helpers = file.path(revision_directory, "R", "distance_r1.R"),
  spectral_engine = file.path(full_directory, "R", "engine.R"),
  gated_distance = file.path(moon_directory, "R", "gated_distance.R")
)
invisible(lapply(source_files, source))

source_hashes <- unname(tools::md5sum(source_files))
names(source_hashes) <- names(source_files)
qmd_md5 <- unname(tools::md5sum(input_file))

# Runtime switches do not define a different statistical analysis.
analysis_contract <- analysis_parameters
analysis_contract$execution$run_experiments <- NULL
analysis_contract$execution$datasets <- NULL
analysis_contract$execution$allow_download <- NULL
analysis_contract$execution$show_progress <- NULL
analysis_contract$execution$output_directory <- NULL
analysis_contract$execution$raw_data_directory <- NULL

configuration_tag <- substr(
  digest::digest(
    list(
      parameters = analysis_contract,
      source_hashes = source_hashes
    ),
    algo = "xxhash64"
  ),
  1L, 12L
)

result_directory <- file.path(
  revision_directory,
  analysis_parameters$execution$output_directory,
  configuration_tag
)
raw_data_directory <- file.path(
  revision_directory,
  analysis_parameters$execution$raw_data_directory
)

kmeans_specification <- list(
  nstart = analysis_parameters$spectral$kmeans_nstart,
  iter_max = analysis_parameters$spectral$kmeans_iter_max,
  algorithm = analysis_parameters$spectral$kmeans_algorithm
)

classification_specification <- list(
  dataset = "SAheart",
  folds = 10L,
  repeats = 5L,
  maxit = 1000L,
  max_weights = 10000L,
  seed_offset = 707L,
  interaction_terms = "ldl:famhist",
  comparison = paste(
    "main effects versus the LDL-by-family-history interaction, motivated",
    "by the clinical interpretation of family history as an effect modifier"
  )
)

saheart_source <- "https://hastie.su.domains/ElemStatLearn/datasets/SAheart.data"
```

The analysis loads the local `manydist` source tree.  These checksums identify
the exact distance and clustering implementations used by the document.

```{r}
#| label: source-manifest

knitr::kable(data.frame(
  component = names(source_files),
  file = unname(source_files),
  md5 = source_hashes,
  row.names = NULL
))
```

# Frozen dataset registry

The registry below records the empirical role of every variable before any
method is fitted.  Labels never enter a dissimilarity or an algorithm-level
benchmark.  Continuous missing values are median-imputed, categorical missing
values receive an explicit `Missing` level, constant predictors are removed,
and continuous predictors are standardized.  The same completed analysis frame
is supplied to all methods.

```{r}
#| label: dataset-registry

dataset_registry <- data.frame(
  dataset = c(
    "Star", "Automobile", "Australian_credit", "Palmer_penguins",
    "Statlog_heart", "SAheart"
  ),
  display_name = c(
    "Star", "Automobile", "Australian Credit Approval", "Palmer Penguins",
    "Statlog Heart", "South African Heart"
  ),
  source = c(
    "FPDclustering::Star (CRAN copy of the Kaggle Star data)",
    analysis_parameters$sources$Automobile,
    analysis_parameters$sources$Australian_credit,
    "palmerpenguins::penguins",
    analysis_parameters$sources$Statlog_heart,
    saheart_source
  ),
  label = c(
    "Type (Red and Brown Dwarf merged)",
    "symboling (-2 merged with -1)",
    "credit approval class",
    "species",
    "presence of heart disease",
    "coronary heart disease"
  ),
  design_note = c(
    "Four physical measurements; colour and spectral class",
    "Engine location excluded because only three cars have rear engines",
    "Six continuous and eight nominal predictors",
    "Year excluded; four morphometric and two nominal predictors",
    "Six continuous and six nominal predictors; ordered slope excluded",
    paste(
      "Eight numerical cardiovascular risk factors and family history;",
      "LDL-by-family-history is the supervised interaction diagnostic"
    )
  ),
  stringsAsFactors = FALSE
)

knitr::kable(dataset_registry)
```

South African Heart contains one categorical predictor, family-history
status.  Conditional TVD requires another categorical variable against which
to form conditional profiles; with a single categorical predictor,
`manydist` therefore uses matching dissimilarity for that baseline block.  The
case study is included specifically to examine the numerical--categorical
interaction, whose construction and multiplicity gate are unchanged.

## Data acquisition and dataset-specific roles

```{r}
#| label: dataset-loaders

r1rd_normalize_name <- function(x) {
  gsub("[^a-z0-9]+", "_", tolower(x))
}

r1rd_package_data <- function(package, object) {
  if (!requireNamespace(package, quietly = TRUE)) {
    stop("Dataset `", object, "` requires package `", package, "`.")
  }
  environment <- new.env(parent = emptyenv())
  suppressWarnings(utils::data(list = object, package = package, envir = environment))
  if (!exists(object, envir = environment, inherits = FALSE)) {
    stop("Could not load `", package, "::", object, "`.")
  }
  get(object, envir = environment, inherits = FALSE)
}

r1rd_cached_source <- function(dataset, extension = "dat", url = NULL) {
  dir.create(raw_data_directory, recursive = TRUE, showWarnings = FALSE)
  destination <- file.path(raw_data_directory, paste0(dataset, ".", extension))
  if (file.exists(destination)) return(destination)
  if (!isTRUE(analysis_parameters$execution$allow_download)) {
    stop(
      "Missing cached source for `", dataset, "`: ", destination,
      ". Set execution$allow_download = TRUE for the first execution render."
    )
  }
  url <- url %||% analysis_parameters$sources[[dataset]]
  if (is.null(url) || !nzchar(url)) {
    stop("No download URL is registered for `", dataset, "`.")
  }
  status <- utils::download.file(url, destination, mode = "wb", quiet = FALSE)
  if (!identical(status, 0L) || !file.exists(destination)) {
    stop("Download failed for `", dataset, "` from ", url, ".")
  }
  destination
}

r1rd_column <- function(data, candidates, dataset) {
  normalized <- r1rd_normalize_name(names(data))
  candidate_names <- r1rd_normalize_name(candidates)
  match_index <- match(candidate_names, normalized, nomatch = 0L)
  match_index <- match_index[match_index > 0L]
  if (!length(match_index)) {
    stop(
      "None of the expected columns for `", dataset, "` were found: ",
      paste(candidates, collapse = ", "), ". Observed names: ",
      paste(names(data), collapse = ", ")
    )
  }
  names(data)[match_index[[1L]]]
}

r1rd_load_star <- function() {
  raw <- as.data.frame(r1rd_package_data("FPDclustering", "Star"))
  names(raw) <- r1rd_normalize_name(names(raw))
  selected <- c(
    k = r1rd_column(raw, c("K", "temperature"), "Star"),
    lum = r1rd_column(raw, c("Lum", "luminosity"), "Star"),
    rad = r1rd_column(raw, c("Rad", "radius"), "Star"),
    mag = r1rd_column(raw, c("Mag", "absolute_magnitude"), "Star"),
    col = r1rd_column(raw, c("Col", "color", "colour"), "Star"),
    spect = r1rd_column(raw, c("Spect", "spectral_class"), "Star"),
    type = r1rd_column(raw, c("Type", "star_type"), "Star")
  )
  raw <- raw[unname(selected)]
  names(raw) <- names(selected)
  truth <- as.character(raw$type)
  numeric_truth <- suppressWarnings(as.integer(truth))
  if (all(!is.na(numeric_truth)) && all(numeric_truth %in% 0:5)) {
    truth[numeric_truth %in% c(0L, 1L)] <- "Red_or_Brown_Dwarf"
  } else {
    dwarf <- grepl("red.*dwarf|brown.*dwarf", truth, ignore.case = TRUE)
    truth[dwarf] <- "Red_or_Brown_Dwarf"
  }
  list(
    x = data.frame(
      K = as.numeric(raw$k), Lum = as.numeric(raw$lum),
      Rad = as.numeric(raw$rad), Mag = as.numeric(raw$mag),
      Col = factor(raw$col), Spect = factor(raw$spect),
      check.names = FALSE
    ),
    truth = droplevels(factor(truth)),
    provenance = list(package = "FPDclustering", object = "Star")
  )
}

r1rd_load_automobile <- function() {
  path <- r1rd_cached_source("Automobile")
  column_names <- c(
    "symboling", "normalized_losses", "make", "fuel_type", "aspiration",
    "num_doors", "body_style", "drive_wheels", "engine_location",
    "wheel_base", "length", "width", "height", "curb_weight",
    "engine_type", "num_cylinders", "engine_size", "fuel_system", "bore",
    "stroke", "compression_ratio", "horsepower", "peak_rpm", "city_mpg",
    "highway_mpg", "price"
  )
  raw <- utils::read.csv(
    path, header = FALSE, col.names = column_names, na.strings = "?",
    stringsAsFactors = FALSE, strip.white = TRUE
  )
  numeric_names <- c(
    "normalized_losses", "wheel_base", "length", "width", "height",
    "curb_weight", "engine_size", "bore", "stroke", "compression_ratio",
    "horsepower", "peak_rpm", "city_mpg", "highway_mpg", "price"
  )
  categorical_names <- c(
    "make", "fuel_type", "aspiration", "num_doors", "body_style",
    "drive_wheels", "engine_type", "num_cylinders", "fuel_system"
  )
  raw[numeric_names] <- lapply(raw[numeric_names], as.numeric)
  raw[categorical_names] <- lapply(raw[categorical_names], factor)
  truth <- as.character(raw$symboling)
  truth[truth == "-2"] <- "-1"
  list(
    x = raw[c(numeric_names, categorical_names)],
    truth = droplevels(factor(truth)),
    provenance = list(url = analysis_parameters$sources$Automobile, file = path)
  )
}

r1rd_load_australian <- function() {
  path <- r1rd_cached_source("Australian_credit")
  raw <- utils::read.table(path, header = FALSE, stringsAsFactors = FALSE)
  if (ncol(raw) != 15L) stop("Australian Credit source must contain 15 columns.")
  names(raw) <- paste0("A", seq_len(15L))
  categorical_index <- c(1L, 4L, 5L, 6L, 8L, 9L, 11L, 12L)
  numeric_index <- c(2L, 3L, 7L, 10L, 13L, 14L)
  raw[categorical_index] <- lapply(raw[categorical_index], factor)
  raw[numeric_index] <- lapply(raw[numeric_index], as.numeric)
  list(
    x = raw[c(numeric_index, categorical_index)],
    truth = droplevels(factor(raw[[15L]])),
    provenance = list(url = analysis_parameters$sources$Australian_credit, file = path)
  )
}

r1rd_load_penguins <- function() {
  if (!requireNamespace("palmerpenguins", quietly = TRUE)) {
    stop("Dataset `Palmer_penguins` requires package `palmerpenguins`.")
  }
  raw <- as.data.frame(palmerpenguins::penguins)
  list(
    x = data.frame(
      bill_length_mm = raw$bill_length_mm,
      bill_depth_mm = raw$bill_depth_mm,
      flipper_length_mm = raw$flipper_length_mm,
      body_mass_g = raw$body_mass_g,
      island = factor(raw$island),
      sex = factor(raw$sex),
      check.names = FALSE
    ),
    truth = droplevels(factor(raw$species)),
    provenance = list(package = "palmerpenguins", object = "penguins")
  )
}

r1rd_load_heart <- function() {
  path <- r1rd_cached_source("Statlog_heart")
  raw <- utils::read.table(path, header = FALSE, stringsAsFactors = FALSE)
  if (ncol(raw) != 14L) stop("Statlog Heart source must contain 14 columns.")
  names(raw) <- paste0("V", seq_len(14L))
  numeric_index <- c(1L, 4L, 5L, 8L, 10L, 12L)
  categorical_index <- c(2L, 3L, 6L, 7L, 9L, 13L)
  raw[numeric_index] <- lapply(raw[numeric_index], as.numeric)
  raw[categorical_index] <- lapply(raw[categorical_index], factor)
  list(
    x = raw[c(numeric_index, categorical_index)],
    truth = droplevels(factor(raw[[14L]])),
    provenance = list(url = analysis_parameters$sources$Statlog_heart, file = path)
  )
}

r1rd_load_saheart <- function() {
  path <- r1rd_cached_source("SAheart", extension = "csv", url = saheart_source)
  raw <- utils::read.csv(path, stringsAsFactors = FALSE, strip.white = TRUE)
  required <- c(
    "sbp", "tobacco", "ldl", "adiposity", "famhist", "typea",
    "obesity", "alcohol", "age", "chd"
  )
  if (!all(required %in% names(raw))) {
    stop("South African Heart source does not contain the expected columns.")
  }
  numeric_names <- c(
    "sbp", "tobacco", "ldl", "adiposity", "typea", "obesity",
    "alcohol", "age"
  )
  raw[numeric_names] <- lapply(raw[numeric_names], as.numeric)
  list(
    x = data.frame(
      raw[numeric_names], famhist = factor(raw$famhist), check.names = FALSE
    ),
    truth = droplevels(factor(raw$chd)),
    provenance = list(
      url = saheart_source,
      source = "Hastie, Tibshirani, and Friedman, Elements of Statistical Learning",
      file = path
    )
  )
}

r1rd_load_dataset <- function(dataset) {
  switch(
    dataset,
    Star = r1rd_load_star(),
    Automobile = r1rd_load_automobile(),
    Australian_credit = r1rd_load_australian(),
    Palmer_penguins = r1rd_load_penguins(),
    Statlog_heart = r1rd_load_heart(),
    SAheart = r1rd_load_saheart(),
    stop("Unknown dataset: ", dataset)
  )
}
```

## Shared preprocessing

```{r}
#| label: preprocessing

r1rd_prepare_data <- function(loaded, dataset) {
  x <- as.data.frame(loaded$x, check.names = FALSE)
  truth <- droplevels(factor(loaded$truth))
  if (nrow(x) != length(truth)) stop("Predictors and labels have different lengths.")

  keep <- !is.na(truth)
  x <- x[keep, , drop = FALSE]
  truth <- droplevels(truth[keep])

  numeric_names <- names(x)[vapply(x, is.numeric, logical(1))]
  categorical_names <- names(x)[vapply(
    x, function(value) is.factor(value) || is.character(value), logical(1)
  )]
  if (length(c(numeric_names, categorical_names)) != ncol(x)) {
    stop("Every predictor must be numerical or nominal categorical in `", dataset, "`.")
  }

  numeric_missing <- integer(length(numeric_names))
  names(numeric_missing) <- numeric_names
  for (name in numeric_names) {
    value <- as.numeric(x[[name]])
    missing <- !is.finite(value)
    numeric_missing[[name]] <- sum(missing)
    median_value <- stats::median(value[!missing], na.rm = TRUE)
    if (!is.finite(median_value)) stop("All values are missing in `", name, "`.")
    value[missing] <- median_value
    x[[name]] <- value
  }

  categorical_missing <- integer(length(categorical_names))
  names(categorical_missing) <- categorical_names
  for (name in categorical_names) {
    value <- as.character(x[[name]])
    missing <- is.na(value) | !nzchar(trimws(value))
    categorical_missing[[name]] <- sum(missing)
    value[missing] <- analysis_parameters$preprocessing$missing_level
    x[[name]] <- droplevels(factor(value))
  }

  constant <- names(x)[vapply(x, function(value) {
    length(unique(value)) < 2L
  }, logical(1))]
  if (length(constant)) x <- x[setdiff(names(x), constant)]

  numeric_names <- names(x)[vapply(x, is.numeric, logical(1))]
  categorical_names <- names(x)[vapply(x, is.factor, logical(1))]
  if (!length(numeric_names) || !length(categorical_names)) {
    stop("`", dataset, "` must retain numerical and categorical predictors.")
  }

  if (isTRUE(analysis_parameters$preprocessing$standardize_numeric)) {
    x[numeric_names] <- lapply(x[numeric_names], function(value) {
      scaled <- as.numeric(scale(value))
      if (any(!is.finite(scaled))) stop("Numerical standardization failed.")
      scaled
    })
  }
  x[categorical_names] <- lapply(x[categorical_names], droplevels)

  list(
    x = x,
    truth = truth,
    provenance = loaded$provenance,
    preprocessing = list(
      numeric_missing = numeric_missing,
      categorical_missing = categorical_missing,
      removed_constant_predictors = constant
    )
  )
}
```

# Common dissimilarity and clustering functions

All six pairwise dissimilarities are passed to the same normalized spectral
clustering routine.  The Gaussian bandwidth is the median of the positive
off-diagonal dissimilarities, and the row-normalized spectral embedding has
dimension $K$, the known number of reference classes.  Its final k-means step
uses 100 internal starts and at most 100 iterations.

```{r}
#| label: common-functions

r1rd_global_gaussian <- function(distance) {
  distance <- full_validate_distance(distance)
  positive <- distance[upper.tri(distance) & distance > 0]
  if (!length(positive)) stop("Distance has no positive off-diagonal entries.")
  sigma <- stats::median(positive)
  if (!is.finite(sigma) || sigma <= 0) stop("Invalid Gaussian bandwidth.")
  affinity <- exp(-(distance^2) / (2 * sigma^2))
  diag(affinity) <- 0
  list(matrix = affinity, sigma = sigma)
}

r1rd_cluster_distance <- function(distance, truth, seed) {
  affinity <- r1rd_global_gaussian(distance)
  fit <- full_spectral_cluster(
    affinity$matrix,
    k = nlevels(truth),
    seed = seed,
    kmeans_specification = kmeans_specification
  )
  list(
    cluster = fit$cluster,
    ARI = mclust::adjustedRandIndex(fit$cluster, truth),
    objective = fit$objective,
    iterations = fit$iterations,
    sigma = affinity$sigma
  )
}

r1rd_alignment_ratio <- function(distance, partition) {
  distance <- full_validate_distance(distance)
  upper <- upper.tri(distance)
  same <- outer(partition, partition, `==`)[upper]
  values <- distance[upper]
  within <- mean(values[same])
  between <- mean(values[!same])
  if (is.finite(within) && within > 0) between / within else NA_real_
}

r1rd_distance_spearman <- function(first, second) {
  first <- full_validate_distance(first)
  second <- full_validate_distance(second)
  if (!identical(dim(first), dim(second))) {
    stop("Distance matrices must have identical dimensions.")
  }
  upper <- upper.tri(first)
  if (!stats::sd(first[upper]) || !stats::sd(second[upper])) return(NA_real_)
  suppressWarnings(stats::cor(
    first[upper], second[upper], method = "spearman"
  ))
}

r1rd_knn_overlap <- function(first, second, proportion) {
  first <- full_validate_distance(first)
  second <- full_validate_distance(second)
  if (!identical(dim(first), dim(second))) {
    stop("Distance matrices must have identical dimensions.")
  }
  n <- nrow(first)
  k <- max(1L, min(n - 1L, ceiling(proportion * (n - 1L))))
  overlap <- vapply(seq_len(n), function(index) {
    first_order <- order(first[index, ], na.last = NA)
    second_order <- order(second[index, ], na.last = NA)
    first_neighbours <- head(first_order[first_order != index], k)
    second_neighbours <- head(second_order[second_order != index], k)
    length(intersect(first_neighbours, second_neighbours)) / k
  }, numeric(1))
  mean(overlap)
}

r1rd_stratified_folds <- function(truth, folds, seed) {
  truth <- droplevels(factor(truth))
  if (folds < 2L || folds > min(table(truth))) {
    stop("Invalid number of stratified folds.")
  }
  set.seed(seed)
  assignment <- integer(length(truth))
  for (level in levels(truth)) {
    index <- which(truth == level)
    assignment[index] <- sample(rep(seq_len(folds), length.out = length(index)))
  }
  assignment
}

r1rd_classification_metrics <- function(truth, predicted, probabilities) {
  truth <- droplevels(factor(truth))
  predicted <- factor(predicted, levels = levels(truth))
  confusion <- table(truth, predicted)
  recall <- diag(confusion) / rowSums(confusion)
  precision <- diag(confusion) / colSums(confusion)
  f1 <- 2 * precision * recall / (precision + recall)
  f1[!is.finite(f1)] <- 0

  probabilities <- as.matrix(probabilities)
  probabilities <- probabilities[, levels(truth), drop = FALSE]
  observed_probability <- probabilities[cbind(
    seq_along(truth), match(truth, levels(truth))
  )]
  observed_probability <- pmax(pmin(observed_probability, 1 - 1e-15), 1e-15)
  c(
    accuracy = mean(predicted == truth),
    balanced_accuracy = mean(recall),
    macro_F1 = mean(f1),
    log_loss = -mean(log(observed_probability))
  )
}

r1rd_run_classification_diagnostic <- function(data, truth, seed) {
  numeric_names <- names(data)[vapply(data, is.numeric, logical(1))]
  categorical_names <- names(data)[vapply(data, is.factor, logical(1))]
  if (!length(numeric_names) || !length(categorical_names)) {
    stop("The classification diagnostic requires both predictor types.")
  }

  model_data <- data
  model_data$.truth <- droplevels(factor(truth))
  main_formula <- stats::reformulate(names(data), response = ".truth")
  interaction_terms <- classification_specification$interaction_terms %||%
    as.vector(outer(numeric_names, categorical_names, paste, sep = ":"))
  unknown_interaction_variables <- setdiff(
    unique(unlist(strsplit(interaction_terms, ":", fixed = TRUE))),
    names(data)
  )
  if (length(unknown_interaction_variables)) {
    stop(
      "Unknown variables in the supervised interaction specification: ",
      paste(unknown_interaction_variables, collapse = ", ")
    )
  }
  interaction_formula <- stats::reformulate(
    c(names(data), interaction_terms), response = ".truth"
  )

  fit_model <- function(formula, training_data) {
    nnet::multinom(
      formula,
      data = training_data,
      trace = FALSE,
      maxit = classification_specification$maxit,
      MaxNWts = classification_specification$max_weights
    )
  }

  repeat_rows <- list()
  row_index <- 0L
  classes <- levels(model_data$.truth)
  model_formulas <- list(
    main_effects = main_formula,
    numeric_categorical_interactions = interaction_formula
  )

  for (repeat_index in seq_len(classification_specification$repeats)) {
    fold <- r1rd_stratified_folds(
      model_data$.truth,
      classification_specification$folds,
      seed + repeat_index
    )
    for (model_name in names(model_formulas)) {
      predicted <- rep(NA_character_, nrow(model_data))
      probabilities <- matrix(
        NA_real_, nrow(model_data), length(classes),
        dimnames = list(NULL, classes)
      )
      for (fold_index in seq_len(classification_specification$folds)) {
        test <- fold == fold_index
        fit <- fit_model(model_formulas[[model_name]], model_data[!test, ])
        fold_probability <- stats::predict(
          fit, newdata = model_data[test, , drop = FALSE], type = "probs"
        )
        if (is.null(dim(fold_probability))) {
          fold_probability <- cbind(
            1 - fold_probability, fold_probability
          )
          colnames(fold_probability) <- classes
        } else {
          fold_probability <- as.matrix(fold_probability)
        }
        probabilities[test, colnames(fold_probability)] <- fold_probability
        predicted[test] <- classes[max.col(fold_probability, ties.method = "first")]
      }
      metrics <- r1rd_classification_metrics(
        model_data$.truth, predicted, probabilities
      )
      row_index <- row_index + 1L
      repeat_rows[[row_index]] <- data.frame(
        repeat_id = repeat_index,
        model = model_name,
        accuracy = unname(metrics[["accuracy"]]),
        balanced_accuracy = unname(metrics[["balanced_accuracy"]]),
        macro_F1 = unname(metrics[["macro_F1"]]),
        log_loss = unname(metrics[["log_loss"]]),
        stringsAsFactors = FALSE
      )
    }
  }
  repeat_results <- do.call(rbind, repeat_rows)
  performance <- do.call(rbind, lapply(names(model_formulas), function(model_name) {
    observed <- repeat_results[repeat_results$model == model_name, , drop = FALSE]
    data.frame(
      model = model_name,
      accuracy_mean = mean(observed$accuracy),
      accuracy_sd = stats::sd(observed$accuracy),
      balanced_accuracy_mean = mean(observed$balanced_accuracy),
      balanced_accuracy_sd = stats::sd(observed$balanced_accuracy),
      macro_F1_mean = mean(observed$macro_F1),
      macro_F1_sd = stats::sd(observed$macro_F1),
      log_loss_mean = mean(observed$log_loss),
      log_loss_sd = stats::sd(observed$log_loss),
      stringsAsFactors = FALSE
    )
  }))
  rownames(performance) <- NULL

  full_main <- fit_model(main_formula, model_data)
  full_interaction <- fit_model(interaction_formula, model_data)
  main_loglik <- stats::logLik(full_main)
  interaction_loglik <- stats::logLik(full_interaction)
  likelihood_ratio <- 2 * (as.numeric(interaction_loglik) - as.numeric(main_loglik))
  degrees_freedom <- attr(interaction_loglik, "df") - attr(main_loglik, "df")
  main_performance <- performance[performance$model == "main_effects", ]
  interaction_performance <- performance[
    performance$model == "numeric_categorical_interactions", 
  ]
  omnibus <- data.frame(
    main_effects_df = attr(main_loglik, "df"),
    interaction_model_df = attr(interaction_loglik, "df"),
    likelihood_ratio = likelihood_ratio,
    degrees_freedom = degrees_freedom,
    p_value = stats::pchisq(likelihood_ratio, degrees_freedom, lower.tail = FALSE),
    accuracy_gain = interaction_performance$accuracy_mean -
      main_performance$accuracy_mean,
    balanced_accuracy_gain = interaction_performance$balanced_accuracy_mean -
      main_performance$balanced_accuracy_mean,
    macro_F1_gain = interaction_performance$macro_F1_mean -
      main_performance$macro_F1_mean,
    log_loss_reduction = main_performance$log_loss_mean -
      interaction_performance$log_loss_mean,
    majority_class_accuracy = max(prop.table(table(model_data$.truth))),
    main_convergence = full_main$convergence,
    interaction_convergence = full_interaction$convergence,
    stringsAsFactors = FALSE
  )

  list(
    performance = performance,
    omnibus = omnibus,
    repeat_results = repeat_results,
    formulas = list(
      main_effects = paste(deparse(main_formula), collapse = " "),
      numeric_categorical_interactions = paste(
        deparse(interaction_formula), collapse = " "
      )
    ),
    specification = classification_specification
  )
}

r1rd_equal_width_factor <- function(x, bins) {
  observed <- range(x, finite = TRUE)
  if (!all(is.finite(observed)) || observed[[1L]] == observed[[2L]]) {
    return(factor(rep("constant", length(x))))
  }
  factor(cut(
    x,
    breaks = seq(observed[[1L]], observed[[2L]], length.out = bins + 1L),
    include.lowest = TRUE,
    ordered_result = FALSE
  ))
}

r1rd_minmax_scale <- function(x) {
  observed <- range(x, finite = TRUE)
  if (!all(is.finite(observed)) || observed[[1L]] == observed[[2L]]) {
    return(rep(0, length(x)))
  }
  (x - observed[[1L]]) / (observed[[2L]] - observed[[1L]])
}

# Pairwise form of Ahmad and Dey (2007), Equation (3.3.1), matching the
# implementation used in R1_simulations.qmd.
r1rd_ahmad_dey_distance <- function(data, bins = 5L) {
  bins <- as.integer(bins)
  numeric_columns <- vapply(data, is.numeric, logical(1))
  categorical_columns <- vapply(data, is.factor, logical(1))
  numeric_data <- data[, numeric_columns, drop = FALSE]
  categorical_data <- data[, categorical_columns, drop = FALSE]

  discretized_numeric <- lapply(
    numeric_data, r1rd_equal_width_factor, bins = bins
  )
  names(discretized_numeric) <- names(numeric_data)
  discrete_profiles <- data.frame(
    categorical_data, discretized_numeric, check.names = FALSE
  )
  discrete_profiles[] <- lapply(discrete_profiles, function(x) {
    droplevels(factor(x))
  })
  if (ncol(discrete_profiles) < 2L) {
    stop("Ahmad--Dey conditional profiles require at least two variables.")
  }

  delta <- as.matrix(manydist:::cat_delta(
    discrete_profiles, method_cat = "tvd"
  )$tvd)
  level_counts <- vapply(discrete_profiles, nlevels, integer(1))
  stops <- cumsum(level_counts)
  starts <- c(1L, head(stops, -1L) + 1L)
  positions <- Map(seq.int, starts, stops)

  n <- nrow(data)
  categorical_distance <- matrix(0, n, n)
  for (column_index in seq_len(ncol(categorical_data))) {
    levels_index <- positions[[column_index]]
    observed_index <- as.integer(discrete_profiles[[column_index]])
    contribution <- delta[levels_index, levels_index, drop = FALSE][
      observed_index, observed_index, drop = FALSE
    ]
    categorical_distance <- categorical_distance + contribution^2
  }

  numeric_distance <- matrix(0, n, n)
  if (ncol(numeric_data)) {
    numeric_profile_index <- ncol(categorical_data) + seq_len(ncol(numeric_data))
    numeric_weights <- vapply(numeric_profile_index, function(column_index) {
      block <- delta[
        positions[[column_index]], positions[[column_index]], drop = FALSE
      ]
      if (nrow(block) < 2L) 0 else mean(block[upper.tri(block)])
    }, numeric(1))
    scaled_numeric <- as.data.frame(lapply(numeric_data, r1rd_minmax_scale))
    weighted_numeric <- sweep(as.matrix(scaled_numeric), 2L, numeric_weights, `*`)
    numeric_distance <- as.matrix(stats::dist(weighted_numeric))^2
  }

  distance <- numeric_distance + categorical_distance
  diag(distance) <- 0
  full_validate_distance(distance, "Ahmad--Dey")
}

# The external Modified Gower port currently drops dimensions internally when
# exactly one categorical column is present.  This algebraically equivalent
# wrapper preserves that column as a matrix for the South African Heart case.
r1rd_modified_gower_single_categorical <- function(data) {
  numeric_columns <- vapply(data, is.numeric, logical(1))
  numeric_data <- data[, numeric_columns, drop = FALSE]
  categorical_data <- data[, !numeric_columns, drop = FALSE]
  if (ncol(categorical_data) != 1L) {
    stop("The single-categorical Modified Gower wrapper received another design.")
  }

  numerical_dispersion <- vapply(numeric_data, function(value) {
    denominator <- stats::IQR(value)
    if (!is.finite(denominator) || denominator <= 0) return(0)
    numerator <- sum(abs(outer(value, value, `-`))[upper.tri(diag(length(value)))])
    numerator / denominator
  }, numeric(1))
  category <- categorical_data[[1L]]
  categorical_mismatch <- sum(outer(category, category, `!=`)[
    upper.tri(diag(length(category)))
  ])
  categorical_dispersion <- 2 * categorical_mismatch / nlevels(category)
  self_adaptive_weight <- mean(numerical_dispersion) / categorical_dispersion

  mutual_information_weight <- manydist:::mg_normalized_MI(
    numeric_data, categorical_data
  )
  feature_weight <- c(rep(1, ncol(numeric_data)), self_adaptive_weight) *
    mutual_information_weight
  feature_weight[!is.finite(feature_weight)] <- 0
  if (!any(feature_weight > 0)) {
    stop("Modified Gower produced no positive feature weights.")
  }
  feature_weight <- feature_weight / sum(feature_weight)

  full_validate_distance(
    as.matrix(manydist:::mg_gower_dist_modify(
      data.x = data,
      var.weights = feature_weight,
      robcb = "iqr",
      KR.corr = TRUE
    )),
    "Modified Gower"
  )
}

r1rd_distance_set <- function(data, gated_fit) {
  categorical_count <- sum(vapply(data, is.factor, logical(1)))
  list(
    D_mint = gated_fit$distance,
    D_m = gated_fit$distance_no_interaction,
    Gower = as.matrix(manydist::mdist(data, preset = "gower")$distance),
    Modified_Gower = if (categorical_count == 1L) {
      r1rd_modified_gower_single_categorical(data)
    } else {
      as.matrix(manydist::mdist(data, preset = "mod_gower")$distance)
    },
    Euclidean_onehot = as.matrix(
      manydist::mdist(data, preset = "euclidean")$distance
    ),
    Ahmad_Dey = r1rd_ahmad_dey_distance(
      data, bins = analysis_parameters$distance$ahmad_dey_bins
    )
  )
}

r1rd_run_kprototypes <- function(data, truth, seed) {
  if (!requireNamespace("clustMixType", quietly = TRUE)) {
    stop("k-prototypes requires package `clustMixType`.")
  }
  specification <- analysis_parameters$algorithm_benchmarks$kprototypes
  set.seed(seed)
  fit <- clustMixType::kproto(
    data,
    k = nlevels(truth),
    lambda = specification$lambda,
    type = specification$type,
    iter.max = specification$iter_max,
    nstart = specification$nstart,
    na.rm = "no",
    keep.data = FALSE,
    verbose = FALSE
  )
  list(
    cluster = fit$cluster,
    ARI = mclust::adjustedRandIndex(fit$cluster, truth),
    objective = fit$tot.withinss %||% sum(fit$withinss),
    iterations = fit$iter %||% NA_integer_,
    lambda = fit$lambda %||% NA_real_
  )
}

r1rd_run_kamila <- function(data, truth, seed) {
  if (!requireNamespace("kamila", quietly = TRUE)) {
    stop("KAMILA requires package `kamila`.")
  }
  specification <- analysis_parameters$algorithm_benchmarks$kamila
  numeric_data <- data[vapply(data, is.numeric, logical(1))]
  categorical_data <- data[vapply(data, is.factor, logical(1))]
  set.seed(seed)
  fit <- kamila::kamila(
    conVar = numeric_data,
    catFactor = categorical_data,
    numClust = nlevels(truth),
    numInit = specification$num_init,
    maxIter = specification$max_iter,
    conInitMethod = specification$con_init_method,
    catBw = specification$cat_bw,
    verbose = FALSE,
    calcNumClust = "none"
  )
  list(
    cluster = fit$finalMemb,
    ARI = mclust::adjustedRandIndex(fit$finalMemb, truth),
    objective = fit$finalObj %||% fit$finalLogLik %||% NA_real_,
    # `numIter` contains one value for every internal initialization.  The
    # clustering and objective above describe the single solution selected by
    # KAMILA, so retain one scalar diagnostic rather than expanding the result
    # table to `numInit` rows.
    iterations = if (length(fit$numIter)) {
      max(as.integer(fit$numIter), na.rm = TRUE)
    } else {
      NA_integer_
    }
  )
}
```

# Dataset-level analysis

```{r}
#| label: analysis-functions

method_labels <- c(
  D_mint = "D_mint",
  D_m = "D_m",
  Gower = "Gower",
  Modified_Gower = "Modified Gower",
  Euclidean_onehot = "Euclidean one-hot",
  Ahmad_Dey = "Ahmad--Dey",
  Kprototypes = "k-prototypes",
  KAMILA = "KAMILA"
)

r1rd_artifact_file <- function(dataset) {
  file.path(result_directory, paste0(dataset, ".rds"))
}

r1rd_result_row <- function(dataset, method, family, fit = NULL, seconds = NA_real_,
                            error = NA_character_) {
  data.frame(
    dataset = dataset,
    method = method,
    method_label = unname(method_labels[[method]]),
    family = family,
    ARI = if (is.null(fit)) NA_real_ else fit$ARI,
    objective = if (is.null(fit)) NA_real_ else fit$objective %||% NA_real_,
    iterations = if (is.null(fit)) NA_integer_ else fit$iterations %||% NA_integer_,
    elapsed_seconds = seconds,
    error = error,
    stringsAsFactors = FALSE
  )
}

r1rd_safe_fit <- function(expression) {
  started <- proc.time()[[3L]]
  result <- tryCatch(
    list(value = force(expression), error = NA_character_),
    error = function(e) list(value = NULL, error = conditionMessage(e))
  )
  result$seconds <- proc.time()[[3L]] - started
  result
}

r1rd_neighbour_sensitivity <- function(data, truth, dataset, primary_fit,
                                       primary_partition, seed) {
  if (!isTRUE(analysis_parameters$neighbour_sensitivity$enabled)) {
    return(list(results = data.frame(), gates = data.frame()))
  }
  proportions <- sort(unique(c(
    analysis_parameters$neighbour_sensitivity$proportions,
    analysis_parameters$distance$prop_nn
  )))
  result_rows <- list()
  gate_rows <- list()

  for (index in seq_along(proportions)) {
    prop_nn <- proportions[[index]]
    is_primary <- isTRUE(all.equal(prop_nn, analysis_parameters$distance$prop_nn))
    fit <- if (is_primary) primary_fit else scmix_gated_active_distance(
      data = data,
      gamma = analysis_parameters$distance$gamma,
      prop_nn = prop_nn,
      score = analysis_parameters$distance$score,
      decision = analysis_parameters$distance$decision,
      gate_permutations = analysis_parameters$distance$gate_permutations,
      gate_alpha = analysis_parameters$distance$gate_alpha,
      gate_seed = seed + analysis_parameters$distance$gate_seed_offset
    )
    partition <- if (is_primary) primary_partition else {
      r1rd_cluster_distance(fit$distance, truth, seed)$cluster
    }
    result_rows[[index]] <- data.frame(
      dataset = dataset,
      prop_nn = prop_nn,
      ARI = mclust::adjustedRandIndex(partition, truth),
      partition_ARI_vs_primary = mclust::adjustedRandIndex(
        partition, primary_partition
      ),
      active_count = fit$diagnostics$active_count,
      active_variables = paste(fit$diagnostics$active_names, collapse = ", "),
      stringsAsFactors = FALSE
    )
    gate_rows[[index]] <- data.frame(
      dataset = dataset,
      prop_nn = prop_nn,
      variable = names(fit$gate$detected),
      detected = unname(fit$gate$detected),
      multiplicity_adjusted_p_value = unname(fit$gate$p_value),
      statistic = unname(fit$gate$statistic),
      stringsAsFactors = FALSE
    )
  }
  list(
    results = do.call(rbind, result_rows),
    gates = do.call(rbind, gate_rows)
  )
}

r1rd_run_dataset <- function(dataset, dataset_index) {
  output_file <- r1rd_artifact_file(dataset)
  if (file.exists(output_file)) return(output_file)
  dir.create(dirname(output_file), recursive = TRUE, showWarnings = FALSE)

  started_at <- Sys.time()
  seed <- analysis_parameters$execution$seed_base + dataset_index * 1000L
  prepared <- r1rd_prepare_data(r1rd_load_dataset(dataset), dataset)
  data <- prepared$x
  truth <- prepared$truth

  gated_fit <- scmix_gated_active_distance(
    data = data,
    gamma = analysis_parameters$distance$gamma,
    prop_nn = analysis_parameters$distance$prop_nn,
    score = analysis_parameters$distance$score,
    decision = analysis_parameters$distance$decision,
    gate_permutations = analysis_parameters$distance$gate_permutations,
    gate_alpha = analysis_parameters$distance$gate_alpha,
    gate_seed = seed + analysis_parameters$distance$gate_seed_offset
  )

  distances <- r1rd_distance_set(data, gated_fit)
  result_rows <- list()
  partitions <- list()
  row_index <- 0L

  for (method in names(distances)) {
    if (!isTRUE(analysis_parameters$methods[[method]])) next
    fitted <- r1rd_safe_fit(r1rd_cluster_distance(distances[[method]], truth, seed))
    row_index <- row_index + 1L
    result_rows[[row_index]] <- r1rd_result_row(
      dataset, method, "spectral dissimilarity", fitted$value,
      fitted$seconds, fitted$error
    )
    if (!is.null(fitted$value)) partitions[[method]] <- fitted$value$cluster
  }

  if (isTRUE(analysis_parameters$methods[["Kprototypes"]])) {
    fitted <- r1rd_safe_fit(r1rd_run_kprototypes(data, truth, seed + 101L))
    row_index <- row_index + 1L
    result_rows[[row_index]] <- r1rd_result_row(
      dataset, "Kprototypes", "algorithm benchmark", fitted$value,
      fitted$seconds, fitted$error
    )
    if (!is.null(fitted$value)) partitions$Kprototypes <- fitted$value$cluster
  }
  if (isTRUE(analysis_parameters$methods[["KAMILA"]])) {
    fitted <- r1rd_safe_fit(r1rd_run_kamila(data, truth, seed + 202L))
    row_index <- row_index + 1L
    result_rows[[row_index]] <- r1rd_result_row(
      dataset, "KAMILA", "algorithm benchmark", fitted$value,
      fitted$seconds, fitted$error
    )
    if (!is.null(fitted$value)) partitions$KAMILA <- fitted$value$cluster
  }

  results <- do.call(rbind, result_rows)
  primary_partition <- partitions$D_mint
  baseline_partition <- partitions$D_m
  if (is.null(primary_partition) || is.null(baseline_partition)) {
    stop("Both D_mint and D_m must succeed for the interaction diagnostics.")
  }

  dmint_ari <- results$ARI[results$method == "D_mint"]
  dm_ari <- results$ARI[results$method == "D_m"]
  partition_ari <- mclust::adjustedRandIndex(
    primary_partition, baseline_partition
  )
  alignment <- if (gated_fit$diagnostics$active_count > 0L) {
    r1rd_alignment_ratio(
      gated_fit$components$interaction_active_average,
      primary_partition
    )
  } else {
    NA_real_
  }
  interaction_baseline_spearman <- interaction_numeric_spearman <-
    interaction_categorical_spearman <- NA_real_
  neighbourhood_overlap <- 1
  if (gated_fit$diagnostics$active_count > 0L) {
    active_interaction <- gated_fit$components$interaction_active_average
    interaction_baseline_spearman <- r1rd_distance_spearman(
      active_interaction, gated_fit$distance_no_interaction
    )
    interaction_numeric_spearman <- r1rd_distance_spearman(
      active_interaction, gated_fit$components$numeric
    )
    interaction_categorical_spearman <- r1rd_distance_spearman(
      active_interaction, gated_fit$components$categorical
    )
    neighbourhood_overlap <- r1rd_knn_overlap(
      gated_fit$distance_no_interaction,
      gated_fit$distance,
      analysis_parameters$distance$prop_nn
    )
  }

  sensitivity <- r1rd_neighbour_sensitivity(
    data, truth, dataset, gated_fit, primary_partition, seed
  )
  classification <- if (identical(
    dataset, classification_specification$dataset
  )) {
    r1rd_run_classification_diagnostic(
      data, truth, seed + classification_specification$seed_offset
    )
  } else {
    NULL
  }
  gate_table <- data.frame(
    dataset = dataset,
    variable = names(gated_fit$gate$detected),
    detected = unname(gated_fit$gate$detected),
    multiplicity_adjusted_p_value = unname(gated_fit$gate$p_value),
    statistic = unname(gated_fit$gate$statistic),
    stringsAsFactors = FALSE
  )

  artifact <- list(
    schema_version = 2L,
    analysis_version = analysis_parameters$analysis_version,
    configuration_tag = configuration_tag,
    dataset = dataset,
    results = results,
    gate = gate_table,
    neighbour_sensitivity = sensitivity$results,
    neighbour_sensitivity_gates = sensitivity$gates,
    classification = classification,
    diagnostics = data.frame(
      dataset = dataset,
      n = nrow(data),
      p_numeric = sum(vapply(data, is.numeric, logical(1))),
      q_categorical = sum(vapply(data, is.factor, logical(1))),
      K = nlevels(truth),
      active_count = gated_fit$diagnostics$active_count,
      active_variables = paste(gated_fit$diagnostics$active_names, collapse = ", "),
      ARI_D_mint_minus_D_m = dmint_ari - dm_ari,
      partition_ARI_D_mint_vs_D_m = partition_ari,
      partition_changed = is.finite(partition_ari) && partition_ari < 1 - 1e-12,
      R_align = alignment,
      interaction_baseline_spearman = interaction_baseline_spearman,
      interaction_numeric_spearman = interaction_numeric_spearman,
      interaction_categorical_spearman = interaction_categorical_spearman,
      neighbourhood_overlap_D_m_vs_D_mint = neighbourhood_overlap,
      interaction_added_mean = gated_fit$diagnostics$added_mean_distinct,
      stringsAsFactors = FALSE
    ),
    preprocessing = prepared$preprocessing,
    data_provenance = prepared$provenance,
    parameters = analysis_parameters,
    source_hashes = source_hashes,
    qmd_md5 = qmd_md5,
    package_versions = vapply(
      c("manydist", "mclust", "clustMixType", "kamila", "nnet"),
      function(package) {
        if (requireNamespace(package, quietly = TRUE)) {
          as.character(utils::packageVersion(package))
        } else {
          NA_character_
        }
      },
      character(1)
    ),
    started_at = format(started_at, tz = "UTC", usetz = TRUE),
    finished_at = format(Sys.time(), tz = "UTC", usetz = TRUE),
    session_info = capture.output(utils::sessionInfo())
  )

  temporary_file <- tempfile("R1-real-", tmpdir = dirname(output_file), fileext = ".rds")
  saveRDS(artifact, temporary_file, version = 3L)
  if (!file.rename(temporary_file, output_file)) {
    unlink(temporary_file)
    stop("Could not atomically save artifact: ", output_file)
  }
  output_file
}
```

# Optional execution

```{r}
#| label: execute-datasets

selected_datasets <- intersect(
  dataset_registry$dataset,
  analysis_parameters$execution$datasets
)
unknown_datasets <- setdiff(
  analysis_parameters$execution$datasets,
  dataset_registry$dataset
)
if (length(unknown_datasets)) {
  stop("Unknown selected datasets: ", paste(unknown_datasets, collapse = ", "))
}

selected_files <- vapply(selected_datasets, r1rd_artifact_file, character(1))
pending <- selected_datasets[!file.exists(selected_files)]

if (isTRUE(analysis_parameters$execution$run_experiments) && length(pending)) {
  benchmark_packages <- c("clustMixType", "kamila")
  missing_benchmarks <- benchmark_packages[!vapply(
    benchmark_packages, requireNamespace, logical(1), quietly = TRUE
  )]
  if (length(missing_benchmarks)) {
    stop(
      "Install the enabled benchmark packages before execution: ",
      paste(missing_benchmarks, collapse = ", ")
    )
  }
  if ("Star" %in% pending && !requireNamespace("FPDclustering", quietly = TRUE)) {
    stop("Install `FPDclustering` before running the Star analysis.")
  }
  if ("Palmer_penguins" %in% pending &&
      !requireNamespace("palmerpenguins", quietly = TRUE)) {
    stop("Install `palmerpenguins` before running the Penguins analysis.")
  }

  for (dataset in pending) {
    if (isTRUE(analysis_parameters$execution$show_progress)) {
      message("Running real-data analysis: ", dataset)
    }
    r1rd_run_dataset(dataset, match(dataset, dataset_registry$dataset))
  }
} else if (!isTRUE(analysis_parameters$execution$run_experiments)) {
  message(
    "Safe render: no analyses launched. Completed artifacts: ",
    sum(file.exists(selected_files)), "/", length(selected_files), "."
  )
}
```

# Consolidated results

```{r}
#| label: collect-artifacts

all_files <- vapply(dataset_registry$dataset, r1rd_artifact_file, character(1))
completed_files <- all_files[file.exists(all_files)]
artifacts <- lapply(completed_files, readRDS)

# Artifacts produced by the first execution of this document stored KAMILA's
# per-initialization `numIter` vector as repeated, otherwise identical result
# rows.  Collapse any such rows on collection.  This leaves the fitted
# clustering and ARI unchanged and makes old and newly produced artifacts obey
# the intended one-row-per-dataset-and-method reporting contract.
r1rd_collapse_result_rows <- function(results) {
  groups <- split(
    seq_len(nrow(results)),
    interaction(results$dataset, results$method, drop = TRUE)
  )
  rows <- lapply(groups, function(index) {
    row <- results[index[[1L]], , drop = FALSE]
    observed_iterations <- results$iterations[index]
    if (any(is.finite(observed_iterations))) {
      row$iterations <- max(observed_iterations, na.rm = TRUE)
    }
    row
  })
  output <- do.call(rbind, rows)
  rownames(output) <- NULL
  output
}

r1rd_bind_rows_fill <- function(rows) {
  rows <- Filter(function(value) !is.null(value) && nrow(value), rows)
  if (!length(rows)) return(data.frame())
  all_names <- unique(unlist(lapply(rows, names), use.names = FALSE))
  rows <- lapply(rows, function(value) {
    missing_names <- setdiff(all_names, names(value))
    for (name in missing_names) value[[name]] <- NA
    value[all_names]
  })
  output <- do.call(rbind, rows)
  rownames(output) <- NULL
  output
}

if (length(artifacts)) {
  result_table <- r1rd_collapse_result_rows(do.call(
    rbind, lapply(artifacts, `[[`, "results")
  ))
  diagnostic_table <- r1rd_bind_rows_fill(lapply(
    artifacts, `[[`, "diagnostics"
  ))
  gate_table <- r1rd_bind_rows_fill(lapply(artifacts, `[[`, "gate"))
  sensitivity_table <- r1rd_bind_rows_fill(
    lapply(artifacts, `[[`, "neighbour_sensitivity")
  )
  classification_performance_table <- r1rd_bind_rows_fill(lapply(
    artifacts,
    function(artifact) {
      if (is.null(artifact$classification)) return(NULL)
      data.frame(
        dataset = artifact$dataset,
        artifact$classification$performance,
        check.names = FALSE
      )
    }
  ))
  classification_test_table <- r1rd_bind_rows_fill(lapply(
    artifacts,
    function(artifact) {
      if (is.null(artifact$classification)) return(NULL)
      data.frame(
        dataset = artifact$dataset,
        artifact$classification$omnibus,
        check.names = FALSE
      )
    }
  ))
} else {
  result_table <- diagnostic_table <- gate_table <- sensitivity_table <-
    classification_performance_table <- classification_test_table <- data.frame()
}
```

## Dataset characteristics

```{r}
#| label: dataset-characteristics

if (nrow(diagnostic_table)) {
  characteristics <- diagnostic_table[
    match(dataset_registry$dataset, diagnostic_table$dataset, nomatch = 0L),
    c("dataset", "n", "p_numeric", "q_categorical", "K"),
    drop = FALSE
  ]
  characteristics <- merge(
    dataset_registry[c("dataset", "display_name", "label")],
    characteristics,
    by = "dataset",
    all.x = TRUE,
    sort = FALSE
  )
  knitr::kable(characteristics, row.names = FALSE)
} else {
  cat("No dataset artifacts are available yet.")
}
```

## Consolidated ARI table

```{r}
#| label: consolidated-ari

if (nrow(result_table)) {
  ari_long <- result_table[c("dataset", "method_label", "ARI")]
  ari_wide <- reshape(
    ari_long,
    idvar = "dataset",
    timevar = "method_label",
    direction = "wide"
  )
  names(ari_wide) <- sub("^ARI\\.", "", names(ari_wide))
  ari_wide <- ari_wide[match(
    dataset_registry$dataset, ari_wide$dataset, nomatch = 0L
  ), , drop = FALSE]
  desired_columns <- c("dataset", unname(method_labels))
  ari_wide <- ari_wide[c(
    intersect(desired_columns, names(ari_wide)),
    setdiff(names(ari_wide), desired_columns)
  )]
  numeric_columns <- vapply(ari_wide, is.numeric, logical(1))
  ari_wide[numeric_columns] <- lapply(
    ari_wide[numeric_columns], round, digits = 3L
  )
  knitr::kable(ari_wide, row.names = FALSE)
} else {
  cat("No ARI results are available yet.")
}
```

## Supervised interaction diagnostic

The South African Heart labels are used here only as an external diagnostic.
A logistic classifier containing main effects is compared with one that also
allows the effect of LDL cholesterol to vary with family-history status.  This
clinically interpretable numerical-by-categorical interaction is specified
independently of the distance construction.  Performance is estimated by repeated
stratified 10-fold cross-validation.  The likelihood-ratio test assesses the
added interaction, while the cross-validated differences establish whether
its contribution generalizes out of sample.  The labels and fitted classifiers
are not used by the interaction gate or by any clustering method.

```{r}
#| label: supervised-interaction-performance

if (nrow(classification_performance_table)) {
  shown <- classification_performance_table
  model_labels <- c(
    main_effects = "Main effects",
    numeric_categorical_interactions = "Main effects + LDL by family history"
  )
  shown$model <- unname(model_labels[shown$model])
  numeric_columns <- vapply(shown, is.numeric, logical(1))
  shown[numeric_columns] <- lapply(shown[numeric_columns], round, digits = 3L)
  knitr::kable(shown, row.names = FALSE)
} else {
  cat("No supervised interaction diagnostic is available yet.")
}
```

```{r}
#| label: supervised-interaction-test

if (nrow(classification_test_table)) {
  shown <- classification_test_table
  shown$p_value <- format.pval(shown$p_value, digits = 3L, eps = 1e-4)
  numeric_columns <- vapply(shown, is.numeric, logical(1))
  shown[numeric_columns] <- lapply(shown[numeric_columns], round, digits = 3L)
  knitr::kable(shown, row.names = FALSE)
} else {
  cat("No supervised interaction test is available yet.")
}
```

## Targeted-case interpretation

```{r}
#| label: targeted-case-summary
#| results: asis

if (nrow(diagnostic_table) && nrow(classification_test_table)) {
  case_diagnostic <- diagnostic_table[
    diagnostic_table$dataset == classification_specification$dataset,
    , drop = FALSE
  ]
  case_results <- result_table[
    result_table$dataset == classification_specification$dataset,
    c("method", "ARI"), drop = FALSE
  ]
  case_test <- classification_test_table[
    classification_test_table$dataset == classification_specification$dataset,
    , drop = FALSE
  ]
  dm_ari <- case_results$ARI[case_results$method == "D_m"]
  dmint_ari <- case_results$ARI[case_results$method == "D_mint"]
  cat(sprintf(
    paste0(
      "For South African Heart, the multiplicity-adjusted gate selects family ",
      "history ($p=%.3f$). Adding its interaction contribution raises ARI from ",
      "$%.3f$ for $D_m$ to $%.3f$ for $D_{\\mathrm{mint}}$. The resulting ",
      "interaction dissimilarity has $R_{\\mathrm{align}}=%.2f$, while the ",
      "mean nearest-neighbour overlap between the two distances is %.2f. ",
      "The LDL-by-family-history term is supported independently by the ",
      "classifier likelihood-ratio test ($p=%.4f$) and improves its repeated-",
      "cross-validation accuracy by %.3f.\n\n"
    ),
    min(gate_table$multiplicity_adjusted_p_value[
      gate_table$dataset == classification_specification$dataset
    ]),
    dm_ari,
    dmint_ari,
    case_diagnostic$R_align,
    case_diagnostic$neighbourhood_overlap_D_m_vs_D_mint,
    case_test$p_value,
    case_test$accuracy_gain
  ))
  cat(paste0(
    "Because family history is binary and is the only categorical predictor, ",
    "its observation-level interaction dissimilarity is rank-equivalent to ",
    "categorical matching. Here the gate supplies a numerical-data-driven ",
    "justification for reweighting that categorical separation. The case is ",
    "therefore a targeted illustration of the proposed diagnostics, not an ",
    "unbiased estimate of the method's average advantage across applications."
  ))
} else {
  cat("The targeted-case artifact is not available yet.")
}
```

## Interaction diagnostics

These diagnostics distinguish a positive interaction gate from an empirical
benefit in the final clustering.  `R_align` is the between-cluster to
within-cluster mean ratio of the active-average interaction dissimilarity,
computed using the partition returned by $D_{\mathrm{mint}}$.  The three
Spearman coefficients compare the active interaction dissimilarity with the
baseline, numerical, and categorical dissimilarities.  Neighbourhood overlap
is the mean proportion of the $\lceil\pi_{nn}(n-1)\rceil$ nearest neighbours
shared by $D_m$ and $D_{\mathrm{mint}}$.  These measures are descriptive: they
show whether the interaction geometry is coherent with the other blocks and
whether it changes local geometry without imposing an additional selection
threshold.

```{r}
#| label: interaction-diagnostics

if (nrow(diagnostic_table)) {
  diagnostic_columns <- c(
    "dataset", "active_count", "active_variables", "ARI_D_mint_minus_D_m",
    "partition_ARI_D_mint_vs_D_m", "partition_changed", "R_align",
    "interaction_baseline_spearman", "interaction_numeric_spearman",
    "interaction_categorical_spearman",
    "neighbourhood_overlap_D_m_vs_D_mint",
    "interaction_added_mean"
  )
  for (name in setdiff(diagnostic_columns, names(diagnostic_table))) {
    diagnostic_table[[name]] <- NA
  }
  shown <- diagnostic_table[diagnostic_columns]
  numeric_columns <- vapply(shown, is.numeric, logical(1))
  shown[numeric_columns] <- lapply(shown[numeric_columns], round, digits = 3L)
  knitr::kable(shown, row.names = FALSE)
} else {
  cat("No interaction diagnostics are available yet.")
}
```

```{r}
#| label: gate-detail

if (nrow(gate_table)) {
  shown <- gate_table
  shown$multiplicity_adjusted_p_value <- round(
    shown$multiplicity_adjusted_p_value, 4L
  )
  shown$statistic <- round(shown$statistic, 4L)
  knitr::kable(shown, row.names = FALSE)
} else {
  cat("No gate results are available yet.")
}
```

## Neighbour-proportion sensitivity

```{r}
#| label: neighbour-sensitivity

if (nrow(sensitivity_table)) {
  shown <- sensitivity_table
  shown$ARI <- round(shown$ARI, 3L)
  shown$partition_ARI_vs_primary <- round(
    shown$partition_ARI_vs_primary, 3L
  )
  knitr::kable(shown, row.names = FALSE)
} else {
  cat("No neighbour-sensitivity results are available yet.")
}
```

## Errors and completion audit

```{r}
#| label: completion-audit

completion <- data.frame(
  dataset = dataset_registry$dataset,
  artifact = unname(all_files),
  complete = file.exists(all_files),
  stringsAsFactors = FALSE
)
knitr::kable(completion, row.names = FALSE)

if (nrow(result_table)) {
  failures <- result_table[!is.na(result_table$error), c(
    "dataset", "method_label", "error"
  ), drop = FALSE]
  if (nrow(failures)) {
    knitr::kable(failures, row.names = FALSE)
  } else {
    cat("All enabled methods completed for every available artifact.")
  }
}
```

# Practical interpretation of the interaction term

The gate answers whether a numerical--categorical interaction is detectable;
it does not by itself establish that the interaction should be added to the
clustering dissimilarity.  The empirical results are therefore interpreted in
the following order.

1. If no variable passes the multiplicity-adjusted gate,
   $D_{\mathrm{mint}}=D_m$ exactly and there is no empirical reason to add an
   interaction term.
2. If the active set changes substantially across the pre-specified
   $\pi_{nn}$ grid, the interaction evidence is regarded as unstable.
3. If the active set is stable, the partition ARI between $D_m$ and
   $D_{\mathrm{mint}}$ measures whether the detected interaction has a
   material effect on the clustering, independently of the external labels.
4. When reference labels are available for evaluation, the ARI difference
   establishes whether that material change improves external agreement.
   Label information is never used by the gate or by either dissimilarity.
5. In the targeted South African Heart case study, the separate
   classifier comparison establishes whether numerical--categorical
   interactions improve supervised discrimination.  High classification
   performance with modest clustering ARI indicates that the classes are
   predictable but do not form fully separated unsupervised groups; it does
   not make the clustering labels part of the distance construction.
6. A detected but externally unhelpful interaction is reported as substantive
   structure that is not aligned with the target partition under the chosen
   clustering procedure. It is not evidence that interactions are universally
   harmful, just as a favourable result is not evidence that they should
   always be included.

In applications without external labels, the gate and its neighbourhood
sensitivity should be supplemented by label-free evidence such as resampling
stability, changes in the spectral embedding, and established domain
knowledge.  The interaction term should not be retained solely because the
gate rejects its null hypothesis.

# Reporting contract for the manuscript

The manuscript table should be copied from the consolidated ARI table only
after all six completion flags are `TRUE` and the error audit is empty.  The
text should report the active variables and the $D_{\mathrm{mint}}-D_m$ ARI
difference for every dataset.  Evidence for an interaction effect requires
more than a passed gate: the interaction term should also change the partition
or improve its agreement with the external labels.  A passed gate with
unchanged or worse ARI is reported as a detected association that did not
improve that empirical clustering.  For the targeted case study, the
cross-validated classifier comparison and omnibus interaction test are
reported alongside ARI to distinguish supervised predictability from
unsupervised cluster recovery.

```{r}
#| label: session-information

utils::sessionInfo()
```
