---
title: "Revised simulations for spectral clustering of mixed data"
subtitle: "Reproducible supplementary material for the first revision"
author: "Supplementary simulation code"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: false
---

This document defines the complete revised synthetic-data study.  The three
experiments are presented in their intended manuscript order:

1. a controlled nonlinear interaction experiment with a matched null;
2. the code-faithful reconstruction of the original cross-type experiment;
3. a paired multilevel structured-block experiment.

The first code chunk contains the quantities that are expected to be edited
when changing the design.  The remaining chunks should normally remain
unchanged.  Task artifacts are configuration-hashed and are never overwritten
by an ordinary re-render.

# Analysis parameters

```{r}
#| label: analysis-parameters
#| echo: true

analysis_parameters <- list(
  analysis_version = "R1-synthetic-v1",

  execution = list(
    # Change to TRUE only when results should actually be produced.
    run_experiments = FALSE,
    # Select any subset without changing the statistical configuration or
    # result directory. Completed experiments are collected together.
    experiments = c("Simulation_1", "Simulation_2", "Simulation_3"),
    # `replicates` defines the final study. `replicates_to_run` selects the
    # current batch and may be changed between renders without changing the
    # result directory. Existing task artifacts are skipped automatically.
    replicates = 20L,
    replicates_to_run = 1:20,
    workers = 1L,
    show_progress = TRUE,
    output_directory = "results/R1_simulations",
    seed_stride = 100L,
    cluster_seed_offset = 11L
  ),

  distance = list(
    preset = "custom",
    method_num = "pc_scores",
    method_cat = "tvd",
    commensurable = FALSE,
    interaction = FALSE,
    gamma = 1,
    prop_nn = 0.10,
    score = "ba",
    decision = "prior_corrected",
    gate_permutations = 499L,
    gate_alpha = 0.01,
    gate_seed_offset = 31L,
    # Ahmad--Dey (2007) used five equal-width intervals unless stated
    # otherwise when computing numerical attribute significance.
    ahmad_dey_bins = 5L
  ),

  neighbour_sensitivity = list(
    # The confirmatory analysis uses prop_nn = 0.10 above. This targeted
    # grid addresses the reviewers' request without repeating every external
    # comparator at every neighbour proportion.
    enabled = TRUE,
    proportions = c(0.02, 0.05, 0.10, 0.20),
    experiment_1_configuration = "reference",
    experiment_2_sample_size = 500L,
    experiment_3_n_per_group = 120L,
    experiment_3_profile_strength = 0.35
  ),

  spectral = list(
    affinity = "global_gaussian",
    primary_bandwidth_multiplier = 1,
    sensitivity_bandwidth_multipliers = c(0.5, 1, 2),
    kmeans_nstart = 100L,
    kmeans_iter_max = 100L,
    kmeans_algorithm = "Hartigan-Wong"
  ),

  methods = c(
    Gated_active_average_SC = TRUE,
    Package_no_interaction_SC = TRUE,
    Gower_SC = TRUE,
    Modified_Gower_SC = TRUE,
    Euclidean_onehot_SC = TRUE,
    Ahmad_Dey_SC = TRUE,
    # Optional robustness comparator; it was not requested by the reviewers.
    DKSS_SC = FALSE
  ),

  experiment_1 = list(
    label = "Controlled nonlinear interaction and matched null",
    seed_base = 26100000L,
    n_per_moon = 180L,
    moon_noise_sd = 0.08,
    sensor_noise_sd = 0.02,
    band_probabilities = c(A = 0.10, B = 0.45, C = 0.45),
    p_numeric_max = 10L,
    q_categorical_max = 5L,
    configurations = data.frame(
      configuration = c(
        "reference", "numeric_medium", "numeric_high",
        "categorical_high", "joint_high"
      ),
      p_numeric = c(2L, 6L, 10L, 2L, 10L),
      q_categorical = c(2L, 2L, 2L, 5L, 5L),
      stringsAsFactors = FALSE
    ),
    interaction_states = c("signal", "null"),
    bandwidth_configuration = "reference"
  ),

  experiment_2 = list(
    label = "Original cross-type simulation reconstructed from executable code",
    seed_base = 26200000L,
    sample_sizes = c(500L, 1000L),
    p_numeric = 6L,
    q_categorical = 4L,
    bandwidth_sample_size = 500L
  ),

  experiment_3 = list(
    label = "Paired multilevel structured blocks",
    seed_base = 26300000L,
    primary_n_per_group = 120L,
    sample_sizes_per_group = c(120L, 240L),
    primary_profile_strength = 0.35,
    profile_strengths = c(0.20, 0.35, 0.50),
    p_numeric = 12L,
    categorical_levels = rep(c(2L, 3L, 4L), 4L),
    numeric_loading = 0.72,
    numeric_mean_scale = 0.65,
    categorical_model = "nominal_association_blocks",
    categorical_copy_probability = 0.85,
    arms = c("aligned", "decoupled")
  )
)
```

Setting `run_experiments = FALSE` makes an ordinary render safe: it constructs
the task plan and reads any artifacts already present, but does not launch
simulation jobs.  To produce the results, change that single value to `TRUE`.
The `workers` value controls task-level parallelism.
The `execution$experiments` vector controls only which missing task artifacts
are produced by a render. It is deliberately excluded from the configuration
hash, so Simulation 1, 2, and 3 can be run separately and accumulated in the
same result directory. Collection and summary chunks always inspect the full
three-experiment plan.

Likewise, `execution$replicates` defines the final 20-replicate study, whereas
`execution$replicates_to_run` selects only the current batch. The default first
batch is `1:5`; subsequent renders can use `6:10`, `11:15`, and `16:20`, or
cumulative selections such as `1:10`. The batch selector is excluded from the
configuration hash, and every existing task artifact is skipped, so all
batches accumulate safely in the same result directory and the summaries
report all completed replicates.

With `execution$show_progress = TRUE`, an execution render reports the number
of selected, previously completed, and remaining task artifacts. With one
worker, the progress bar advances after every task; with multiple workers, it
advances after each parallel wave. The bar counts tasks rather than estimated
clock time because scenarios have different computational costs.

The primary neighbour proportion is fixed in advance at
`distance$prop_nn = 0.10`. The targeted sensitivity analysis evaluates
`neighbour_sensitivity$proportions` only for the gated method and only in the
representative conditions declared above. For categorical levels $a$ and
$b$, the realized neighbourhood size is

$$
k_{ab}=\max\{1,\lfloor \pi_{nn}(m_{ab}-1)\rfloor\},
$$

where $m_{ab}$ is the number of observations whose category is $a$ or
$b$. Thus the same proportion adapts to the available sample size for each
level pair rather than imposing a common fixed number of neighbours.

# Reproducible setup

```{r}
#| label: setup
#| include: true

`%||%` <- function(x, y) if (is.null(x)) y else x

input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) "")
if (!length(input_file) || is.na(input_file[[1L]]) || !nzchar(input_file[[1L]])) {
  input_file <- file.path(
    getwd(),
    "spectral_cluster_experiments", "spectral_stco_revision_r1",
    "R1_simulations.qmd"
  )
}
input_file <- input_file[[1L]]
input_file <- normalizePath(input_file, mustWork = TRUE)
revision_directory <- dirname(input_file)
repository_root <- dirname(dirname(revision_directory))
full_directory <- file.path(revision_directory, "full")
moon_directory <- file.path(revision_directory, "moon_scaling")
structured_directory <- file.path(revision_directory, "structured_blocks_pilot")

required_packages <- c("devtools", "digest", "knitr", "mclust", "mvtnorm", "yaml")
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages)) {
  stop("Missing required packages: ", paste(missing_packages, collapse = ", "))
}

devtools::load_all(file.path(repository_root, "manydist"), quiet = TRUE)

source_files <- c(
  bridge = file.path(revision_directory, "R", "manydist_bridge.R"),
  gate_helpers = file.path(revision_directory, "R", "distance_r1.R"),
  full_scenarios = file.path(full_directory, "R", "scenarios.R"),
  full_engine = file.path(full_directory, "R", "engine.R"),
  moon_generator = file.path(moon_directory, "R", "generator.R"),
  gated_distance = file.path(moon_directory, "R", "gated_distance.R"),
  structured_generator = file.path(structured_directory, "R", "generator.R")
)
invisible(lapply(source_files, source))

source_hashes <- unname(tools::md5sum(source_files))
names(source_hashes) <- names(source_files)

# Runtime choices do not define a new statistical analysis.
analysis_contract <- analysis_parameters
analysis_contract$execution$run_experiments <- NULL
analysis_contract$execution$experiments <- NULL
analysis_contract$execution$replicates_to_run <- NULL
analysis_contract$execution$workers <- NULL
analysis_contract$execution$show_progress <- NULL
analysis_contract$execution$output_directory <- NULL
configuration_tag <- substr(
  digest::digest(
    list(parameters = analysis_contract, source_hashes = source_hashes),
    algo = "xxhash64"
  ),
  1L,
  12L
)

result_directory <- file.path(
  revision_directory,
  analysis_parameters$execution$output_directory,
  configuration_tag
)

kmeans_specification <- list(
  nstart = analysis_parameters$spectral$kmeans_nstart,
  iter_max = analysis_parameters$spectral$kmeans_iter_max,
  algorithm = analysis_parameters$spectral$kmeans_algorithm
)
```

The analysis uses the local `manydist` source tree.  The following checksums
identify every sourced implementation file.

```{r}
#| label: source-manifest

knitr::kable(data.frame(
  component = names(source_files),
  file = unname(source_files),
  md5 = source_hashes,
  row.names = NULL
))
```

# Common distance and clustering construction

The package baseline is

```{r}
#| label: baseline-specification
#| eval: false

manydist::mdist(
  x = data,
  preset = "custom",
  method_num = "pc_scores",
  method_cat = "tvd",
  commensurable = FALSE,
  interaction = FALSE,
  ncomp = number_of_numeric_variables
)
```

For active gate set $A$, the paper-specific extension is

$$
D_{\mathrm{R1}}
=D_{\mathrm{base}}
+\gamma\,\overline D_{\mathrm{num}}
  \left\{|A|^{-1}\sum_{j\in A}D_{\mathrm{int},j}\right\}.
$$

If $A$ is empty, the returned matrix is bit-for-bit identical to the package
baseline.  All spectral methods use the Ng--Jordan--Weiss normalized spectral
construction with a global Gaussian affinity.  The primary bandwidth is the
median positive off-diagonal distance.

The synthetic study deliberately holds the clustering algorithm fixed.  Every
enabled method supplies a pairwise dissimilarity to the same spectral routine.
Centroid-based algorithms such as k-prototypes and KAMILA are therefore not
mixed into the primary simulation tables; if reported elsewhere, they are
labelled separately as algorithm-level benchmarks.

The Ahmad--Dey comparator uses the pairwise dissimilarity in Equation (3.3.1)
of [Ahmad and Dey (2007)](https://doi.org/10.1016/j.datak.2007.03.016), rather
than their k-means-type optimization.  Numerical variables are min--max
normalized and divided into five equal-width intervals solely to estimate
their association-based significance weights.  Categorical level distances
are the average total-variation distances between conditional profiles over
the remaining categorical variables and discretized numerical variables.  The
resulting pairwise matrix is then clustered by the same spectral routine used
for every other method.  This preserves the shared TVD building block while
isolating differences in numerical representation, weighting, and the gated
cross-type augmentation.

```{r}
#| label: common-functions

r1_sanitize_id <- function(x) {
  gsub("[^A-Za-z0-9_-]+", "-", as.character(x))
}

r1_global_gaussian <- function(distance, multiplier = 1) {
  distance <- full_validate_distance(distance)
  positive <- distance[upper.tri(distance) & distance > 0]
  if (!length(positive)) stop("Distance has no positive off-diagonal values.")
  sigma <- as.numeric(multiplier) * stats::median(positive)
  if (!is.finite(sigma) || sigma <= 0) stop("Invalid Gaussian bandwidth.")
  affinity <- exp(-(distance^2) / (2 * sigma^2))
  diag(affinity) <- 0
  list(matrix = affinity, sigma = sigma)
}

r1_cluster_distance <- function(distance, truth, seed, bandwidth_multiplier = 1) {
  affinity <- r1_global_gaussian(distance, bandwidth_multiplier)
  fit <- full_spectral_cluster(
    affinity$matrix,
    k = nlevels(droplevels(factor(truth))),
    seed = seed,
    kmeans_specification = kmeans_specification
  )
  list(
    cluster = fit$cluster,
    ARI = mclust::adjustedRandIndex(fit$cluster, truth),
    sigma = affinity$sigma,
    objective = fit$objective,
    iterations = fit$iterations,
    ifault = fit$ifault
  )
}

r1_alignment_ratio <- function(distance, partition) {
  distance <- full_validate_distance(distance)
  upper <- upper.tri(distance)
  same <- outer(partition, partition, `==`)[upper]
  values <- distance[upper]
  within <- mean(values[same])
  between <- mean(values[!same])
  if (within > 0) between / within else NA_real_
}

r1_enabled <- function(method) {
  isTRUE(unname(analysis_parameters$methods[[method]]))
}

r1_equal_width_factor <- function(x, bins) {
  x <- as.numeric(x)
  observed <- range(x, finite = TRUE)
  if (!all(is.finite(observed)) || observed[[1L]] == observed[[2L]]) {
    return(factor(rep("constant", length(x))))
  }
  factor(cut(
    x,
    breaks = seq(observed[[1L]], observed[[2L]], length.out = bins + 1L),
    include.lowest = TRUE,
    ordered_result = FALSE
  ))
}

r1_minmax_scale <- function(x) {
  x <- as.numeric(x)
  observed <- range(x, finite = TRUE)
  if (!all(is.finite(observed)) || observed[[1L]] == observed[[2L]]) {
    return(rep(0, length(x)))
  }
  (x - observed[[1L]]) / (observed[[2L]] - observed[[1L]])
}

# Pairwise form of Ahmad and Dey (2007), Equation (3.3.1).  The categorical
# level dissimilarities are total-variation distances between conditional
# profiles. `manydist` computes those profiles and level deltas; this wrapper
# adds the paper's five-bin numerical significance weights and squared
# pairwise aggregation before passing the result to spectral clustering.
r1_ahmad_dey_distance <- function(data, bins = 5L) {
  bins <- as.integer(bins)
  if (!is.finite(bins) || bins < 2L) stop("`bins` must be at least two.")

  numeric_columns <- vapply(data, is.numeric, logical(1))
  categorical_columns <- vapply(data, is.factor, logical(1))
  numeric_data <- data[, numeric_columns, drop = FALSE]
  categorical_data <- data[, categorical_columns, drop = FALSE]

  discretized_numeric <- lapply(numeric_data, r1_equal_width_factor, bins = bins)
  names(discretized_numeric) <- names(numeric_data)
  discrete_profiles <- data.frame(
    categorical_data,
    discretized_numeric,
    check.names = FALSE
  )
  discrete_profiles[] <- lapply(discrete_profiles, function(x) {
    droplevels(factor(x))
  })

  if (!ncol(discrete_profiles)) {
    stop("Ahmad--Dey distance requires at least one input variable.")
  }
  if (ncol(discrete_profiles) == 1L) {
    stop("Ahmad--Dey conditional profiles require at least two variables.")
  }

  delta_fit <- manydist:::cat_delta(
    discrete_profiles,
    method_cat = "tvd"
  )
  delta <- as.matrix(delta_fit$tvd)
  level_counts <- vapply(discrete_profiles, nlevels, integer(1))
  stops <- cumsum(level_counts)
  starts <- c(1L, head(stops, -1L) + 1L)
  positions <- Map(seq.int, starts, stops)

  n <- nrow(data)
  categorical_distance <- matrix(0, n, n)
  if (ncol(categorical_data)) {
    for (column_index in seq_len(ncol(categorical_data))) {
      levels_index <- positions[[column_index]]
      observed_index <- as.integer(discrete_profiles[[column_index]])
      contribution <- delta[levels_index, levels_index, drop = FALSE][
        observed_index,
        observed_index,
        drop = FALSE
      ]
      categorical_distance <- categorical_distance + contribution^2
    }
  }

  numeric_distance <- matrix(0, n, n)
  numeric_weights <- numeric(0)
  if (ncol(numeric_data)) {
    numeric_profile_index <- ncol(categorical_data) + seq_len(ncol(numeric_data))
    numeric_weights <- vapply(numeric_profile_index, function(column_index) {
      block <- delta[
        positions[[column_index]],
        positions[[column_index]],
        drop = FALSE
      ]
      if (nrow(block) < 2L) 0 else mean(block[upper.tri(block)])
    }, numeric(1))
    scaled_numeric <- as.data.frame(lapply(numeric_data, r1_minmax_scale))
    weighted_numeric <- sweep(
      as.matrix(scaled_numeric),
      2L,
      numeric_weights,
      `*`
    )
    numeric_distance <- as.matrix(stats::dist(weighted_numeric))^2
  }

  distance <- numeric_distance + categorical_distance
  diag(distance) <- 0
  attributes(distance)$numeric_weights <- numeric_weights
  full_validate_distance(distance, "Ahmad--Dey")
}
```

# Deterministic task plan

Signal/null states in Experiment 1 and aligned/decoupled arms in Experiment 3
use the same data seed.  This makes their comparisons paired rather than
comparisons of unrelated Monte Carlo samples.

```{r}
#| label: build-task-plan

r1_build_task_plan <- function(parameters) {
  replicates <- seq_len(as.integer(parameters$execution$replicates))
  stride <- as.integer(parameters$execution$seed_stride)
  neighbour_sensitivity <- parameters$neighbour_sensitivity
  tasks <- list()
  task_index <- 0L

  experiment_1 <- parameters$experiment_1
  for (configuration_index in seq_len(nrow(experiment_1$configurations))) {
    configuration <- experiment_1$configurations[configuration_index, , drop = FALSE]
    for (replicate in replicates) {
      data_seed <- experiment_1$seed_base +
        configuration_index * 100000L + replicate * stride
      for (state in experiment_1$interaction_states) {
        task_index <- task_index + 1L
        tasks[[task_index]] <- list(
          experiment = "Simulation_1",
          condition = as.character(configuration$configuration),
          state = state,
          replicate = replicate,
          data_seed = data_seed,
          p_numeric = as.integer(configuration$p_numeric),
          q_categorical = as.integer(configuration$q_categorical),
          bandwidth_sensitive = identical(
            as.character(configuration$configuration),
            experiment_1$bandwidth_configuration
          ),
          neighbour_sensitive = isTRUE(neighbour_sensitivity$enabled) &&
            identical(
              as.character(configuration$configuration),
              neighbour_sensitivity$experiment_1_configuration
            )
        )
      }
    }
  }

  experiment_2 <- parameters$experiment_2
  for (size_index in seq_along(experiment_2$sample_sizes)) {
    n <- as.integer(experiment_2$sample_sizes[[size_index]])
    for (replicate in replicates) {
      task_index <- task_index + 1L
      tasks[[task_index]] <- list(
        experiment = "Simulation_2",
        condition = paste0("n", n),
        state = "interaction",
        replicate = replicate,
        data_seed = experiment_2$seed_base +
          size_index * 100000L + replicate * stride,
        n = n,
        bandwidth_sensitive = n == experiment_2$bandwidth_sample_size,
        neighbour_sensitive = isTRUE(neighbour_sensitivity$enabled) &&
          n == neighbour_sensitivity$experiment_2_sample_size
      )
    }
  }

  experiment_3 <- parameters$experiment_3
  condition_grid <- unique(rbind(
    data.frame(
      n_per_group = experiment_3$sample_sizes_per_group,
      profile_strength = experiment_3$primary_profile_strength
    ),
    data.frame(
      n_per_group = experiment_3$primary_n_per_group,
      profile_strength = experiment_3$profile_strengths
    )
  ))
  rownames(condition_grid) <- NULL

  for (condition_index in seq_len(nrow(condition_grid))) {
    n_per_group <- as.integer(condition_grid$n_per_group[[condition_index]])
    profile_strength <- as.numeric(condition_grid$profile_strength[[condition_index]])
    condition <- paste0(
      "npg", n_per_group,
      "_profile", format(profile_strength, nsmall = 2, trim = TRUE)
    )
    for (replicate in replicates) {
      data_seed <- experiment_3$seed_base +
        condition_index * 100000L + replicate * stride
      for (arm in experiment_3$arms) {
        task_index <- task_index + 1L
        tasks[[task_index]] <- list(
          experiment = "Simulation_3",
          condition = condition,
          state = arm,
          replicate = replicate,
          data_seed = data_seed,
          n_per_group = n_per_group,
          profile_strength = profile_strength,
          bandwidth_sensitive = n_per_group == experiment_3$primary_n_per_group &&
            isTRUE(all.equal(
              profile_strength,
              experiment_3$primary_profile_strength
            )),
          neighbour_sensitive = isTRUE(neighbour_sensitivity$enabled) &&
            n_per_group == neighbour_sensitivity$experiment_3_n_per_group &&
            isTRUE(all.equal(
              profile_strength,
              neighbour_sensitivity$experiment_3_profile_strength
            ))
        )
      }
    }
  }

  for (index in seq_along(tasks)) {
    task <- tasks[[index]]
    task$task_id <- index
    task$key <- paste(
      task$experiment,
      r1_sanitize_id(task$condition),
      r1_sanitize_id(task$state),
      sprintf("rep%03d", task$replicate),
      sep = "__"
    )
    tasks[[index]] <- task
  }
  tasks
}

task_plan <- r1_build_task_plan(analysis_parameters)
valid_experiments <- unique(vapply(
  task_plan,
  `[[`,
  character(1),
  "experiment"
))
selected_experiments <- unique(as.character(
  analysis_parameters$execution$experiments
))
unknown_experiments <- setdiff(selected_experiments, valid_experiments)
if (!length(selected_experiments) || length(unknown_experiments)) {
  stop(
    "`execution$experiments` must be a non-empty subset of: ",
    paste(valid_experiments, collapse = ", "),
    if (length(unknown_experiments)) {
      paste0(". Unknown values: ", paste(unknown_experiments, collapse = ", "))
    } else {
      "."
    },
    call. = FALSE
  )
}
requested_replicates <- analysis_parameters$execution$replicates_to_run
if (!is.numeric(requested_replicates) || !length(requested_replicates) ||
    anyNA(requested_replicates) ||
    any(requested_replicates != as.integer(requested_replicates)) ||
    any(requested_replicates < 1L) ||
    any(requested_replicates > analysis_parameters$execution$replicates)) {
  stop(
    "`execution$replicates_to_run` must contain integers between 1 and ",
    analysis_parameters$execution$replicates,
    ".",
    call. = FALSE
  )
}
selected_replicates <- sort(unique(as.integer(requested_replicates)))
execution_task_plan <- Filter(
  function(task) {
    task$experiment %in% selected_experiments &&
      task$replicate %in% selected_replicates
  },
  task_plan
)
task_table <- do.call(rbind, lapply(task_plan, function(task) {
  data.frame(
    task_id = task$task_id,
    experiment = task$experiment,
    condition = task$condition,
    state = task$state,
    replicate = task$replicate,
    data_seed = task$data_seed,
    bandwidth_sensitive = task$bandwidth_sensitive,
    neighbour_sensitive = task$neighbour_sensitive,
    selected_for_execution = task$experiment %in% selected_experiments &&
      task$replicate %in% selected_replicates,
    stringsAsFactors = FALSE
  )
}))

task_table$planned_tasks <- 1L
task_table$selected_tasks <- as.integer(task_table$selected_for_execution)
task_counts <- stats::aggregate(
  cbind(planned_tasks, selected_tasks) ~ experiment + condition + state,
  task_table,
  sum
)
knitr::kable(task_counts)
```

# Data generation

## Simulation 1: nonlinear interaction mechanism

The first experiment extends the interaction-moons example.  Six generating
groups are the combinations of two moon curves and three arc bands.  Extra
numerical variables are noisy projections of the same latent geometry; extra
categorical variables are exactly balanced nuisance factors.  The matched
null permutes the complete categorical rows, preserving both marginal blocks
while removing their cross-type alignment.

## Simulation 2: original executable construction

The second experiment follows the historical `simulation xBIG.R` code.  It
uses six numerical variables, four categorical variables, and three groups
defined by the coefficient associated with the first two categorical
variables.  Numerical variables `V3`, `V2`, and `V1` are constructed
recursively from later numerical coordinates.  The fourth and third
categorical variables are retained as nuisance variables.

## Simulation 3: multilevel structured blocks

The final experiment has three numerical covariance/loading regimes and three
nominal association regimes. Its 12 categorical variables consist of four
binary, four three-level, and four four-level factors. The aligned and
decoupled arms reuse exactly the same rows within each block; only their
matching changes.

```{r}
#| label: generate-task-data

r1_generate_task_data <- function(task) {
  if (identical(task$experiment, "Simulation_1")) {
    specification <- analysis_parameters$experiment_1
    generated <- moon_scaling_generate(
      seed = task$data_seed,
      n_per_moon = specification$n_per_moon,
      p_numeric = task$p_numeric,
      q_categorical = task$q_categorical,
      interaction_state = task$state,
      moon_noise_sd = specification$moon_noise_sd,
      sensor_noise_sd = specification$sensor_noise_sd,
      band_probabilities = specification$band_probabilities,
      p_numeric_max = specification$p_numeric_max,
      q_categorical_max = specification$q_categorical_max
    )
    return(list(
      x = generated$data,
      truth = droplevels(factor(generated$truth)),
      diagnostics = generated$diagnostics
    ))
  }

  if (identical(task$experiment, "Simulation_2")) {
    specification <- analysis_parameters$experiment_2
    return(full_make_legacy_simulation_2(
      parameters = list(
        n = task$n,
        p_numeric = specification$p_numeric,
        q_categorical = specification$q_categorical
      ),
      seed = task$data_seed
    ))
  }

  if (identical(task$experiment, "Simulation_3")) {
    specification <- analysis_parameters$experiment_3
    pair <- structured_blocks_make_pair(
      specification = list(
        n_per_group = task$n_per_group,
        p_numeric = specification$p_numeric,
        categorical_levels = specification$categorical_levels,
        numeric_loading = specification$numeric_loading,
        categorical_model = specification$categorical_model,
        categorical_copy_probability =
          specification$categorical_copy_probability,
        categorical_profile_strength = task$profile_strength,
        numeric_mean_scale = specification$numeric_mean_scale
      ),
      seed = task$data_seed
    )
    selected <- pair[[task$state]]
    return(list(
      x = selected$x,
      truth = droplevels(factor(selected$truth)),
      diagnostics = c(
        list(
          generator = "structured_multilevel_blocks_v1",
          arm = selected$arm,
          categorical_regime_truth_ARI = mclust::adjustedRandIndex(
            selected$categorical_regime,
            selected$truth
          )
        ),
        pair$diagnostics
      )
    ))
  }

  stop("Unknown experiment: ", task$experiment, call. = FALSE)
}
```

# Production of task-level results

Each task produces one RDS artifact containing the method results, gate
decisions, estimated-partition diagnostics, generation metadata, source
hashes, and session provenance.  Existing artifacts under the same
configuration tag are skipped.

```{r}
#| label: task-runner-functions

r1_method_labels <- c(
  Gated_active_average_SC = "Gated active average",
  Package_no_interaction_SC = "Package no interaction",
  Gower_SC = "Gower",
  Modified_Gower_SC = "Modified Gower",
  Euclidean_onehot_SC = "Euclidean one-hot",
  Ahmad_Dey_SC = "Ahmad--Dey distance",
  DKSS_SC = "DKSS"
)

r1_task_file <- function(task) {
  file.path(result_directory, paste0(task$key, ".rds"))
}

r1_distance_methods <- function(data, gated_fit) {
  methods <- list()
  errors <- list()
  seconds <- numeric()

  add_distance <- function(name, expression) {
    started <- proc.time()[[3L]]
    fitted <- tryCatch(
      list(value = full_validate_distance(force(expression), name), error = NA_character_),
      error = function(e) list(value = NULL, error = conditionMessage(e))
    )
    methods[[name]] <<- fitted$value
    errors[[name]] <<- fitted$error
    seconds[[name]] <<- proc.time()[[3L]] - started
  }

  if (r1_enabled("Gated_active_average_SC")) {
    add_distance("Gated_active_average_SC", gated_fit$distance)
  }
  if (r1_enabled("Package_no_interaction_SC")) {
    add_distance(
      "Package_no_interaction_SC",
      gated_fit$distance_no_interaction
    )
  }
  if (r1_enabled("Gower_SC")) {
    add_distance(
      "Gower_SC",
      as.matrix(manydist::mdist(data, preset = "gower")$distance)
    )
  }
  if (r1_enabled("Modified_Gower_SC")) {
    add_distance(
      "Modified_Gower_SC",
      as.matrix(manydist::mdist(data, preset = "mod_gower")$distance)
    )
  }
  if (r1_enabled("Euclidean_onehot_SC")) {
    add_distance(
      "Euclidean_onehot_SC",
      as.matrix(manydist::mdist(data, preset = "euclidean")$distance)
    )
  }
  if (r1_enabled("Ahmad_Dey_SC")) {
    add_distance(
      "Ahmad_Dey_SC",
      r1_ahmad_dey_distance(
        data,
        bins = analysis_parameters$distance$ahmad_dey_bins
      )
    )
  }
  if (r1_enabled("DKSS_SC")) {
    add_distance(
      "DKSS_SC",
      as.matrix(manydist::mdist(data, preset = "dkss")$distance)
    )
  }

  list(distances = methods, errors = errors, seconds = seconds)
}

r1_result_row <- function(task, method, bandwidth_multiplier, score = NULL,
                          distance_seconds = NA_real_, clustering_seconds = NA_real_,
                          error = NA_character_) {
  data.frame(
    task_key = task$key,
    experiment = task$experiment,
    condition = task$condition,
    state = task$state,
    replicate = task$replicate,
    data_seed = task$data_seed,
    method = method,
    method_label = unname(r1_method_labels[[method]]),
    bandwidth_multiplier = bandwidth_multiplier,
    ARI = if (is.null(score)) NA_real_ else score$ARI,
    affinity_sigma = if (is.null(score)) NA_real_ else score$sigma,
    objective = if (is.null(score)) NA_real_ else score$objective,
    iterations = if (is.null(score)) NA_integer_ else score$iterations,
    ifault = if (is.null(score)) NA_integer_ else score$ifault,
    distance_seconds = distance_seconds,
    clustering_seconds = clustering_seconds,
    error = error,
    stringsAsFactors = FALSE
  )
}

r1_pairwise_k_values <- function(data, prop_nn) {
  categorical <- data[vapply(data, is.factor, logical(1))]
  values <- unlist(lapply(categorical, function(labels) {
    pairwise_k <- .scmix_pairwise_k(labels, prop_nn)
    pairwise_k[upper.tri(pairwise_k)]
  }), use.names = FALSE)
  as.integer(values[is.finite(values)])
}

r1_upper_correlation <- function(candidate, reference) {
  candidate_values <- candidate[upper.tri(candidate)]
  reference_values <- reference[upper.tri(reference)]
  if (isTRUE(all.equal(candidate_values, reference_values, tolerance = 1e-12))) {
    return(1)
  }
  if (!length(candidate_values) ||
      stats::sd(candidate_values) <= sqrt(.Machine$double.eps) ||
      stats::sd(reference_values) <= sqrt(.Machine$double.eps)) {
    return(NA_real_)
  }
  stats::cor(candidate_values, reference_values)
}

r1_relative_frobenius <- function(candidate, reference) {
  denominator <- norm(reference, type = "F")
  numerator <- norm(candidate - reference, type = "F")
  if (denominator <= sqrt(.Machine$double.eps)) {
    return(if (numerator <= sqrt(.Machine$double.eps)) 0 else Inf)
  }
  numerator / denominator
}

r1_gate_jaccard <- function(candidate, reference) {
  candidate_active <- names(candidate)[candidate]
  reference_active <- names(reference)[reference]
  active_union <- union(candidate_active, reference_active)
  if (!length(active_union)) return(1)
  length(intersect(candidate_active, reference_active)) / length(active_union)
}

r1_run_neighbour_sensitivity <- function(
    task,
    data,
    truth,
    primary_fit,
    primary_partition,
    primary_fit_seconds
) {
  specification <- analysis_parameters$neighbour_sensitivity
  empty <- list(results = data.frame(), gates = data.frame())
  if (!isTRUE(specification$enabled) || !isTRUE(task$neighbour_sensitive)) {
    return(empty)
  }

  distance_specification <- analysis_parameters$distance
  primary_prop <- distance_specification$prop_nn
  proportions <- sort(unique(c(specification$proportions, primary_prop)))
  cluster_seed <- task$data_seed +
    analysis_parameters$execution$cluster_seed_offset
  gate_seed <- task$data_seed + distance_specification$gate_seed_offset

  fits <- vector("list", length(proportions))
  partitions <- vector("list", length(proportions))
  fit_seconds <- rep(NA_real_, length(proportions))
  errors <- rep(NA_character_, length(proportions))

  for (index in seq_along(proportions)) {
    prop_nn <- proportions[[index]]
    is_primary <- isTRUE(all.equal(prop_nn, primary_prop))

    if (is_primary) {
      fitted <- list(value = primary_fit, error = NA_character_)
      fit_seconds[[index]] <- primary_fit_seconds
    } else {
      started <- proc.time()[[3L]]
      fitted <- tryCatch(
        list(
          value = scmix_gated_active_distance(
            data = data,
            gamma = distance_specification$gamma,
            prop_nn = prop_nn,
            score = distance_specification$score,
            decision = distance_specification$decision,
            gate_permutations = distance_specification$gate_permutations,
            gate_alpha = distance_specification$gate_alpha,
            # The common seed gives every proportion the same permutations.
            gate_seed = gate_seed
          ),
          error = NA_character_
        ),
        error = function(e) list(value = NULL, error = conditionMessage(e))
      )
      fit_seconds[[index]] <- proc.time()[[3L]] - started
    }

    fits[[index]] <- fitted$value
    errors[[index]] <- fitted$error
    if (is.null(fitted$value)) next

    if (is_primary && !is.null(primary_partition)) {
      partitions[[index]] <- primary_partition
    } else {
      clustered <- tryCatch(
        r1_cluster_distance(
          fitted$value$distance,
          truth,
          cluster_seed,
          bandwidth_multiplier =
            analysis_parameters$spectral$primary_bandwidth_multiplier
        ),
        error = function(e) e
      )
      if (inherits(clustered, "error")) {
        errors[[index]] <- conditionMessage(clustered)
      } else {
        partitions[[index]] <- clustered$cluster
      }
    }
  }

  primary_index <- which(vapply(
    proportions,
    function(value) isTRUE(all.equal(value, primary_prop)),
    logical(1)
  ))[[1L]]
  reference_fit <- fits[[primary_index]]
  reference_partition <- partitions[[primary_index]]

  result_rows <- vector("list", length(proportions))
  gate_rows <- list()
  gate_index <- 0L

  for (index in seq_along(proportions)) {
    prop_nn <- proportions[[index]]
    fit <- fits[[index]]
    partition <- partitions[[index]]
    k_values <- r1_pairwise_k_values(data, prop_nn)

    result_rows[[index]] <- data.frame(
      task_key = task$key,
      experiment = task$experiment,
      condition = task$condition,
      state = task$state,
      replicate = task$replicate,
      data_seed = task$data_seed,
      prop_nn = prop_nn,
      primary_prop = isTRUE(all.equal(prop_nn, primary_prop)),
      realized_k_min = if (length(k_values)) min(k_values) else NA_integer_,
      realized_k_median = if (length(k_values)) stats::median(k_values) else NA_real_,
      realized_k_max = if (length(k_values)) max(k_values) else NA_integer_,
      ARI = if (is.null(partition)) NA_real_ else
        mclust::adjustedRandIndex(partition, truth),
      active_count = if (is.null(fit)) NA_integer_ else
        fit$diagnostics$active_count,
      exact_baseline_recovery = if (is.null(fit)) NA else
        identical(fit$distance, fit$distance_no_interaction),
      gate_jaccard_to_primary = if (is.null(fit)) NA_real_ else
        r1_gate_jaccard(fit$gate$detected, reference_fit$gate$detected),
      partition_ARI_to_primary = if (
        is.null(partition) || is.null(reference_partition)
      ) NA_real_ else mclust::adjustedRandIndex(partition, reference_partition),
      interaction_correlation_to_primary = if (is.null(fit)) NA_real_ else
        r1_upper_correlation(
          fit$components$interaction_active_average,
          reference_fit$components$interaction_active_average
        ),
      interaction_relative_difference_to_primary = if (is.null(fit)) NA_real_ else
        r1_relative_frobenius(
          fit$components$interaction_active_average,
          reference_fit$components$interaction_active_average
        ),
      final_distance_correlation_to_primary = if (is.null(fit)) NA_real_ else
        r1_upper_correlation(fit$distance, reference_fit$distance),
      final_distance_relative_difference_to_primary = if (is.null(fit)) NA_real_ else
        r1_relative_frobenius(fit$distance, reference_fit$distance),
      distance_seconds = fit_seconds[[index]],
      error = errors[[index]],
      stringsAsFactors = FALSE
    )

    if (!is.null(fit)) {
      gate_index <- gate_index + 1L
      gate_rows[[gate_index]] <- data.frame(
        task_key = task$key,
        experiment = task$experiment,
        condition = task$condition,
        state = task$state,
        replicate = task$replicate,
        prop_nn = prop_nn,
        variable = names(fit$gate$detected),
        detected = unname(fit$gate$detected),
        adjusted_p_value = unname(fit$gate$p_value),
        statistic = unname(fit$gate$statistic),
        stringsAsFactors = FALSE
      )
    }
  }

  list(
    results = do.call(rbind, result_rows),
    gates = if (length(gate_rows)) do.call(rbind, gate_rows) else data.frame()
  )
}

r1_run_task <- function(task) {
  output_file <- r1_task_file(task)
  if (file.exists(output_file)) return(output_file)
  dir.create(dirname(output_file), recursive = TRUE, showWarnings = FALSE)

  started_at <- Sys.time()
  generated <- r1_generate_task_data(task)
  data <- generated$x
  truth <- droplevels(factor(generated$truth))
  distance_specification <- analysis_parameters$distance

  gated_started <- proc.time()[[3L]]
  gated_fit <- scmix_gated_active_distance(
    data = data,
    gamma = distance_specification$gamma,
    prop_nn = distance_specification$prop_nn,
    score = distance_specification$score,
    decision = distance_specification$decision,
    gate_permutations = distance_specification$gate_permutations,
    gate_alpha = distance_specification$gate_alpha,
    gate_seed = task$data_seed + distance_specification$gate_seed_offset
  )
  gated_seconds <- proc.time()[[3L]] - gated_started

  constructed <- r1_distance_methods(data, gated_fit)
  bandwidths <- if (isTRUE(task$bandwidth_sensitive)) {
    analysis_parameters$spectral$sensitivity_bandwidth_multipliers
  } else {
    analysis_parameters$spectral$primary_bandwidth_multiplier
  }

  rows <- list()
  partitions <- list()
  row_index <- 0L
  cluster_seed <- task$data_seed +
    analysis_parameters$execution$cluster_seed_offset

  for (method in names(constructed$distances)) {
    distance <- constructed$distances[[method]]
    for (bandwidth in bandwidths) {
      clustering_started <- proc.time()[[3L]]
      fitted <- tryCatch(
        list(
          value = r1_cluster_distance(
            distance,
            truth,
            cluster_seed,
            bandwidth_multiplier = bandwidth
          ),
          error = NA_character_
        ),
        error = function(e) list(value = NULL, error = conditionMessage(e))
      )
      clustering_seconds <- proc.time()[[3L]] - clustering_started
      row_index <- row_index + 1L
      rows[[row_index]] <- r1_result_row(
        task,
        method,
        bandwidth,
        score = fitted$value,
        distance_seconds = constructed$seconds[[method]],
        clustering_seconds = clustering_seconds,
        error = fitted$error
      )
      if (isTRUE(all.equal(
        bandwidth,
        analysis_parameters$spectral$primary_bandwidth_multiplier
      )) && !is.null(fitted$value)) {
        partitions[[method]] <- fitted$value$cluster
      }
    }
  }

  failed_distances <- names(constructed$errors)[vapply(
    constructed$errors,
    function(error) length(error) && !is.na(error),
    logical(1)
  )]
  for (method in failed_distances) {
    row_index <- row_index + 1L
    rows[[row_index]] <- r1_result_row(
      task,
      method,
      analysis_parameters$spectral$primary_bandwidth_multiplier,
      distance_seconds = constructed$seconds[[method]],
      error = constructed$errors[[method]]
    )
  }

  exact_baseline_recovery <- identical(
    gated_fit$distance,
    gated_fit$distance_no_interaction
  )
  baseline_gated_partition_ARI <- if (
    all(c("Gated_active_average_SC", "Package_no_interaction_SC") %in%
        names(partitions))
  ) {
    mclust::adjustedRandIndex(
      partitions$Gated_active_average_SC,
      partitions$Package_no_interaction_SC
    )
  } else {
    NA_real_
  }
  interaction_alignment <- if (
    gated_fit$diagnostics$active_count > 0 &&
      "Gated_active_average_SC" %in% names(partitions)
  ) {
    r1_alignment_ratio(
      gated_fit$components$interaction_active_average,
      partitions$Gated_active_average_SC
    )
  } else {
    NA_real_
  }

  gate_table <- data.frame(
    variable = names(gated_fit$gate$detected),
    detected = unname(gated_fit$gate$detected),
    adjusted_p_value = unname(gated_fit$gate$p_value),
    statistic = unname(gated_fit$gate$statistic),
    stringsAsFactors = FALSE
  )

  neighbour_sensitivity <- r1_run_neighbour_sensitivity(
    task = task,
    data = data,
    truth = truth,
    primary_fit = gated_fit,
    primary_partition = partitions[["Gated_active_average_SC"]],
    primary_fit_seconds = gated_seconds
  )

  artifact <- list(
    schema_version = 1L,
    analysis_version = analysis_parameters$analysis_version,
    configuration_tag = configuration_tag,
    task = task,
    generator_diagnostics = generated$diagnostics,
    results = do.call(rbind, rows),
    gate = gate_table,
    neighbour_sensitivity = neighbour_sensitivity$results,
    neighbour_sensitivity_gates = neighbour_sensitivity$gates,
    diagnostics = data.frame(
      task_key = task$key,
      experiment = task$experiment,
      condition = task$condition,
      state = task$state,
      replicate = task$replicate,
      data_seed = task$data_seed,
      n = nrow(data),
      p_numeric = sum(vapply(data, is.numeric, logical(1))),
      q_categorical = sum(vapply(data, is.factor, logical(1))),
      active_count = gated_fit$diagnostics$active_count,
      exact_baseline_recovery = exact_baseline_recovery,
      baseline_gated_partition_ARI = baseline_gated_partition_ARI,
      interaction_alignment_ratio = interaction_alignment,
      interaction_added_mean = gated_fit$diagnostics$added_mean_distinct,
      final_distance_mean = gated_fit$diagnostics$final_mean_distinct,
      gated_distance_seconds = gated_seconds,
      stringsAsFactors = FALSE
    ),
    source_hashes = source_hashes,
    qmd_md5 = unname(tools::md5sum(input_file)),
    manydist_version = as.character(utils::packageVersion("manydist")),
    started_at = format(started_at, tz = "UTC", usetz = TRUE),
    finished_at = format(Sys.time(), tz = "UTC", usetz = TRUE),
    session_info = capture.output(utils::sessionInfo())
  )

  temporary_file <- tempfile("R1-task-", tmpdir = dirname(output_file), fileext = ".rds")
  saveRDS(artifact, temporary_file, version = 3L)
  if (file.exists(output_file)) {
    unlink(temporary_file)
    return(output_file)
  }
  if (!file.rename(temporary_file, output_file)) {
    unlink(temporary_file)
    stop("Could not atomically save task artifact: ", output_file)
  }
  output_file
}
```

## Optional execution

```{r}
#| label: execute-tasks

if (isTRUE(analysis_parameters$execution$run_experiments)) {
  workers <- max(1L, as.integer(analysis_parameters$execution$workers))
  if (.Platform$OS.type == "windows" && workers > 1L) {
    warning("Falling back to one worker on Windows in this document.")
    workers <- 1L
  }

  selected_complete <- vapply(
    execution_task_plan,
    function(task) file.exists(r1_task_file(task)),
    logical(1)
  )
  remaining_task_plan <- execution_task_plan[!selected_complete]
  cat(
    "Selected tasks: ", length(execution_task_plan),
    "; already complete: ", sum(selected_complete),
    "; remaining: ", length(remaining_task_plan), ".\n",
    sep = ""
  )

  r1_execute_with_progress <- function(tasks, workers, show_progress) {
    if (!length(tasks)) return(character())

    progress_bar <- if (isTRUE(show_progress)) {
      utils::txtProgressBar(
        min = 0L,
        max = length(tasks),
        initial = 0L,
        style = 3L,
        file = stderr()
      )
    } else {
      NULL
    }
    on.exit({
      if (!is.null(progress_bar)) close(progress_bar)
    }, add = TRUE)

    produced <- vector("list", length(tasks))
    completed <- 0L

    if (workers == 1L) {
      for (index in seq_along(tasks)) {
        produced[[index]] <- r1_run_task(tasks[[index]])
        completed <- completed + 1L
        if (!is.null(progress_bar)) {
          utils::setTxtProgressBar(progress_bar, completed)
        }
      }
    } else {
      # Wave-based scheduling provides reliable progress without adding a
      # package dependency. Task durations differ, so updates occur whenever
      # all tasks in the current wave have finished.
      waves <- split(
        seq_along(tasks),
        ceiling(seq_along(tasks) / workers)
      )
      for (wave in waves) {
        wave_results <- parallel::mclapply(
          tasks[wave],
          r1_run_task,
          mc.cores = min(workers, length(wave)),
          mc.preschedule = FALSE
        )
        failed <- vapply(wave_results, inherits, logical(1), "try-error")
        if (any(failed)) {
          stop(as.character(wave_results[[which(failed)[[1L]]]]), call. = FALSE)
        }
        produced[wave] <- wave_results
        completed <- completed + length(wave)
        if (!is.null(progress_bar)) {
          utils::setTxtProgressBar(progress_bar, completed)
        }
      }
    }

    unlist(produced, use.names = FALSE)
  }

  produced_files <- r1_execute_with_progress(
    remaining_task_plan,
    workers = workers,
    show_progress = analysis_parameters$execution$show_progress
  )
  invisible(produced_files)
} else {
  cat(
    "Simulation execution is disabled. Set `run_experiments = TRUE` in the ",
    "first code chunk to produce missing task artifacts for: ",
    paste(selected_experiments, collapse = ", "),
    "; replicates: ",
    paste(selected_replicates, collapse = ", "),
    ".\n",
    sep = ""
  )
}
```

# Collection and paper-ready summaries

```{r}
#| label: collect-results

available_files <- vapply(task_plan, r1_task_file, character(1))
available_files <- available_files[file.exists(available_files)]
artifacts <- lapply(available_files, readRDS)

results <- if (length(artifacts)) {
  do.call(rbind, lapply(artifacts, `[[`, "results"))
} else {
  data.frame()
}
diagnostics <- if (length(artifacts)) {
  do.call(rbind, lapply(artifacts, `[[`, "diagnostics"))
} else {
  data.frame()
}
gate_results <- if (length(artifacts)) {
  do.call(rbind, lapply(artifacts, function(artifact) {
    cbind(
      artifact$gate,
      task_key = artifact$task$key,
      experiment = artifact$task$experiment,
      condition = artifact$task$condition,
      state = artifact$task$state,
      replicate = artifact$task$replicate,
      stringsAsFactors = FALSE
    )
  }))
} else {
  data.frame()
}

neighbour_sensitivity_tables <- Filter(
  function(table) is.data.frame(table) && nrow(table),
  lapply(artifacts, `[[`, "neighbour_sensitivity")
)
neighbour_sensitivity_results <- if (length(neighbour_sensitivity_tables)) {
  do.call(rbind, neighbour_sensitivity_tables)
} else {
  data.frame()
}

neighbour_sensitivity_gate_tables <- Filter(
  function(table) is.data.frame(table) && nrow(table),
  lapply(artifacts, `[[`, "neighbour_sensitivity_gates")
)
neighbour_sensitivity_gates <- if (length(neighbour_sensitivity_gate_tables)) {
  do.call(rbind, neighbour_sensitivity_gate_tables)
} else {
  data.frame()
}

completion <- data.frame(
  configuration_tag = configuration_tag,
  selected_experiments = paste(selected_experiments, collapse = ", "),
  selected_replicates = paste(selected_replicates, collapse = ", "),
  planned_tasks = length(task_plan),
  selected_tasks = length(execution_task_plan),
  completed_tasks = length(artifacts),
  completed_selected_tasks = sum(vapply(
    execution_task_plan,
    function(task) file.exists(r1_task_file(task)),
    logical(1)
  )),
  completion_fraction = length(artifacts) / length(task_plan),
  selected_completion_fraction = sum(vapply(
    execution_task_plan,
    function(task) file.exists(r1_task_file(task)),
    logical(1)
  )) / length(execution_task_plan)
)
knitr::kable(completion, digits = 3)
```

The following tables are populated automatically once artifacts exist under
the current configuration tag.

## Primary ARI summaries

```{r}
#| label: primary-ari-summary

if (!nrow(results)) {
  cat("No result artifacts are available for this configuration.\n")
} else {
  primary <- results[
    is.na(results$bandwidth_multiplier) |
      abs(results$bandwidth_multiplier -
            analysis_parameters$spectral$primary_bandwidth_multiplier) < 1e-12,
    , drop = FALSE
  ]
  successful <- primary[is.na(primary$error) & is.finite(primary$ARI), , drop = FALSE]
  ari_summary <- stats::aggregate(
    ARI ~ experiment + condition + state + method_label,
    successful,
    function(values) c(
      mean = mean(values),
      sd = stats::sd(values),
      q025 = stats::quantile(values, 0.025),
      q975 = stats::quantile(values, 0.975),
      n = length(values)
    )
  )
  ari_summary <- data.frame(
    ari_summary[c("experiment", "condition", "state", "method_label")],
    ari_summary$ARI,
    row.names = NULL,
    check.names = FALSE
  )
  knitr::kable(ari_summary, digits = 3)
}
```

## Paired contribution of the interaction term

```{r}
#| label: paired-interaction-contrast

if (!nrow(results)) {
  cat("No paired contrasts are available yet.\n")
} else {
  primary_pair <- results[
    results$method %in% c(
      "Gated_active_average_SC",
      "Package_no_interaction_SC"
    ) &
      !is.na(results$bandwidth_multiplier) &
      abs(results$bandwidth_multiplier - 1) < 1e-12 &
      is.na(results$error),
    c("experiment", "condition", "state", "replicate", "method", "ARI")
  ]
  paired <- reshape(
    primary_pair,
    idvar = c("experiment", "condition", "state", "replicate"),
    timevar = "method",
    direction = "wide"
  )
  paired$gated_minus_baseline <-
    paired$ARI.Gated_active_average_SC -
    paired$ARI.Package_no_interaction_SC
  paired_summary <- stats::aggregate(
    gated_minus_baseline ~ experiment + condition + state,
    paired,
    function(values) c(
      mean = mean(values),
      sd = stats::sd(values),
      positive_fraction = mean(values > 0),
      zero_fraction = mean(values == 0),
      n = length(values)
    )
  )
  paired_summary <- data.frame(
    paired_summary[c("experiment", "condition", "state")],
    paired_summary$gated_minus_baseline,
    row.names = NULL,
    check.names = FALSE
  )
  knitr::kable(paired_summary, digits = 3)
}
```

## Gate activation, exact nesting, and post-hoc alignment

```{r}
#| label: gate-summary

if (!nrow(diagnostics)) {
  cat("No gate diagnostics are available yet.\n")
} else {
  gate_groups <- diagnostics[c("experiment", "condition", "state")]
  gate_summary_parts <- lapply(
    c(
      "active_count",
      "exact_baseline_recovery",
      "baseline_gated_partition_ARI",
      "interaction_alignment_ratio"
    ),
    function(variable) {
      part <- stats::aggregate(
        diagnostics[[variable]],
        gate_groups,
        function(values) mean(values, na.rm = TRUE)
      )
      names(part)[[ncol(part)]] <- variable
      part
    }
  )
  gate_summary <- Reduce(
    function(left, right) merge(
      left,
      right,
      by = c("experiment", "condition", "state"),
      all = TRUE,
      sort = FALSE
    ),
    gate_summary_parts
  )
  gate_summary$interaction_alignment_ratio[
    is.nan(gate_summary$interaction_alignment_ratio)
  ] <- NA_real_
  names(gate_summary)[names(gate_summary) == "exact_baseline_recovery"] <-
    "exact_nesting_fraction"
  knitr::kable(gate_summary, digits = 3)
}
```

The alignment statistic is descriptive.  It measures whether the active
interaction distance is larger between than within the identified gated
clusters.  It is not used to select interactions, tune `gamma`, or claim
recovery of an unknown external truth.

## Global-bandwidth sensitivity

```{r}
#| label: bandwidth-summary

if (!nrow(results)) {
  cat("No bandwidth results are available yet.\n")
} else {
  bandwidth_results <- results[
    !is.na(results$bandwidth_multiplier) & is.na(results$error),
    , drop = FALSE
  ]
  bandwidth_summary <- stats::aggregate(
    ARI ~ experiment + condition + state + method_label + bandwidth_multiplier,
    bandwidth_results,
    function(values) c(mean = mean(values), sd = stats::sd(values), n = length(values))
  )
  bandwidth_summary <- data.frame(
    bandwidth_summary[
      c("experiment", "condition", "state", "method_label", "bandwidth_multiplier")
    ],
    bandwidth_summary$ARI,
    row.names = NULL,
    check.names = FALSE
  )
  knitr::kable(bandwidth_summary, digits = 3)
}
```

## Nearest-neighbour proportion sensitivity

The primary analysis fixes $\pi_{nn}=0.10$ before examining the results.
The sensitivity grid is evaluated on the same generated datasets, with the
same clustering seed and the same sequence of label permutations at every
neighbour proportion. It is therefore a paired robustness analysis rather
than a tuning exercise. External competitors are not recomputed because
their distances do not depend on $\pi_{nn}$.

```{r}
#| label: neighbour-proportion-summary

r1_mean_na <- function(values) {
  if (all(is.na(values))) NA_real_ else mean(values, na.rm = TRUE)
}

if (!nrow(neighbour_sensitivity_results)) {
  cat("No nearest-neighbour sensitivity results are available yet.\n")
} else {
  neighbour_summary <- stats::aggregate(
    cbind(
      realized_k_min,
      realized_k_median,
      realized_k_max,
      ARI,
      active_count,
      exact_baseline_recovery,
      gate_jaccard_to_primary,
      partition_ARI_to_primary,
      interaction_correlation_to_primary,
      interaction_relative_difference_to_primary,
      final_distance_correlation_to_primary,
      final_distance_relative_difference_to_primary
    ) ~ experiment + condition + state + prop_nn,
    neighbour_sensitivity_results,
    r1_mean_na,
    na.action = stats::na.pass
  )
  knitr::kable(neighbour_summary, digits = 3)
}
```

The realized $k_{ab}$ columns make the proportion interpretable for each
sample-size and category-frequency regime. Gate Jaccard similarity, the
interaction-distance comparison, and partition ARI are all relative to the
pre-specified 10% analysis. Exact agreement is not required; the purpose is
to establish whether the substantive interaction findings and clustering
conclusions persist over a reasonable local-neighbourhood range.

```{r}
#| label: neighbour-proportion-gates

if (!nrow(neighbour_sensitivity_gates)) {
  cat("No nearest-neighbour gate results are available yet.\n")
} else {
  neighbour_gate_summary <- stats::aggregate(
    cbind(detected, adjusted_p_value, statistic) ~
      experiment + condition + state + prop_nn + variable,
    neighbour_sensitivity_gates,
    r1_mean_na,
    na.action = stats::na.pass
  )
  names(neighbour_gate_summary)[
    names(neighbour_gate_summary) == "detected"
  ] <- "detection_fraction"
  knitr::kable(neighbour_gate_summary, digits = 3)
}
```

## Method failures

```{r}
#| label: method-failures

if (!nrow(results) || !any(!is.na(results$error))) {
  cat("No method failures are recorded for the available artifacts.\n")
} else {
  failures <- unique(results[
    !is.na(results$error),
    c("experiment", "condition", "state", "replicate", "method", "error")
  ])
  knitr::kable(failures)
}
```

# Interpretation map for the revised manuscript

The experiments answer complementary questions.

- **Simulation 1** is the mechanism study.  It asks whether the active-average
  construction helps when the cross-type signal defines the groups and
  whether it nests exactly in the matched null.
- **Simulation 2** preserves continuity with the submitted work by rerunning
  the historical executable generator under the revised distance.
- **Simulation 3** introduces correlated numerical variables and associated
  multilevel nominal variables.  Its paired decoupling distinguishes methods
  that exploit genuine block alignment from methods that indiscriminately
  overweight the categorical block.

The confirmatory estimand is the paired ARI difference between the gated
method and its pure-package no-interaction baseline.  Comparisons with external
methods remain important, but the proposed construction is not claimed to
dominate every competitor in every data-generating mechanism.

# Session information

```{r}
#| label: session-information

sessionInfo()
```
